---
title: "What's New in CVXR 1.8.x"
author: Anqi Fu and Balasubramanian Narasimhan
---

## Complete Rewrite Using S7

CVXR 1.8.x is a ground-up rewrite using R's S7 object system, designed
to be isomorphic with CVXPY 1.8.1 for long-term maintainability. This
page summarizes the key changes from CVXR 1.x that may affect users.

## New Features

- **S7 class system** replaces S4 for all expression, constraint, and
  problem classes. Significantly faster construction and method dispatch.
- **13 solvers**: CLARABEL (default), SCS, OSQP, HiGHS, MOSEK, Gurobi,
  GLPK, GLPK_MI, ECOS, ECOS_BB, CPLEX, CVXOPT, and PIQP.
- **Mixed-integer programming** via GLPK_MI, ECOS_BB, Gurobi, CPLEX,
  or HiGHS (`boolean = TRUE` or `integer = TRUE` in `Variable()`).
- **Parameter support** via `Parameter()` class and `EvalParams` reduction.
- **50+ atom classes** covering LP, QP, SOCP, SDP, exponential cone, and
  power cone problems.
- **DPP** (Disciplined Parameterized Programming) for efficient
  parameter re-solve with compilation caching.
- **DGP** (Disciplined Geometric Programming) via `psolve(prob, gp = TRUE)`.
- **DQCP** (Disciplined Quasiconvex Programming) via `psolve(prob, qcp = TRUE)`.
- **Complex variable support** via `Variable(n, complex = TRUE)`.
- **Warm-start support** for 6 solvers (OSQP, SCS, Gurobi, MOSEK, CLARABEL, HiGHS).
- **Matrix package interoperability** via `as_cvxr_expr()`. Matrix
  package objects (`dgCMatrix`, `dgeMatrix`, `dsCMatrix`, `ddiMatrix`,
  `sparseVector`) use S4 dispatch which preempts S7/S3, so they cannot
  be used directly with CVXR operators. Wrapping with `as_cvxr_expr()`
  converts them to CVXR `Constant` objects while preserving sparsity
  (unlike `as.matrix()` which densifies). Base R `matrix` and `numeric`
  objects work natively without wrapping.

## Breaking Changes from CVXR 1.x

### New solve interface

The primary solve function is now `psolve()`, which returns the optimal
value directly:

```{r, eval = FALSE}
library(CVXR)
x <- Variable(2)
prob <- Problem(Minimize(sum_squares(x)), list(x >= 1))
opt_val <- psolve(prob)       # returns optimal value directly
x_val <- value(x)             # extract variable value
prob_status <- status(prob)   # check status
```

The old `solve()` still works but returns a compatibility list:

```{r, eval = FALSE}
result <- solve(prob)
result$value       # optimal value
result$getValue(x) # variable value (deprecated)
result$status      # problem status
```

### API changes

| Old API | New API |
|---------|---------|
| `solve(problem)` | `psolve(problem)` |
| `result$getValue(x)` | `value(x)` |
| `result$value` | return value of `psolve()` |
| `result$status` | `status(problem)` |
| `result$getDualValue(con)` | `dual_value(con)` |
| `problem_status(prob)` | `status(prob)` |
| `problem_solution(prob)` | `solution(prob)` |
| `get_problem_data(prob, solver)` | `problem_data(prob, solver)` |

### Axis parameter changes

The `axis` parameter now uses R's `apply()` convention (1-based indexing):

| Old CVXR | New CVXR | Meaning |
|----------|----------|---------|
| `axis = 1` | `axis = 1` | Row-wise reduction (unchanged) |
| `axis = 2` | `axis = 2` | Column-wise reduction (unchanged) |
| `axis = NA` | `axis = NULL` | All entries |

Passing `axis = 0` now produces an informative error with migration guidance.

### PSD constraints

PSD constraints use `PSD(A - B)` instead of `A %>>% B` (though `%>>%`
and `%<<%` operators are still available for backward compatibility).

### Solver changes

- **Removed**: CBC
- **Added**: HiGHS (LP, QP, MILP), Gurobi (LP, QP, SOCP, MIP), CVXOPT
  (LP, SOCP), PIQP (QP)
- **Default solver**: CLARABEL (replaces ECOS)

### Supported solvers

| Solver | R Package | Type | Problem Classes |
|--------|-----------|------|-----------------|
| CLARABEL | `clarabel` (>= 0.11.2) | Conic | LP, QP, SOCP, SDP, ExpCone, PowCone |
| SCS | `scs` (>= 3.2.7) | Conic | LP, QP, SOCP, SDP, ExpCone, PowCone |
| MOSEK | `Rmosek` (>= 11.1.1) | Conic | LP, QP, SOCP, SDP, ExpCone, PowCone |
| ECOS | `ECOSolveR` (>= 0.6) | Conic | LP, SOCP, ExpCone |
| ECOS_BB | `ECOSolveR` (>= 0.6) | Conic | LP, SOCP, ExpCone + MI |
| GUROBI | `gurobi` (>= 13.0.1) | Conic/QP | LP, QP, SOCP, MI |
| GLPK | `Rglpk` (>= 0.6.5.1) | Conic | LP |
| GLPK_MI | `Rglpk` (>= 0.6.5.1) | Conic | LP, MILP |
| HIGHS | `highs` (>= 1.12.0.3) | Conic/QP | LP, QP, MILP |
| CVXOPT | `cccp` (>= 0.3.3) | Conic | LP, SOCP |
| OSQP | `osqp` (>= 1.0.0) | QP | LP, QP |
| CPLEX | `Rcplex` (>= 0.3.8) | QP | LP, QP, MI |
| PIQP | `piqp` (>= 0.6.2) | QP | LP, QP |

## New Atoms and Functions

### Convenience atoms

| Function | Description |
|----------|-------------|
| `ptp(x, axis, keepdims)` | Peak-to-peak (range): `max(x) - min(x)` |
| `cvxr_mean(x, axis, keepdims)` | Arithmetic mean along an axis |
| `cvxr_std(x, axis, keepdims, ddof)` | Standard deviation |
| `cvxr_var(x, axis, keepdims, ddof)` | Variance |
| `vdot(x, y)` | Vector dot product (inner product) |
| `cvxr_outer(x, y)` | Outer product of two vectors |
| `inv_prod(x)` | Reciprocal of product of entries |
| `loggamma(x)` | Elementwise log of gamma function |
| `log_normcdf(x)` | Elementwise log of standard normal CDF |
| `cummax_expr(x, axis)` | Cumulative maximum along an axis |
| `dotsort(X, W)` | Weighted sorted dot product |

### Boolean logic atoms

For mixed-integer programming: `Not()`, `And()`, `Or()`, `Xor()`,
`implies()`, `iff()`.

### Other new atoms

- `perspective(f, s)` for perspective functions
- `FiniteSet(expr, values)` constraint for discrete optimization
- `ceil_expr()`, `floor_expr()` for DQCP problems

## Backward-Compatibility Aliases

- `tv()` is an alias for `total_variation()`
- `norm2(x)` is an alias for `p_norm(x, 2)`
- `multiply(x, y)` is an alias for elementwise multiplication
- `installed_solvers()` lists available solver packages
- Old `solve()` still works and returns a compatibility list
- Old function names (`problem_status`, `getValue`, etc.) still work
  but emit once-per-session deprecation warnings

## Migration Guide

To migrate code from CVXR 1.x to 1.8.x:

1. Replace `result <- solve(problem)` with `opt_val <- psolve(problem)`
2. Replace `result$getValue(x)` with `value(x)`
3. Replace `result$value` with the return value from `psolve()`
4. Replace `result$status` with `status(problem)`
5. Replace `result$getDualValue(con)` with `dual_value(con)`
6. Replace `add_to_solver_blacklist()` calls (no longer needed)
7. Update solver names: `"ECOS"` → `"CLARABEL"`, `"GLPK"` → `"HIGHS"`
8. Update `axis` arguments: `axis = NA` → `axis = NULL` (row/column axis values 1 and 2 are unchanged)
9. Replace `A %>>% B` with `PSD(A - B)` if desired
10. Wrap Matrix package objects with `as_cvxr_expr()` before using
    them in CVXR expressions (e.g., `as_cvxr_expr(A) %*% x` instead
    of `A %*% x` when `A` is a `dgCMatrix` or other Matrix class).
    This preserves sparsity. Base R matrices need no wrapping.

## Known Limitations

- Matrix package objects (`dgCMatrix`, `dgeMatrix`, `ddiMatrix`,
  `sparseVector`) cannot be used directly with CVXR operators due to
  S4 dispatch preempting S7/S3. Wrap with `as_cvxr_expr()` first.
  Base R `matrix`/`numeric` work natively. This is an R dispatch
  limitation requiring upstream changes in S7 and/or Matrix.
- Warm-start for HiGHS is blocked (R `highs` package lacks
  `setSolution()` API).
- Derivative/sensitivity API deferred (requires `diffcp`, no R
  equivalent).

## References
