---
title: Convex Optimization in R
author: Anqi Fu and Balasubramanian Narasimhan
date: '2017-10-30'
slug: cvxr
bibliography: bibtex/cvxr_refs.bib
type: homepage
---

```{r, echo = FALSE}
## Check pre-requisite packages for generating site
required_pkgs <- c(
    "here",
    "testthat",
    "CVXR",
    "reticulate",
    "lpSolveAPI",
    "Rglpk",
    "ElemStatLearn",
    "kableExtra",
    "stringr",
    "ggplot2",
    "devtools",
    "l1tf",
    "glmnet",
    "boot",
    "tidyr",
    "profvis",
    "ggforce",
    "nnls",
    "markovchain",
    "quantreg",
    "isotone",
    "MASS",
    "RColorBrewer",
    "grid",
    "Matrix",
    "expm",
    "dplyr",
    "survey")
missing_pkgs <- setdiff(required_pkgs, row.names(installed.packages()))
if (length(missing_pkgs) > 0) {
    print(missing_pkgs)
    stop("Missing above packages; install them.")
}
```

## What is `CVXR`?

`CVXR` is an R package that provides an object-oriented modeling
language for convex optimization, similar to `CVX`, `CVXPY`, `YALMIP`,
and `Convex.jl`. It allows the user to formulate convex optimization
problems in a natural mathematical syntax rather than the restrictive
standard form required by most solvers. The user specifies an
objective and set of constraints by combining constants, variables,
and parameters using a library of functions with known mathematical
properties. `CVXR` then applies
signed [disciplined convex programming (DCP)](/post/cvxr_dcp) to
verify the problem’s convexity. Once verified, the problem is
converted into standard form using graph implementations and
passed to a quadratic solver such as [OSQP](https://osqp.org/), or a cone solver such
as [ECOS](https://github.com/embotech/ecos)
or [SCS](https://github.com/cvxgrp/scs), or even commericial solvers,
if installed.

## Citing `CVXR`

`CVXR` is published in @cvxr2020
<[doi:10.18637/jss.v094.i14](https://dx.doi.org/10.18637/jss.v094.i14)>. A
BibTeX citation entry using citation key `cvxr2020` is below.

```
@Article{cvxr2020,
    title = {{CVXR}: An {R} Package for Disciplined Convex Optimization},
    author = {Anqi Fu and Balasubramanian Narasimhan and Stephen Boyd},
    journal = {Journal of Statistical Software},
    year = {2020},
    volume = {94},
    number = {14},
    pages = {1--34},
    doi = {10.18637/jss.v094.i14},
  }
```

## Supported Solvers

```{r, echo = FALSE}
`%>%`  <- magrittr::`%>%`
check <- "\U2713"
d  <- tibble::tibble(
                  Solver = c("CBC", "GLPK", "GLPK_MI", "OSQP", "CPLEX", "ECOS", "ECOS_BB", "GUROBI", "MOSEK", "SCS"),
                  `R package` = c("[rcbc](https://github.com/dirkschumacher/rcbc)",
                                  "[`Rglpk`](https://cran.r-project.org/package=Rglpk)",
                                  "[`Rglpk`](https://cran.r-project.org/package=Rglpk)",
                                  "[osqp](https://www.osqp.org)",
                                  "[Rcplex](https://cran.r-project.org/package=Rcplex)",
                                  "[ECOSolveR](https://cran.r-project.org/package=ECOSolveR)",
                                  "[ECOSolveR](https://cran.r-project.org/package=ECOSolveR)",
                                  "[gurobi](https://www.gurobi.com/documentation/9.0/refman/ins_the_r_package.html)",
                                  "[Rmosek](https://cran.r-project.org/package=Rmosek)",
                                  "[scs](https://cran.r-project.org/package=scs)"),
                  LP = c(check, check, check, check, check, check, " " , check, check, check),
                  QP = c(" " , " " , " " , check, check, check, " " , check, check, check),
                  SOCP = c(" " , " " , " " , " " , check, check, " " , check, check, check),
                  SDP = c(" " , " " , " " , " " , " " , " " , " " , " " , check, check),
                  EXP = c(" " , " " , " " , " " , " " , " " , " " , " " , check, check),
                  MIP = c(check, " " , check, " " , " " , " " , check, check, check, check)
              )
knitr::kable(d, format = "html") %>%
    kableExtra::kable_styling("striped") %>%
    kableExtra::column_spec(1:8, background = "#ececec")
                  
```

For more detail, see [Using other solvers](/cvxr_examples/cvxr_using-other-solvers/).

## News

For the latest news, please see the [Package
Documentation](https://www.cvxgrp.org/CVXR). 

`CVXR` continues to be developed on
[Github](https://github.com/cvxgrp/CVXR). Version 1.0 was a major
release implementing the reductions framework described in
@rewriting:2018. More details of all the changes can be found in
[What's new in 1.0](/cvxr_examples/whats_new_1.0/).
  

## Installing `CVXR`

`CVXR` is installed like any other R package from
[CRAN](https://cloud.r-project.org).

## Documentation

Two videos provide a good starting point:

- [Anqi Fu's talk](https://www.youtube.com/watch?v=MyglbtnmQ8A) at
 [useR!  2018](http://user2018.r-project.org)
- [Anqi's
talk](http://channel9.msdn.com/Events/useR-international-R-User-conference/useR2016/CVXR-An-R-Package-for-Modeling-Convex-Optimization-Problems)
at [useR!  2016](http://user2016.r-project.org) introducing `CVXR`. 

Materials form our [useR!
2019](http://user2019.r-project.org/tutorials/) including exercises
and solutions are available as a [bookdown
project](https://github.com/bnaras/cvxr_tutorial). Follow the
instructions in the README.

An introductory vignette is installed with the package and can be
viewed using `vignette("cvxr_intro", package = "CVXR")`.

A large set of examples, many ported from `CVXPY`, are available on
this site in the [Examples](/post/cvxr_examples/) section.

## Acknowledgements

We are grateful to [Steven
Diamond](https://web.stanford.edu/~stevend2/), [John
Miller](https://people.eecs.berkeley.edu/~miller_john/), and Paul
Kunsberg Rosenfield for their contributions to the software’s
development. In particular, we are indebted to Steven Diamond for his
work on [CVXPY](http://www.cvxpy.org/en/latest/). Most of `CVXR`'s
code, documentation, and examples were ported from his Python library.

## About this site

Much of the documentation on this site was modified
from [`CVXPY`](https://cvxpy.org) in a semi-automated way. 

This site was constructed
using [R markdown documents](http://rmarkdown.rstudio.com/), the
wonderful [blogdown](https://cloud.r-package.org/package="blogdown")
package by [Yihui Xie](https://yihui.name/), and
the
[Hugo Alabaster theme](https://themes.gohugo.io/theme/hugo-alabaster-theme/). The
pages are served out of
the [CVXR docs](https://github.com/bnaras/cvxr_docs) repository.

## References
