---
title: "Lasso Regression"
author: "Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: ml_lasso
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
```

## Introduction

Lasso regression is, like ridge regression, a **shrinkage** method. It
differs from ridge regression in its choice of penalty: lasso imposes
an $\ell_1$ **penalty** on the parameters $\beta$. That is, lasso
finds an assignment to $\beta$ that minimizes the function

$$
f(\beta) = \|X\beta - Y\|_2^2 + \lambda \|\beta\|_1,
$$

where $\lambda$ is a hyperparameter and, as usual, $X$ is the
training data and $Y$ the observations. The $\ell_1$ penalty
encourages **sparsity** in the learned parameters, and, as we will
see, can drive many coefficients to zero. In this sense, lasso is a
continuous **feature selection** method.

In this example, we show how to fit a lasso model using `CVXR`, how
to evaluate the model, and how to tune the hyperparameter $\lambda$.

## Writing the Objective Function

We can decompose the **objective function** as the sum of a **least
squares loss function** and an $\ell_1$ **regularizer**.

In `CVXR`, we express the loss as `sum_squares(X %*% beta - Y)` (or
equivalently `p_norm(X %*% beta - Y, 2)^2`) and the regularizer as
`p_norm(beta, 1)`. We also define a helper to compute the mean
squared error (MSE) on any dataset.

```{r}
loss_fn <- function(X, Y, beta) sum_squares(X %*% beta - Y)
regularizer <- function(beta) p_norm(beta, 1)
objective_fn <- function(X, Y, beta, lambd) {
    loss_fn(X, Y, beta) + lambd * regularizer(beta)
}
mse <- function(X, Y, beta) {
    (1.0 / nrow(X)) * value(loss_fn(X, Y, beta))
}
```

## Generating Data

We generate training examples and observations that are linearly
related; we make the relationship *sparse*, and we will see how lasso
approximately recovers it.

```{r}
set.seed(1)
m <- 100
n <- 20
sigma <- 5
density <- 0.2

## Generate sparse true coefficients
beta_star <- rnorm(n)
idxs <- sample.int(n, size = as.integer((1 - density) * n), replace = FALSE)
beta_star[idxs] <- 0

## Generate data matrix and observations
X <- matrix(rnorm(m * n), nrow = m, ncol = n)
Y <- X %*% beta_star + rnorm(m, sd = sigma)

## Split into train and test sets
X_train <- X[1:50, ]
Y_train <- Y[1:50]
X_test  <- X[51:100, ]
Y_test  <- Y[51:100]
```

## Fitting the Model

All we need to do to fit the model is create a `CVXR` problem where
the objective is to minimize the objective function defined above. We
solve this for many values of $\lambda$ to obtain the regularization
path.

```{r}
beta <- Variable(n)

lambd_values <- 10^seq(-2, 3, length.out = 50)
train_errors <- numeric(length(lambd_values))
test_errors  <- numeric(length(lambd_values))
beta_values  <- vector("list", length(lambd_values))

for (i in seq_along(lambd_values)) {
    problem <- Problem(Minimize(objective_fn(X_train, Y_train, beta,
                                             lambd_values[i])))
    result <- psolve(problem)
    train_errors[i] <- mse(X_train, Y_train, beta)
    test_errors[i]  <- mse(X_test, Y_test, beta)
    beta_values[[i]] <- value(beta)
}
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(train_errors = train_errors,
                 test_errors = test_errors,
                 beta_values = beta_values),
            file = file.path(testdata_dir, "result1.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    saved <- readRDS(file = file.path(testdata_dir, "result1.RDS"))
    expect_equal(train_errors, saved$train_errors, tolerance = 1e-5)
    expect_equal(test_errors, saved$test_errors, tolerance = 1e-5)
}
```

## Evaluating the Model

Just as we saw for ridge regression, regularization improves
generalizability. The plot below shows that, up to a point, increasing
$\lambda$ reduces test error while increasing training error---the
classic bias-variance trade-off.

```{r}
plot(lambd_values, train_errors, type = "l", col = "blue",
     log = "x", xlab = expression(lambda),
     ylab = "Mean Squared Error (MSE)",
     main = "Mean Squared Error (MSE)",
     ylim = range(c(train_errors, test_errors)))
lines(lambd_values, test_errors, col = "red")
legend("topleft", legend = c("Train error", "Test error"),
       col = c("blue", "red"), lty = 1)
```

## Regularization Path and Feature Selection

As $\lambda$ increases, the parameters are driven to $0$. By
$\lambda \approx 10$, approximately 80 percent of the coefficients
are *exactly* zero. This parallels the fact that $\beta^*$ was
generated such that 80 percent of its entries were zero. The features
corresponding to the slowest decaying coefficients can be interpreted
as the most important ones.

**Qualitatively, lasso differs from ridge in that the former often
drives parameters to exactly zero, whereas the latter shrinks
parameters but does not usually zero them out. That is, lasso results
in sparse models; ridge (usually) does not.**

```{r}
beta_mat <- do.call(cbind, beta_values)
matplot(lambd_values, t(beta_mat), type = "l", log = "x",
        xlab = expression(lambda), ylab = expression(beta[i]),
        main = "Regularization Path")
```

## Session Info

```{r}
sessionInfo()
```

## References
