---
title: "Logistic Regression with l1 Regularization"
author: "Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: ml_logistic
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
```

## Introduction

In this example, we use `CVXR` to train a logistic regression
classifier with $\ell_1$ regularization. We are given data
$(x_i, y_i)$, $i = 1, \ldots, m$. The $x_i \in \mathbf{R}^n$ are
feature vectors, while the $y_i \in \{0, 1\}$ are associated boolean
classes.

Our goal is to construct a linear classifier
$\hat{y} = \mathbb{1}[\beta^T x > 0]$, which is $1$ when
$\beta^T x$ is positive and $0$ otherwise. We model the posterior
probabilities of the classes given the data linearly, with

$$
\log \frac{\Pr(Y = 1 \mid X = x)}{\Pr(Y = 0 \mid X = x)} = \beta^T x.
$$

This implies that

$$
\Pr(Y = 1 \mid X = x) = \frac{\exp(\beta^T x)}{1 + \exp(\beta^T x)}, \quad
\Pr(Y = 0 \mid X = x) = \frac{1}{1 + \exp(\beta^T x)}.
$$

We fit $\beta$ by maximizing the log-likelihood of the data, plus a
regularization term $\lambda \|\beta\|_1$ with $\lambda > 0$:

$$
\ell(\beta) = \sum_{i=1}^{m} y_i \beta^T x_i - \log(1 + \exp(\beta^T x_i)) - \lambda \|\beta\|_1.
$$

Because $\ell$ is a concave function of $\beta$, this is a convex
optimization problem.

## Generating Data

In the following code we generate data with $n = 50$ features by
randomly choosing $x_i$ and supplying a sparse
$\beta_{\text{true}} \in \mathbf{R}^n$. We then set
$y_i = \mathbb{1}[\beta_{\text{true}}^T x_i + z_i > 0]$, where the
$z_i$ are i.i.d. normal random variables. We divide the data into
training and test sets with $m = 50$ examples each.

```{r}
set.seed(1)
n <- 50
m <- 50

sigmoid <- function(z) 1 / (1 + exp(-z))

beta_true <- c(1, 0.5, -0.5, rep(0, n - 3))
X <- (matrix(runif(m * n), nrow = m, ncol = n) - 0.5) * 10
Y <- round(sigmoid(X %*% beta_true + rnorm(m, sd = 0.5)))

X_test <- (matrix(runif(2 * m * n), nrow = 2 * m, ncol = n) - 0.5) * 10
Y_test <- round(sigmoid(X_test %*% beta_true + rnorm(2 * m, sd = 0.5)))
```

## Formulating the Problem

We next formulate the optimization problem using `CVXR`. The
`logistic` atom in `CVXR` computes $f(z) = \log(1 + e^z)$, which we
use to express the log-likelihood.

```{r}
beta <- Variable(n)
log_likelihood <- sum_entries(
    multiply(Y, X %*% beta) - logistic(X %*% beta)
)
```

## Fitting the Model

We solve the optimization problem for a range of $\lambda$ to compute
a trade-off curve. We then plot the train and test error over the
trade-off curve. A reasonable choice of $\lambda$ is the value that
minimizes the test error.

```{r}
## Classification error function
calc_error <- function(scores, labels) {
    preds <- ifelse(scores > 0, 1, 0)
    sum(abs(preds - labels)) / length(labels)
}
```

```{r}
trials <- 100
train_error <- numeric(trials)
test_error  <- numeric(trials)
lambda_vals <- 10^seq(-2, 0, length.out = trials)
beta_vals   <- vector("list", trials)

for (i in seq_len(trials)) {
    problem <- Problem(Maximize(log_likelihood / m -
                                lambda_vals[i] * p_norm(beta, 1)))
    result <- psolve(problem)
    beta_hat <- value(beta)
    train_error[i] <- calc_error(X %*% beta_hat, Y)
    test_error[i]  <- calc_error(X_test %*% beta_hat, Y_test)
    beta_vals[[i]] <- beta_hat
}
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(train_error = train_error,
                 test_error = test_error,
                 beta_vals = beta_vals),
            file = file.path(testdata_dir, "result1.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    saved <- readRDS(file = file.path(testdata_dir, "result1.RDS"))
    expect_equal(train_error, saved$train_error, tolerance = 1e-5)
    expect_equal(test_error, saved$test_error, tolerance = 1e-5)
}
```

```{r}
plot(lambda_vals, train_error, type = "l", col = "blue",
     log = "x", xlab = expression(lambda),
     ylab = "Classification Error",
     main = "Train vs. Test Error",
     ylim = range(c(train_error, test_error)))
lines(lambda_vals, test_error, col = "red")
legend("topleft", legend = c("Train error", "Test error"),
       col = c("blue", "red"), lty = 1)
```

## Regularization Path

We also plot the regularization path, or the $\beta_i$ versus
$\lambda$. Notice that a few features remain non-zero longer for
larger $\lambda$ than the rest, which suggests that these features
are the most important.

```{r}
beta_mat <- do.call(cbind, beta_vals)
matplot(lambda_vals, t(beta_mat), type = "l", log = "x",
        xlab = expression(lambda), ylab = expression(beta[i]),
        main = "Regularization Path")
```

## Comparing True and Reconstructed Coefficients

We plot the true $\beta$ versus the reconstructed $\beta$, as chosen
to minimize error on the test set. The non-zero coefficients are
reconstructed with good accuracy. There are a few values in the
reconstructed $\beta$ that are non-zero but should be zero.

```{r}
idx <- which.min(test_error)
plot(beta_true, type = "l", col = "blue",
     xlab = "i", ylab = expression(beta[i]),
     main = "True vs. Reconstructed Coefficients",
     ylim = range(c(beta_true, beta_vals[[idx]])))
lines(beta_vals[[idx]], col = "red")
legend("topright", legend = c(expression("True " * beta),
                              expression("Reconstructed " * beta)),
       col = c("blue", "red"), lty = 1)
```

## Session Info

```{r}
sessionInfo()
```

## References
