---
title: "Ridge Regression"
author: "CVXPY Developers and Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: ml_ridge
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
library(ggplot2)
library(tidyr)
source(here::here("_check_status.R"))
```

## Introduction

Ridge regression is a regression technique that is quite similar to
unadorned least squares linear regression: simply adding an $\ell_2$
**penalty** on the parameters $\beta$ to the objective function for
linear regression yields the objective function for ridge regression.

Our goal is to find an assignment to $\beta$ that minimizes the
function

$$
f(\beta) = \|X\beta - Y\|_2^2 + \lambda \|\beta\|_2^2,
$$

where $\lambda$ is a hyperparameter and, as usual, $X$ is the
training data and $Y$ the observations. In practice, we tune
$\lambda$ until we find a model that generalizes well to the test
data.

Ridge regression is an example of a **shrinkage method**: compared to
least squares, it shrinks the parameter estimates in the hopes of
**reducing variance, improving prediction accuracy, and aiding
interpretation**.

In this example, we show how to fit a ridge regression model using
`CVXR`, how to evaluate the model, and how to tune the
hyperparameter $\lambda$.

## Writing the Objective Function

We can decompose the **objective function** as the sum of a **least
squares loss function** and an $\ell_2$ **regularizer**.

```{r}
loss_fn <- function(X, Y, beta) p_norm(X %*% beta - Y, 2)^2
regularizer <- function(beta) p_norm(beta, 2)^2
objective_fn <- function(X, Y, beta, lambd) {
    loss_fn(X, Y, beta) + lambd * regularizer(beta)
}
mse <- function(X, Y, beta) {
    (1.0 / nrow(X)) * value(loss_fn(X, Y, beta))
}
```

## Generating Data

Because ridge regression encourages the parameter estimates to be
small, it tends to lead to models with **less variance** than those
fit with vanilla linear regression. We generate a small dataset that
will illustrate this.

```{r}
set.seed(1)
m <- 100
n <- 20
sigma <- 5

## Generate true coefficients
beta_star <- rnorm(n)

## Generate an ill-conditioned data matrix
X <- matrix(rnorm(m * n), nrow = m, ncol = n)

## Corrupt the observations with additive Gaussian noise
Y <- X %*% beta_star + rnorm(m, sd = sigma)

## Split into train and test sets
X_train <- X[1:50, ]
Y_train <- Y[1:50]
X_test  <- X[51:100, ]
Y_test  <- Y[51:100]
```

## Fitting the Model

All we need to do to fit the model is create a `CVXR` problem where
the objective is to minimize the objective function defined above. We
solve this for many values of $\lambda$ to obtain the regularization
path.

```{r}
beta <- Variable(n)

lambd_values <- 10^seq(-2, 2, length.out = 50)
train_errors <- numeric(length(lambd_values))
test_errors  <- numeric(length(lambd_values))
beta_values  <- vector("list", length(lambd_values))

for (i in seq_along(lambd_values)) {
    problem <- Problem(Minimize(objective_fn(X_train, Y_train, beta,
                                             lambd_values[i])))
    result <- psolve(problem, solver = "CLARABEL")
    check_solver_status(problem)
    train_errors[i] <- mse(X_train, Y_train, beta)
    test_errors[i]  <- mse(X_test, Y_test, beta)
    beta_values[[i]] <- value(beta)
}
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(train_errors = train_errors,
                 test_errors = test_errors,
                 beta_values = beta_values),
            file = file.path(testdata_dir, "result1.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    saved <- readRDS(file = file.path(testdata_dir, "result1.RDS"))
    expect_equal(train_errors, saved$train_errors, tolerance = 1e-5)
    expect_equal(test_errors, saved$test_errors, tolerance = 1e-5)
}
```

## Evaluating the Model

Notice that, up to a point, penalizing the size of the parameters
reduces test error at the cost of increasing the training error,
trading off higher bias for lower variance; in other words, this
indicates that, for our example, a properly tuned ridge regression
**generalizes better** than a least squares linear regression.

```{r}
df_error <- data.frame(lambda = lambd_values,
                       Train = train_errors,
                       Test = test_errors) |>
    pivot_longer(-lambda, names_to = "Set", values_to = "MSE")
ggplot(df_error, aes(x = lambda, y = MSE, color = Set)) +
    geom_line() +
    scale_x_log10() +
    scale_color_manual(values = c(Train = "blue", Test = "red")) +
    labs(x = expression(lambda), y = "Mean Squared Error (MSE)",
         title = "Mean Squared Error (MSE)") +
    theme_minimal()
```

## Regularization Path

As expected, increasing $\lambda$ drives the parameters towards $0$.
In a real-world example, those parameters that approach zero slower
than others might correspond to the more **informative** features. It
is in this sense that ridge regression can be considered **model
selection**.

```{r}
beta_mat <- do.call(cbind, beta_values)
df_path <- data.frame(lambda = lambd_values, t(beta_mat)) |>
    pivot_longer(-lambda, names_to = "Coefficient", values_to = "Value")
ggplot(df_path, aes(x = lambda, y = Value, color = Coefficient)) +
    geom_line() +
    scale_x_log10() +
    labs(x = expression(lambda), y = expression(beta[i]),
         title = "Regularization Path") +
    theme_minimal() +
    guides(color = "none")
```

## Session Info

```{r, echo = FALSE}
sessionInfo()
```

## References
