---
title: "Support Vector Machine with l1 Regularization"
author: "CVXPY Developers and Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: ml_svm
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
library(ggplot2)
library(tidyr)
source(here::here("_check_status.R"))
```

## Introduction

In this example we use `CVXR` to train an SVM classifier with
$\ell_1$-regularization. We are given data $(x_i, y_i)$,
$i = 1, \ldots, m$. The $x_i \in \mathbf{R}^n$ are feature vectors,
while the $y_i \in \{\pm 1\}$ are associated boolean outcomes. Our
goal is to construct a good linear classifier
$\hat{y} = \text{sign}(\beta^T x - v)$. We find the parameters
$\beta, v$ by minimizing the (convex) function

$$
f(\beta, v) = \frac{1}{m} \sum_i \left(1 - y_i (\beta^T x_i - v)\right)_+ + \lambda \|\beta\|_1.
$$

The first term is the average hinge loss. The second term shrinks the
coefficients in $\beta$ and encourages sparsity. The scalar
$\lambda \geq 0$ is a (regularization) parameter. Minimizing
$f(\beta, v)$ simultaneously selects features and fits the classifier.

## Example

In the following code we generate data with $n = 20$ features by
randomly choosing $x_i$ and a sparse
$\beta_{\text{true}} \in \mathbf{R}^n$. We then set
$y_i = \text{sign}(\beta_{\text{true}}^T x_i - v_{\text{true}} - z_i)$,
where the $z_i$ are i.i.d. normal random variables. We divide the
data into training and test sets with $m = 1000$ examples each.

```{r}
set.seed(1)
n <- 20
m <- 1000
TEST <- m
DENSITY <- 0.2

beta_true <- rnorm(n)
idxs <- sample.int(n, size = as.integer((1 - DENSITY) * n), replace = FALSE)
beta_true[idxs] <- 0
offset <- 0
sigma <- 45

X <- matrix(rnorm(m * n, sd = 5), nrow = m, ncol = n)
Y <- as.vector(sign(X %*% beta_true + offset + rnorm(m, sd = sigma)))
X_test <- matrix(rnorm(TEST * n, sd = 5), nrow = TEST, ncol = n)
Y_test <- as.vector(sign(X_test %*% beta_true + offset + rnorm(TEST, sd = sigma)))
```

We next formulate the optimization problem using `CVXR`.

```{r}
beta  <- Variable(n)
v     <- Variable()
loss  <- sum_entries(pos(1 - multiply(Y, X %*% beta - v)))
reg   <- p_norm(beta, 1)
```

We solve the optimization problem for a range of $\lambda$ to compute
a trade-off curve. We then plot the train and test error over the
trade-off curve. A reasonable choice of $\lambda$ is the value that
minimizes the test error.

```{r}
TRIALS <- 100
train_error <- numeric(TRIALS)
test_error  <- numeric(TRIALS)
lambda_vals <- 10^seq(-2, 0, length.out = TRIALS)
beta_vals   <- vector("list", TRIALS)

for (i in seq_len(TRIALS)) {
    prob <- Problem(Minimize(loss / m + lambda_vals[i] * reg))
    result <- psolve(prob)
    check_solver_status(prob)
    beta_hat <- drop(value(beta))
    v_hat    <- drop(value(v))
    train_error[i] <- sum(sign(X %*% beta_true + offset) !=
                          sign(X %*% beta_hat - v_hat)) / m
    test_error[i]  <- sum(sign(X_test %*% beta_true + offset) !=
                          sign(X_test %*% beta_hat - v_hat)) / TEST
    beta_vals[[i]] <- beta_hat
}
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(train_error = train_error,
                 test_error = test_error,
                 beta_vals = beta_vals),
            file = file.path(testdata_dir, "result1.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    saved <- readRDS(file = file.path(testdata_dir, "result1.RDS"))
    expect_equal(train_error, saved$train_error, tolerance = 1e-5)
    expect_equal(test_error, saved$test_error, tolerance = 1e-5)
}
```

```{r}
df_error <- data.frame(lambda = lambda_vals,
                       Train = train_error,
                       Test = test_error) |>
    pivot_longer(-lambda, names_to = "Set", values_to = "Error")
ggplot(df_error, aes(x = lambda, y = Error, color = Set)) +
    geom_line() +
    scale_x_log10() +
    scale_color_manual(values = c(Train = "blue", Test = "red")) +
    labs(x = expression(lambda), y = "Classification Error",
         title = "Train vs. Test Error") +
    theme_minimal()
```

## Regularization Path

We also plot the regularization path, or the $\beta_i$ versus
$\lambda$. Notice that the $\beta_i$ do not necessarily decrease
monotonically as $\lambda$ increases. Four features remain non-zero
longer for larger $\lambda$ than the rest, which suggests that these
features are the most important. In fact $\beta_{\text{true}}$ had 4
non-zero values.

```{r}
beta_mat <- do.call(cbind, lapply(beta_vals, as.vector))
df_path <- data.frame(lambda = lambda_vals, t(beta_mat)) |>
    pivot_longer(-lambda, names_to = "Coefficient", values_to = "Value")
ggplot(df_path, aes(x = lambda, y = Value, color = Coefficient)) +
    geom_line() +
    scale_x_log10() +
    labs(x = expression(lambda), y = expression(beta[i]),
         title = "Regularization Path") +
    theme_minimal() +
    guides(color = "none")
```

## Session Info

```{r, echo = FALSE}
sessionInfo()
```

## References
