---
title: "Derivatives Fundamentals"
author: "Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: deriv_fundamentals
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
```

## Introduction

::: {.callout-note}
The derivative functionality (`backward`, `derivative`, `gradient`,
`delta`) is planned for a future release of `CVXR`. The code below
illustrates the intended API but is not yet executable.
:::

This example introduces the fundamentals of computing the derivative
of the solution map to optimization problems. The derivative can be
used for **sensitivity analysis**, to see how a solution would change
given small changes to the parameters, and to compute **gradients** of
scalar-valued functions of the solution.

We will consider a simple disciplined geometric program (DGP). The
geometric program under consideration is

$$
\begin{array}{ll}
\text{minimize} & 1/(xyz) \\
\text{subject to} & a(xy + xz + yz) \leq b \\
& x \geq y^c,
\end{array}
$$

where $x \in \mathbf{R}_{++}$, $y \in \mathbf{R}_{++}$, and
$z \in \mathbf{R}_{++}$ are the variables, and
$a \in \mathbf{R}_{++}$, $b \in \mathbf{R}_{++}$, and
$c \in \mathbf{R}$ are the parameters. The vector

$$
\alpha = \begin{bmatrix} a \\ b \\ c \end{bmatrix}
$$

is the vector of parameters.

## Setting Up the Problem

```{r}
x <- Variable(pos = TRUE)
y <- Variable(pos = TRUE)
z <- Variable(pos = TRUE)

a <- Parameter(pos = TRUE)
b <- Parameter(pos = TRUE)
cc <- Parameter()  # 'c' is reserved in R

objective_fn <- 1 / (x * y * z)
objective <- Minimize(objective_fn)
constraints <- list(a * (x * y + x * z + y * z) <= b, x >= y^cc)
problem <- Problem(objective, constraints)
```

We verify that the problem is a DGP.

```{r}
is_dgp(problem)
```

## Solving the Problem

Next, we solve the problem, setting the parameters $a$, $b$, and $c$
to $2$, $1$, and $0.5$.

```{r, eval = FALSE}
value(a) <- 2.0
value(b) <- 1.0
value(cc) <- 0.5
result <- psolve(problem, gp = TRUE, requires_grad = TRUE)

cat(sprintf("x = %f\n", value(x)))
cat(sprintf("y = %f\n", value(y)))
cat(sprintf("z = %f\n", value(z)))
```

Notice the keyword argument `requires_grad = TRUE`; this is necessary
to subsequently compute derivatives.

## Solution Map

The **solution map** of the above problem is a function

$$
\mathcal{S} : \mathbf{R}^2_{++} \times \mathbf{R} \to \mathbf{R}^3_{++}
$$

which maps the parameter vector to the vector of optimal solutions

$$
\mathcal{S}(\alpha) = \begin{bmatrix} x(\alpha) \\ y(\alpha) \\ z(\alpha) \end{bmatrix}.
$$

Here, $x(\alpha)$, $y(\alpha)$, and $z(\alpha)$ are the optimal
values of the variables corresponding to the parameter vector.

As an example, we just saw that

$$
\mathcal{S}((2.0, 1.0, 0.5)) \approx \begin{bmatrix} 0.5612 \\ 0.3150 \\ 0.3690 \end{bmatrix}.
$$

## Sensitivity Analysis

When the solution map is differentiable, we can use its derivative

$$
\mathsf{D}\mathcal{S}(\alpha) \in \mathbf{R}^{3 \times 3}
$$

to perform a **sensitivity analysis**, which studies how the solution
would change given small changes to the parameters.

Suppose we perturb the parameters by a vector of small magnitude
$\mathsf{d}\alpha \in \mathbf{R}^3$. We can approximate the change
$\Delta$ in the solution due to the perturbation using the derivative,
as

$$
\Delta = \mathcal{S}(\alpha + \mathsf{d}\alpha) - \mathcal{S}(\alpha) \approx \mathsf{D}\mathcal{S}(\alpha) \mathsf{d}\alpha.
$$

We can compute this in `CVXR` as follows. Partition the perturbation
as

$$
\mathsf{d}\alpha = \begin{bmatrix} \mathsf{d}a \\ \mathsf{d}b \\ \mathsf{d}c \end{bmatrix}.
$$

We set the `delta` attributes of the parameters to their
perturbations, and then call the `derivative` method.

```{r, eval = FALSE}
da <- 1e-2
db <- 1e-2
dc <- 1e-2

delta(a) <- da
delta(b) <- db
delta(cc) <- dc
derivative(problem)
```

The `derivative` method populates the `delta` attributes of the
variables as a side-effect, with the predicted change in the variable.
We can compare the predictions to the actual solution of the perturbed
problem.

```{r, eval = FALSE}
x_hat <- value(x) + delta(x)
y_hat <- value(y) + delta(y)
z_hat <- value(z) + delta(z)

value(a) <- value(a) + da
value(b) <- value(b) + db
value(cc) <- value(cc) + dc

psolve(problem, gp = TRUE)
cat(sprintf("x: predicted %.5f  actual %.5f\n", x_hat, value(x)))
cat(sprintf("y: predicted %.5f  actual %.5f\n", y_hat, value(y)))
cat(sprintf("z: predicted %.5f  actual %.5f\n", z_hat, value(z)))

## Restore original parameter values
value(a) <- value(a) - da
value(b) <- value(b) - db
value(cc) <- value(cc) - dc
```

In this case, the predictions and the actual solutions are fairly
close.

## Gradient

We can compute the gradient of a scalar-valued function of the
solution with respect to the parameters. Let
$f : \mathbf{R}^{3} \to \mathbf{R}$, and suppose we want to compute
the gradient of the composition $f \circ \mathcal{S}$. By the chain
rule,

$$
\nabla f(\mathcal{S}(\alpha)) = \mathsf{D}^T\mathcal{S}(\alpha) \begin{bmatrix} \mathsf{d}x \\ \mathsf{d}y \\ \mathsf{d}z \end{bmatrix},
$$

where $\mathsf{D}^T\mathcal{S}$ is the adjoint (or transpose) of the
derivative operator, and $\mathsf{d}x$, $\mathsf{d}y$, and
$\mathsf{d}z$ are the partial derivatives of $f$ with respect to its
arguments.

We can compute the gradient in `CVXR`, using the `backward` method.
As an example, suppose

$$
f(x, y, z) = \frac{1}{2}(x^2 + y^2 + z^2),
$$

so that $\mathsf{d}x = x$, $\mathsf{d}y = y$, and
$\mathsf{d}z = z$. Let
$\mathsf{d}\alpha = \nabla f(\mathcal{S}(\alpha))$, and suppose we
subtract $\eta \, \mathsf{d}\alpha$ from the parameter, where $\eta$
is a positive constant. Using the following code, we can compare
$f(\mathcal{S}(\alpha - \eta \, \mathsf{d}\alpha))$ with the value
predicted by the gradient,

$$
f(\mathcal{S}(\alpha - \eta \, \mathsf{d}\alpha)) \approx f(\mathcal{S}(\alpha)) - \eta \, \mathsf{d}\alpha^T \mathsf{d}\alpha.
$$

```{r, eval = FALSE}
result <- psolve(problem, gp = TRUE, requires_grad = TRUE)

f_val <- function(xv, yv, zv) 0.5 * (xv^2 + yv^2 + zv^2)

original <- f_val(value(x), value(y), value(z))

gradient(x) <- value(x)
gradient(y) <- value(y)
gradient(z) <- value(z)
backward(problem)

eta <- 0.5
da_grad <- gradient(a)
db_grad <- gradient(b)
dc_grad <- gradient(cc)
dalpha <- c(da_grad, db_grad, dc_grad)
predicted <- original - eta * sum(dalpha^2)

value(a) <- value(a) - eta * da_grad
value(b) <- value(b) - eta * db_grad
value(cc) <- value(cc) - eta * dc_grad
psolve(problem, gp = TRUE)
actual <- f_val(value(x), value(y), value(z))

cat(sprintf("original %.5f  predicted %.5f  actual %.5f\n",
            original, predicted, actual))
```

## Session Info

```{r}
sessionInfo()
```

## References
