---
title: "Queuing Design"
author: "CVXPY Developers and Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: deriv_queuing_design
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
```

## Introduction

::: {.callout-warning}
The derivative functionality (`backward`, `derivative`, `gradient`,
`delta`) is **not implemented** in `CVXR` at this time. The sensitivity
analysis and perturbation sections below illustrate the intended API
based on CVXPY but are not executable.
:::

In this example, we consider the design of an $M/M/N$ queuing system,
and we study the sensitivity of the design with respect to various
parameters, such as the maximum total delay and total service rate.
This example was described in the paper [Differentiating through
Log-Log Convex Programs](http://web.stanford.edu/~boyd/papers/pdf/diff_llcvx.pdf).

We consider the optimization of a (Markovian) queuing system, with
$N$ queues. A queuing system is a collection of queues, in which
queued items wait to be served; the queued items might be threads in
an operating system, or packets in an input or output buffer of a
networking system. A natural goal is to minimize the service load of
the system, given constraints on various properties of the queuing
system, such as limits on the maximum delay or latency. In this
example, we formulate this design problem as a log-log convex program
(LLCP), and compute the sensitivity of the design variables with
respect to the parameters.

## Problem Formulation

We assume that items arriving at the $i$th queue are generated by a
Poisson process with rate $\lambda_i$, and that the service times for
the $i$th queue follow an exponential distribution with parameter
$\mu_i$, for $i = 1, \ldots, N$. The service load of the queuing
system is a function
$\ell : \mathbf{R}^{N}_{++} \times \mathbf{R}^{N}_{++} \to \mathbf{R}^{N}_{++}$
of the arrival rate vector $\lambda$ and the service rate vector
$\mu$, with components

$$
\ell_i(\lambda, \mu) = \frac{\mu_i}{\lambda_i}, \quad i = 1, \ldots, N.
$$

(This is the reciprocal of the traffic load, which is usually denoted
by $\rho$.) Similarly, the queue occupancy, the average delay, and
the total delay of the system are (respectively) functions $q$, $w$,
and $d$ of $\lambda$ and $\mu$, with components

$$
q_i(\lambda, \mu) = \frac{\ell_i(\lambda, \mu)^{-2}}{1 - \ell_i(\lambda, \mu)^{-1}}, \quad
w_i(\lambda, \mu) = \frac{q_i(\lambda, \mu)}{\lambda_i} + \frac{1}{\mu_i}, \quad
d_i(\lambda, \mu) = \frac{1}{\mu_i - \lambda_i}.
$$

The queuing system has limits on the queue occupancy, average queuing
delay, and total delay, which must satisfy

$$
q(\lambda, \mu) \leq q_{\max}, \quad w(\lambda, \mu) \leq w_{\max}, \quad d(\lambda, \mu) \leq d_{\max},
$$

where $q_{\max}$, $w_{\max}$, and $d_{\max} \in \mathbf{R}^{N}_{++}$
are parameters and the inequalities are meant elementwise.
Additionally, the arrival rate vector $\lambda$ must be at least
$\lambda_{\min} \in \mathbf{R}^{N}_{++}$, and the sum of the service
rates must be no greater than $\mu_{\max} \in \mathbf{R}_{++}$.

Our design problem is to choose the arrival rates and service times to
minimize a weighted sum of the service loads,
$\gamma^T \ell(\lambda, \mu)$, where
$\gamma \in \mathbf{R}^{N}_{++}$ is the weight vector, while
satisfying the constraints. The problem is

$$
\begin{array}{ll}
\text{minimize} & \gamma^T \ell(\lambda, \mu) \\
\text{subject to} & q(\lambda, \mu) \leq q_{\max} \\
& w(\lambda, \mu) \leq w_{\max} \\
& d(\lambda, \mu) \leq d_{\max} \\
& \lambda \geq \lambda_{\min}, \quad \sum_{i=1}^{N} \mu_i \leq \mu_{\max}.
\end{array}
$$

Here, $\lambda, \mu \in \mathbf{R}^{N}_{++}$ are the variables and
$\gamma, q_{\max}, w_{\max}, d_{\max}, \lambda_{\min} \in \mathbf{R}^{N}_{++}$
and $\mu_{\max} \in \mathbf{R}_{++}$ are the parameters. This problem
is an LLCP.

## Setting Up the Problem

```{r}
N <- 2

mu  <- Variable(N, pos = TRUE, name = "mu")
lam <- Variable(N, pos = TRUE, name = "lambda")
ell <- Variable(N, pos = TRUE, name = "ell")

w_max   <- Parameter(N, pos = TRUE, name = "w_max",   value = c(2.5, 3.0))
d_max   <- Parameter(N, pos = TRUE, name = "d_max",   value = c(2.0, 2.0))
q_max   <- Parameter(N, pos = TRUE, name = "q_max",   value = c(4.0, 5.0))
lam_min <- Parameter(N, pos = TRUE, name = "lam_min", value = c(0.5, 0.8))
mu_max  <- Parameter(1, pos = TRUE, name = "mu_max",  value = 3.0)
gamma   <- Parameter(N, pos = TRUE, name = "gamma",   value = c(1.0, 2.0))

## Queuing system functions
lq <- (ell)^(-2) / one_minus_pos(ell^(-1))
q  <- lq
w  <- lq / lam + 1 / mu
d  <- 1 / diff_pos(mu, lam)

constraints <- list(
    w <= w_max,
    d <= d_max,
    q <= q_max,
    lam >= lam_min,
    sum_entries(mu) <= mu_max,
    ell == mu / lam
)

objective_fn <- t(gamma) %*% ell

problem <- Problem(Minimize(objective_fn), constraints)
```

## Solving the Problem

```{r, eval = FALSE}
result <- psolve(problem, requires_grad = TRUE, gp = TRUE,
                 eps = 1e-14, max_iters = 10000, mode = "dense")
cat(sprintf("Optimal value: %f\n", result))
```

The solution is printed below.

```{r, eval = FALSE}
cat(sprintf("mu:  [%f, %f]\n", value(mu)[1], value(mu)[2]))
cat(sprintf("lam: [%f, %f]\n", value(lam)[1], value(lam)[2]))
cat(sprintf("ell: [%f, %f]\n", value(ell)[1], value(ell)[2]))
```

## Sensitivities

We compute the derivative of each variable with respect to the
parameters. One takeaway is that the solution is highly sensitive to
the values of $d_{\max}$ and $\mu_{\max}$, and that increasing these
parameters would decrease the service loads, especially on the first
queue.

```{r, eval = FALSE}
result <- psolve(problem, requires_grad = TRUE, gp = TRUE,
                 eps = 1e-14, max_iters = 10000, mode = "dense")

params_list <- list(gamma = gamma, w_max = w_max, d_max = d_max,
                    q_max = q_max, lam_min = lam_min, mu_max = mu_max)

for (var in list(lam, mu, ell)) {
    cat(sprintf("Variable: %s\n", name(var)))

    cat("Gradient with respect to first component\n")
    gradient(var) <- c(1.0, 0.0)
    backward(problem)
    for (nm in names(params_list)) {
        p <- params_list[[nm]]
        g <- gradient(p)
        if (length(g) == 2) {
            cat(sprintf("  %s: %.3g, %.3g\n", nm, g[1], g[2]))
        } else {
            cat(sprintf("  %s: %.3g\n", nm, g))
        }
    }

    cat("Gradient with respect to second component\n")
    gradient(var) <- c(0.0, 1.0)
    backward(problem)
    for (nm in names(params_list)) {
        p <- params_list[[nm]]
        g <- gradient(p)
        if (length(g) == 2) {
            cat(sprintf("  %s: %.3g, %.3g\n", nm, g[1], g[2]))
        } else {
            cat(sprintf("  %s: %.3g\n", nm, g))
        }
    }

    gradient(var) <- c(0.0, 0.0)
    cat("\n")
}
```

## Perturbation Analysis

Next, we perturb each parameter by a small amount, and compare the
prediction of a first-order approximation to the solution of the
perturbed problem to the true solution.

```{r, eval = FALSE}
result <- psolve(problem, requires_grad = TRUE, gp = TRUE,
                 eps = 1e-14, max_iters = 10000, mode = "dense")

mu_value  <- value(mu)
lam_value <- value(lam)

delta_frac <- 0.01
for (nm in names(params_list)) {
    p <- params_list[[nm]]
    delta(p) <- value(p) * delta_frac
}

derivative(problem)

lam_pred <- (delta(lam) / lam_value) * 100
mu_pred  <- (delta(mu) / mu_value) * 100

cat(sprintf("lam predicted (percent change): [%.4f, %.4f]\n", lam_pred[1], lam_pred[2]))
cat(sprintf("mu  predicted (percent change): [%.4f, %.4f]\n", mu_pred[1], mu_pred[2]))

## Now solve the perturbed problem
for (nm in names(params_list)) {
    p <- params_list[[nm]]
    value(p) <- value(p) + delta(p)
}
psolve(problem, gp = TRUE, eps = 1e-14, max_iters = 10000)

lam_actual <- ((value(lam) - lam_value) / lam_value) * 100
mu_actual  <- ((value(mu) - mu_value) / mu_value) * 100

cat(sprintf("lam actual    (percent change): [%.4f, %.4f]\n", lam_actual[1], lam_actual[2]))
cat(sprintf("mu  actual    (percent change): [%.4f, %.4f]\n", mu_actual[1], mu_actual[2]))
```

## Session Info

```{r, echo = FALSE}
sessionInfo()
```

## References
