---
title: "Structured Prediction"
author: "CVXPY Developers and Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: deriv_structured_prediction
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
```

## Introduction

::: {.callout-warning}
The derivative functionality (`backward`, `derivative`, `gradient`,
`delta`) is **not implemented** in `CVXR` at this time. The gradient
descent training loop below illustrates the intended API based on CVXPY
but is not executable.
:::

In this example, we fit a regression model to structured data, using a
log-log convex program (LLCP). The training dataset $\mathcal{D}$
contains $N$ input-output pairs $(x, y)$, where
$x \in \mathbf{R}^{n}_{++}$ is an input and
$y \in \mathbf{R}^{m}_{++}$ is an output. The entries of each output
$y$ are sorted in ascending order, meaning
$y_1 \leq y_2 \leq \cdots \leq y_m$.

Our regression model
$\phi : \mathbf{R}^{n}_{++} \to \mathbf{R}^{m}_{++}$ takes as input
a vector $x \in \mathbf{R}^{n}_{++}$, and solves an LLCP to produce a
prediction $\hat{y} \in \mathbf{R}^{m}_{++}$. In particular, the
solution of the LLCP is the model's prediction. The model is of the
form

$$
\begin{array}{lll}
\phi(x) = &
\text{argmin} & \mathbf{1}^T (z/y + y/z) \\
& \text{subject to} & y_i \leq y_{i+1}, \quad i = 1, \ldots, m-1 \\
&& z_i = c_i x_1^{A_{i1}} x_2^{A_{i2}} \cdots x_n^{A_{in}}, \quad i = 1, \ldots, m.
\end{array}
$$

Here, the minimization is over $y \in \mathbf{R}^{m}_{++}$ and an
auxiliary variable $z \in \mathbf{R}^{m}_{++}$, $\phi(x)$ is the
optimal value of $y$, and the parameters are
$c \in \mathbf{R}^{m}_{++}$ and $A \in \mathbf{R}^{m \times n}$. The
ratios in the objective are meant elementwise. Given a vector $x$,
this model finds a sorted vector $\hat{y}$ whose entries are close to
monomial functions of $x$ (which are the entries of $z$), as measured
by the fractional error.

The training loss $\mathcal{L}(\phi)$ of the model on the training
set is the mean squared loss

$$
\mathcal{L}(\phi) = \frac{1}{N} \sum_{(x, y) \in \mathcal{D}} \|y - \phi(x)\|_2^2.
$$

We fit the parameters $c$ and $A$ by an iterative projected gradient
descent method on $\mathcal{L}(\phi)$. In each iteration, we first
compute predictions $\phi(x)$ for each input in the training set;
this requires solving $N$ LLCPs. Next, we evaluate the training loss
$\mathcal{L}(\phi)$. To update the parameters, we compute the
gradient $\nabla \mathcal{L}(\phi)$ of the training loss with respect
to the parameters $c$ and $A$. This requires differentiating through
the solution map of the LLCP.

This example is described in the paper [Differentiating through
Log-Log Convex Programs](http://web.stanford.edu/~boyd/papers/pdf/diff_llcvx.pdf).

## Data Generation

We generate synthetic data by choosing a true $A$ and $c$, generating
lognormal inputs, and solving the LLCP with the true parameters to
produce sorted outputs.

```{r}
n <- 20   # input dimension
m <- 10   # output dimension
N <- 100  # number of training pairs
N_val <- 50  # number of validation pairs

set.seed(243)

## True parameters
A_true <- matrix(rnorm(m * n), nrow = m, ncol = n) / 10
c_true <- abs(rnorm(m))
```

```{r, eval = FALSE}
generate_data <- function(num_points, seed) {
    set.seed(seed)

    ## Generate lognormal inputs
    inputs <- matrix(exp(rnorm(num_points * n)), nrow = num_points, ncol = n)

    ## For each input, solve the LLCP with true parameters to get output
    input_p <- Parameter(n, pos = TRUE)
    prediction <- multiply(c_true, gmatmul(A_true, input_p))
    y_var <- Variable(m, pos = TRUE)
    obj_fn <- sum_entries(prediction / y_var + y_var / prediction)
    cons <- lapply(seq_len(m - 1), function(i) y_var[i] <= y_var[i + 1])
    prob <- Problem(Minimize(obj_fn), cons)

    outputs <- matrix(0, nrow = num_points, ncol = m)
    for (i in seq_len(num_points)) {
        value(input_p) <- inputs[i, ]
        psolve(prob, gp = TRUE)
        outputs[i, ] <- value(y_var)
    }
    list(inputs = inputs, outputs = outputs)
}
```

```{r, eval = FALSE}
train_data <- generate_data(N, 243)
train_inputs  <- train_data$inputs
train_outputs <- train_data$outputs
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(train_outputs_1 = train_outputs[1, ]),
            file = file.path(testdata_dir, "result1.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    saved <- readRDS(file = file.path(testdata_dir, "result1.RDS"))
    expect_equal(train_outputs[1, ], saved$train_outputs_1, tolerance = 1e-4)
}
```

```{r, eval = FALSE}
plot(train_outputs[1, ], type = "o", pch = 16,
     xlab = "i", ylab = expression(y[i]),
     main = "First Training Output (sorted)")
```

```{r, eval = FALSE}
val_data <- generate_data(N_val, 0)
val_inputs  <- val_data$inputs
val_outputs <- val_data$outputs
```

## Monomial Fit to Each Component

We initialize the parameters in our LLCP model by fitting monomials
to the training data, without enforcing the monotonicity constraint.
This is a standard least squares problem in log space.

```{r, eval = FALSE}
log_outputs <- t(log(train_outputs))  # m x N
log_inputs  <- t(log(train_inputs))   # n x N

## Solve: log_outputs ≈ theta^T %*% log_inputs + log_c %*% ones^T
## Variables: theta (n x m), log_c (m x 1)
theta <- Variable(c(n, m))
log_c <- Variable(c(m, 1))
offsets <- do.call(cbind, replicate(N, log_c, simplify = FALSE))  # m x N
cp_preds <- t(theta) %*% log_inputs + offsets   # m x N
lstsq_obj <- (1 / N) * sum_squares(cp_preds - log_outputs)
lstsq_problem <- Problem(Minimize(lstsq_obj))

lstsq_result <- psolve(lstsq_problem, verbose = TRUE)
cat(sprintf("Least squares fit: %f\n", lstsq_result))
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(lstsq_result = lstsq_result),
            file = file.path(testdata_dir, "result2.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    saved <- readRDS(file = file.path(testdata_dir, "result2.RDS"))
    expect_equal(lstsq_result, saved$lstsq_result, tolerance = 1e-3)
}
```

```{r, eval = FALSE}
## Extract initialized parameters
A_init <- t(value(theta))    # m x n
c_init <- exp(value(log_c))  # m x 1
```

## Fitting with the LLCP Model

We now set up the parameterized LLCP and iteratively fit the
parameters using gradient descent with the `backward` method.

```{r, eval = FALSE}
A_param <- Parameter(c(m, n))
c_param <- Parameter(m, pos = TRUE)
x_slack <- Variable(n, pos = TRUE)
x_param <- Parameter(n, pos = TRUE)
y <- Variable(m, pos = TRUE)

prediction <- multiply(c_param, gmatmul(A_param, x_slack))
obj_fn <- sum_entries(prediction / y + y / prediction)
cons <- c(list(x_slack == x_param),
          lapply(seq_len(m - 1), function(i) y[i] <= y[i + 1]))
problem <- Problem(Minimize(obj_fn), cons)

cat(sprintf("Is DGP: %s\n", is_dgp(problem)))
```

We run projected gradient descent. In each epoch, for every training
example, we solve the LLCP, compute the gradient of the squared loss
with respect to $A$ and $c$ via `backward`, and accumulate the
gradients. We then update the parameters.

```{r, eval = FALSE}
## Initialize parameters from the monomial fit
A_current <- A_init
c_current <- as.vector(c_init)

lr <- 5e-2  # learning rate
n_epochs <- 10

train_losses <- numeric(n_epochs)
val_losses   <- numeric(n_epochs)

for (epoch in seq_len(n_epochs)) {
    ## -- Training pass --
    A_grad_accum <- matrix(0, m, n)
    c_grad_accum <- numeric(m)
    train_loss <- 0

    for (i in seq_len(N)) {
        value(A_param) <- A_current
        value(c_param) <- c_current
        value(x_param) <- train_inputs[i, ]

        psolve(problem, gp = TRUE, requires_grad = TRUE)
        y_pred <- value(y)
        residual <- y_pred - train_outputs[i, ]
        train_loss <- train_loss + sum(residual^2)

        ## Set gradient of y to d(loss)/d(y) = 2 * residual / N
        gradient(y) <- 2 * residual / N
        backward(problem)

        A_grad_accum <- A_grad_accum + gradient(A_param)
        c_grad_accum <- c_grad_accum + gradient(c_param)
    }
    train_losses[epoch] <- train_loss / N

    ## -- Validation pass (no gradients needed) --
    val_loss <- 0
    for (i in seq_len(N_val)) {
        value(A_param) <- A_current
        value(c_param) <- c_current
        value(x_param) <- val_inputs[i, ]
        psolve(problem, gp = TRUE)
        y_pred <- value(y)
        val_loss <- val_loss + sum((y_pred - val_outputs[i, ])^2)
    }
    val_losses[epoch] <- val_loss / N_val

    cat(sprintf("(epoch %d) train / val (%.4f / %.4f)\n",
                epoch, train_losses[epoch], val_losses[epoch]))

    ## -- Update parameters --
    A_current <- A_current - lr * A_grad_accum
    c_current <- c_current - lr * c_grad_accum
    ## Project c to stay positive
    c_current <- pmax(c_current, 1e-8)
}
```

## Comparing Predictions

We compare the LLCP model predictions (with fitted parameters) to the
true outputs on a validation example.

```{r, eval = FALSE}
## Compute LLCP predictions on a validation example
val_idx <- 1
value(A_param) <- A_current
value(c_param) <- c_current
value(x_param) <- val_inputs[val_idx, ]
psolve(problem, gp = TRUE)
llcp_pred <- value(y)

## Compute least-squares monomial predictions (no monotonicity)
lstsq_pred <- as.vector(c_init) *
    exp(A_init %*% log(val_inputs[val_idx, ]))

plot(val_outputs[val_idx, ], type = "o", pch = 16, col = "orange",
     xlab = "i", ylab = expression(y[i]),
     main = "Validation Example: True vs. Predictions",
     ylim = range(c(val_outputs[val_idx, ], llcp_pred, lstsq_pred)))
lines(llcp_pred, type = "o", pch = 17, col = "teal")
lines(lstsq_pred, type = "o", pch = 15, col = "gray", lty = 2)
legend("topleft",
       legend = c("True", "LLCP", "Least Squares"),
       col = c("orange", "teal", "gray"),
       pch = c(16, 17, 15), lty = c(1, 1, 2))
```

## Session Info

```{r, echo = FALSE}
sessionInfo()
```

## References
