---
title: DGP Fundamentals
author: "Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: dgp_fundamentals
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
```

## Introduction

This example will introduce you to the fundamentals of **disciplined
geometric programming** (**DGP**), which lets you formulate and solve
_log-log convex programs_ (LLCPs) in `CVXR`.

LLCPs are problems that become convex after the variables, objective
functions, and constraint functions are replaced with their logs, an
operation that we refer to as a log-log transformation. LLCPs
generalize geometric programming.

## 1. Log-Log Curvature

Just as every Expression in `CVXR` has a curvature (constant, affine,
convex, concave, or unknown), every Expression also has a log-log
curvature.

A function $f : D \subseteq \mathbf{R}^{n}_{++} \to \mathbf{R}$ is
said to be **log-log convex** if the function $F(u) = \log f(e^u)$,
with domain $\{u \in \mathbf{R}^n : e^u \in D\}$, is convex; the
function $F$ is called the log-log transformation of $f$. The function
$f$ is **log-log concave** if $F$ is concave, and it is **log-log
affine** if $F$ is affine.

Notice that if a function has log-log curvature, then its domain and
range can only include positive numbers.

For an Expression to have known log-log curvature, all of the
Constants, Variables, and Parameters it refers to must be elementwise
positive.

### Variables and Parameters

Variables and parameters must be positive, i.e., they must be
constructed with the option `pos = TRUE`:

```{r}
v <- Variable(pos = TRUE)
cat("Positive variable is log-log affine:", CVXR:::is_log_log_affine(v), "\n")

p <- Parameter(pos = TRUE)
cat("Positive parameter is log-log affine:", CVXR:::is_log_log_affine(p), "\n")
```

### Functions from Geometric Programming

A function $f(x)$ is log-log affine if and only if it is given by

$$
f(x) = cx_1^{a_1}x_2^{a_2} \ldots x_n^{a_n},
$$

where $c > 0$ and the $a_i$ are real numbers. In the context of
geometric programming, such a function is called a monomial.

```{r}
x <- Variable(3, pos = TRUE)
cc <- 2.0
a <- c(0.5, 2.0, 1.8)

monomial <- cc * power(x[1], a[1]) * power(x[2], a[2]) * power(x[3], a[3])
cat("Monomial is log-log affine:", CVXR:::is_log_log_affine(monomial), "\n")
```

A sum of monomial functions is log-log convex; in the context of
geometric programming, such a function is called a posynomial. There
are functions that are not posynomials that are still log-log convex.

```{r}
x <- Variable(pos = TRUE)
y <- Variable(pos = TRUE)

constant <- 2.0
monomial <- constant * x * y
posynomial <- monomial + power(x, 1.5) * power(y, -1)
reciprocal <- power(posynomial, -1)

cat("monomial is log-log affine:", CVXR:::is_log_log_affine(monomial), "\n")
cat("posynomial is log-log convex:", CVXR:::is_log_log_convex(posynomial), "\n")
cat("reciprocal is log-log concave:", CVXR:::is_log_log_concave(reciprocal), "\n")
```

## 2. Log-Log Curvature Ruleset

`CVXR` has a library of atomic functions with known log-log curvature
and monotonicity. It uses this information to tag every Expression
with a log-log curvature. The composition rules are:

A function $f(\text{expr}_1, \text{expr}_2, \ldots, \text{expr}_n)$
is log-log convex if $f$ is a log-log convex function and for each
$\text{expr}_i$ one of the following conditions holds:

- $f$ is increasing in argument $i$ and $\text{expr}_i$ is log-log convex.
- $f$ is decreasing in argument $i$ and $\text{expr}_i$ is log-log concave.
- $\text{expr}_i$ is log-log affine.

Similar rules apply for log-log concave and log-log affine
expressions.

If an Expression satisfies the composition rule, we colloquially say
that the Expression "is DGP." You can check the log-log curvature of
an expression using `CVXR:::is_log_log_convex()`, `CVXR:::is_log_log_concave()`, and
`CVXR:::is_log_log_affine()`. You can check whether a Problem is DGP by
calling `CVXR:::is_dgp()`.

```{r}
x <- Variable(pos = TRUE)
y <- Variable(pos = TRUE)

monomial <- 2.0 * x * y
posynomial <- monomial + power(x, 1.5) * power(y, -1)

cat("monomial is log-log affine:", CVXR:::is_log_log_affine(monomial), "\n")
cat("posynomial is log-log convex:", CVXR:::is_log_log_convex(posynomial), "\n")
```

## 3. DGP Problems

An LLCP is an optimization problem of the form

$$
\begin{array}{ll}
\mbox{minimize} & f_0(x) \\
\mbox{subject to} & f_i(x) \leq \tilde{f_i}, \quad i=1, \ldots, m\\
& g_i(x) = \tilde{g_i}, \quad i=1, \ldots, p,
\end{array}
$$

where the functions $f_i$ are log-log convex, $\tilde{f_i}$ are
log-log concave, and the functions $g_i$ and $\tilde{g_i}$ are
log-log affine.

A problem is DGP if additionally all the functions are DGP. You can
check whether a `CVXR` Problem is DGP by calling `CVXR:::is_dgp()`
on the problem object.

```{r}
x <- Variable(pos = TRUE)
y <- Variable(pos = TRUE)
z <- Variable(pos = TRUE)

objective_fn <- x * y * z
constraints <- list(
  4 * x * y * z + 2 * x * z <= 10,
  x <= 2 * y,
  y <= 2 * x,
  z >= 1
)
problem <- Problem(Maximize(objective_fn), constraints)
cat("Is this problem DGP?", CVXR:::is_dgp(problem), "\n")
```

### Solving DGP Problems

You can solve a DGP Problem by calling `psolve()` with `gp = TRUE`.

```{r}
result <- psolve(problem, gp = TRUE)
cat("Optimal value:", result, "\n")
cat("x:", value(x), "\n")
cat("y:", value(y), "\n")
cat("z:", value(z), "\n")
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(result = result, x = value(x), y = value(y), z = value(z)),
            file = file.path(testdata_dir, "dgp_fundamentals1.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    saved <- readRDS(file = file.path(testdata_dir, "dgp_fundamentals1.RDS"))
    expect_identical(result, saved$result)
    expect_identical(value(x), saved$x)
    expect_identical(value(y), saved$y)
    expect_identical(value(z), saved$z)
}
```

If you forget to supply `gp = TRUE`, an error will be raised because
the problem is not DCP-compliant.

## 4. Next Steps

### Atoms

`CVXR` has a large library of log-log convex functions, including
common functions like `exp()`, `log()`, and the difference between two
numbers via `one_minus_pos()`. Check out the DGP tutorial page for the
full list of atoms.

### References

For a reference on DGP, consult the following paper:
<https://web.stanford.edu/~boyd/papers/dgp.html>

## Session Info

```{r}
sessionInfo()
```
