---
title: Maximizing the Volume of a Box
author: "CVXPY Developers and Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: dgp_max_volume_box
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
source(here::here("_check_status.R"))
```

## Introduction

_This example is adapted from Boyd, Kim, Vandenberghe, and Hassibi,_
"[A Tutorial on Geometric Programming](https://web.stanford.edu/~boyd/papers/pdf/gp_tutorial.pdf)".

In this example, we maximize the volume of a box with height $h$,
width $w$, and depth $d$, with limits on the wall area $2(hw + hd)$
and the floor area $wd$, subject to bounds on the aspect ratios $h/w$
and $d/w$. The optimization problem is

$$
\begin{array}{ll}
\mbox{maximize} & hwd \\
\mbox{subject to} & 2(hw + hd) \leq A_{\text{wall}}, \\
& wd \leq A_{\text{flr}}, \\
& \alpha \leq h/w \leq \beta, \\
& \gamma \leq d/w \leq \delta.
\end{array}
$$

## Problem Formulation

```{r}
# Problem data
A_wall <- 100
A_flr <- 10
alpha <- 0.5
beta <- 2
gamma <- 0.5
delta <- 2

h <- Variable(pos = TRUE)
w <- Variable(pos = TRUE)
d <- Variable(pos = TRUE)

volume <- h * w * d
wall_area <- 2 * (h * w + h * d)
flr_area <- w * d
hw_ratio <- h / w
dw_ratio <- d / w

constraints <- list(
  wall_area <= A_wall,
  flr_area <= A_flr,
  hw_ratio >= alpha,
  hw_ratio <= beta,
  dw_ratio >= gamma,
  dw_ratio <= delta
)

problem <- Problem(Maximize(volume), constraints)
cat("Is problem DGP?", is_dgp(problem), "\n")
```

## Solution

```{r, results = "hold"}
result <- psolve(problem, gp = TRUE)
check_solver_status(problem)
cat("Optimal value (volume):", result, "\n")
cat("h:", value(h), "\n")
cat("w:", value(w), "\n")
cat("d:", value(d), "\n")
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(result = result, h = value(h), w = value(w), d = value(d)),
            file = file.path(testdata_dir, "max_volume_box1.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    saved <- readRDS(file = file.path(testdata_dir, "max_volume_box1.RDS"))
    expect_identical(result, saved$result)
    expect_identical(value(h), saved$h)
    expect_identical(value(w), saved$w)
    expect_identical(value(d), saved$d)
}
```

## Sensitivity Analysis via Dual Values

The dual values provide sensitivity information. A 1% increase in the
allowed wall area should yield approximately a proportional increase
in the maximum volume value.

```{r, results = "hold"}
cat("Dual value for wall area constraint:", dual_value(constraints[[1]]), "\n")
cat("Dual value for floor area constraint:", dual_value(constraints[[2]]), "\n")
```

The dual value for the wall area constraint is approximately 0.83,
meaning a 1% increase in allowed wall space yields approximately a
0.83% increase in the maximum volume. The dual value for the floor
area constraint is approximately 0.67, meaning a 1% increase in
allowed floor space yields approximately a 0.67% increase in maximum
volume.

## Session Info

```{r, echo = FALSE}
sessionInfo()
```

## References

- Boyd, S., Kim, S.-J., Vandenberghe, L., Hassibi, A. (2007). A
  Tutorial on Geometric Programming. _Optimization and Engineering_,
  8(1), 67--127.
