---
title: Perron-Frobenius Matrix Completion
author: "Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: dgp_pf_matrix_completion
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
```

## Introduction

The DGP atom library has several functions of positive matrices,
including the trace, (matrix) product, sum, Perron-Frobenius
eigenvalue, and $(I - X)^{-1}$ (eye-minus-inverse). In this example,
we use some of these atoms to formulate and solve an interesting
matrix completion problem.

In this problem, we are given some entries of an elementwise positive
matrix $A$, and the goal is to choose the missing entries so as to
minimize the Perron-Frobenius eigenvalue or spectral radius. Letting
$\Omega$ denote the set of indices $(i, j)$ for which $A_{ij}$ is
known, the optimization problem is

$$
\begin{array}{ll}
\mbox{minimize} & \lambda_{\text{pf}}(X) \\
\mbox{subject to} & \prod_{(i, j) \not\in \Omega} X_{ij} = 1 \\
& X_{ij} = A_{ij}, \, (i, j) \in \Omega,
\end{array}
$$

which is a log-log convex program.

## Problem Data

We consider the partially known matrix

$$
A = \begin{bmatrix}
1.0 & ? &  1.9 \\
? & 0.8 &  ? \\
3.2 & 5.9&  ?
\end{bmatrix},
$$

where the question marks denote the missing entries.

## Problem Formulation and Solution

```{r}
n <- 3

# Known entries: (row, col) pairs (1-indexed in R)
known_rows <- c(1, 1, 2, 3, 3)
known_cols <- c(1, 3, 2, 1, 2)
known_values <- c(1.0, 1.9, 0.8, 3.2, 5.9)

X <- Variable(c(n, n), pos = TRUE)
objective_fn <- pf_eigenvalue(X)

constraints <- list()
for (k in seq_along(known_values)) {
  constraints <- c(constraints, list(X[known_rows[k], known_cols[k]] == known_values[k]))
}

# Constraint on the product of unknown entries
# Unknown entries are: (1,2), (2,1), (2,3), (3,3)
constraints <- c(constraints, list(
  X[1, 2] * X[2, 1] * X[2, 3] * X[3, 3] == 1.0
))

problem <- Problem(Minimize(objective_fn), constraints)
result <- psolve(problem, gp = TRUE)
cat("Optimal value (spectral radius):", result, "\n")
cat("X:\n")
print(value(X))
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(result = result, X = value(X)),
            file = file.path(testdata_dir, "pf_matrix_completion1.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    saved <- readRDS(file = file.path(testdata_dir, "pf_matrix_completion1.RDS"))
    expect_identical(result, saved$result)
    expect_identical(value(X), saved$X)
}
```

The optimal spectral radius is approximately 4.70. The solver finds
values for the unknown entries that minimize this spectral radius
while satisfying the constraint that the product of the unknown
entries equals 1.

## Session Info

```{r}
sessionInfo()
```

## References

- Agrawal, A., Diamond, S., Boyd, S. (2019). Disciplined Geometric
  Programming. _Optimization Letters_, 13(5), 961--976.
