---
title: Power Control
author: "CVXPY Developers and Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: dgp_power_control
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
source(here::here("_check_status.R"))
```

## Introduction

_This example is adapted from Boyd, Kim, Vandenberghe, and Hassibi,_
"[A Tutorial on Geometric Programming](https://web.stanford.edu/~boyd/papers/pdf/gp_tutorial.pdf)."

_The problem data is adapted from the corresponding example in CVX's
example library (Almir Mutapcic)._

This example formulates and solves a power control problem for
communication systems, in which the goal is to minimize the total
transmitter power across $n$ transmitters, each transmitting positive
power levels $P_1, P_2, \ldots, P_n$ to $n$ receivers, labeled $1,
\ldots, n$, with receiver $i$ receiving signal from transmitter $i$.

The power received from transmitter $j$ at receiver $i$ is $G_{ij}
P_j$, where $G_{ij} > 0$ represents the path gain from transmitter
$j$ to receiver $i$. The signal power at receiver $i$ is $G_{ii}
P_i$, and the interference power at receiver $i$ is $\sum_{k \neq i}
G_{ik} P_k$. The noise power at receiver $i$ is $\sigma_i$, and the
signal to interference-plus-noise ratio (SINR) of the $i$-th
receiver-transmitter pair is

$$
S_i = \frac{G_{ii}P_i}{\sigma_i + \sum_{k \neq i} G_{ik}P_k}.
$$

The transmitters and receivers are constrained to have a minimum SINR
$S^{\min}$, and the $P_i$ are bounded between $P_i^{\min}$ and
$P_i^{\max}$. This gives the problem

$$
\begin{array}{ll}
\mbox{minimize} & P_1 + \cdots + P_n \\
\mbox{subject to} & P_i^{\min} \leq P_i \leq P_i^{\max}, \\
& 1/S^{\min} \geq \frac{\sigma_i + \sum_{k \neq i} G_{ik}P_k}{G_{ii}P_i}.
\end{array}
$$

## Problem Data

```{r}
n <- 5                        # number of transmitters and receivers
sigma <- rep(0.5, n)          # noise power at the receiver i
p_min <- rep(0.1, n)          # minimum power at the transmitter i
p_max <- rep(5, n)            # maximum power at the transmitter i
sinr_min <- 0.2               # threshold SINR for each receiver

# Path gain matrix
G <- matrix(
  c(1.0, 0.1, 0.2, 0.1, 0.05,
    0.1, 1.0, 0.1, 0.1, 0.05,
    0.2, 0.1, 1.0, 0.2, 0.2,
    0.1, 0.1, 0.2, 1.0, 0.1,
    0.05, 0.05, 0.2, 0.1, 1.0),
  nrow = 5, ncol = 5, byrow = TRUE
)
```

## Problem Formulation

```{r}
p <- Variable(n, pos = TRUE)
objective <- Minimize(sum_entries(p))

# Compute interference plus noise for each receiver
S_p <- list()
for (i in 1:n) {
  interference_terms <- list()
  for (k in 1:n) {
    if (k != i) {
      interference_terms <- c(interference_terms, list(G[i, k] * p[k]))
    }
  }
  # Sum all interference terms
  interf <- interference_terms[[1]]
  if (length(interference_terms) > 1) {
    for (j in 2:length(interference_terms)) {
      interf <- interf + interference_terms[[j]]
    }
  }
  S_p[[i]] <- sigma[i] + interf
}

# Build the signal power vector and the inverse SINR constraint
signal_power <- list()
inverse_sinr <- list()
for (i in 1:n) {
  signal_power[[i]] <- G[i, i] * p[i]
  inverse_sinr[[i]] <- S_p[[i]] / signal_power[[i]]
}

constraints <- list(
  p >= p_min,
  p <= p_max
)

for (i in 1:n) {
  constraints <- c(constraints, list(inverse_sinr[[i]] <= 1 / sinr_min))
}

problem <- Problem(objective, constraints)
cat("Is problem DGP?", is_dgp(problem), "\n")
```

## Solution

```{r, results = "hold"}
result <- psolve(problem, gp = TRUE)
check_solver_status(problem)
cat("Optimal total power:", result, "\n")
cat("Optimal power levels:", value(p), "\n")
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(result = result, p = value(p)),
            file = file.path(testdata_dir, "power_control1.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    saved <- readRDS(file = file.path(testdata_dir, "power_control1.RDS"))
    expect_identical(result, saved$result)
    expect_identical(value(p), saved$p)
}
```

We can verify that all the SINR constraints are active at the optimal
solution. The inverse SINR values should all be equal to
$1/S^{\min}$ at the optimum:

```{r}
for (i in 1:n) {
  cat("Inverse SINR for receiver", i, ":", value(inverse_sinr[[i]]), "\n")
}
```

## Session Info

```{r, echo = FALSE}
sessionInfo()
```

## References

- Boyd, S., Kim, S.-J., Vandenberghe, L., Hassibi, A. (2007). A
  Tutorial on Geometric Programming. _Optimization and Engineering_,
  8(1), 67--127.
