---
title: Rank-One Nonnegative Matrix Factorization
author: "Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: dgp_rank_one_nmf
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
```

## Introduction

The DGP atom library has several functions of positive matrices,
including the trace, (matrix) product, sum, Perron-Frobenius
eigenvalue, and $(I - X)^{-1}$ (eye-minus-inverse). In this example,
we use some of these atoms to approximate a partially known
elementwise positive matrix as the outer product of two positive
vectors.

We would like to approximate $A$ as the outer product of two positive
vectors $x$ and $y$, with $x$ normalized so that the product of its
entries equals $1$. Our criterion is the average relative deviation
between the entries of $A$ and $xy^T$, that is,

$$
\frac{1}{mn} \sum_{i=1}^{m} \sum_{j=1}^{n} R(A_{ij}, x_i y_j),
$$

where $R$ is the relative deviation of two positive numbers, defined as

$$
R(a, b) = \max\{a/b, b/a\} - 1.
$$

The corresponding optimization problem is

$$
\begin{array}{ll}
\mbox{minimize} & \frac{1}{mn} \sum_{i=1}^{m} \sum_{j=1}^{n} R(X_{ij}, x_i y_j) \\
\mbox{subject to} & x_1 x_2 \cdots x_m = 1 \\
& X_{ij} = A_{ij}, \quad \text{for } (i, j) \in \Omega,
\end{array}
$$

with variables $X \in \mathbf{R}^{m \times n}_{++}$, $x \in
\mathbf{R}^{m}_{++}$, and $y \in \mathbf{R}^{n}_{++}$. We can cast
this problem as an equivalent generalized geometric program by
discarding the $-1$ from the relative deviations.

## Problem Data

We consider the partially known matrix

$$
A = \begin{bmatrix}
1.0 & ? &  1.9 \\
? & 0.8 &  ? \\
3.2 & 5.9&  ?
\end{bmatrix},
$$

where the question marks denote the missing entries.

## Problem Formulation and Solution

```{r}
m <- 3
n <- 3

X <- Variable(c(m, n), pos = TRUE)
x <- Variable(m, pos = TRUE)
y <- Variable(n, pos = TRUE)

# Construct the outer product matrix: each row i is x[i] * y^T
outer_product <- vstack(t(x[1] * y), t(x[2] * y), t(x[3] * y))

# Relative deviations (without the -1): max(X/outer_product, outer_product/X)
relative_deviations <- max_elemwise(
  multiply(X, power(outer_product, -1)),
  multiply(power(X, -1), outer_product)
)

objective <- sum_entries(relative_deviations)

constraints <- list(
  X[1, 1] == 1.0,
  X[1, 3] == 1.9,
  X[2, 2] == 0.8,
  X[3, 1] == 3.2,
  X[3, 2] == 5.9,
  x[1] * x[2] * x[3] == 1.0
)

problem <- Problem(Minimize(objective), constraints)
result <- psolve(problem, gp = TRUE)

cat(sprintf("Optimal value (average relative deviation): %g\n",
            1.0 / (m * n) * (result - m * n)))
cat("Outer product approximation:\n")
print(value(outer_product))
cat("x:", value(x), "\n")
cat("y:", value(y), "\n")
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(result = result, x = value(x), y = value(y),
                 outer_product = value(outer_product)),
            file = file.path(testdata_dir, "rank_one_nmf1.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    saved <- readRDS(file = file.path(testdata_dir, "rank_one_nmf1.RDS"))
    expect_equal(result, saved$result, tolerance = 1e-5)
    expect_equal(value(x), saved$x, tolerance = 1e-5)
    expect_equal(value(y), saved$y, tolerance = 1e-5)
}
```

The optimal average relative deviation is essentially zero, meaning
the outer product $xy^T$ matches $A$ exactly at the known entries. The
solver fills in the unknown entries of $X$ to be consistent with the
rank-one outer product structure.

## Session Info

```{r}
sessionInfo()
```

## References

- Agrawal, A., Diamond, S., Boyd, S. (2019). Disciplined Geometric
  Programming. _Optimization Letters_, 13(5), 961--976.
