---
title: Stochastically Ordered Survival Functions
author: "Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: dgp_survival_ordering
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
library(ggplot2)
library(survival)      # Kaplan-Meier, for validation
source(here::here("_check_status.R"))
theme_set(theme_bw())

## Oropharynx carcinoma data (tstage, nstage, status, days), bundled with
## this example. Originally the Kalbfleisch-Prentice data, as distributed in
## the invGauss package's d.oropha.rec (times converted from years to days).
oropharynx <- readRDS(here::here("examples/dgp/oropharynx.rds"))
```

## Introduction

In [Estimating Stochastically Ordered Survival Functions via
Geometric Programming](https://doi.org/10.1198/jcgs.2009.06140), Lim
et al. give a nice application of geometric programming to nonparametric
estimation.

Suppose we have survival functions $S_1, \ldots, S_N$ for $N$
populations and prior knowledge that they are **stochastically
ordered**. Two forms are common [@LimKimWang:2009]:

- **Simple** ordering, written $S_a \preceq S_b$: population $a$ dies
  stochastically earlier, i.e. $S_a(t) \le S_b(t)$ for every $t$.
- **Uniform** ordering: the ratio $S_a(t)/S_b(t)$ is monotone in $t$ ---
  a stronger requirement that compares the two populations *conditionally
  on survival so far*, equivalent to an ordering of the hazard rates
  [@DykstraKocharRobertson:1991].

The unconstrained nonparametric maximum-likelihood estimates (NPMLEs)
--- Kaplan--Meier curves and their interval-censored analogues ---
often *violate* an ordering we believe on scientific grounds, crossing
where they should not. @LimKimWang:2009 observed that imposing the
ordering turns the NPMLE into a **geometric program** (GP), and hence,
in modern terms, a **disciplined geometric program** (DGP)
[@AgrawalDiamondBoyd:2019] that `CVXR` can solve directly via
`psolve(prob, gp = TRUE)`.

We take up the easily-understood **right-censored** case first (Section 3
of the paper), then the **interval-censored** case (Section 4).

## Geometric programs in one paragraph

A **monomial** is a function $c\,x_1^{a_1}\cdots x_n^{a_n}$ with $c > 0$,
real exponents $a_i$, and positive variables $x_i$; a sum of monomials is
a **posynomial**. A geometric program minimizes a posynomial subject to
constraints of the form (posynomial $\le 1$) and (monomial $= 1$). The
logarithmic change of variables $x_i = e^{y_i}$ makes every such problem
convex. `CVXR` recognizes this structure automatically: positive
variables are declared with `pos = TRUE`, the objective and each
constraint carry a *log-log curvature*, and `is_dgp()` checks that they
compose into a valid program.

# Right censoring

With right-censored data each subject either fails at an observed time or
is known only to survive past its censoring time. Consider $N$
populations. For population $i$, let $t_{i1} < \cdots < t_{i m_i}$ be its
distinct observed **failure times**, and at $t_{ij}$ let
$$
n_{ij} = \#\{\text{subjects of population } i \text{ at risk just before } t_{ij}\},
\qquad
d_{ij} = \#\{\text{failures at } t_{ij}\}.
$$

### The likelihood is a monomial

Define the **discrete conditional survival probability**
$$
p_{ij} = P(Y_i > t_{ij} \mid Y_i > t_{i,j-1}) = \frac{S_i(t_{ij})}{S_i(t_{i,j-1})},
\qquad q_{ij} = 1 - p_{ij},
$$
so that the survival function is the running product (the Kaplan--Meier
form)
$$
S_i(t_{ij}) = \prod_{r=1}^{j} p_{ir}.
$$
Conditionally on the $n_{ij}$ subjects at risk, the $d_{ij}$ failures are
independent Bernoulli events: each subject fails at this instant with
probability $q_{ij}$ and survives it with probability $p_{ij}$. So the
$t_{ij}$ term contributes $q_{ij}^{d_{ij}} p_{ij}^{n_{ij} - d_{ij}}$, and
the full likelihood is the **monomial**
$$
\mathcal L = \prod_{i=1}^{N} \prod_{j=1}^{m_i} p_{ij}^{\,n_{ij} - d_{ij}}\, q_{ij}^{\,d_{ij}} .
$$
Because a monomial is log-log affine, the objective fits directly into a
GP; the modeling work is in the constraints: that the $p_{ij}, q_{ij}$ are
probabilities, that survival is nonincreasing, and that the populations are
ordered. As we will see, the sum-to-one constraint alone recovers the
ordinary NPMLE (Kaplan--Meier), and the ordering monomials add the
stochastic ordering.

### Constraints

**Sum-to-one (relaxed).** Since $p_{ij}$ and $q_{ij}$ are complementary
probabilities, $p_{ij} + q_{ij} = 1$. A GP admits only *monomial*
equalities, and $p_{ij} + q_{ij}$ is a posynomial, so we relax to the
posynomial inequality
$$
p_{ij} + q_{ij} \le 1 .
$$
This is exact at the optimum: both exponents $n_{ij} - d_{ij}$ and
$d_{ij}$ are nonnegative, so $\mathcal L$ increases in each $p_{ij}$ and
$q_{ij}$; wherever a variable is informative the optimizer pushes it up
until the constraint binds.

**Monotonicity is automatic.** A survival curve must be nonincreasing.
Here that is free: $S_i(t_{ij}) = \prod_{r \le j} p_{ir}$ is a product of
factors $p_{ir} \le 1$, so it decreases in $j$ with no extra constraint.
(This is a genuine convenience of the conditional parameterization; the
interval-censored model below is not so lucky.)

**Simple ordering.** "$S_a$ stochastically smaller than $S_b$" means
$S_a(t) \le S_b(t)$ at every $t$. Both survival curves are step functions
that only change at failure times, so it suffices to enforce the
inequality at each pooled failure time $t$:
$$
\prod_{r \,:\, t_{ar} \le t} p_{ar}
\;\le\;
\prod_{r \,:\, t_{br} \le t} p_{br}
\quad\Longleftrightarrow\quad
\Bigl(\textstyle\prod_{t_{ar} \le t} p_{ar}\Bigr)
\Bigl(\textstyle\prod_{t_{br} \le t} p_{br}\Bigr)^{-1} \le 1,
$$
a monomial $\le 1$. (Uniform ordering, when the two populations share a
grid, is even simpler: $p_{aj}\, p_{bj}^{-1} \le 1$ at each $j$ --- an
ordering of the one-step conditional survival probabilities, i.e. of the
discrete hazards.)

Everything is a monomial except $p_{ij} + q_{ij} \le 1$ (a posynomial
$\le 1$), so the problem is a DGP.

### The oropharynx data

We use the oropharyngeal-carcinoma trial of Kalbfleisch and Prentice
(`r nrow(oropharynx)` patients, bundled with this example as
`oropharynx.rds`). Following @LimKimWang:2009 we group patients by tumor
stage `tstage` (T) and nodal stage `nstage` (N), and model time to death.

```{r rc-counts}
## discrete-hazard counts for one (T, N) group
grp_counts <- function(T, N) {
  s  <- subset(oropharynx, tstage == T & nstage == N)
  dt <- sort(unique(s$days[s$status == 1]))            # distinct failure times
  out <- data.frame(t = dt,
             d = sapply(dt, function(x) sum(s$days == x & s$status == 1)),  # failures
             n = sapply(dt, function(x) sum(s$days >= x)))                  # at risk
  attr(out, "tmax") <- max(s$days)   # last observation, for the flat Kaplan-Meier tail
  out
}
```

### CVXR implementation

`fit_rc()` builds the monomial likelihood, the relaxed sum-to-one
constraints, and one simple-ordering monomial per pooled failure time for
each requested pair. A pair `c(a, b)` requests $S_a \le S_b$.

```{r rc-fit}
fit_rc <- function(groups, order = list()) {
  G <- length(groups)
  p <- lapply(groups, function(g) Variable(nrow(g), pos = TRUE))
  q <- lapply(groups, function(g) Variable(nrow(g), pos = TRUE))
  mono <- function(v, a) Reduce(`*`, lapply(which(a != 0), function(j) v[j]^a[j]))
  Scum <- function(i, ix) if (length(ix)) prod_entries(p[[i]][ix]) else 1   # empty prod = 1

  objective <- Reduce(`*`, lapply(seq_len(G), function(i)
                 mono(p[[i]], groups[[i]]$n - groups[[i]]$d) * mono(q[[i]], groups[[i]]$d)))
  constraints <- lapply(seq_len(G), function(i) p[[i]] + q[[i]] <= 1)

  for (o in order) {                     # S_a(t) <= S_b(t) at every pooled failure time
    a <- o[1]; b <- o[2]
    for (tk in sort(unique(c(groups[[a]]$t, groups[[b]]$t)))) {
      ia <- which(groups[[a]]$t <= tk); ib <- which(groups[[b]]$t <= tk)
      constraints <- c(constraints, Scum(a, ia) * Scum(b, ib)^(-1) <= 1)
    }
  }
  prob <- Problem(Maximize(objective), constraints)
  val  <- psolve(prob, gp = TRUE)
  check_solver_status(prob)
  list(is_dgp = is_dgp(prob), logLik = log(val),
       S = lapply(seq_len(G), function(i) cumprod(value(p[[i]]))))   # S = running product
}
```

A quick sanity check: with no ordering, the GP must return the ordinary
Kaplan--Meier estimator.

```{r rc-km-check}
g <- grp_counts(3, 3)
fit <- fit_rc(list(g))
km  <- survfit(Surv(days, status) ~ 1, data = subset(oropharynx, tstage == 3 & nstage == 3))
max(abs(fit$S[[1]] - summary(km, times = g$t)$surv))     # ~ 0: GP == Kaplan-Meier
```

### Simple ordering (paper Figure 1)

The paper's first example takes the three T = 3 groups differing in nodal
involvement, $(T,N) = (3,1), (3,2), (3,3)$, and imposes $S_3 \preceq S_2
\preceq S_1$ (more nodal involvement, worse survival).

```{r rc-fig1, fig.width = 8, fig.height = 3.3}
g1 <- list("(3,1)" = grp_counts(3, 1),
           "(3,2)" = grp_counts(3, 2),
           "(3,3)" = grp_counts(3, 3))

rc_free  <- fit_rc(g1)
rc_order <- fit_rc(g1, list(c(3, 2), c(2, 1)))     # S3 <= S2 <= S1

step_rc <- function(fit, groups, panel)
  do.call(rbind, lapply(seq_along(groups), function(i) {
    g <- groups[[i]]; S <- fit$S[[i]]; tmax <- attr(g, "tmax")
    t <- c(0, g$t); s <- c(1, S)
    if (tmax > max(g$t)) { t <- c(t, tmax); s <- c(s, tail(S, 1)) }  # flat KM tail to last obs
    data.frame(t = t, S = s, grp = paste0("(T,N)=", names(groups)[i]), panel = panel)
  }))

df1 <- rbind(step_rc(rc_free,  g1, "(a) Kaplan-Meier (unconstrained)"),
             step_rc(rc_order, g1, "(b) ordered: S3 <= S2 <= S1"))

ggplot(df1, aes(t, S, colour = grp)) + geom_step(linewidth = 0.6) +
  facet_wrap(~panel) + coord_cartesian(xlim = c(0, 1850), ylim = c(0, 1)) +
  labs(x = "time (days)", y = "survival probability", colour = NULL) +
  theme(legend.position = "inside", legend.position.inside = c(0.9, 0.8),
        legend.background = element_blank())
```

The unconstrained Kaplan--Meier curves cross --- $(3,2)$ sits above
$(3,1)$ for much of the range, contradicting $S_2 \le S_1$. Imposing the
ordering removes the crossings, reproducing Figure 1 of @LimKimWang:2009.

```{r rc-fig1-ll}
c(unconstrained = round(rc_free$logLik, 2), ordered = round(rc_order$logLik, 2))
```

### Partial ordering (paper Figure 2)

Ordering need not be a total order. For the four groups $(3,0), (3,3),
(4,0), (4,3)$ --- populations $1$–$4$ --- larger T or larger N should
worsen survival, giving the *partial* order $S_4 \preceq S_2 \preceq S_1$
and $S_4 \preceq S_3 \preceq S_1$, while $S_2$ and $S_3$ are left
incomparable. Each "$\preceq$" is just another list of pairs.

```{r rc-fig2, fig.width = 8, fig.height = 3.3}
g2 <- list("(3,0)" = grp_counts(3, 0), "(3,3)" = grp_counts(3, 3),
           "(4,0)" = grp_counts(4, 0), "(4,3)" = grp_counts(4, 3))

pf_free  <- fit_rc(g2)
pf_order <- fit_rc(g2, list(c(4, 2), c(2, 1), c(4, 3), c(3, 1)))

df2 <- rbind(step_rc(pf_free,  g2, "(a) Kaplan-Meier (unconstrained)"),
             step_rc(pf_order, g2, "(b) partial order"))

ggplot(df2, aes(t, S, colour = grp)) + geom_step(linewidth = 0.6) +
  facet_wrap(~panel) + coord_cartesian(xlim = c(0, 1850), ylim = c(0, 1)) +
  labs(x = "time (days)", y = "survival probability", colour = NULL) +
  theme(legend.position = "inside", legend.position.inside = c(0.9, 0.75),
        legend.background = element_blank())
```

```{r rc-fig2-ll}
c(unconstrained = round(pf_free$logLik, 2), partial_order = round(pf_order$logLik, 2))
```

# Interval censoring

Right-censored data tell us a failure time *exactly* or *bounds it below*.
**Interval-censored** data only bound it: the failure is known to lie
between two examination times. The simplest form is **"case 1" interval
censoring**, also called **current-status data**: each subject is
examined exactly *once*, at a time $C$, and all we learn is the single bit
$\delta = \mathbf 1\{Y \le C\}$ --- whether the event has already
happened. The failure time is thus known only to lie in $(0, C]$ (if
$\delta = 1$) or in $(C, \infty)$ (if $\delta = 0$). ("Case 2" data have
two examinations per subject and "case $k$" data have $k$; see
@Sun:2006.) We treat case 1, as in Section 4.2 of @LimKimWang:2009.

### The likelihood is again a monomial

Consider $N$ populations, pool all examination times into a common grid
$t_1 < \cdots < t_m$, and for population $i$ at $t_j$ let
$$
n_{ij} = \#\{\text{subjects of population } i \text{ examined at } t_j\},
\qquad
r_{ij} = \#\{\text{those with the event already present}\}.
$$
Because a subject examined at $t_j$ has the event already present with
probability $P(Y_i \le t_j) = 1 - S_i(t_j)$, and these are independent,
$r_{ij} \sim \mathrm{Binomial}(n_{ij}, 1 - S_i(t_j))$. This time we
parameterize *directly* with the survival values,
$$
q_{ij} = S_i(t_j) \in (0, 1], \qquad p_{ij} = 1 - S_i(t_j),
$$
and the likelihood is the **monomial**
$$
\mathcal L = \prod_{i=1}^{N} \prod_{j=1}^{m} p_{ij}^{\,r_{ij}}\, q_{ij}^{\,n_{ij} - r_{ij}} .
$$

### Constraints

**Sum-to-one (relaxed)** as before: $p_{ij} + q_{ij} \le 1$, tight at the
optimum for the same nonnegative-exponent reason.

**Monotonicity is now explicit.** Survival must be nonincreasing.
Following the authors' code we impose this on $p_{ij} = 1 - S_i(t_j)$,
making *it* nondecreasing, and read survival back as $S_i(t_j) = 1 -
p_{ij}$:
$$
S_i(t_j) \ge S_i(t_{j+1})
\quad\Longleftrightarrow\quad
p_{ij}\, p_{i,j+1}^{-1} \le 1,
$$
a monomial $\le 1$ for $j = 1, \ldots, m - 1$. This choice matters for the
*free* early points. A population with no examinations yet has $p_{ij}$
unconstrained by the data; because $p$ increases from near 0, those values
sit at ~0, so $S = 1 - p$ stays at $1$ until the first event --- the
standard survival convention, and exactly what the paper's code produces
with no ad-hoc adjustment. (Imposing monotonicity on $q = S$ instead and
reading $S = q$ is algebraically equivalent but leaves those same free
points centered *below* 1.)

**Ordering.** Both populations share the grid $t_1, \ldots, t_m$.

- **Simple**, $S_a \preceq S_b$ (population $a$ smaller): $1 - p_{aj} \le 1
  - p_{bj} \Leftrightarrow p_{aj} \ge p_{bj}$, i.e. the monomial $p_{bj}\,
  p_{aj}^{-1} \le 1$ at each $t_j$.
- **Uniform**, "$S_a$ uniformly larger than $S_b$": the ratio
  $S_a(t)/S_b(t)$ is nondecreasing, i.e. $\dfrac{S_a(t_j)}{S_b(t_j)} \le
  \dfrac{S_a(t_{j+1})}{S_b(t_{j+1})}$. This is a monomial only in the
  *survival* variables $q_{ij} = S_i(t_j)$; cross-multiplying (all terms
  positive),
  $$
  q_{aj}\, q_{b,j+1} \le q_{a,j+1}\, q_{bj}
  \quad\Longleftrightarrow\quad
  q_{aj}\, q_{b,j+1}\, q_{a,j+1}^{-1}\, q_{bj}^{-1} \le 1.
  $$

So the simple-ordering fit works in $p$ (survival $1 - p$), while uniform
ordering works in the survival variables $q$ directly. Uniform ordering
implies simple ordering.

### The lung-tumor data

The data are the germ-free-mouse experiment of @HoelWalburg:1972,
tabulated as Table 1.3 in @Sun:2006. At death each mouse was autopsied for
a lung tumor. Because the tumor is *occult* --- present or not at death,
and not the cause of death --- each animal yields a single current-status
observation: the death time is the examination time $C$, and $\delta = 1$
if a tumor was found. The two arms are

| Population | Environment | Mice | Found with tumor |
|---|---|---:|---:|
| 1 | conventional | 96 | 27 |
| 2 | germ-free | 47 | 35 |

(The two arms hold $96 + 47 = 143$ animals in the data distributed with
the paper; the study enrolled 144.) Pooling the death times gives $m =
127$ distinct examination times (45–1008 days). The germ-free mice have
the higher tumor prevalence, so their tumor-free survival sits *below* the
conventional arm. Following @LimKimWang:2009's Figure 4 we impose that
germ-free tumor onset is stochastically earlier, $S_2(t) \le S_1(t)$, which
the unconstrained estimates violate at several times. (The *text* of §4.2
states the reverse, $S_1 \le S_2$, but the paper's supplementary code
imposes $S_2 \le S_1$ and its figure plots germ-free below --- so we follow
the code.)

```{r ic-data, echo = FALSE}
time <- c(
  45, 198, 215, 217, 257, 262, 266, 371, 381, 412, 431, 447, 454, 459, 475,
  477, 479, 484, 485, 500, 502, 503, 505, 508, 515, 516, 524, 531, 539, 541,
  546, 553, 556, 563, 565, 570, 572, 575, 577, 582, 585, 588, 594, 600, 601,
  603, 608, 609, 614, 616, 624, 632, 638, 642, 644, 647, 650, 651, 653, 656,
  659, 660, 662, 663, 667, 672, 673, 677, 679, 689, 692, 693, 695, 698, 702,
  709, 710, 718, 720, 721, 723, 728, 731, 752, 760, 762, 773, 775, 777, 779,
  781, 782, 785, 789, 795, 808, 810, 811, 814, 815, 817, 839, 842, 846, 851,
  871, 873, 876, 880, 886, 888, 890, 894, 896, 911, 913, 914, 914, 916, 921,
  921, 926, 936, 942, 945, 986, 1008)
r1 <- c(
  0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 1, 0, 0, 1, 0, 0, 0, 0, 0, 1,
  0, 0, 0, 1, 0, 0, 0, 0, 1, 1, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 1, 0, 0, 0, 1,
  1, 0, 0, 0, 0, 0, 1, 1, 0, 1, 1, 0, 0, 0, 0, 1, 0, 0, 1, 0, 0, 0, 0, 1, 1,
  1, 0, 0, 0, 0, 1, 0, 1, 0, 0, 0, 0, 1, 0, 1, 0, 0, 0, 0, 1, 0, 0, 1, 0, 0,
  0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0,
  0, 0)
n1 <- c(
  1, 1, 1, 1, 1, 1, 1, 1, 1, 0, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1,
  1, 0, 1, 1, 1, 0, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 0, 1, 2,
  1, 2, 1, 3, 2, 2, 1, 1, 1, 1, 2, 1, 1, 1, 2, 1, 2, 1, 1, 1, 0, 1, 0, 1, 1,
  1, 0, 1, 1, 1, 1, 1, 1, 0, 1, 1, 1, 1, 1, 1, 0, 0, 0, 0, 1, 0, 0, 1, 0, 1,
  0, 1, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0,
  0, 0)
r2 <- c(
  0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0,
  0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0,
  0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 2, 0, 0, 0, 0,
  0, 1, 0, 0, 0, 0, 0, 0, 1, 0, 0, 1, 0, 0, 0, 1, 1, 0, 1, 0, 1, 1, 0, 1, 0,
  0, 0, 1, 1, 1, 1, 1, 1, 0, 0, 2, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 0, 1,
  0, 1)
n2 <- c(
  0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0,
  0, 1, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0,
  0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 2, 0, 1, 0, 0,
  0, 1, 0, 0, 0, 0, 0, 0, 1, 0, 0, 1, 0, 0, 0, 1, 1, 1, 1, 0, 1, 1, 0, 2, 0,
  1, 0, 1, 1, 2, 1, 1, 1, 1, 0, 2, 1, 1, 1, 1, 2, 1, 1, 1, 1, 1, 1, 1, 1, 1,
  1, 1)
m <- length(time)
```

### CVXR implementation

Following the authors' code, the builder parameterizes by $p = 1 - S$ made
monotone *increasing*, reads survival back as $1 - p$, and takes the simple
ordering as a function of the two `p` vectors.

```{r ic-fit}
monomial <- function(v, a) Reduce(`*`, lapply(which(a != 0), function(j) v[j]^a[j]))

fit_ic <- function(order_fun = function(p1, p2) list()) {
  p1 <- Variable(m, pos = TRUE); q1 <- Variable(m, pos = TRUE)  # population 1
  p2 <- Variable(m, pos = TRUE); q2 <- Variable(m, pos = TRUE)  # population 2

  objective <- monomial(p1, r1) * monomial(q1, n1 - r1) *
               monomial(p2, r2) * monomial(q2, n2 - r2)

  constraints <- c(
    list(p1 + q1 <= 1, p2 + q2 <= 1),                     # sum-to-one (relaxed)
    list(p1[1:(m-1)] <= p1[2:m], p2[1:(m-1)] <= p2[2:m]), # p increasing => S = 1-p decreasing
    order_fun(p1, p2))                                    # ordering monomials

  prob <- Problem(Maximize(objective), constraints)
  val  <- psolve(prob, gp = TRUE)
  check_solver_status(prob)
  list(is_dgp = is_dgp(prob), logLik = log(val), S1 = 1 - value(p1), S2 = 1 - value(p2))
}

no_order     <- function(p1, p2) list()
simple_order <- function(p1, p2) list(p1 <= p2)   # p1 <= p2  <=>  S1 >= S2 (germ-free below)
```

### Simple ordering (paper Figure 4)

```{r ic-fig4, fig.width = 8, fig.height = 3.2}
ic_free   <- fit_ic(no_order)
ic_simple <- fit_ic(simple_order)

step_ic <- function(fit, panel)
  rbind(data.frame(t = c(0, time), S = c(1, fit$S1), grp = "Conventional Env."),
        data.frame(t = c(0, time), S = c(1, fit$S2), grp = "Germ-free Env.")) |>
    transform(panel = panel)

df4 <- rbind(step_ic(ic_free,   "(a) not ordered"),
             step_ic(ic_simple, "(b) ordered: S2 <= S1"))

ggplot(df4, aes(t, S, colour = grp, linetype = grp)) +
  geom_step(linewidth = 0.5) + facet_wrap(~panel) +
  scale_colour_manual(values = c("Conventional Env." = "blue",
                                 "Germ-free Env." = "darkgreen")) +
  coord_cartesian(xlim = c(0, 1200), ylim = c(0, 1)) +
  labs(x = "time", y = "survival probability", colour = NULL, linetype = NULL) +
  theme(legend.position = "inside", legend.position.inside = c(0.79, 0.96),
        legend.justification.inside = c(0, 1), legend.background = element_blank())
```

The unconstrained estimates keep conventional above germ-free but cross at
several times where the germ-free curve pokes above. Imposing $S_2 \le
S_1$ removes those crossings --- germ-free stays cleanly below conventional
--- reproducing Figure 4 of @LimKimWang:2009. Both arms sit flat at $1$
until their first tumor, as in the paper's figure; this falls out of the
$p = 1-S$ parameterization with no plotting convention applied.

```{r ic-fig4-viol}
c(unconstrained_violations = sum(ic_free$S2   - ic_free$S1   > 1e-6),
  simple_violations        = sum(ic_simple$S2 - ic_simple$S1 > 1e-6))
```

### Uniform ordering

Uniform ordering forces the ratio $S_1/S_2$ to be monotone, so the curves
may only diverge. Since it is a monomial only in the survival variables
$q$, this fit is built in $q$ (made monotone decreasing) rather than in
$p$, and the survival is simply $q$ itself. Imposed with conventional
uniformly larger, it keeps conventional cleanly above germ-free with a
widening gap; the uniform constraint pins the early free values, so the
curves again sit flat at 1 before each first tumor.

```{r ic-uniform, fig.width = 8, fig.height = 3.2}
fit_ic_uniform <- function() {
  p1 <- Variable(m, pos = TRUE); q1 <- Variable(m, pos = TRUE)
  p2 <- Variable(m, pos = TRUE); q2 <- Variable(m, pos = TRUE)
  objective <- monomial(p1, r1) * monomial(q1, n1 - r1) *
               monomial(p2, r2) * monomial(q2, n2 - r2)
  constraints <- c(
    list(p1 + q1 <= 1, p2 + q2 <= 1),
    list(q1[2:m] <= q1[1:(m-1)], q2[2:m] <= q2[1:(m-1)]),   # q monotone decreasing
    list(q1[1:(m-1)] * q2[2:m] <= q1[2:m] * q2[1:(m-1)]))   # S1/S2 nondecreasing
  prob <- Problem(Maximize(objective), constraints)
  val  <- psolve(prob, gp = TRUE); check_solver_status(prob)
  list(logLik = log(val), S1 = value(q1), S2 = value(q2))
}
ic_uniform <- fit_ic_uniform()

dfu <- rbind(step_ic(ic_free,    "x") |> transform(fit = "unconstrained"),
             step_ic(ic_uniform, "x") |> transform(fit = "uniform"))

ggplot(dfu, aes(t, S, colour = grp, linewidth = fit, alpha = fit)) +
  geom_step() +
  scale_colour_manual(values = c("Conventional Env." = "blue",
                                 "Germ-free Env." = "darkgreen")) +
  scale_linewidth_manual(values = c(unconstrained = 0.4, uniform = 1.0)) +
  scale_alpha_manual(values = c(unconstrained = 0.4, uniform = 1)) +
  coord_cartesian(xlim = c(0, 1200), ylim = c(0, 1)) +
  labs(x = "time", y = "survival probability",
       colour = NULL, linewidth = NULL, alpha = NULL) +
  theme(legend.position = "inside", legend.position.inside = c(0.72, 0.96),
        legend.justification.inside = c(0, 1), legend.background = element_blank())
```

```{r ic-uniform-check}
c(unconstrained_ratio_monotone = all(diff(ic_free$S1    / ic_free$S2)    >= -1e-6),
  uniform_ratio_monotone       = all(diff(ic_uniform$S1 / ic_uniform$S2) >= -1e-6))
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(rc = rc_order$logLik, pf = pf_order$logLik,
                 ic = ic_simple$logLik, un = ic_uniform$logLik),
            file = file.path(testdata_dir, "survival_ordering.RDS"))
} else {
    ref <- readRDS(file = file.path(testdata_dir, "survival_ordering.RDS"))
    expect_equal(rc_order$logLik, ref$rc, tolerance = 1e-3)
    expect_equal(ic_simple$logLik, ref$ic, tolerance = 1e-3)
    expect_equal(ic_uniform$logLik, ref$un, tolerance = 1e-3)
}
```

# Notes

- The *unconstrained* NPMLE needs no GP at all: `survival::survfit()`
  computes it directly --- Kaplan--Meier for the right-censored groups, and
  the Turnbull (isotonic) estimator for the interval-censored case via
  `Surv(L, R, type = "interval2")`. The GP with no ordering reproduces it
  (the Kaplan--Meier check above), so the GP earns its keep only for the
  *ordering* constraints, which `survfit()` cannot impose; and the
  constrained log-likelihood is never above the unconstrained one.
- The two censoring types use different parameterizations, and each gets a
  different constraint for free. The right-censored form uses the
  conditional survival $p_{ij} = S_i(t_j)/S_i(t_{j-1})$ with $S = \prod p$:
  monotonicity is automatic (each $p \le 1$), but simple ordering is a
  running-product inequality. The interval form uses $p_{ij} = 1 - S_i(t_j)$
  made monotone *increasing* with $S = 1 - p$: monotonicity is an explicit
  constraint, but simple ordering is a single monomial at each time point
  (and uniform ordering a monomial in the survival variables $q$).
- `CVXR`'s `power()` atom takes a scalar exponent, so a vector power such
  as `q1^r1` errors; the monomial is built as a product of scalar-power
  terms (the `monomial()`/`mono()` helpers).
- At the optimum $p + q = 1$ binds wherever the variable is informative.
  Where an exponent vanishes (e.g. an arm before its first examination) the
  survival value is *not* pinned by the data --- a flat, non-unique stretch
  of the NPMLE. Which representative the solver returns depends on the
  parameterization: with $p = 1-S$ monotone increasing those free values sit
  at ~0, so $S = 1-p$ stays at 1 (the standard convention). This is a
  property of the parameterization, not a plotting fix-up --- and it matches
  the paper's supplementary code, which likewise reads $S = 1 - p$.

## Session Information

```{r}
sessionInfo()
```

## References {.unnumbered}
