---
title: Disciplined Geometric Programming
author: "Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: dgp_tutorial
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
source(here::here("_check_status.R"))
```

## Introduction

Disciplined geometric programming (DGP) is an analog of DCP for
_log-log convex_ functions, that is, functions of positive variables
that are convex with respect to the geometric mean instead of the
arithmetic mean.

While DCP is a ruleset for constructing convex programs, DGP is a
ruleset for log-log convex programs (LLCPs), which are problems that
are convex after the variables, objective functions, and constraint
functions are replaced with their logs, an operation that we refer to
as a _log-log_ transformation. Every geometric program (GP) and
generalized geometric program (GGP) is an LLCP, but there are LLCPs
that are neither GPs nor GGPs.

`CVXR` lets you form and solve DGP problems, just as it does for DCP
problems. For example, the following code solves a simple geometric
program:

```{r}
# DGP requires Variables to be declared positive via `pos=TRUE`.
x <- Variable(pos = TRUE)
y <- Variable(pos = TRUE)
z <- Variable(pos = TRUE)

objective_fn <- x * y * z
constraints <- list(
  4 * x * y * z + 2 * x * z <= 10,
  x <= 2 * y,
  y <= 2 * x,
  z >= 1
)
problem <- Problem(Maximize(objective_fn), constraints)
result <- psolve(problem, gp = TRUE)
check_solver_status(problem)
cat("Optimal value:", result, "\n")
cat("x:", value(x), "\n")
cat("y:", value(y), "\n")
cat("z:", value(z), "\n")
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(result = result, x = value(x), y = value(y), z = value(z)),
            file = file.path(testdata_dir, "dgp_tutorial1.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    saved <- readRDS(file = file.path(testdata_dir, "dgp_tutorial1.RDS"))
    expect_identical(result, saved$result)
    expect_identical(value(x), saved$x)
    expect_identical(value(y), saved$y)
    expect_identical(value(z), saved$z)
}
```

Note that to solve DGP problems, you must pass the option `gp = TRUE`
to `psolve()`.

## Log-Log Curvature

Just as every Expression in `CVXR` has a curvature (constant, affine,
convex, concave, or unknown), every Expression also has a log-log
curvature.

A function $f : D \subseteq \mathbf{R}^n_{++} \to \mathbf{R}$ is
said to be log-log convex if the function $F(u) = \log f(e^u)$, with
domain $\{u \in \mathbf{R}^n : e^u \in D\}$, is convex (where
$\mathbf{R}^n_{++}$ denotes the set of positive reals and the
logarithm and exponential are meant elementwise); the function $F$ is
called the log-log transformation of $f$. The function $f$ is log-log
concave if $F$ is concave, and it is log-log affine if $F$ is affine.

Every log-log affine function has the form

$$
f(x) = cx_1^{a_1}x_2^{a_2} \ldots x_n^{a_n}
$$

where $x$ is in $\mathbf{R}^n_{++}$, the $a_i$ are real numbers, and
$c$ is a positive scalar. In the context of geometric programming,
such a function is called a monomial function. A sum of monomials,
known as a posynomial function in geometric programming, is a log-log
convex function.

| Log-Log Curvature | Meaning |
|---|---|
| log-log constant | $F$ is a constant (so $f$ is a positive constant) |
| log-log affine | $F(\theta u + (1-\theta)v) = \theta F(u) + (1-\theta)F(v)$ for all $u, v, \theta \in [0,1]$ |
| log-log convex | $F(\theta u + (1-\theta)v) \leq \theta F(u) + (1-\theta)F(v)$ for all $u, v, \theta \in [0,1]$ |
| log-log concave | $F(\theta u + (1-\theta)v) \geq \theta F(u) + (1-\theta)F(v)$ for all $u, v, \theta \in [0,1]$ |
| unknown | DGP analysis cannot determine the curvature |

`CVXR`'s log-log curvature analysis can flag Expressions as unknown
even when they are log-log convex or log-log concave. Note that any
log-log constant expression is also log-log affine, and any log-log
affine expression is log-log convex and log-log concave.

## Log-Log Curvature Rules

For an Expression to have known log-log curvature, all of the
Constants, Variables, and Parameters it refers to must be elementwise
positive. Variables and Parameters are positive only if the keyword
argument `pos = TRUE` is supplied to their constructors (e.g.,
`x <- Variable(pos = TRUE)`). To summarize, when formulating a DGP
problem, _all Constants should be elementwise positive, and all
Variables and Parameters must be constructed with the attribute
`pos = TRUE`_.

DGP analysis is exactly analogous to DCP analysis. It is based on a
library of atoms (functions) with known monotonicity and log-log
curvature and a single composition rule.

A function $f(\text{expr}_1, \text{expr}_2, \ldots, \text{expr}_n)$
is log-log convex if $f$ is a log-log convex function and for each
$\text{expr}_i$ one of the following conditions holds:

- $f$ is increasing in argument $i$ and $\text{expr}_i$ is log-log convex.
- $f$ is decreasing in argument $i$ and $\text{expr}_i$ is log-log concave.
- $\text{expr}_i$ is log-log affine.

A function $f(\text{expr}_1, \text{expr}_2, \ldots, \text{expr}_n)$
is log-log concave if $f$ is a log-log concave function and for each
$\text{expr}_i$ one of the following conditions holds:

- $f$ is increasing in argument $i$ and $\text{expr}_i$ is log-log concave.
- $f$ is decreasing in argument $i$ and $\text{expr}_i$ is log-log convex.
- $\text{expr}_i$ is log-log affine.

A function $f(\text{expr}_1, \text{expr}_2, \ldots, \text{expr}_n)$
is log-log affine if $f$ is a log-log affine function and each
$\text{expr}_i$ is log-log affine.

## DGP Problems

A Problem is constructed from an objective and a list of constraints.
If a problem follows the DGP rules, it is guaranteed to be an LLCP
and solvable by `CVXR`. The DGP rules require that the problem
objective have one of two forms:

- `Minimize(log-log convex)`
- `Maximize(log-log concave)`

The only valid constraints under the DGP rules are:

- log-log affine `==` log-log affine
- log-log convex `<=` log-log concave
- log-log concave `>=` log-log convex

Here are some examples of DGP problems:

```{r}
# DGP requires Variables to be declared positive via `pos=TRUE`.
x <- Variable(pos = TRUE)
y <- Variable(pos = TRUE)
z <- Variable(pos = TRUE)

objective_fn <- x * y * z
constraints <- list(
  4 * x * y * z + 2 * x * z <= 10,
  x <= 2 * y,
  y <= 2 * x,
  z >= 1
)
problem <- Problem(Maximize(objective_fn), constraints)
cat("Is problem DGP?", is_dgp(problem), "\n")
```

You can solve a DGP Problem by calling `psolve()` with `gp = TRUE`:

```{r}
result <- psolve(problem, gp = TRUE)
check_solver_status(problem)
cat("Optimal value:", result, "\n")
cat("x:", value(x), "\n")
cat("y:", value(y), "\n")
cat("z:", value(z), "\n")
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(result = result, x = value(x), y = value(y), z = value(z)),
            file = file.path(testdata_dir, "dgp_tutorial2.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    saved <- readRDS(file = file.path(testdata_dir, "dgp_tutorial2.RDS"))
    expect_identical(result, saved$result)
    expect_identical(value(x), saved$x)
    expect_identical(value(y), saved$y)
    expect_identical(value(z), saved$z)
}
```

## DGP Atoms

The DGP atom library includes atomic functions with known log-log
curvature and monotonicity. `CVXR` uses the function information and
the DGP rules to mark expressions with a log-log curvature. Note that
every DGP expression is positive.

### Infix Operators

The infix operators `+`, `*`, `/` are treated as atoms. The operators
`*` and `/` are log-log affine functions. The operator `+` is log-log
convex in both its arguments.

### Scalar Functions

| Function | Log-Log Curvature | Monotonicity |
|---|---|---|
| `geo_mean(x)` | log-log affine | increasing |
| `sum_entries(X)` | log-log convex | increasing |
| `prod_entries(X)` | log-log affine | increasing |
| `pf_eigenvalue(X)` | log-log convex | increasing |

### Elementwise Functions

| Function | Log-Log Curvature | Monotonicity |
|---|---|---|
| `exp(x)` | log-log convex | increasing |
| `log(x)` (domain $x > 1$) | log-log concave | increasing |
| `power(x, p)` | log-log affine | incr. for $p > 0$; decr. for $p < 0$ |
| `sqrt(x)` | log-log affine | increasing |
| `multiply(x, y)` | log-log affine | increasing |
| `maximum(x, y)` | log-log convex | increasing |
| `minimum(x, y)` | log-log concave | increasing |
| `one_minus_pos(x)` | log-log concave | decreasing |

### Matrix Functions

| Function | Log-Log Curvature | Monotonicity |
|---|---|---|
| `eye_minus_inv(X)` | log-log convex | increasing |

For a complete reference on DGP, see the paper at
<https://web.stanford.edu/~boyd/papers/dgp.html>.

## Session Info

```{r}
sessionInfo()
```

## References

- Agrawal, A., Diamond, S., Boyd, S. (2019). Disciplined Geometric
  Programming. _Optimization Letters_, 13(5), 961--976.
