---
title: Isotonic Regression
author: Anqi Fu and Balasubramanian Narasimhan
date: '2017-11-01'
slug: cvxr_isotonic-regression
bibliography: ../bibtex/cvxr_refs.bib
link-citations: true
categories: []
tags: []
---

## Introduction

[Isotonic regression](https://en.wikipedia.org/wiki/Isotonic_regression) is
regression with monotonic constraints. There are several packages in R
to fit isotonic regression models. In this example, we
consider [`isotone`](https://cran.r-project.org/package=isotone) which
uses a pooled-adjacent-violators algorithm (PAVA) and active set
methods to perform the fit.

## Pituitary Data Example

We will use data from the `isotone` package (see @isotone) on the size
of pituitary fissures for 11 subjects between 8 and 14 years of age.

```{r}
library(isotone)
data("pituitary")
str(pituitary)
```

Since the size is expected to increase with age, an isotonic fit is suggested.

```{r}
res_p <- with(pituitary, gpava(age, size))
```

The `cvxr` formulation expresses this pretty much in the mathematical form. 

```{r}
suppressMessages(suppressWarnings(library(cvxr)))
x_p <- with(pituitary, {
      n <- length(size)
      x <- Variable(n)
      objective <- Minimize(p_norm(size - x, 2))
      constraint <- list(diff(x) >= 0)
      problem <- Problem(objective, constraint)
      result <- solve(problem)
      result$getValue(x)
})
```

We define a variable `x` of size `n`, the number of observations. The objective to be minimized is the least-squares error (`Pnorm`). The constraint specifies monotonicity and it uses the `diff` function familiar to R users (see `?diff`), except that it now refers to elements of the `x`. A problem is created and called via `solve`. The resulting optimal value for `x` is retrieved using `getValue(x)`. 

As the output below shows, the results are very close.
```{r}
cbind(res_p$x, x_p)
```

## Handling Ties

Package `isotone` provides additional methods for handling tied data besides the default `ties = "primary"` method; `ties = "secondary"` enforces equality within ties, and `ties = "tertiary"`
enforces monotonicity on the means. (The latter may cause individual fits to be non-monotonic.)

```{r}
res_s <- with(pituitary, gpava(age, size, ties = "secondary"))
res_t <- with(pituitary, gpava(age, size, ties = "tertiary"))
```

In `cvxr`, the secondary method just requires an additional constraint to enforce
equality within tied values; no other modification is necessary. We do that below by figuring out the tied observation indices using `base::split` and force those `x` values to be equal (i.e. `diff` == 0).

```{r}
x_s <- with(pituitary, {
    n <- length(size)
    x <- Variable(n)
    objective <- Minimize(p_norm(size - x, 2))
    secondary_constraints <- lapply(base::split(x = seq_len(n),
                                                f = age),
                                    function(i) diff(x[i]) == 0)
    constraint <- c(diff(x) >= 0,
                    secondary_constraints)
    problem <- Problem(objective, constraint)
    solve(problem)$getValue(x)
})
```

Finally, the tertiary method requires computing the block means to be computed for use
in enforcing monotonicity. 
```{r}
x_t <- with(pituitary, {
    n <- length(size)
    x <- Variable(n)
    objective <- Minimize(p_norm(size - x, 2))
    blocks <- base::split(x = seq_len(n),
                          f = pituitary$age)
    block_means <- lapply(blocks, function(i) {
        v <- numeric(n)
        v[i] <- 1.0 / length(i)
        matrix(v, nrow = 1) %*% x
    })
    block_mean_vector <- do.call(vstack, block_means)
    constraint <- list(diff(block_mean_vector) >= 0)
    problem <- Problem(objective, constraint)
    solve(problem)$getValue(x)
})
```
In the above, we make use of the `VStack` function to create a single vector of the block means.

We can how all three solutions compare.

```{r, echo = FALSE}
library(kableExtra)
d <- data.frame(res_p$x, x_p, res_s$x, x_s, res_t$x, x_t)
names(d) <- c("Isotone (P)", "CVXR (P)", "Isotone (S)", "CVXR(S)",
              "Isotone (T)", "CVXR (T)")
knitr::kable(d, digits = 2, format = "html", caption = "P = primary, S = Secondary, T = Tertiary") %>%
    column_spec(1, background = "#ececec", ) %>%
    column_spec(2, background = "#ececec") %>%
    column_spec(3, background = "#ececec") %>%
    column_spec(4, background = "#ececec") %>%
    column_spec(5, background = "#ececec") %>%
    column_spec(6, background = "#ececec")
    
```

## Notes

Other arbitrary desired behavior can be specified for tied
observations using the above as a guide.

## References
