---
title: Kelly Gambling
author: Anqi Fu and Balasubramanian Narasimhan
date: '2017-11-02'
slug: cvxr_kelly-strategy
bibliography: ../bibtex/cvxr_refs.bib
link-citations: true
categories: []
tags: []
---

## Introduction

In Kelly gambling [@kelly], we are given the opportunity to bet on $n$
possible outcomes, which yield a random non-negative return of $r \in
{\mathbf R}_+^n$. The return $r$ takes on exactly $K$ values
$r_1,\ldots,r_K$ with known probabilities $\pi_1,\ldots,\pi_K$. This
gamble is repeated over $T$ periods. In a given period $t$, let $b_i
\geq 0$ denote the fraction of our wealth bet on outcome $i$. Assuming
the $n$th outcome is equivalent to not wagering (it returns one with
certainty), the fractions must satisfy $\sum_{i=1}^n b_i = 1$. Thus,
at the end of the period, our cumulative wealth is $w_t =
(r^Tb)w_{t-1}$. Our goal is to maximize the average growth rate with
respect to $b \in {\mathbf R}^n$: 


\begin{array}{ll}
\underset{b}{\mbox{maximize}} & \sum_{j=1}^K \pi_j\log(r_j^Tb) \\
\mbox{subject to} & b \geq 0, \quad \sum_{i=1}^n b_i = 1.  
\end{array}


## Example

We solve the Kelly gambling problem for $K = 100$ and $n = 20$. The
probabilities $\pi_j \sim \mbox{Uniform}(0,1)$, and the potential
returns $r_j \sim \mbox{Uniform}(0.5,1.5)$ except for $r_n = {\mathbf 1}$,
which represents the payoff from not wagering. With an initial wealth
of $w_0 = 1$, we simulate the growth trajectory of our Kelly optimal
bets over $P = 100$ periods, assuming returns are i.i.d. over time. In
the following code, `rets` is the $K \times n$ matrix of possible
returns with row $r_i$, while `ps` is the vector of return
probabilities $(\pi_1,\ldots,\pi_K)$.


```{r}
suppressMessages(suppressWarnings(library(CVXR)))
set.seed(1)
n <- 20      # Total bets
K <- 100     # Number of possible returns
PERIODS <- 100
TRIALS <- 5

## Generate return probabilities
ps <- runif(K)
ps <- ps/sum(ps)

## Generate matrix of possible returns
rets <- runif(K*(n-1), 0.5, 1.5)
shuff <- sample(1:length(rets), size = length(rets), replace = FALSE)
rets[shuff[1:30]] <- 0    # Set 30 returns to be relatively low
rets[shuff[31:60]] <- 5     # Set 30 returns to be relatively high
rets <- matrix(rets, nrow = K, ncol = n-1)
rets <- cbind(rets, rep(1, K))   # Last column represents not betting

## Solve for Kelly optimal bets
b <- Variable(n)
obj <- Maximize(t(ps) %*% log(rets %*% b))
constraints <- list(sum(b) == 1, b >= 0)
prob <- Problem(obj, constraints)
result <- solve(prob)
bets <- result$getValue(b)

## Naive betting scheme: bet in proportion to expected return
bets_cmp <- matrix(0, nrow = n)
bets_cmp[n] <- 0.15                  # Hold 15% of wealth
rets_avg <- ps %*% rets
## tidx <- order(rets_avg[-n], decreasing = TRUE)[1:9]
tidx <- 1:(n-1)
fracs <- rets_avg[tidx]/sum(rets_avg[tidx])
bets_cmp[tidx] <- fracs*(1-bets_cmp[n])

## Calculate wealth over time
wealth <- matrix(0, nrow = PERIODS, ncol = TRIALS)
wealth_cmp <- matrix(0, nrow = PERIODS, ncol = TRIALS)
for(i in seq_len(TRIALS)) {
    sidx <- sample(K, size = PERIODS, replace = TRUE, prob = ps)
    winnings <- rets[sidx,] %*% bets
    wealth[,i] <- cumprod(winnings)
    
    winnings_cmp <- rets[sidx,] %*% bets_cmp
    wealth_cmp[,i] <- cumprod(winnings_cmp)
}
```

Growth curves for five independent trials are plotted in the figures
below. Red lines represent the wealth each period from
the Kelly bets, while cyan lines are the result of the naive
bets. Clearly, Kelly optimal bets perform better, producing greater
net wealth by the final period. 

```{r}
library(ggplot2)
library(tidyr)
df <- data.frame(seq_len(PERIODS), wealth)
names(df) <- c("x", paste0("kelly", seq_len(TRIALS)))
plot.data1 <- gather(df, key = "trial", value = "wealth",
                     paste0("kelly", seq_len(TRIALS)),
                     factor_key = TRUE)
plot.data1$Strategy <- "Kelly Optimal Bets"
df <- data.frame(seq_len(PERIODS), wealth_cmp)
names(df) <- c("x", paste0("naive", seq_len(TRIALS)))
plot.data2 <- gather(df, key = "trial", value = "wealth",
                     paste0("naive", seq_len(TRIALS)),
                     factor_key = TRUE)
plot.data2$Strategy <- "Naive Bets"
plot.data <- rbind(plot.data1, plot.data2)
ggplot(data = plot.data) +
    geom_line(mapping = aes(x = x, y = wealth, group = trial, color = Strategy)) +
    scale_y_log10() +
    labs(x = "Time", y = "Wealth") +
    theme(legend.position = "top")
```

## Extensions

As observed in some trajectories above, wealth tends to drop by a
significant amount before increasing eventually. One way to reduce
this drawdown risk is to add a convex constraint as described in 
@RCK [5.3]

$$
\log\left(\sum_{j=1}^K \exp(\log\pi_j - \lambda
\log(r_j^Tb))\right) \leq 0 
$$

where $\lambda \geq 0$ is the risk-aversion parameter. With `CVXR`,
this can be accomplished in a single line using the
`log_sum_exp` atom. Other extensions like wealth goals, betting
restrictions, and VaR/CVaR bounds are also readily incorporated.

## References
