---
title: Largest Ball in a Polyhedron in 2D
author: Anqi Fu and Balasubramanian Narasimhan
date: '2017-11-02'
slug: cvxr_2d_ball
bibliography: ../bibtex/cvxr_refs.bib
link-citations: true
categories: []
tags: []
---

## Problem

The following is a problem from @BoydVandenberghe:2004, section 4.3.1.

Find the largest Euclidean ball (i.e. its center and radius) that lies
in a polyhedron described by affine inequalites:

$$
P = {x : a_i'*x <= b_i, i=1,...,m} 
$$

where x is in ${\mathbf R}^2$.

We define variables that determine the polyhedron.

```{r}
a1 <- matrix(c(2,1))
a2 <- matrix(c(2,-1))
a3 <- matrix(c(-1,2))
a4 <- matrix(c(-1,-2))
b <- rep(1,4)
```

Next, we formulate the `CVXR` problem.

```{r}
suppressMessages(suppressWarnings(library(CVXR)))
r <- Variable(name = "radius")
x_c <- Variable(2, name = "center")
obj <- Maximize(r)
constraints <- list(
    t(a1) %*% x_c + p_norm(a1, 2) * r <= b[1],
    t(a2) %*% x_c + p_norm(a2, 2) * r <= b[2],
    t(a3) %*% x_c + p_norm(a3, 2) * r <= b[3],
    t(a4) %*% x_c + p_norm(a4, 2) * r <= b[4]
)
p <- Problem(obj, constraints)
```

All that remains is to solve the problem and read off the solution.

```{r}
result <- solve(p)
radius <- result$getValue(r)
center <- result$getValue(x_c)
cat(sprintf("The radius is %0.5f for an area %0.5f\n", radius, pi * radius^2))    
```

## A Plot

```{r, fig.width=6, fig.height=6}
library(ggplot2)
library(ggforce)
ggplot() +
    geom_abline(slope = -a1[1] / a1[2], intercept = b[1] / a1[2]) +
    geom_abline(slope = -a2[1] / a2[2], intercept = b[2] / a2[2]) +
    geom_abline(slope = -a3[1] / a3[2], intercept = b[3] / a3[2]) +
    geom_abline(slope = -a4[1] / a4[2], intercept = b[4] / a4[2]) +
    geom_circle(mapping = aes(x0 = center[1], y0 = center[2], r = radius), color = "blue") +
    geom_point(mapping = aes(x = center[1], y = center[2]), color = "red", size = 2) +
    geom_line(mapping = aes(x = c(center[1], center[1] - radius), y = c(center[2], center[2])),
              arrow = arrow(length = unit(0.03, "npc"), ends = "first", type = "closed"),
              color = "brown") +
    annotate("text", x = -0.2, y = 0.04, label = sprintf("r = %0.5f", radius)) +
    labs(x = "x", y = "y") +
    xlim(-1, 1) + ylim(-1, 1)
```

## References
