---
title: CVXR Functions
author: Anqi Fu and Balasubramanian Narasimhan
---

## Functions

Here we describe the functions that can be applied to `CVXR`
expressions. `CVXR` uses the function information in this section and
the [Disciplined Convex Programming tools](http://dcp.stanford.edu/)
to mark expressions with a sign and curvature.

### Operators

The infix operators `+`, `-`, `*`, `%*%`, `/` are treated as functions. `+`
and `-` are affine functions. `*` and `/` are affine in `CVXR` because
`expr1*expr2` and `expr1 %*% expr2` are allowed only when one of the expressions is constant
and `expr1/expr2` is allowed only when `expr2` is a scalar constant.

#### Comparison operators

The comparison operators `==`, `<=`, and `>=` are overloaded to create
constraint objects:

```{r, eval = FALSE}
x == y   # Equality constraint
x <= y   # Inequality constraint (convex <= concave)
x >= y   # Inequality constraint (concave >= convex)
```

Note that strict inequalities `<` and `>` are **not** supported, nor
is `!=`.

#### Semidefinite operators

The operators `%>>%` and `%<<%` create positive semidefinite (PSD) and
negative semidefinite (NSD) constraints, respectively:

```{r, eval = FALSE}
X %>>% 0   # X is PSD
X %<<% 0   # X is NSD
```

::: {.callout-warning}
Beware of operator precedence: `A - B %>>% 0` is parsed as
`A - (B %>>% 0)`.  Use `(A - B) %>>% 0` instead.
:::

#### Boolean operators

The operators `&`, `|`, and `!` are overloaded for boolean variables
to create logical constraints:

```{r, eval = FALSE}
x & y    # equivalent to And(x, y)
x | y    # equivalent to Or(x, y)
!x       # equivalent to Not(x)
```

#### R math and summary functions

`CVXR` overloads many standard R functions so they can be applied
directly to expressions:

- **Math**: `abs()`, `exp()`, `log()`, `sqrt()`, `cumsum()`,
  `ceiling()`, `floor()`, `log2()`, `log10()`, `log1p()`
- **Summary**: `sum()`, `max()`, `min()`, `prod()`
- **Complex**: `Re()`, `Im()`, `Conj()`

These dispatch to the corresponding `CVXR` atoms (e.g., `abs(x)`
calls `Abs`, `sum(x)` calls `SumEntries`).

#### Indexing and slicing

All non-scalar expressions can be indexed using the syntax
`expr[i, j]`. Indexing is an affine function. The syntax `expr[i]`
can be used as a shorthand for `expr[i, 1]` when `expr` is a column
vector. Similarly, `expr[i]` is shorthand for `expr[1, i]` when
`expr` is a row vector.

Non-scalar expressions can also be sliced into using the standard R
slicing syntax. For example, `expr[i:j, r]` selects rows `i`
through `j` of column `r` and returns a vector.

`CVXR` supports advanced indexing using lists of indices or boolean
arrays. The semantics are the same as in R. Any time R might return a
numeric vector, `CVXR` returns a column vector.

#### Transpose

The transpose of any expression can be obtained using the syntax
`t(expr)`. Transpose is an affine function.

#### Power

For any `CVXR` expression `expr`, the power operator `expr^p` is
equivalent to the function `power(expr, p)`.

### Scalar functions

A scalar function takes one or more scalars, vectors, or matrices as arguments
and returns a scalar.

```{r, echo = FALSE, message = FALSE}
library(kableExtra)
```

#### Norms

```{r, echo = FALSE}
d <- data.frame(
    Function = c(
        "<code>norm1(x)</code>",
        "<code>norm2(x)</code>",
        "<code>norm_inf(x)</code>",
        "<code>p_norm(x, p)</code>, \\(p \\geq 1\\)",
        "<code>p_norm(x, p)</code>, \\(0 < p < 1\\)",
        "<code>cvxr_norm(X, &quot;fro&quot;)</code>",
        "<code>cvxr_norm(X, &quot;nuc&quot;)</code>",
        "<code>norm_nuc(X)</code>",
        "<code>cvxr_norm(X, 2)</code>",
        "<code>sigma_max(X)</code>",
        "<code>mixed_norm(X, p, q)</code>"
    ),
    Meaning = c(
        "\\(\\sum_i |x_i|\\)",
        "\\(\\sqrt{\\sum_i x_i^2}\\)",
        "\\(\\max_i |x_i|\\)",
        "\\(\\left(\\sum_i |x_i|^p\\right)^{1/p}\\)",
        "\\(\\left(\\sum_i |x_i|^p\\right)^{1/p}\\)",
        "\\(\\sqrt{\\sum_{ij} X_{ij}^2}\\)",
        "\\(\\sum_i \\sigma_i(X)\\)",
        "\\(\\sum_i \\sigma_i(X)\\)",
        "\\(\\sigma_{\\max}(X)\\)",
        "\\(\\sigma_{\\max}(X)\\)",
        "\\(\\left(\\sum_k \\left(\\sum_j |X_{jk}|^p\\right)^{q/p}\\right)^{1/q}\\)"
    ),
    Curvature = c("convex", "convex", "convex", "convex", "concave",
                  "convex", "convex", "convex", "convex", "convex", "convex"),
    Sign = c(rep("\\(\\geq 0\\)", 11)),
    Domain = c(rep("", 4), "\\(x \\geq 0\\)", "", "", "", "", "", ""),
    stringsAsFactors = FALSE
)
knitr::kable(d, "html", escape = FALSE,
             col.names = c("Function", "Meaning", "Curvature", "Sign", "Domain")) |>
    kable_styling("striped") |>
    column_spec(1:5, background = "#ececec")
```

#### Aggregation and reduction

```{r, echo = FALSE}
d <- data.frame(
    Function = c(
        "<code>sum_entries(X)</code>",
        "<code>sum_squares(x)</code>",
        "<code>sum_largest(x, k)</code>",
        "<code>sum_smallest(x, k)</code>",
        "<code>max_entries(x)</code>",
        "<code>min_entries(x)</code>",
        "<code>prod_entries(x)</code>",
        "<code>cvxr_mean(x)</code>",
        "<code>cvxr_var(x)</code>",
        "<code>cvxr_std(x)</code>",
        "<code>ptp(x)</code>",
        "<code>cvar(x, &beta;)</code>",
        "<code>dotsort(x, w)</code>"
    ),
    Meaning = c(
        "\\(\\sum_{ij} X_{ij}\\)",
        "\\(\\sum_i x_i^2\\)",
        "sum of \\(k\\) largest \\(x_i\\)",
        "sum of \\(k\\) smallest \\(x_i\\)",
        "\\(\\max_i x_i\\)",
        "\\(\\min_i x_i\\)",
        "\\(\\prod_i x_i\\)",
        "\\(\\frac{1}{n}\\sum_i x_i\\)",
        "\\(\\frac{1}{n}\\sum_i (x_i - \\bar{x})^2\\)",
        "\\(\\sqrt{\\frac{1}{n}\\sum_i (x_i - \\bar{x})^2}\\)",
        "\\(\\max(x) - \\min(x)\\)",
        "conditional value-at-risk at level \\(\\beta\\)",
        "\\(w_{[1]}^T x_{[1]} + \\cdots + w_{[n]}^T x_{[n]}\\) (sorted)"
    ),
    Curvature = c("affine", "convex", "convex", "concave", "convex",
                  "concave", "log-log concave", "affine", "convex", "convex",
                  "convex", "convex", "convex"),
    Sign = c("unknown", "\\(\\geq 0\\)", "unknown", "unknown", "unknown",
             "unknown", "\\(\\geq 0\\)", "unknown", "\\(\\geq 0\\)",
             "\\(\\geq 0\\)", "\\(\\geq 0\\)", "unknown", "unknown"),
    Domain = c("", "", "\\(k \\in \\{1,\\ldots,n\\}\\)", "\\(k \\in \\{1,\\ldots,n\\}\\)",
               "", "", "\\(x > 0\\)", "", "", "", "",
               "\\(\\beta \\in (0, 1]\\)", "\\(w\\) constant"),
    stringsAsFactors = FALSE
)
knitr::kable(d, "html", escape = FALSE,
             col.names = c("Function", "Meaning", "Curvature", "Sign", "Domain")) |>
    kable_styling("striped") |>
    column_spec(1:5, background = "#ececec")
```

#### Matrix and quadratic functions

```{r, echo = FALSE}
d <- data.frame(
    Function = c(
        "<code>matrix_trace(X)</code>",
        "<code>log_det(X)</code>",
        "<code>lambda_max(X)</code>",
        "<code>lambda_min(X)</code>",
        "<code>lambda_sum_largest(X, k)</code>",
        "<code>lambda_sum_smallest(X, k)</code>",
        "<code>matrix_frac(x, P)</code>",
        "<code>quad_form(x, P)</code>",
        "<code>quad_over_lin(x, y)</code>",
        "<code>tr_inv(X)</code>",
        "<code>geo_mean(x)</code>",
        "<code>harmonic_mean(x)</code>",
        "<code>log_sum_exp(x)</code>",
        "<code>scalar_product(x, c)</code>",
        "<code>vdot(x, y)</code>",
        "<code>tv(x)</code> / <code>total_variation(x)</code>",
        "<code>inv_prod(x)</code>",
        "<code>perspective(f, s)</code>"
    ),
    Meaning = c(
        "\\(\\mathrm{tr}(X)\\)",
        "\\(\\log \\det(X)\\)",
        "\\(\\lambda_{\\max}(X)\\)",
        "\\(\\lambda_{\\min}(X)\\)",
        "sum of \\(k\\) largest eigenvalues of \\(X\\)",
        "sum of \\(k\\) smallest eigenvalues of \\(X\\)",
        "\\(x^T P^{-1} x\\)",
        "\\(x^T P x\\)",
        "\\(\\|x\\|_2^2 / y\\)",
        "\\(\\mathrm{tr}(X^{-1})\\)",
        "\\(\\left(\\prod_i x_i\\right)^{1/n}\\)",
        "\\(\\frac{n}{\\sum_i 1/x_i}\\)",
        "\\(\\log\\left(\\sum_i e^{x_i}\\right)\\)",
        "\\(c^T x\\) where \\(c\\) is constant",
        "\\(x^T y\\)",
        "\\(\\sum_i |x_{i+1} - x_i|\\)",
        "\\(1 / \\prod_i x_i\\)",
        "\\(s \\cdot f(x/s)\\)"
    ),
    Curvature = c("affine", "concave", "convex", "concave",
                  "convex", "concave", "convex",
                  "convex (\\(P \\succeq 0\\))", "convex",
                  "convex", "concave", "concave",
                  "convex", "affine", "affine",
                  "convex", "convex", "convex"),
    Sign = c("unknown", "unknown", "unknown", "unknown",
             "unknown", "unknown", "\\(\\geq 0\\)",
             "\\(\\geq 0\\)", "\\(\\geq 0\\)",
             "\\(\\geq 0\\)", "\\(\\geq 0\\)", "\\(\\geq 0\\)",
             "unknown", "unknown", "unknown",
             "\\(\\geq 0\\)", "\\(\\geq 0\\)", "varies"),
    Domain = c("", "\\(X \\in \\mathbf{S}^n_{++}\\)", "\\(X \\in \\mathbf{S}^n\\)",
               "\\(X \\in \\mathbf{S}^n\\)", "\\(X \\in \\mathbf{S}^n\\)",
               "\\(X \\in \\mathbf{S}^n\\)",
               "\\(P \\succ 0\\)", "\\(P \\succeq 0\\) or \\(P \\preceq 0\\)",
               "\\(y > 0\\)", "\\(X \\succ 0\\)",
               "\\(x \\geq 0\\)", "\\(x > 0\\)",
               "", "", "", "", "\\(x > 0\\)",
               "\\(s > 0\\)"),
    stringsAsFactors = FALSE
)
knitr::kable(d, "html", escape = FALSE,
             col.names = c("Function", "Meaning", "Curvature", "Sign", "Domain")) |>
    kable_styling("striped") |>
    column_spec(1:5, background = "#ececec")
```

#### Clarifications

The domain $\mathbf{S}^n$ refers to the set of symmetric matrices. The
domains $\mathbf{S}^n_+$ and $\mathbf{S}^n_-$ refer to the set of
positive semi-definite and negative semi-definite matrices,
respectively. Similarly, $\mathbf{S}^n_{++}$ and $\mathbf{S}^n_{--}$
refer to the set of positive definite and negative definite matrices,
respectively.

For a vector expression `x`, `norm2(x)` gives the
Euclidean norm. For a matrix expression `X`, `cvxr_norm(X, 2)` and
`sigma_max(X)` give the spectral norm (largest singular value).

The function `cvxr_norm(X, "fro")` is called
the
[Frobenius norm](https://en.wikipedia.org/wiki/Matrix_norm#Frobenius_norm) and
`cvxr_norm(X, "nuc")` (or `norm_nuc(X)`)
the
[nuclear norm](https://en.wikipedia.org/wiki/Matrix_norm#Schatten_norms). The
nuclear norm can also be defined as the sum of `X`'s singular values.

The functions `max_entries` and `min_entries` give the largest and
smallest entry, respectively, in a single expression. These functions
should not be confused with `max_elemwise` and `min_elemwise` (see
[Elementwise functions](#elementwise-functions)). Use `max_elemwise`
and `min_elemwise` to find the maximum or minimum of a list of scalar
expressions.

The function `sum_entries` sums all the entries in a single
expression. The built-in R `sum` should be used to add together a
list of expressions. For example, the following code sums
three expressions:

```{r, eval = FALSE}
expr_sum <- sum(expr1, expr2, expr3)
```

### Functions along an axis

The functions `sum_entries`, `cvxr_norm`, `max_entries`, and `min_entries`
can be applied along an axis.  Given an `m` by `n` expression `expr`,
the syntax `func(expr, axis=1)` applies `func` to each row,
returning a `m` by 1 expression.  The syntax `func(expr, axis=2)`
applies `func` to each column, returning a 1 by `n` expression. For
example, the following code sums along the columns and rows of a
matrix variable:

```{r, eval = FALSE}
X <- Variable(c(5, 4))
row_sums <- sum_entries(X, axis=1) # Has size (5, 1)
col_sums <- sum_entries(X, axis=2) # Has size (1, 4)
```

Note that the use of `axis` differs from its use in `CVXPY` where
`axis=2` implies the columns. In `CVXR`, we align our implementation
with the `base::apply` function. The default in most cases is `axis =
NA`, which treats a matrix as one long vector, basically the same as
apply with `c(1,2)`. The exception is `cumsum_axis` (see below), which cannot take
`axis = NA`; it will throw an error.

### Elementwise functions {#elementwise-functions}

These functions operate on each element of their arguments. For
example, if `X` is a 5 by 4 matrix variable, then `abs(X)` is a 5 by 4
matrix expression. `abs(X)[1, 2]` is equivalent to `abs(X[1, 2])`.

Elementwise functions that take multiple arguments, such as
`max_elemwise` and `multiply`, operate on the corresponding
elements of each argument.  For example, if `X` and `Y` are both 3 by
3 matrix variables, then `max_elemwise(X, Y)` is a 3 by 3 matrix
expression. `max_elemwise(X, Y)[2, 1]` is equivalent to
`max_elemwise(X[2, 1], Y[2, 1])`. This means all arguments must have
the same dimensions or be scalars, which are promoted.

```{r, echo = FALSE}
d <- data.frame(
    Function = c(
        "<code>abs(x)</code>",
        "<code>entr(x)</code>",
        "<code>exp(x)</code>",
        "<code>huber(x, M)</code>",
        "<code>inv_pos(x)</code>",
        "<code>kl_div(x, y)</code>",
        "<code>log(x)</code>",
        "<code>log1p_atom(x)</code>",
        "<code>log_normcdf(x)</code>",
        "<code>logistic(x)</code>",
        "<code>loggamma(x)</code>",
        "<code>max_elemwise(x, y, ...)</code>",
        "<code>min_elemwise(x, y, ...)</code>",
        "<code>multiply(c, x)</code>",
        "<code>neg(x)</code>",
        "<code>pos(x)</code>",
        "<code>diff_pos(x, y)</code>",
        "<code>one_minus_pos(x)</code>",
        "<code>power(x, 0)</code>",
        "<code>power(x, 1)</code>",
        "<code>power(x, p)</code>, \\(p \\geq 1\\)",
        "<code>power(x, p)</code>, \\(0 < p < 1\\)",
        "<code>power(x, p)</code>, \\(p < 0\\)",
        "<code>rel_entr(x, y)</code>",
        "<code>scalene(x, &alpha;, &beta;)</code>",
        "<code>sqrt(x)</code>",
        "<code>square(x)</code>",
        "<code>xexp(x)</code>"
    ),
    Meaning = c(
        "\\(|x|\\)",
        "\\(-x \\log x\\)",
        "\\(e^x\\)",
        "\\(2(\\sqrt{1+(x/M)^2}-1)M^2\\)",
        "\\(1/x\\)",
        "\\(x \\log(x/y) - x + y\\)",
        "\\(\\log x\\)",
        "\\(\\log(1 + x)\\)",
        "\\(\\log \\Phi(x)\\) (\\(\\Phi\\) = standard normal CDF)",
        "\\(\\log(1 + e^x)\\)",
        "\\(\\log \\Gamma(x)\\)",
        "\\(\\max(x, y, \\ldots)\\)",
        "\\(\\min(x, y, \\ldots)\\)",
        "\\(c \\cdot x\\) (\\(c\\) constant)",
        "\\(\\max(-x, 0)\\)",
        "\\(\\max(x, 0)\\)",
        "\\(\\max(x - y, 0)\\)",
        "\\(\\max(1 - x, 0)\\)",
        "\\(1\\)",
        "\\(x\\)",
        "\\(x^p\\)",
        "\\(x^p\\)",
        "\\(x^p\\)",
        "\\(x \\log(x/y)\\)",
        "\\(\\alpha \\cdot \\text{pos}(x) + \\beta \\cdot \\text{neg}(x)\\)",
        "\\(\\sqrt{x}\\)",
        "\\(x^2\\)",
        "\\(x e^x\\)"
    ),
    Curvature = c(
        "convex", "concave", "convex", "convex", "convex",
        "convex", "concave", "concave", "concave", "convex",
        "convex", "convex", "concave", "affine", "convex",
        "convex", "convex", "convex",
        "constant", "affine", "convex", "concave", "convex",
        "convex",
        "convex",
        "concave", "convex", "convex"
    ),
    Sign = c(
        "\\(\\geq 0\\)", "unknown", "\\(\\geq 0\\)", "\\(\\geq 0\\)", "\\(\\geq 0\\)",
        "\\(\\geq 0\\)", "unknown", "unknown", "\\(\\leq 0\\)", "\\(\\geq 0\\)",
        "unknown", "unknown", "unknown", "unknown", "\\(\\geq 0\\)",
        "\\(\\geq 0\\)", "\\(\\geq 0\\)", "\\(\\geq 0\\)",
        "\\(\\geq 0\\)", "unknown", "\\(\\geq 0\\)", "\\(\\geq 0\\)", "\\(\\geq 0\\)",
        "unknown",
        "\\(\\geq 0\\)",
        "\\(\\geq 0\\)", "\\(\\geq 0\\)", "unknown"
    ),
    Domain = c(
        "", "\\(x > 0\\)", "", "\\(M > 0\\)", "\\(x > 0\\)",
        "\\(x > 0, y > 0\\)", "\\(x > 0\\)", "\\(x > -1\\)", "", "",
        "\\(x > 0\\)", "", "", "", "",
        "", "", "",
        "", "", "\\(x \\geq 0\\)", "\\(x \\geq 0\\)", "\\(x > 0\\)",
        "\\(x > 0, y > 0\\)",
        "\\(\\alpha, \\beta \\geq 0\\)",
        "\\(x \\geq 0\\)", "", ""
    ),
    stringsAsFactors = FALSE
)
knitr::kable(d, "html", escape = FALSE,
             col.names = c("Function", "Meaning", "Curvature", "Sign", "Domain")) |>
    kable_styling("striped") |>
    column_spec(1:5, background = "#ececec")
```

#### DQCP elementwise functions

The following functions are quasiconvex or quasiconcave. They can be
used in DQCP problems (pass `qcp = TRUE` to `psolve()`).

```{r, echo = FALSE}
d <- data.frame(
    Function = c("<code>ceil_expr(x)</code>", "<code>floor_expr(x)</code>"),
    Meaning = c("\\(\\lceil x \\rceil\\)", "\\(\\lfloor x \\rfloor\\)"),
    Curvature = c("quasiconvex", "quasiconcave"),
    Sign = c("unknown", "unknown"),
    Domain = c("", ""),
    stringsAsFactors = FALSE
)
knitr::kable(d, "html", escape = FALSE,
             col.names = c("Function", "Meaning", "Curvature", "Sign", "Domain")) |>
    kable_styling("striped") |>
    column_spec(1:5, background = "#ececec")
```

### Vector/matrix functions

A vector/matrix function takes one or more scalars, vectors, or matrices as arguments
and returns a vector or matrix.

```{r, echo = FALSE}
d <- data.frame(
    Function = c(
        "<code>bmat(list(list(X, Y), ...))</code>",
        "<code>conv(c, x)</code>",
        "<code>cumsum_axis(x, axis)</code>",
        "<code>cummax_expr(x, axis)</code>",
        "<code>cvxr_diff(x)</code>",
        "<code>cvxr_outer(x, y)</code>",
        "<code>diag(x)</code> / <code>diag(X)</code>",
        "<code>hstack(x, y, ...)</code>",
        "<code>kronecker(X, Y)</code>",
        "<code>reshape_expr(x, c(m, n))</code>",
        "<code>upper_tri(X)</code>",
        "<code>vec(X)</code>",
        "<code>vec_to_upper_tri(x)</code>",
        "<code>vstack(x, y, ...)</code>"
    ),
    Meaning = c(
        "block matrix from list of lists",
        "\\(c * x\\) (discrete convolution)",
        "cumulative sum along axis",
        "cumulative max along axis",
        "\\(k\\)th order differences",
        "\\(x y^T\\) (outer product)",
        "diagonal vector \\(\\leftrightarrow\\) diagonal matrix",
        "horizontal stack",
        "\\(X \\otimes Y\\) (Kronecker product)",
        "reshape to \\(m \\times n\\)",
        "upper triangular elements as vector",
        "vectorize (column-major)",
        "vector to upper triangular matrix",
        "vertical stack"
    ),
    Curvature = c(
        "affine", "affine", "affine", "convex", "affine",
        "affine (one arg constant)", "affine", "affine", "affine",
        "affine", "affine", "affine", "affine", "affine"
    ),
    Sign = rep("unknown", 14),
    Domain = c("", "\\(c\\) constant", "axis required", "axis required",
               "", "one arg must be constant", "", "", "one arg must be constant",
               "", "\\(X\\) square", "", "", ""),
    stringsAsFactors = FALSE
)
knitr::kable(d, "html", escape = FALSE,
             col.names = c("Function", "Meaning", "Curvature", "Sign", "Domain")) |>
    kable_styling("striped") |>
    column_spec(1:5, background = "#ececec")
```

#### DGP-specific functions

The following functions are for use in Disciplined Geometric
Programming (DGP) problems. They are log-log convex or log-log
concave.

```{r, echo = FALSE}
d <- data.frame(
    Function = c(
        "<code>pf_eigenvalue(X)</code>",
        "<code>eye_minus_inv(X)</code>",
        "<code>resolvent(X, s)</code>",
        "<code>gmatmul(A, x)</code>"
    ),
    Meaning = c(
        "Perron-Frobenius eigenvalue",
        "\\((I - X)^{-1}\\)",
        "\\((sI - X)^{-1}\\)",
        "generalized matrix multiply (DGP)"
    ),
    Curvature = c("log-log convex", "log-log convex", "log-log convex", "log-log affine"),
    Domain = c("\\(X \\geq 0\\)", "spectral radius \\(< 1\\)",
               "spectral radius \\(< s\\)", "\\(A\\) constant, \\(x > 0\\)"),
    stringsAsFactors = FALSE
)
knitr::kable(d, "html", escape = FALSE,
             col.names = c("Function", "Meaning", "Curvature", "Domain")) |>
    kable_styling("striped") |>
    column_spec(1:4, background = "#ececec")
```

### Boolean logic atoms

The following atoms are used for boolean constraints in mixed-integer
problems. They operate on boolean variables.

```{r, echo = FALSE}
d <- data.frame(
    Function = c("<code>And(x, y)</code>", "<code>Or(x, y)</code>", "<code>Not(x)</code>",
                 "<code>Xor(x, y)</code>", "<code>implies(x, y)</code>", "<code>iff(x, y)</code>"),
    Meaning = c("\\(x \\wedge y\\)", "\\(x \\vee y\\)", "\\(\\neg x\\)",
                "\\(x \\oplus y\\)", "\\(x \\Rightarrow y\\)", "\\(x \\Leftrightarrow y\\)"),
    Domain = c(rep("boolean variables", 6)),
    stringsAsFactors = FALSE
)
knitr::kable(d, "html", escape = FALSE,
             col.names = c("Function", "Meaning", "Domain")) |>
    kable_styling("striped") |>
    column_spec(1:3, background = "#ececec")
```

### Constraint constructors

These create constraint objects directly.

```{r, echo = FALSE}
d <- data.frame(
    Function = c("<code>SOC(t, x)</code>", "<code>PSD(X)</code>", "<code>ExpCone(x, y, z)</code>",
                 "<code>PowCone3D(x, y, z, alpha)</code>", "<code>FiniteSet(x, S)</code>"),
    Meaning = c(
        "\\(\\|x\\|_2 \\leq t\\) (second-order cone)",
        "\\(X \\succeq 0\\) (positive semidefinite)",
        "\\(y e^{x/y} \\leq z,\\; y > 0\\) (exponential cone)",
        "\\(\\|z\\| \\leq x^\\alpha y^{1-\\alpha}\\) (power cone)",
        "\\(x \\in S\\) (finite set membership)"
    ),
    stringsAsFactors = FALSE
)
knitr::kable(d, "html", escape = FALSE,
             col.names = c("Function", "Meaning")) |>
    kable_styling("striped") |>
    column_spec(1:2, background = "#ececec")
```

## Session Info

```{r}
sessionInfo()
```
