---
title: Visualization
author: "Anqi Fu, Balasubramanian Narasimhan, Steven Diamond, Stephen Boyd"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: visualization
---

```{r}
#| label: prereqs
#| echo: false
#| message: false
suppressWarnings(suppressMessages(library(CVXR)))
source(here::here("_check_status.R"))
```

The `visualize()` function generates interactive visualizations of
CVXR optimization problems, showing the expression tree with DCP
curvature annotations. It is useful for understanding how CVXR
decomposes a problem and for diagnosing DCP violations.

## Function Signature

```r
visualize(
  problem,
  output = c("text", "json", "html", "latex", "tikz"),
  solver = NULL,
  digits = 4L,
  file   = NULL,
  open   = interactive(),
  doc_base = "https://cvxr.rbind.io/reference/"
)
```

The key arguments:

- `output`: format for the visualization. `"html"` produces a
  self-contained interactive HTML page; `"text"` prints to the console
- `solver`: solver to use for generating standard form and solver
  data. Set to `TRUE` to auto-select a solver, or pass a solver name
  (e.g., `"ECOS"`, `"OSQP"`). When specified, the HTML output
  includes two additional tabs: **Standard Form** and **Solver Data**.
  Omit or set to `NULL` to show only the expression tree and Smith
  forms
- `file`: path to write the output (used with `"html"`)
- `open`: whether to open the file in a browser (default `TRUE` in
  interactive sessions)
- `doc_base`: base URL for documentation links in the visualization

Two helper features for richer output:

- **`latex_name`** argument in `Variable()` and `Parameter()` sets the
  display name in LaTeX-rendered visualizations
- **`set_label(constraint, label)`** attaches a human-readable label
  to a constraint

## Smith Form Notation

The visualization uses $\varphi$ (phi) symbols to represent CVXR
atoms in the _Smith form_ --- the intermediate representation that
shows how each expression is decomposed before conic canonicalization.
Each atom is written as $t_i = \varphi^{\text{tag}}(\text{args})$ in
the Smith form, and relaxed to an inequality ($\geq$ for convex,
$\leq$ for concave) in the relaxed Smith form.

The table below lists the $\varphi$ functions that have dedicated
annotations. Any atom not listed here uses a generic fallback:
$\varphi^{\texttt{ClassName}}$.

### Affine Atoms

These atoms preserve curvature (both convex _and_ concave). Their
Smith form uses equality ($=$) since no relaxation is needed.

| Symbol | CVXR Function | Meaning |
|:------:|:--------------|:--------|
| $\varphi^{+}$ | `+` | Addition |
| $\varphi^{-}$ | `-x` | Negation |
| $\varphi^{\times}$ | `%*%` | Matrix multiplication |
| $\varphi^{\odot}$ | `*` (elementwise) | Hadamard (elementwise) product |
| $\varphi^{\div}$ | `/` | Division by constant |
| $\varphi^{\top}$ | `t()` | Transpose |
| $\varphi^{\text{vec}}$ | `reshape_expr()`, `vec()` | Vectorization / reshape |
| $\varphi^{\Sigma}$ | `sum()`, `sum_entries()` | Sum of entries |

### Convex and Concave Atoms

These atoms have specific curvature. In the relaxed Smith form,
convex atoms use $t_i \geq \varphi(\cdots)$ and concave atoms use
$t_i \leq \varphi(\cdots)$.

| Symbol | CVXR Function | Meaning | Curvature |
|:------:|:--------------|:--------|:---------:|
| $\varphi^{\lvert\cdot\rvert}$ | `abs()` | Absolute value | convex |
| $\varphi^{\lVert\cdot\rVert_2}$ | `p_norm(x, 2)` | Euclidean norm | convex |
| $\varphi^{\lVert\cdot\rVert_p}$ | `p_norm(x, p)` | General $p$-norm | convex ($p \geq 1$) |
| $\varphi^{(\cdot)^2}$ | `power(x, 2)`, `x^2` | Square | convex |
| $\varphi^{(\cdot)^p}$ | `power(x, p)`, `x^p` | General power | varies with $p$ |
| $\varphi^{\text{qol}}$ | `quad_over_lin()` | $\lVert x\rVert_2^2 / y$ | convex |
| $\varphi^{\text{quad}}$ | `quad_form(x, P)` | $x^\top P x$ | convex ($P \succeq 0$) |
| $\varphi^{\max}$ | `max_entries()`, `maximum()` | Maximum | convex |
| $\varphi^{\min}$ | `min_entries()`, `minimum()` | Minimum | concave |

### Leaves and Fallback

| Symbol | Node Type | Description |
|:------:|:----------|:------------|
| $\mathbf{x}$ | `Variable` | Optimization variable |
| $\theta_{\text{name}}$ | `Parameter` | Named parameter (value set at solve time) |
| literal value | `Constant` | Numeric constant (shown as value or $C_{(m,n)}$) |
| $\varphi^{\texttt{ClassName}}$ | any other `Atom` | Generic fallback for atoms without custom annotation |

Atoms that appear in the examples on this page but use the generic
fallback include `entr()`, `huber()`, `log_sum_exp()`, `lambda_max()`,
and `pos()`.

## DCP-Compliant Examples

The examples below show `visualize()` output for problems that
satisfy the DCP rules. Each tree node is annotated with its curvature
and sign using color-coded borders: blue for convex, red for concave,
green for affine, and grey for constant. These colors indicate
curvature, not validity --- a red border on a concave node is
expected, not an error. DCP violations are shown differently, with a
red background and a banner (see the [Non-DCP examples](#non-dcp-examples)
below).

### Simple Norm Minimization

$$
\text{minimize} \; \|x\|_2 \quad \text{subject to} \quad x \geq 1
$$

A basic second-order cone program. The objective `p_norm(x, 2)` is
convex and the constraint is affine, so the problem is DCP-compliant.

```{r}
#| label: ex1-build
x <- Variable(3, name = "x")
prob1 <- Problem(Minimize(p_norm(x, 2)), list(x >= 1))
is_dcp(prob1)
```

```{r}
#| label: ex1-viz
#| echo: false
#| results: hide
visualize(prob1, output = "html", solver = TRUE,
          file = "viz/simple-norm.html", open = FALSE)
```

<iframe src="viz/simple-norm.html" width="100%" height="550" style="border: 1px solid #dee2e6; border-radius: 6px;" loading="lazy"></iframe>

### Least Squares

$$
\text{minimize} \; \|Ax - b\|_2^2
$$

The classic unconstrained least squares problem. `sum_squares()` is a
convex atom applied to an affine argument.

```{r}
#| label: ex2-build
set.seed(42)
A <- matrix(rnorm(12), 4, 3)
b <- rnorm(4)
x2 <- Variable(3, name = "x")
prob2 <- Problem(Minimize(sum_squares(A %*% x2 - b)))
is_dcp(prob2)
```

```{r}
#| label: ex2-viz
#| echo: false
#| results: hide
visualize(prob2, output = "html", solver = TRUE,
          file = "viz/least-squares.html", open = FALSE)
```

<iframe src="viz/least-squares.html" width="100%" height="550" style="border: 1px solid #dee2e6; border-radius: 6px;" loading="lazy"></iframe>

### Portfolio Optimization

$$
\text{maximize} \; \boldsymbol{\mu}^T \mathbf{w} - \gamma \, \mathbf{w}^T \Sigma \mathbf{w}
\quad \text{subject to} \quad \sum w_i = 1, \quad w \geq 0
$$

A Markowitz portfolio problem with a risk-aversion parameter
$\gamma$. This example demonstrates `latex_name` for display
customization and `set_label()` for named constraints. Note that
parameter values must be set (via `value()<-`) before calling
`visualize()` with `solver = TRUE`, so that the problem data matrices
can be compiled for the Standard Form and Solver Data tabs.

```{r}
#| label: ex3-build
w <- Variable(4, name = "w", latex_name = "\\mathbf{w}")
mu <- Parameter(4, name = "mu", latex_name = "\\boldsymbol{\\mu}")
gamma <- Parameter(1, name = "gamma", nonneg = TRUE, latex_name = "\\gamma")
Sigma <- matrix(c(1,.5,.3,.1, .5,1,.4,.2, .3,.4,1,.5, .1,.2,.5,1), 4, 4)
## Set parameter values so the solver tabs can compile problem data
value(mu) <- c(0.12, 0.10, 0.07, 0.03)
value(gamma) <- 0.5
prob3 <- Problem(
  Maximize(t(mu) %*% w - gamma * quad_form(w, Sigma)),
  list(
    set_label(sum(w) == 1, "budget"),
    set_label(w >= 0, "long only")
  )
)
is_dcp(prob3)
```

```{r}
#| label: ex3-viz
#| echo: false
#| results: hide
visualize(prob3, output = "html", solver = TRUE,
          file = "viz/portfolio.html", open = FALSE)
```

<iframe src="viz/portfolio.html" width="100%" height="550" style="border: 1px solid #dee2e6; border-radius: 6px;" loading="lazy"></iframe>

### Entropy Maximization

$$
\text{maximize} \; \sum \text{entr}(x_i) \quad \text{subject to} \quad \sum x_i = 1
$$

Maximum entropy distribution. The `entr()` atom is concave, so
maximizing a sum of concave atoms is DCP-compliant.

```{r}
#| label: ex4-build
x4 <- Variable(5, name = "x")
prob4 <- Problem(Maximize(sum(entr(x4))), list(sum(x4) == 1))
is_dcp(prob4)
```

```{r}
#| label: ex4-viz
#| echo: false
#| results: hide
visualize(prob4, output = "html", solver = TRUE,
          file = "viz/entropy.html", open = FALSE)
```

<iframe src="viz/entropy.html" width="100%" height="550" style="border: 1px solid #dee2e6; border-radius: 6px;" loading="lazy"></iframe>

### Huber Loss

$$
\text{minimize} \; \sum \text{huber}(x_i)
$$

The Huber function is convex; minimizing its sum is DCP-compliant.
This is an unconstrained problem.

```{r}
#| label: ex5-build
x5 <- Variable(3, name = "x")
prob5 <- Problem(Minimize(sum(huber(x5))))
is_dcp(prob5)
```

```{r}
#| label: ex5-viz
#| echo: false
#| results: hide
visualize(prob5, output = "html", solver = TRUE,
          file = "viz/huber.html", open = FALSE)
```

<iframe src="viz/huber.html" width="100%" height="550" style="border: 1px solid #dee2e6; border-radius: 6px;" loading="lazy"></iframe>

### Log-Sum-Exp

$$
\text{minimize} \; \log\sum e^{x_i} \quad \text{subject to} \quad x \geq -1
$$

`log_sum_exp()` is a convex atom. The constraint is affine.

```{r}
#| label: ex6-build
x6 <- Variable(4, name = "x")
prob6 <- Problem(Minimize(log_sum_exp(x6)), list(x6 >= -1))
is_dcp(prob6)
```

```{r}
#| label: ex6-viz
#| echo: false
#| results: hide
visualize(prob6, output = "html", solver = TRUE,
          file = "viz/log-sum-exp.html", open = FALSE)
```

<iframe src="viz/log-sum-exp.html" width="100%" height="550" style="border: 1px solid #dee2e6; border-radius: 6px;" loading="lazy"></iframe>

### Semidefinite Program

$$
\text{minimize} \; \lambda_{\max}(X) \quad \text{subject to} \quad X_{11} = 1
$$

An SDP with a symmetric matrix variable. `lambda_max()` is convex and
the equality constraint is affine.

```{r}
#| label: ex7-build
X7 <- Variable(c(3, 3), name = "X", symmetric = TRUE)
prob7 <- Problem(Minimize(lambda_max(X7)), list(X7[1, 1] == 1))
is_dcp(prob7)
```

```{r}
#| label: ex7-viz
#| echo: false
#| results: hide
visualize(prob7, output = "html", solver = TRUE,
          file = "viz/sdp.html", open = FALSE)
```

<iframe src="viz/sdp.html" width="100%" height="550" style="border: 1px solid #dee2e6; border-radius: 6px;" loading="lazy"></iframe>

### Mixed Atoms

$$
\text{minimize} \; \max_i |x_i| + \sum \text{pos}(x_i)
$$

A combination of `max_entries(abs(x))` (convex) and
`sum_entries(pos(x))` (convex). The sum of convex functions is convex.

```{r}
#| label: ex8-build
x8 <- Variable(3, name = "x")
prob8 <- Problem(Minimize(max_entries(abs(x8)) + sum_entries(pos(x8))))
is_dcp(prob8)
```

```{r}
#| label: ex8-viz
#| echo: false
#| results: hide
visualize(prob8, output = "html", solver = TRUE,
          file = "viz/mixed-atoms.html", open = FALSE)
```

<iframe src="viz/mixed-atoms.html" width="100%" height="550" style="border: 1px solid #dee2e6; border-radius: 6px;" loading="lazy"></iframe>

## Non-DCP Examples

The following problems violate the DCP rules. The visualization
highlights the offending nodes in red, making it easy to pinpoint
where the violation occurs.

### Bilinear Objective

$$
\text{minimize} \; x_1 x_2 \quad \text{subject to} \quad x \geq 0
$$

::: {.callout-warning}
**DCP violation:** The product $x_1 x_2$ is neither convex nor
concave --- it is bilinear. `visualize()` marks the objective node in
red.
:::

```{r}
#| label: ex9-build
x9 <- Variable(2, name = "x")
prob9 <- Problem(Minimize(x9[1] * x9[2]), list(x9 >= 0))
is_dcp(prob9)
```

```{r}
#| label: ex9-viz
#| echo: false
#| results: hide
visualize(prob9, output = "html",
          file = "viz/bilinear.html", open = FALSE)
```

<iframe src="viz/bilinear.html" width="100%" height="550" style="border: 1px solid #dee2e6; border-radius: 6px;" loading="lazy"></iframe>

### Nonaffine Equality Constraint

$$
\text{minimize} \; \mathbf{1}^T x \quad \text{subject to} \quad \|x\|_2 = 1
$$

::: {.callout-warning}
**DCP violation:** Equality constraints must be affine. Here
$\|x\|_2$ is convex, so the equality $\|x\|_2 = 1$ is not
DCP-compliant. The constraint node is highlighted in red.
:::

```{r}
#| label: ex10-build
x10 <- Variable(3, name = "x")
prob10 <- Problem(Minimize(sum(x10)), list(p_norm(x10, 2) == 1))
is_dcp(prob10)
```

```{r}
#| label: ex10-viz
#| echo: false
#| results: hide
visualize(prob10, output = "html",
          file = "viz/nonaffine-eq.html", open = FALSE)
```

<iframe src="viz/nonaffine-eq.html" width="100%" height="550" style="border: 1px solid #dee2e6; border-radius: 6px;" loading="lazy"></iframe>

## Text Output

For quick inspection in the console, use `output = "text"` (the
default):

```{r}
#| label: text-output
#| collapse: true
visualize(prob1, output = "text")
```

## Practical Considerations

`visualize()` is designed for small to moderate problems --- the kind
you build during modeling and debugging. Problems with many
variables, large data matrices, or hundreds of constraints will
produce unwieldy expression trees and oversized conic forms that
obscure rather than illuminate. Keep the problem small and
representative when using this tool.

## Session Info

```{r}
#| label: sessioninfo
sessionInfo()
```
