---
title: Visualization
author: "Anqi Fu, Balasubramanian Narasimhan, Steven Diamond, Stephen Boyd"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: visualization
---

```{r}
#| label: prereqs
#| echo: false
#| message: false
suppressWarnings(suppressMessages(library(CVXR)))
source(here::here("_check_status.R"))
```

The `visualize()` function generates interactive visualizations of
CVXR optimization problems, showing the expression tree with DCP
curvature annotations. It is useful for understanding how CVXR
decomposes a problem and for diagnosing DCP violations.

## Function Signature

```r
visualize(
  problem,
  output = c("text", "json", "html", "latex", "tikz"),
  digits = 4L,
  file   = NULL,
  open   = interactive(),
  doc_base = "https://cvxr.rbind.io/reference/"
)
```

The key arguments:

- `output`: format for the visualization. `"html"` produces a
  self-contained interactive HTML page; `"text"` prints to the console
- `file`: path to write the output (used with `"html"`)
- `open`: whether to open the file in a browser (default `TRUE` in
  interactive sessions)
- `doc_base`: base URL for documentation links in the visualization

Two helper features for richer output:

- **`latex_name`** argument in `Variable()` and `Parameter()` sets the
  display name in LaTeX-rendered visualizations
- **`set_label(constraint, label)`** attaches a human-readable label
  to a constraint

## DCP-Compliant Examples

The examples below show `visualize()` output for problems that
satisfy the DCP rules. Each tree node is annotated with its curvature
and sign using color-coded borders: blue for convex, red for concave,
green for affine, and grey for constant. These colors indicate
curvature, not validity --- a red border on a concave node is
expected, not an error. DCP violations are shown differently, with a
red background and a banner (see the [Non-DCP examples](#non-dcp-examples)
below).

### Simple Norm Minimization

$$
\text{minimize} \; \|x\|_2 \quad \text{subject to} \quad x \geq 1
$$

A basic second-order cone program. The objective `p_norm(x, 2)` is
convex and the constraint is affine, so the problem is DCP-compliant.

```{r}
#| label: ex1-build
x <- Variable(3, name = "x")
prob1 <- Problem(Minimize(p_norm(x, 2)), list(x >= 1))
is_dcp(prob1)
```

```{r}
#| label: ex1-viz
#| echo: false
#| results: hide
visualize(prob1, output = "html",
          file = "viz/simple-norm.html", open = FALSE)
```

<iframe src="viz/simple-norm.html" width="100%" height="550" style="border: 1px solid #dee2e6; border-radius: 6px;" loading="lazy"></iframe>

### Least Squares

$$
\text{minimize} \; \|Ax - b\|_2^2
$$

The classic unconstrained least squares problem. `sum_squares()` is a
convex atom applied to an affine argument.

```{r}
#| label: ex2-build
set.seed(42)
A <- matrix(rnorm(12), 4, 3)
b <- rnorm(4)
x2 <- Variable(3, name = "x")
prob2 <- Problem(Minimize(sum_squares(A %*% x2 - b)))
is_dcp(prob2)
```

```{r}
#| label: ex2-viz
#| echo: false
#| results: hide
visualize(prob2, output = "html",
          file = "viz/least-squares.html", open = FALSE)
```

<iframe src="viz/least-squares.html" width="100%" height="550" style="border: 1px solid #dee2e6; border-radius: 6px;" loading="lazy"></iframe>

### Portfolio Optimization

$$
\text{maximize} \; \boldsymbol{\mu}^T \mathbf{w} - \gamma \, \mathbf{w}^T \Sigma \mathbf{w}
\quad \text{subject to} \quad \sum w_i = 1, \quad w \geq 0
$$

A Markowitz portfolio problem with a risk-aversion parameter
$\gamma$. This example demonstrates `latex_name` for display
customization and `set_label()` for named constraints.

```{r}
#| label: ex3-build
w <- Variable(4, name = "w", latex_name = "\\mathbf{w}")
mu <- Parameter(4, name = "mu", latex_name = "\\boldsymbol{\\mu}")
gamma <- Parameter(1, name = "gamma", nonneg = TRUE, latex_name = "\\gamma")
Sigma <- matrix(c(1,.5,.3,.1, .5,1,.4,.2, .3,.4,1,.5, .1,.2,.5,1), 4, 4)
prob3 <- Problem(
  Maximize(t(mu) %*% w - gamma * quad_form(w, Sigma)),
  list(
    set_label(sum(w) == 1, "budget"),
    set_label(w >= 0, "long only")
  )
)
is_dcp(prob3)
```

```{r}
#| label: ex3-viz
#| echo: false
#| results: hide
visualize(prob3, output = "html",
          file = "viz/portfolio.html", open = FALSE)
```

<iframe src="viz/portfolio.html" width="100%" height="550" style="border: 1px solid #dee2e6; border-radius: 6px;" loading="lazy"></iframe>

### Entropy Maximization

$$
\text{maximize} \; \sum \text{entr}(x_i) \quad \text{subject to} \quad \sum x_i = 1
$$

Maximum entropy distribution. The `entr()` atom is concave, so
maximizing a sum of concave atoms is DCP-compliant.

```{r}
#| label: ex4-build
x4 <- Variable(5, name = "x")
prob4 <- Problem(Maximize(sum(entr(x4))), list(sum(x4) == 1))
is_dcp(prob4)
```

```{r}
#| label: ex4-viz
#| echo: false
#| results: hide
visualize(prob4, output = "html",
          file = "viz/entropy.html", open = FALSE)
```

<iframe src="viz/entropy.html" width="100%" height="550" style="border: 1px solid #dee2e6; border-radius: 6px;" loading="lazy"></iframe>

### Huber Loss

$$
\text{minimize} \; \sum \text{huber}(x_i)
$$

The Huber function is convex; minimizing its sum is DCP-compliant.
This is an unconstrained problem.

```{r}
#| label: ex5-build
x5 <- Variable(3, name = "x")
prob5 <- Problem(Minimize(sum(huber(x5))))
is_dcp(prob5)
```

```{r}
#| label: ex5-viz
#| echo: false
#| results: hide
visualize(prob5, output = "html",
          file = "viz/huber.html", open = FALSE)
```

<iframe src="viz/huber.html" width="100%" height="550" style="border: 1px solid #dee2e6; border-radius: 6px;" loading="lazy"></iframe>

### Log-Sum-Exp

$$
\text{minimize} \; \log\sum e^{x_i} \quad \text{subject to} \quad x \geq -1
$$

`log_sum_exp()` is a convex atom. The constraint is affine.

```{r}
#| label: ex6-build
x6 <- Variable(4, name = "x")
prob6 <- Problem(Minimize(log_sum_exp(x6)), list(x6 >= -1))
is_dcp(prob6)
```

```{r}
#| label: ex6-viz
#| echo: false
#| results: hide
visualize(prob6, output = "html",
          file = "viz/log-sum-exp.html", open = FALSE)
```

<iframe src="viz/log-sum-exp.html" width="100%" height="550" style="border: 1px solid #dee2e6; border-radius: 6px;" loading="lazy"></iframe>

### Semidefinite Program

$$
\text{minimize} \; \lambda_{\max}(X) \quad \text{subject to} \quad X_{11} = 1
$$

An SDP with a symmetric matrix variable. `lambda_max()` is convex and
the equality constraint is affine.

```{r}
#| label: ex7-build
X7 <- Variable(c(3, 3), name = "X", symmetric = TRUE)
prob7 <- Problem(Minimize(lambda_max(X7)), list(X7[1, 1] == 1))
is_dcp(prob7)
```

```{r}
#| label: ex7-viz
#| echo: false
#| results: hide
visualize(prob7, output = "html",
          file = "viz/sdp.html", open = FALSE)
```

<iframe src="viz/sdp.html" width="100%" height="550" style="border: 1px solid #dee2e6; border-radius: 6px;" loading="lazy"></iframe>

### Mixed Atoms

$$
\text{minimize} \; \max_i |x_i| + \sum \text{pos}(x_i)
$$

A combination of `max_entries(abs(x))` (convex) and
`sum_entries(pos(x))` (convex). The sum of convex functions is convex.

```{r}
#| label: ex8-build
x8 <- Variable(3, name = "x")
prob8 <- Problem(Minimize(max_entries(abs(x8)) + sum_entries(pos(x8))))
is_dcp(prob8)
```

```{r}
#| label: ex8-viz
#| echo: false
#| results: hide
visualize(prob8, output = "html",
          file = "viz/mixed-atoms.html", open = FALSE)
```

<iframe src="viz/mixed-atoms.html" width="100%" height="550" style="border: 1px solid #dee2e6; border-radius: 6px;" loading="lazy"></iframe>

## Non-DCP Examples

The following problems violate the DCP rules. The visualization
highlights the offending nodes in red, making it easy to pinpoint
where the violation occurs.

### Bilinear Objective

$$
\text{minimize} \; x_1 x_2 \quad \text{subject to} \quad x \geq 0
$$

::: {.callout-warning}
**DCP violation:** The product $x_1 x_2$ is neither convex nor
concave --- it is bilinear. `visualize()` marks the objective node in
red.
:::

```{r}
#| label: ex9-build
x9 <- Variable(2, name = "x")
prob9 <- Problem(Minimize(x9[1] * x9[2]), list(x9 >= 0))
is_dcp(prob9)
```

```{r}
#| label: ex9-viz
#| echo: false
#| results: hide
visualize(prob9, output = "html",
          file = "viz/bilinear.html", open = FALSE)
```

<iframe src="viz/bilinear.html" width="100%" height="550" style="border: 1px solid #dee2e6; border-radius: 6px;" loading="lazy"></iframe>

### Nonaffine Equality Constraint

$$
\text{minimize} \; \mathbf{1}^T x \quad \text{subject to} \quad \|x\|_2 = 1
$$

::: {.callout-warning}
**DCP violation:** Equality constraints must be affine. Here
$\|x\|_2$ is convex, so the equality $\|x\|_2 = 1$ is not
DCP-compliant. The constraint node is highlighted in red.
:::

```{r}
#| label: ex10-build
x10 <- Variable(3, name = "x")
prob10 <- Problem(Minimize(sum(x10)), list(p_norm(x10, 2) == 1))
is_dcp(prob10)
```

```{r}
#| label: ex10-viz
#| echo: false
#| results: hide
visualize(prob10, output = "html",
          file = "viz/nonaffine-eq.html", open = FALSE)
```

<iframe src="viz/nonaffine-eq.html" width="100%" height="550" style="border: 1px solid #dee2e6; border-radius: 6px;" loading="lazy"></iframe>

## Text Output

For quick inspection in the console, use `output = "text"` (the
default):

```{r}
#| label: text-output
visualize(prob1, output = "text")
```

## Session Info

```{r}
#| label: sessioninfo
sessionInfo()
```
