---
title: Near Isotonic and Near Convex Regression
author: Balasubramanian Narasimhan
date: '2017-11-02'
slug: near-isotonic-and-near-convex-regression
bibliography: cvxr_refs.bib
link-citations: true
categories: []
tags: []
---

The [Carbon Dioxide Information Analysis Center (CDIAC)](http://cdiac.ess-dive.lbl.gov/ftp/trends/temp/jonescru/) provides some data on global warming. The data points are the annual temperature anomalies relative to the 1961--1990 mean.

```{r}
library(readr)
cdiac_url <- "http://cdiac.ess-dive.lbl.gov/ftp/trends/temp/jonescru/global.txt"
CDIAC <- read_table(file = cdiac_url, skip=16)[, c("YEAR", "ANNUAL")]
str(CDIAC)
```

Given a set of data points $y \in  {\mathbf R}^m$,
@{TibshiraniHoefling:2011} fit a nearly-isotonic approximation
$\beta \in {\mathbf R}^m$  by solving

\[
\begin{array}{ll}
	\underset{\beta}{\mbox{minimize}} & \frac{1}{2}\sum_{i=1}^m (y_i - \beta_i)^2 + \lambda \sum_{i=1}^{m-1}(\beta_i - \beta_{i+1})_+,
\end{array}
\]
where $\lambda \geq 0$ is a penalty parameter and $x_+ =\max(x,0)$. Our `cvxr` formulation follows directly as shown below. The `pos` atom evaluates $x_+$ elementwise on the input expression.

```{r}
library(cvxr)
neariso_fit <- function(y, lambda) {
    m <- length(y)
    beta <- Variable(m)
    penalty <- sum(pos(diff(beta)))
    obj <- 0.5 * sum((y - beta)^2) + lambda * penalty
    prob <- Problem(Minimize(obj))
    solve(prob)$getValue(beta)
}
```


```{r}
neariso_fit_data <- function(data, index, lambda) {
    sample <- data[index,]                  # Bootstrap sample of rows
    sample <- sample[order(sample$YEAR),]   # Order ascending by year
    neariso_fit(sample$ANNUAL, lambda)
}
library(boot)
boot.out <- boot(data = CDIAC, statistic = neariso_fit_data, R = 10, lambda = 0.44)

ci.neariso <- t(sapply(seq_len(nrow(CDIAC)),
                            function(i) boot.ci(boot.out = boot.out, conf = 0.95,
                                                type = "norm", index = i)$normal[-1]))
data.neariso <- data.frame(year = CDIAC$YEAR, annual = CDIAC$ANNUAL, est = boot.out$t0,
                              lower = ci.neariso[, 1], upper = ci.neariso[, 2])
```

We can now plot the fit and confidence bands

```{r}
library(ggplot2)
(plot.neariso <- ggplot(data = data.neariso) +
     geom_point(mapping = aes(year, annual), color = "red") +
     geom_line(mapping = aes(year, est), color = "blue") +
     geom_ribbon(mapping = aes(x = year, ymin = lower,ymax = upper),alpha=0.3) +
     labs(x = "Year", y = "Temperature Anomalies")
)
```
