---
title: "Fractional Optimization"
author: "Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: dqcp_concave_fractional
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
source(here::here("_check_status.R"))
```

## Introduction

This example shows how to solve a simple _concave fractional problem_,
in which the objective is to maximize the ratio of a nonnegative
concave function and a positive convex function. Concave fractional
problems are quasiconvex programs (QCPs). They can be specified using
disciplined quasiconvex programming (DQCP), and hence can be solved
using `CVXR`.

## Problem Formulation

Our goal is to maximize the function

$$
\frac{\sqrt{x}}{\exp(x)}.
$$

This function is not concave, but it is quasiconcave, as can be seen
by inspecting its graph.

```{r}
x_vals <- seq(0, 10, length.out = 200)
y_vals <- sqrt(x_vals) / exp(x_vals)
plot(x_vals, y_vals, type = "l", col = "blue",
     xlab = "x", ylab = expression(sqrt(x) / exp(x)),
     main = "Concave Fractional Function")
```

## Solving with DQCP

The below code specifies and solves the QCP using DQCP. The concave
fractional function is DQCP-compliant because the ratio atom is
quasiconcave (actually quasilinear), increasing in the numerator when
the denominator is positive, and decreasing in the denominator when
the numerator is nonnegative.

```{r}
x <- Variable()
concave_fractional_fn <- sqrt(x) / exp(x)
problem <- Problem(Maximize(concave_fractional_fn))
cat("Is DQCP?", is_dqcp(problem), "\n")
result <- psolve(problem, qcp = TRUE)
check_solver_status(problem)
cat("Optimal value:", result, "\n")
cat("x:", value(x), "\n")
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(result = result, x = value(x)),
            file = file.path(testdata_dir, "concave_fractional1.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    saved <- readRDS(file = file.path(testdata_dir, "concave_fractional1.RDS"))
    expect_equal(result, saved$result, tolerance = 1e-3)
    expect_equal(value(x), saved$x, tolerance = 1e-3)
}
```

The optimal value of $x$ is approximately $0.5$, which corresponds to
the peak of the function $\sqrt{x}/\exp(x)$. We can verify this
analytically: setting the derivative to zero,

$$
\frac{d}{dx} \left(\frac{\sqrt{x}}{\exp(x)}\right) = \frac{1}{2\sqrt{x}\exp(x)} - \frac{\sqrt{x}}{\exp(x)} \cdot 1 = \frac{1 - 2x}{2\sqrt{x}\exp(x)} = 0,
$$

gives $x = 1/2$.

## Session Info

```{r}
sessionInfo()
```

## References

- Agrawal, A., Boyd, S. (2020). Disciplined Quasiconvex Programming.
  _Optimization Letters_, 14, 1643--1657.
