---
title: "Aerospace Design via Quasiconvex Optimization"
author: "CVXPY Developers and Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: dqcp_hypersonic_shape
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
library(ggplot2)
source(here::here("_check_status.R"))
```

## Introduction

Consider a triangle, or a wedge, located within a hypersonic flow. A
standard aerospace design optimization problem is to design the wedge
to maximize the lift-to-drag ratio (L/D) (or conversely minimize the
D/L ratio), subject to certain geometric constraints. In this example,
the wedge is known to have a constant hypotenuse, and our job is to
choose its width and height.

## Problem Formulation

The drag-to-lift ratio is given by

$$
\frac{D}{L} = \frac{c_d}{c_l},
$$

where $c_d$ and $c_l$ are drag and lift coefficients, respectively,
that are obtained by integrating the projection of the pressure
coefficient in directions parallel to, and perpendicular to, the body.

We will assume the pressure coefficient is given by the Newtonian
sine-squared law for whetted areas of the body,

$$
c_p = 2(\hat{v} \cdot \hat{n})^2
$$

and elsewhere $c_p = 0$. Here, $\hat{v}$ is the free stream
direction, which for simplicity we assume is parallel to the body so
that $\hat{v} = \langle 1, 0 \rangle$, and $\hat{n}$ is the local
unit normal. For a wedge defined by width $\Delta x$ and height
$\Delta y$,

$$
\hat{n} = \langle -\Delta y / s, -\Delta x / s \rangle
$$

where $s$ is the hypotenuse length. Therefore,

$$
c_p = 2\left(\frac{-\Delta y}{s}\right)^2 = \frac{2 \Delta y^2}{s^2}.
$$

The lift and drag coefficients are given by

$$
\begin{aligned}
c_d &= \frac{1}{c}\int_0^s -c_p \hat{n}_x \, ds \\
c_l &= \frac{1}{c}\int_0^s -c_p \hat{n}_y \, ds
\end{aligned}
$$

where $c$ is the reference chord length of the body. Given that
$\hat{n}$ and therefore $c_p$ are constant over the whetted surface of
the body,

$$
\begin{aligned}
c_d &= \frac{s}{c} \frac{2 \Delta y^2}{s^2} \frac{\Delta y}{s} \\
c_l &= \frac{s}{c} \frac{2 \Delta y^2}{s^2} \frac{\Delta x}{s}
\end{aligned}
$$

Assuming $s = 1$, so that
$\Delta y = \sqrt{1 - \Delta x^2}$, plugging in the above into the
equation for $D/L$, we obtain

$$
\frac{D}{L} = \frac{\Delta y}{\Delta x} = \frac{\sqrt{1 - \Delta x^2}}{\Delta x} = \sqrt{\frac{1}{\Delta x^2} - 1}.
$$

This function is representable as a DQCP quasilinear function.

## Plotting the Objective

We can visualize the D/L ratio as a function of $\Delta x$:

```{r}
x_vals <- seq(0.25, 1, length.out = 201)
obj_vals <- sqrt(1 / x_vals^2 - 1)
ggplot(data.frame(x = x_vals, y = obj_vals), aes(x, y)) +
    geom_line(color = "blue") +
    labs(x = expression(Delta * x), y = "D/L Ratio",
         title = "Drag-to-Lift Ratio vs. Wedge Width") +
    theme_minimal()
```

## DQCP Formulation

We now express the objective in `CVXR` and verify its curvature:

```{r}
x <- Variable(pos = TRUE)
obj <- sqrt(inv_pos(power(x, 2)) - 1)
cat("This objective function is", curvature(obj), "\n")
```

Minimizing this objective function subject to constraints representing
payload requirements is a standard aerospace design problem. In this
case we consider the constraint that the wedge must be able to contain
a rectangle of given length and width internally along its hypotenuse.
This is representable as a convex constraint.

```{r}
a <- 0.05  # height of rectangle (should be at most b)
b <- 0.65  # width of rectangle
constraint <- list(a * inv_pos(x) - (1 - b) * sqrt(1 - power(x, 2)) <= 0)
```

## Solving the Problem

```{r, results = "hold"}
prob <- Problem(Minimize(obj), constraint)
result <- psolve(prob, qcp = TRUE)
check_solver_status(prob)
ld_ratio <- 1 / result
x_opt <- value(x)
y_opt <- sqrt(1 - x_opt^2)
cat("Final L/D Ratio:", ld_ratio, "\n")
cat("Final width of wedge:", x_opt, "\n")
cat("Final height of wedge:", y_opt, "\n")
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(result = result, x = x_opt, y = y_opt, ld_ratio = ld_ratio),
            file = file.path(testdata_dir, "hypersonic1.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    saved <- readRDS(file = file.path(testdata_dir, "hypersonic1.RDS"))
    expect_equal(result, saved$result, tolerance = 1e-2)
    expect_equal(x_opt, saved$x, tolerance = 1e-2)
    expect_equal(y_opt, saved$y, tolerance = 1e-2)
}
```

## Verification Plot

Once the solution has been found, we can create a plot to verify that
the rectangle is inscribed within the wedge.

```{r}
y <- sqrt(1 - x_opt^2)
lambda1 <- a * x_opt / y

## Wedge vertices
wedge <- data.frame(
    x = c(0, x_opt, x_opt, 0),
    y = c(0, 0, -y, 0)
)

## Rectangle vertices
pt1 <- c(lambda1 * x_opt, -lambda1 * y)
pt2 <- c((lambda1 + b) * x_opt, -(lambda1 + b) * y)
pt3 <- c((lambda1 + b) * x_opt + a * y, -(lambda1 + b) * y + a * x_opt)
pt4 <- c(lambda1 * x_opt + a * y, -lambda1 * y + a * x_opt)
rect <- data.frame(
    x = c(pt1[1], pt2[1], pt3[1], pt4[1], pt1[1]),
    y = c(pt1[2], pt2[2], pt3[2], pt4[2], pt1[2])
)

ggplot() +
    geom_path(data = wedge, aes(x, y), color = "blue", linewidth = 1) +
    geom_path(data = rect, aes(x, y), color = "red", linewidth = 1) +
    coord_fixed() +
    labs(title = "Wedge with Inscribed Rectangle") +
    theme_minimal()
```

## Session Info

```{r, echo = FALSE}
sessionInfo()
```

## References

- Agrawal, A., Boyd, S. (2020). Disciplined Quasiconvex Programming.
  _Optimization Letters_, 14, 1643--1657.
