---
title: "Minimum-Length Least Squares"
author: "Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: dqcp_min_length_ls
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
source(here::here("_check_status.R"))
```

## Introduction

This example shows how to solve a _minimum-length least squares_
problem, which finds a minimum-length vector $x \in \mathbf{R}^n$
achieving small mean-square error (MSE) for a particular least-squares
problem:

$$
\begin{array}{ll}
\mbox{minimize} & \mathrm{len}(x) \\
\mbox{subject to} & \frac{1}{n}\|Ax - b\|_2^2 \leq \epsilon,
\end{array}
$$

where the variable is $x$ and the problem data are $n$, $A$, $b$, and
$\epsilon$.

Here, $\mathrm{len}(x)$ denotes the index of the last nonzero element
in $x$. This is a quasiconvex function, and the overall problem is a
quasiconvex program (QCP). It can be specified using disciplined
quasiconvex programming (DQCP) and therefore solved using `CVXR`.

## Problem Data

We first construct the problem data. We generate a random $n \times n$
matrix $A$ and a "true" solution $x^*$, then compute $b = Ax^*$.

```{r}
n <- 10
set.seed(1)
A <- matrix(rnorm(n * n), n, n)
x_star <- rnorm(n)
b <- A %*% x_star
epsilon <- 1e-2
```

## Solving the QCP

Now we construct and solve the QCP using DQCP. The objective
`length_expr(x)` computes the index of the last nonzero element in
$x$, and the constraint requires the MSE to be at most $\epsilon$.

```{r}
x <- Variable(n)
mse <- sum_squares(A %*% x - b) / n
problem <- Problem(Minimize(length_expr(x)), list(mse <= epsilon))
cat("Is problem DQCP?", is_dqcp(problem), "\n")
result <- psolve(problem, qcp = TRUE)
check_solver_status(problem)
cat("Found a solution, with length:", result, "\n")
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(result = result, x = value(x), mse = value(mse)),
            file = file.path(testdata_dir, "min_length_ls1.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    saved <- readRDS(file = file.path(testdata_dir, "min_length_ls1.RDS"))
    expect_equal(result, saved$result, tolerance = 1e-2)
}
```

The result tells us that the optimizer found a solution whose last
nonzero element is at a position less than $n$, meaning some trailing
entries of $x$ are (approximately) zero. Let us inspect the MSE and
the solution:

```{r}
cat("MSE:", value(mse), "\n")
```

```{r}
cat("x:", value(x), "\n")
```

For comparison, here is the "true" solution $x^*$ from which the data
was generated:

```{r}
cat("x_star:", as.numeric(x_star), "\n")
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(mse = value(mse), x_val = value(x)),
            file = file.path(testdata_dir, "min_length_ls2.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    saved <- readRDS(file = file.path(testdata_dir, "min_length_ls2.RDS"))
    expect_equal(value(mse), saved$mse, tolerance = 1e-2)
}
```

The minimum-length solution achieves the MSE bound while setting some
trailing components to zero, effectively finding a sparser
representation. This demonstrates how quasiconvex optimization can be
used to find solutions with structural properties (here, sparsity in a
specific sense) that go beyond what standard convex optimization
provides.

## Session Info

```{r}
sessionInfo()
```

## References

- Agrawal, A., Boyd, S. (2020). Disciplined Quasiconvex Programming.
  _Optimization Letters_, 14, 1643--1657.
