---
title: "Disciplined Quasiconvex Programming"
author: "Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: dqcp_tutorial
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
```

## Introduction

Disciplined quasiconvex programming (DQCP) is a generalization of DCP
for quasiconvex functions. Quasiconvexity generalizes convexity: a
function $f$ is quasiconvex if and only if its domain is a convex set
and its sublevel sets $\{x : f(x) \leq t\}$ are convex, for all $t$.

While DCP is a ruleset for constructing convex programs, DQCP is a
ruleset for quasiconvex programs (QCPs), which are optimization
problems in which the objective is to minimize a quasiconvex function
over a convex set. The convex set can be specified using equalities of
affine functions and inequalities of convex and concave functions,
just as in DCP; additionally, DQCP permits inequalities of the form
$f(x) \leq t$, where $f(x)$ is a quasiconvex expression and $t$ is
constant, and $f(x) \geq t$, where $f(x)$ is quasiconcave and $t$ is
constant. Every disciplined convex program is a disciplined
quasiconvex program, but the converse is not true.

`CVXR` lets you form and solve DQCP problems, just as it does for DCP
problems. For example, the following code solves a simple QCP:

```{r}
x <- Variable()
y <- Variable(pos = TRUE)
objective_fn <- -sqrt(x) / y
problem <- Problem(Minimize(objective_fn), list(exp(x) <= y))
result <- psolve(problem, qcp = TRUE)
cat("Is DQCP?", is_dqcp(problem), "\n")
cat("Optimal value:", result, "\n")
cat("x:", value(x), "\n")
cat("y:", value(y), "\n")
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(result = result, x = value(x), y = value(y)),
            file = file.path(testdata_dir, "dqcp_tutorial1.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    saved <- readRDS(file = file.path(testdata_dir, "dqcp_tutorial1.RDS"))
    expect_equal(result, saved$result, tolerance = 1e-3)
    expect_equal(value(x), saved$x, tolerance = 1e-3)
    expect_equal(value(y), saved$y, tolerance = 1e-3)
}
```

Note that to solve DQCP problems, you must pass the option
`qcp = TRUE` to `psolve()`.

## Curvature

DQCP adds two new types of curvature to `CVXR`: _quasiconvex_ and
_quasiconcave_. A function $f$ is quasiconvex if and only if its
domain is a convex set and its sublevel sets
$\{x : f(x) \leq t\}$ are convex, for all $t$; $f$ is quasiconcave
if $-f$ is quasiconvex. Every convex function is also quasiconvex,
and every concave function is also quasiconcave; the converses of
these statements are not true. An expression that is both quasiconvex
and quasiconcave is called _quasilinear_.

`CVXR`'s curvature analysis can flag Expressions as unknown even when
they are quasiconvex or quasiconcave, but it will never mistakenly
flag an expression as quasiconvex or quasiconcave.

The curvature of an Expression is stored in its `curvature()`
attribute. For example:

```{r}
x <- Variable(3)
w <- ceil_expr(x)
cat("Curvature of ceil_expr(x):", curvature(w), "\n")

y <- Variable(nonneg = TRUE)
z <- Variable(nonneg = TRUE)
product <- multiply(y, z)
cat("Curvature of multiply(y, z) (nonneg):", curvature(product), "\n")
```

You can also check the curvature of an Expression by calling
`is_quasiconvex()` and `is_quasiconcave()`. For example,
`is_quasiconvex(y)` and `is_quasiconcave(z)` would evaluate to
`TRUE`. You can check if an expression is quasilinear by calling
`is_quasilinear()`.

## Composition Rules

DQCP analysis is based on applying a general composition theorem from
convex analysis to each expression. An expression is verifiably
quasiconvex under DQCP if it is one of the following:

- convex (under DCP);
- a quasiconvex atom, applied to a variable or constant;
- the maximum of quasiconvex expressions;
- an increasing function of a quasiconvex expression, or a decreasing
  function of a quasiconcave expression;
- an expression of the form $f(e_1, e_2, \ldots, e_n)$ such that
  (1) $f$ is a quasiconvex atom, and (2) for each $i$, $f$ is
  increasing in argument $i$ and $e_i$ is convex, $f$ is decreasing
  in argument $i$ and $e_i$ is concave, or $e_i$ is affine.

An expression is quasiconcave under DQCP if it is one of the
following:

- concave (under DCP);
- a quasiconcave atom, applied to a variable or constant;
- the minimum of quasiconcave expressions;
- an increasing function of a quasiconcave expression, or a
  decreasing function of a quasiconvex expression;
- an expression of the form $f(e_1, e_2, \ldots, e_n)$ such that
  (1) $f$ is a quasiconcave atom, and (2) for each $i$, $f$ is
  increasing in argument $i$ and $e_i$ is concave, $f$ is decreasing
  in argument $i$ and $e_i$ is convex, or $e_i$ is affine.

Whether an atom is quasiconvex or quasiconcave may depend on the signs
of its arguments. For example, the scalar product $xy$ is
quasiconcave when $x$ and $y$ are either both nonnegative or both
nonpositive, and quasiconvex when one of the arguments is nonnegative
and the other is nonpositive.

If an Expression satisfies the above rules, we say that the Expression
"is DQCP." You can check whether an Expression is DQCP by calling
`is_dqcp()`. For example:

```{r}
x <- Variable(pos = TRUE)
y <- Variable(pos = TRUE)

product <- multiply(x, y)

cat("Is quasiconcave?", is_quasiconcave(product), "\n")
cat("Is DQCP?", is_dqcp(product), "\n")
```

An Expression is DQCP precisely when it has known curvature, which
means at least one of `is_constant()`, `is_affine()`, `is_convex()`,
`is_concave()`, `is_quasiconvex()`, or `is_quasiconcave()` returns
`TRUE`.

## DQCP Problems

A Problem is constructed from an objective and a list of constraints.
If a problem follows the DQCP rules, it is guaranteed to be a QCP and
solvable by `CVXR` (if a solution to the problem exists). The DQCP
rules require that the problem objective have one of two forms:

- `Minimize(quasiconvex)`
- `Maximize(quasiconcave)`

The only valid constraints under the DQCP rules are:

- affine `==` affine
- convex `<=` concave
- concave `>=` convex
- quasiconvex `<=` constant
- quasiconcave `>=` constant

You can check that a problem, constraint, or objective satisfies the
DQCP rules by calling `is_dqcp()`. Here are some examples of DQCP and
non-DQCP problems:

```{r}
## The sign of variables affects curvature analysis.
x <- Variable(nonneg = TRUE)
concave_fractional_fn <- x * sqrt(x)
constraint <- list(ceil_expr(x) <= 10)
problem <- Problem(Maximize(concave_fractional_fn), constraint)
cat("Is concave_fractional_fn quasiconcave?",
    is_quasiconcave(concave_fractional_fn), "\n")
cat("Is constraint DQCP?", is_dqcp(constraint[[1]]), "\n")
cat("Is problem DQCP?", is_dqcp(problem), "\n")
```

```{r}
w <- Variable()
fn <- w * sqrt(w)
problem2 <- Problem(Maximize(fn))
cat("Is fn DQCP?", is_dqcp(fn), "\n")
cat("Is problem DQCP?", is_dqcp(problem2), "\n")
```

`CVXR` will raise an error if you call `psolve(problem, qcp = TRUE)`
on a non-DQCP problem.

## DQCP Atoms

Quasiconvex and quasiconcave expressions can be constructed using
convex and concave atoms, using the curvature rules given above. This
section describes new semantics for some existing atoms under DQCP and
introduces new atoms that are quasiconvex or quasiconcave (but not
convex or concave). Many of these new atoms are integer-valued.

### Ratio

The infix operator `/` is an atom, denoting ratio. This atom is both
quasiconvex and quasiconcave when the denominator is known to be
either nonnegative or nonpositive. The ratio `x/y` is increasing in
`x` when `y` is nonnegative, increasing in `y` when `x` is
nonpositive, decreasing in `x` when `y` is nonpositive, and
decreasing in `y` when `x` is nonnegative.

The ratio atom can be used with the composition rule to construct
interesting quasiconvex and quasiconcave expressions. For example, the
ratio of a nonnegative concave function and a positive convex function
is quasiconcave, and the ratio of a nonnegative convex function and a
positive concave function is quasiconvex. Similarly, the ratio of two
affine functions is quasilinear when the denominator is positive.

### Scalar Product

The scalar product `*` is quasiconvex when one of its arguments is
nonnegative and the other is nonpositive, and it is quasiconcave when
its arguments are both nonnegative or both nonpositive. Hence, by the
composition rule, the product of two nonnegative concave functions is
quasiconcave, and the product of a nonnegative concave function and a
nonpositive convex function is quasiconvex.

### Ceiling and Floor

The atoms `ceil_expr(x)` and `floor_expr(x)` are quasilinear and
increasing in their arguments.

## Solving DQCP Problems

A DQCP problem can be solved by calling `psolve(problem, qcp = TRUE)`.
`CVXR` uses a bisection method on the optimal value of the problem to
solve QCPs, and it will automatically find an upper and lower bound
for the bisection. You can optionally provide your own upper and lower
bound when solving a QCP, which can sometimes be helpful. You can
provide these bounds via the keyword arguments `low` and `high`; for
example, `psolve(problem, qcp = TRUE, low = 12, high = 17)` would
limit the bisection to objective values between 12 and 17.

Bisection involves solving a sequence of optimization problems. If
your problem is ill-conditioned, or if you are unlucky, a solver might
fail to solve one of these subproblems, which will result in an error.
If this happens, you can try using a different solver via the `solver`
keyword argument. For example,
`psolve(problem, qcp = TRUE, solver = "SCS")`. To obtain verbose
output describing the bisection, supply the keyword argument
`verbose = TRUE` to the solve method:
`psolve(problem, qcp = TRUE, verbose = TRUE)`.

## Session Info

```{r}
sessionInfo()
```

## References

- Agrawal, A., Boyd, S. (2020). Disciplined Quasiconvex Programming.
  _Optimization Letters_, 14, 1643--1657.
