---
title: "Sensitivity Analysis and Gradients"
author: "CVXPY Developers and Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: dpp_sensitivity
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
```

## Introduction

::: {.callout-warning}
The derivative functionality (`backward`, `derivative`, `gradient`,
`delta`, `requires_grad`) is **not implemented** in `CVXR` at this
time. The code below illustrates the intended API based on CVXPY but is
not executable.
:::

An optimization problem can be viewed as a function mapping parameters
to solutions. This solution map is sometimes differentiable. `CVXR`
has built-in support for computing the derivative of the optimal
variable values of a problem with respect to small perturbations of
the parameters (i.e., the `Parameter` instances appearing in a
problem).

The `Problem` class exposes two methods related to computing the
derivative:

- `derivative()` evaluates the derivative given perturbations to the
  parameters. This lets you calculate how the solution to a problem
  would change given small changes to the parameters, without
  re-solving the problem.
- `backward()` evaluates the adjoint of the derivative, computing
  the gradient of the solution with respect to the parameters. This
  can be useful when combined with automatic differentiation
  software.

The `derivative()` and `backward()` methods are only meaningful when
the problem contains parameters. In order for a problem to be
differentiable, it must be DPP-compliant. `CVXR` can compute the
derivative of any DPP-compliant DCP or DGP problem. At
non-differentiable points, `CVXR` computes a heuristic quantity.

## Example: A Trivial Quadratic

As a first example, we solve a trivial problem with an analytical
solution, to illustrate the usage of `backward()` and `derivative()`.
We construct a problem with a scalar variable `x` and a scalar
parameter `p`. The problem is to minimize the quadratic
$(x - 2p)^2$.

```{r}
x <- Variable()
p <- Parameter()
quadratic <- power(x - 2 * p, 2)
problem <- Problem(Minimize(quadratic))
```

Next, we solve the problem for the particular value of $p = 3$. Note
that when solving the problem, we supply the keyword argument
`requires_grad = TRUE` to `psolve()`.

```{r, eval = FALSE}
value(p) <- 3.0
result <- psolve(problem, requires_grad = TRUE)
cat("Optimal value:", result, "\n")
cat("x:", value(x), "\n")
```

### Using `backward()`

Having solved the problem with `requires_grad = TRUE`, we can now use
`backward()` to differentiate through the problem. We compute the
gradient of the solution with respect to its parameter by calling
`backward()`. As a side-effect, `backward()` populates the
`gradient()` attribute on all parameters with the gradient of the
solution with respect to that parameter.

```{r, eval = FALSE}
backward(problem)
cat("The gradient is", gradient(p), "\n")
```

In this case, the problem has the trivial analytical solution $x^* =
2p$, and the gradient $dx^*/dp$ is therefore just $2$. So, as
expected, the gradient is 2.0.

### Using `derivative()`

Next, we use `derivative()` to see how a small change in `p` would
affect the solution `x`. We perturb `p` by $10^{-5}$ by setting
`delta(p) <- 1e-5`, and calling `derivative()` will populate the
`delta()` attribute of `x` with the change in `x` predicted by a
first-order approximation (which is $\frac{dx}{dp} \cdot \Delta p$).

```{r, eval = FALSE}
delta(p) <- 1e-5
derivative(problem)
cat("x delta is", delta(x), "\n")
```

In this case the solution is trivial and its derivative is just $2p$,
so we know that the delta in $x$ should be $2 \times 10^{-5}$. As
expected, the output confirms this.

We emphasize that this example is trivial, because it has a trivial
analytical solution with a trivial derivative. The `backward()` and
`derivative()` methods are useful because the vast majority of convex
optimization problems do not have analytical solutions: in these
cases, `CVXR` can compute solutions and their derivatives, even though
it would be impossible to derive them by hand.

## When to Use `backward()` vs. `derivative()`

- **`backward()`** should be used when you need the gradient of (a
  scalar-valued function of) the solution with respect to the
  parameters.
- **`derivative()`** should be used for sensitivity analysis, i.e.,
  when you want to know how the solution would change if one or more
  parameters were changed.

When there are multiple variables, it is much more efficient to
compute sensitivities using `derivative()` than it would be to compute
the entire Jacobian (which can be done by calling `backward()`
multiple times, once for each standard basis vector).

## A Note on `backward()` with Multiple Variables

In this simple example, the variable `x` was a scalar, so `backward()`
computed the gradient of `x` with respect to `p`. When there is more
than one scalar variable, by default, `backward()` computes the
gradient of the _sum_ of the optimal variable values with respect to
the parameters.

More generally, `backward()` can be used to compute the gradient of a
scalar-valued function $f$ of the optimal variables, with respect to
the parameters. If $x(p)$ denotes the optimal value of the variable
(which might be a vector or a matrix) for a particular value of the
parameter $p$ and $f(x(p))$ is a scalar, then `backward()` can be
used to compute the gradient of $f$ with respect to $p$. Let $x^* =
x(p)$, and say the derivative of $f$ with respect to $x^*$ is $dx$.
To compute the derivative of $f$ with respect to $p$, before calling
`backward(problem)`, just set `gradient(x) <- dx`.

## Example: Least Squares with Regularization

Here we demonstrate sensitivity analysis on a more practical problem:
a regularized least-squares problem where we want to understand how
the solution changes with respect to the regularization parameter.

```{r, eval = FALSE}
set.seed(42)
n <- 5
m <- 3
A <- matrix(rnorm(n * m), n, m)
b <- rnorm(n)

x <- Variable(m)
lam <- Parameter(nonneg = TRUE)

objective <- sum_squares(A %*% x - b) + lam * sum_squares(x)
problem <- Problem(Minimize(objective))

value(lam) <- 1.0
result <- psolve(problem, requires_grad = TRUE)
cat("Optimal value:", result, "\n")
cat("x:", value(x), "\n")
```

Now compute the gradient of the solution with respect to the
regularization parameter:

```{r, eval = FALSE}
backward(problem)
cat("Gradient of solution w.r.t. lambda:", gradient(lam), "\n")
```

And use `derivative()` to predict the effect of a small perturbation
in $\lambda$:

```{r, eval = FALSE}
delta(lam) <- 0.01
derivative(problem)
cat("Predicted change in x:", delta(x), "\n")
```

## Session Info

```{r, echo = FALSE}
sessionInfo()
```

## References

- Agrawal, A., Barratt, S., Boyd, S., Busseti, E., Moursi, W. M.
  (2019). Differentiating through a cone program.
  _Journal of Applied and Numerical Optimization_, 1(2), 107--115.
- Agrawal, A., Verschueren, R., Diamond, S., Boyd, S. (2018). A
  rewriting system for convex optimization problems.
  _Journal of Control and Decision_, 5(1), 42--60.
