---
title: "Disciplined Parametrized Programming"
author: "Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: dpp_tutorial
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
source(here::here("_check_status.R"))
```

## Introduction

Parameters are symbolic representations of constants. Using
parameters lets you modify the values of constants without
reconstructing the entire problem. When your parametrized problem is
constructed according to _Disciplined Parametrized Programming (DPP)_,
solving it repeatedly for different values of the parameters can be
much faster than repeatedly solving a new problem.

You should read this tutorial if you intend to solve a DCP or DGP
problem many times, for different values of the numerical data, or if
you want to differentiate through the solution map of a DCP or DGP
problem.

## What is DPP?

DPP is a ruleset for producing parametrized DCP- or DGP-compliant
problems that `CVXR` can re-canonicalize very quickly. The first time
a DPP-compliant problem is solved, `CVXR` compiles it and caches the
mapping from parameters to problem data. As a result, subsequent
rewritings of DPP problems can be substantially faster. `CVXR` allows
you to solve parametrized problems that are not DPP, but you will not
see a speed-up when doing so.

## The DPP Ruleset

DPP places mild restrictions on how parameters can enter expressions
in DCP and DGP problems.

### DPP for DCP Problems

In DPP, an expression is said to be _parameter-affine_ if it does not
involve variables and is affine in its parameters, and it is
_parameter-free_ if it does not have parameters. DPP introduces two
restrictions to DCP:

1. Under DPP, all parameters are classified as affine, just like
   variables.
2. Under DPP, the product of two expressions is affine when at least
   one of the expressions is constant, or when one of the expressions
   is parameter-affine and the other is parameter-free.

An expression is DPP-compliant if it is DCP-compliant subject to these
two restrictions. You can check whether an expression or problem is
DPP-compliant by calling `is_dpp()`. For example:

```{r}
m <- 3
n <- 2
x <- Variable(c(n, 1))
F <- Parameter(c(m, n))
G <- Parameter(c(m, n))
g <- Parameter(c(m, 1))
gamma <- Parameter(nonneg = TRUE)

objective <- p_norm(( F + G ) %*% x - g) + gamma * p_norm(x)
cat("Is DPP?", is_dpp(objective), "\n")
```

We can walk through the DPP analysis to understand why `objective` is
DPP-compliant. The product `(F + G) %*% x` is affine under DPP,
because `F + G` is parameter-affine and `x` is parameter-free. The
difference `(F + G) %*% x - g` is affine because the addition atom is
affine and both `(F + G) %*% x` and `-g` are affine. Likewise
`gamma * p_norm(x)` is affine under DPP because `gamma` is
parameter-affine and `p_norm(x)` is parameter-free. The final
objective is then affine under DPP because addition is affine.

Some expressions are DCP-compliant but not DPP-compliant. For example,
DPP forbids taking the product of two parametrized expressions:

```{r}
x <- Variable()
gamma <- Parameter(nonneg = TRUE)
problem <- Problem(Minimize(gamma * gamma * x), list(x >= 1))
cat("Is DPP?", is_dpp(problem), "\n")
cat("Is DCP?", is_dcp(problem), "\n")
```

Just as it is possible to rewrite non-DCP problems in DCP-compliant
ways, it is also possible to re-express non-DPP problems in
DPP-compliant ways. For example, the above problem can be equivalently
written as:

```{r}
x <- Variable()
y <- Variable()
gamma <- Parameter(nonneg = TRUE)
problem <- Problem(Minimize(gamma * y), list(y == gamma * x))
cat("Is DPP?", is_dpp(problem), "\n")
cat("Is DCP?", is_dcp(problem), "\n")
```

In other cases, you can represent non-DPP transformations of
parameters by doing them outside of the DSL, e.g., in R. For example,
if `P` is a parameter and `x` is a variable, `quad_form(x, P)` is not
DPP. You can represent a parametric quadratic form like so:

```{r}
n <- 4
set.seed(1)
L <- matrix(rnorm(n * n), n, n)
P <- t(L) %*% L
P_sqrt <- Parameter(c(n, n))
x <- Variable(c(n, 1))
quad_form_expr <- sum_squares(P_sqrt %*% x)
value(P_sqrt) <- expm::sqrtm(P)
cat("Is DPP?", is_dpp(quad_form_expr), "\n")
```

As another example, the quotient `expr / p` is not DPP-compliant when
`p` is a parameter, but this can be rewritten as `expr * p_tilde`,
where `p_tilde` is a parameter that represents `1/p`.

### DPP for DGP Problems

Just as DGP is the log-log analogue of DCP, DPP for DGP is the
log-log analog of DPP for DCP. DPP introduces two restrictions to
DGP:

1. Under DPP, all positive parameters are classified as
   log-log-affine, just like positive variables.
2. Under DPP, the power atom `power(x, p)` (with base `x` and
   exponent `p`) is log-log affine as long as `x` and `p` are not
   both parametrized.

Note that for powers, the exponent `p` must be either a numerical
constant or a parameter; attempting to construct a power atom in which
the exponent is a compound expression will result in an error.

If a parameter appears in a DGP problem as an exponent, it can have
any sign. If a parameter appears elsewhere in a DGP problem, _it must
be positive_, i.e., it must be constructed with
`Parameter(pos = TRUE)`.

For example, consider the monomial
`c * power(x, a) * power(y, b)` where `c` is a positive parameter
and `a`, `b` are parameters. The expressions `power(x, a)` and
`power(y, b)` are log-log affine, since `x` and `y` do not contain
parameters. The parameter `c` is log-log affine because it is
positive, and the monomial expression is log-log affine because the
product of log-log affine expressions is also log-log affine. This
makes the expression DPP-compliant.

Some expressions are DGP-compliant but not DPP-compliant. For example,
DPP forbids raising a parametrized expression to a power:
`power(power(x, a), a)` is DGP but not DPP because both the base
(`power(x, a)`) and the exponent (`a`) involve the same parameter.

You can represent non-DPP transformations of parameters by doing them
outside of `CVXR`, e.g., in R. For example, if `a_val <- 2.0`, you
could precompute `b <- Parameter(value = a_val^2)` and write
`power(x, b)` instead.

## Repeatedly Solving a DPP Problem


The following example demonstrates how parameters can speed up
repeated solves of a DPP-compliant DCP problem. We set up a
regularized least-squares problem and solve it for a range of
regularization values.

```{r}
n <- 15
m <- 10
set.seed(1)
A <- matrix(rnorm(n * m), n, m)
b <- rnorm(n)
## gamma must be nonneg due to DCP rules
gamma <- Parameter(nonneg = TRUE)

x <- Variable(m)
error <- sum_squares(A %*% x - b)
obj <- Minimize(error + gamma * p_norm(x, 1))
problem <- Problem(obj)
cat("Is DPP?", is_dpp(problem), "\n")
```

```{r}
gamma_vals <- 10^seq(-4, 1, length.out = 50)
dpp_times <- numeric(length(gamma_vals))
new_problem_times <- numeric(length(gamma_vals))

for (i in seq_along(gamma_vals)) {
    value(gamma) <- gamma_vals[i]
    t1 <- proc.time()
    result <- psolve(problem)
    check_solver_status(problem)
    t2 <- proc.time()
    dpp_times[i] <- (t2 - t1)["elapsed"]

    new_problem <- Problem(obj)
    t3 <- proc.time()
    new_result <- psolve(new_problem)
    check_solver_status(new_problem)
    t4 <- proc.time()
    new_problem_times[i] <- (t4 - t3)["elapsed"]
}
```

We can visualize the timing comparison:

```{r}
plot(gamma_vals, dpp_times, type = "l", col = "blue", log = "x",
     xlab = expression(gamma), ylab = "Time (s)",
     main = "DPP Re-solve vs. New Problem")
lines(gamma_vals, new_problem_times, col = "red")
legend("topright",
       legend = c("Re-solving a DPP problem", "Solving a new problem"),
       col = c("blue", "red"), lty = 1)
```

Similar speed-ups can be obtained for DGP problems.

## Session Info

```{r}
sessionInfo()
```

## References

- Agrawal, A., Barratt, S., Boyd, S., Busseti, E., Moursi, W. M.
  (2019). Differentiating through a cone program.
  _Journal of Applied and Numerical Optimization_, 1(2), 107--115.
- Agrawal, A., Verschueren, R., Diamond, S., Boyd, S. (2018). A
  rewriting system for convex optimization problems.
  _Journal of Control and Decision_, 5(1), 42--60.
