---
title: The Catenary Problem
author: Anqi Fu and Balasubramanian Narasimhan
date: '2017-11-02'
slug: cvxr_catenary
bibliography: ../bibtex/cvxr_refs.bib
link-citations: true
categories: []
tags: []
params:
  mode: save
  check_dir: test_data
  data_dir: catenary
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$check_dir, params$data_dir) 
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
library(ggplot2)
```

## Introduction

A chain with uniformly distributed mass hangs from the endpoints
$(0,1)$ and $(1,1)$ on a 2-D plane. Gravitational force acts in the
negative $y$ direction. Our goal is to find the shape of the chain in
equilibrium, which is equivalent to determining the $(x,y)$
coordinates of every point along its curve when its potential energy
is minimized.

This is the famous [catenary
problem](https://en.wikipedia.org/wiki/Catenary).


## A Discrete Version

To formulate as an optimization problem, we parameterize the chain by
its arc length and divide it into $m$ discrete links. The length of
each link must be no more than $h > 0$. Since mass is uniform, the
total potential energy is simply the sum of the
$y$-coordinates. Therefore, our (discretized) problem is

$$ 
\begin{array}{ll} \underset{x,y}{\mbox{minimize}} & \sum_{i=1}^m y_i \\ \mbox{subject to} & x_1 = 0, \quad y_1 = 1, \quad x_m = 1,
\quad y_m = 1 \\ & (x_{i+1} - x_i)^2 + (y_{i+1} - y_i)^2 \leq h^2,
\quad i = 1,\ldots,m-1 \end{array} 
$$

with variables $x \in {\mathbf R}^m$ and $y \in {\mathbf R}^m$. This
discretized version which has been studied by @catenary was suggested
to us by [Hans Werner
Borchers](http://hwborchers.lima-city.de/Presents/catenary.html).

The basic catenary problem has a well-known analytical solution
(see @GelfandFomin:1963) which we can easily verify with `CVXR`. 

```{r}
## Problem data
m <- 101
L <- 2
h <- L / (m - 1)

## Form objective
x <- Variable(m)
y <- Variable(m)
objective <- Minimize(sum(y))

## Form constraints
constraints <- list(x[1] == 0, y[1] == 1,
                    x[m] == 1, y[m] == 1,
                    diff(x)^2 + diff(y)^2 <= h^2)

## Solve the catenary problem
prob <- Problem(objective, constraints)
result <- solve(prob)
```

We can now plot it and compare it with the ideal solution. Below we
use alpha blending and differing line thickness to show the ideal in red and the computed solution in blue.


```{r, fig.cap = "Analytic (red) and computed solution (blue) to the catenary problem"}
xs <- result$getValue(x)
ys <- result$getValue(y)

catenary <- ggplot(data.frame(x = xs, y = ys)) +
    geom_line(mapping = aes(x = x, y = y), color = "blue", size = 1) +
    geom_point(data = data.frame(x = c(xs[1], ys[1]), y = c(xs[m], ys[m])),
               mapping = aes(x = x, y = y), color = "red") 

ideal <- function(x) { 0.22964 *cosh((x -0.5) / 0.22964) - 0.02603 }

catenary + stat_function(fun = ideal , colour = "brown", alpha = 0.5, size = 3)
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save"), error = params$mode %in% c("test", "save")}
cat("Testthat Results: No output is good\n")
if (params$mode == "save") {
    saveRDS(list(xs = xs, ys = ys), file = file.path(testdata_dir, "catenary1.RDS"))
} else {
    catenary1 <- readRDS(file = file.path(testdata_dir, "catenary1.RDS"))
    expect_identical(xs, catenary1$xs)
    expect_identical(ys, catenary1$ys)
}
```

## Additional Ground Constraints

A more interesting situation arises when the ground is not flat. Let
$g \in {\mathbf R}^m$ be the elevation vector (relative to the $x$-axis),
and suppose the right endpoint of our chain has been lowered by
$\Delta y_m = 0.5$. The analytical solution in this case would be
difficult to calculate. However, we need only add two lines to our
constraint definition,

```{r, eval = FALSE}
constr[[4]] <- (y[m] == 0.5)
constr <- c(constr, y >= g)
```
to obtain the new result.

Below, we define $g$ as a staircase function and solve the problem.

```{r}
## Lower right endpoint and add staircase structure
ground <- sapply(seq(0, 1, length.out = m), function(x) {
    if(x < 0.2)
        return(0.6)
    else if(x >= 0.2 && x < 0.4)
        return(0.4)
    else if(x >= 0.4 && x < 0.6)
        return(0.2)
    else
        return(0)
})
constraints <- c(constraints, y >= ground)
constraints[[4]] <- (y[m] == 0.5)
prob <- Problem(objective, constraints)
result <- solve(prob)
```
to obtain the new result. 

The figure below shows the solution of this modified catenary problem
for $m = 101$ and $h = 0.04$. The chain is shown hanging in blue,
bounded below by the red staircase structure, which represents the
ground.

```{r, fig.cap="Asymmetric catenary problem with ground constraints."}
xs <- result$getValue(x)
ys <- result$getValue(y)

ggplot(data.frame(x = xs, y = ys)) +
    geom_line(mapping = aes(x = x, y = y), color = "blue") +
    geom_point(data = data.frame(x = c(xs[1], ys[1]), y = c(xs[m], ys[m])),
               mapping = aes(x = x, y = y), color = "red") +
    geom_line(data.frame(x = xs, y = ground),
              mapping = aes(x = x, y = y), color = "brown")
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save"), error = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(xs = xs, ys = ys), file = file.path(testdata_dir, "catenary2.RDS"))
} else {
    cat("Testthat Results: No output is good\n")    
    catenary2 <- readRDS(file = file.path(testdata_dir, "catenary2.RDS"))
    expect_identical(xs, catenary2$xs)
    expect_identical(ys, catenary2$ys)
}
```



## Session Info

```{r}
sessionInfo()
```

## Source

[R Markdown](https://github.com/bnaras/cvxr_docs/blob/master/content/cvxr_examples/catenary.Rmd)

## References
