---
title: Censored Regression
author: Anqi Fu and Balasubramanian Narasimhan
date: '2017-11-02'
slug: cvxr_censored-regression
bibliography: ../bibtex/cvxr_refs.bib
link-citations: true
categories: []
tags: []
params:
  mode: ignore
  testdata_dir: test_data
  data_dir: censored
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir) 
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
library(ggplot2)
```

## Introduction

Data collected from an experimental study is sometimes censored, so
that only partial information is known about a subset of
observations. For instance, when measuring the lifespan of mice, we
may find a number of subjects live beyond the duration of the
project. Thus, all we know is the lower bound on their lifespan. This
right censoring can be incorporated into a regression model via convex
optimization.

Suppose that only $K$ of our observations $(x_i,y_i)$ are fully
observed, and the remaining are censored such that we observe $x_i$,
but only know $y_i \geq D$ for $i = K+1,\ldots,m$ and some constant $D
\in {\mathbf R}$. We can build an OLS model using the uncensored data,
restricting the fitted values $\hat y_i = x_i^T\beta$ to lie above $D$
for the censored observations:

\[
\begin{array}{ll} \underset{\beta}{\mbox{minimize}} & \sum_{i=1}^K (y_i - x_i^T\beta)^2 \\
\mbox{subject to} & x_i^T\beta \geq D, \quad i = K+1,\ldots,m.
\end{array}
\]

This avoids the bias introduced by standard OLS, while still utilizing
all of the data points in the regression. The constraint requires only
one more line in `CVXR`.

## Example

We will generate synthetic data for this example, censoring
observations beyond a value $D$.

```{r}
## Problem data
n <- 30
M <- 50
K <- 200

set.seed(n * M * K)
X <- matrix(stats::rnorm(K * n), nrow = K, ncol = n)
beta_true <- matrix(stats::rnorm(n), nrow = n, ncol = 1)
y <- X %*% beta_true + 0.3 * sqrt(n) * stats::rnorm(K)

## Order variables based on y
idx <- order(y, decreasing = FALSE)
y_ordered <- y[idx]
X_ordered <- X[idx, ]

## Find cutoff and censor
D <- (y_ordered[M] + y_ordered[M + 1]) / 2
censored <- (y_ordered > D)
y_censored <- pmin(y_ordered, D)
```

We now fit regular OLS, OLS using just the censored data and finally
the censored regression.

```{r}
## Regular OLS
beta <- Variable(n)
obj <- sum((y_censored - X_ordered %*% beta)^2)
prob <- Problem(Minimize(obj))
result <- solve(prob)
beta_ols <- result$getValue(beta)

## OLS using uncensored data
obj <- sum((y_censored[1:M] - X_ordered[1:M,] %*% beta)^2)
prob <- Problem(Minimize(obj))
result <- solve(prob)
beta_unc <- result$getValue(beta)

## Censored regression
obj <- sum((y_censored[1:M] - X_ordered[1:M,] %*% beta)^2)
constr <- list(X_ordered[(M+1):K,] %*% beta >= D)
prob <- Problem(Minimize(obj), constr)
result <- solve(prob)
beta_cens <- result$getValue(beta)
```

Here's are some plots comparing the results. The blue diamond points
are estimates.

```{r}
plot_results <- function(beta_res, title) {
    d <- data.frame(index = seq_len(K),
                    y_ordered = y_ordered,
                    y_fit = as.numeric(X_ordered %*% beta_res),
                    censored = as.factor(censored))
    ggplot(data = d) +
        geom_point(mapping = aes(x = index, y = y_ordered, color = censored)) +
        geom_point(mapping = aes(x = index, y = y_fit), color = "blue", shape = 23) +
        geom_abline(intercept = D, slope = 0, lty = "dashed") +
        labs(x = "Observations", y = "y") +
        ggtitle(title)
}
```

```{r}
plot_results(beta_ols, "Regular OLS.")
```

```{r}
plot_results(beta_unc, "OLS using uncensored data.")
```

```{r}
plot_results(beta_cens, "Censored Regression.")
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save"), error = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(beta_ols = beta_ols, beta_unc = beta_unc, beta_cens = beta_cens), file = file.path(testdata_dir, "censored.RDS"))
} else {
    cat("Testthat Results: No output is good\n")    
    censored <- readRDS(file = file.path(testdata_dir, "censored.RDS"))
    expect_identical(beta_ols, censored$beta_ols)
    expect_identical(beta_unc, censored$beta_unc)
    expect_identical(beta_cens, censored$beta_cens)
}
```

## Session Info

```{r}
sessionInfo()
```

## Source

[R Markdown](https://github.com/bnaras/cvxr_docs/blob/master/content/cvxr_examples/censored-regression.Rmd)

## References
