---
title: Integer Programming
author: Anqi Fu and Balasubramanian Narasimhan
date: '2018-05-15'
slug: cvxr_integer-programming
bibliography: ../bibtex/cvxr_refs.bib
link-citations: true
categories: []
tags: []
params:
  mode: ignore
  testdata_dir: test_data_1.0
  data_dir: integer-programming
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir) 
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
library(kableExtra)
library(ggplot2)
draw_office_layout <- function(office_name = c("WC", paste("Office", 5:7), paste("Office", 1:4))) {
    ## layout is a vector laid out as a 2x4 matrix with lower row first
    d <- data.frame(
        x1 = c(seq(from = 0, to = 6, by = 2), seq(from = 0, to = 6, by = 2)),
        x2 = c(seq(from = 2, to = 8, by = 2), seq(from = 2, to = 8, by = 2)),
        y1 = c(rep(0, 4), rep(1.5, 4)),
        y2 = c(rep(1, 4), rep(2.5, 4)),
        type = c('WC', rep('Office', 7)),
        name = office_name
    )
    d2 <- data.frame(x1 = c(2.75, 4.5, 6.5),
                     x2 = c(3.25, 5.5, 7.5),
                     y1 = rep(0, 3),
                     y2 = rep(0.02,3))
    p <- ggplot() +
        scale_x_continuous(name = "x") +
        scale_y_continuous(name = "y") +
        geom_rect(data = d,
                  mapping = aes(xmin = x1, xmax = x2,
                                ymin = y1, ymax = y2, fill = type), color = "black") +
        geom_text(data = d,
                  mapping = aes(x = x1 + (x2 - x1) / 2, y = y1 + (y2 - y1) / 2,
                                label = name), size = 4) +
        geom_text(mapping = aes(x = 4, y = 1.25, label = "Hallway")) +
        geom_rect(data = d2,
                  mapping = aes(xmin = x1, xmax = x2, ymin = y1, ymax = y2),
                  fill = "white", alpha = 1) +
        theme(axis.line = element_blank(),
              axis.text.x = element_blank(),
              axis.text.y = element_blank(),
              axis.ticks = element_blank(),
              axis.title.x = element_blank(),
              axis.title.y = element_blank(),
              legend.position = "none",
              panel.background = element_blank(),
              panel.border = element_blank(),
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank(),
              plot.background = element_blank())
    p
}
```

Consider the following optimization problem.

$$
\begin{array}{ll}
\mbox{Maximize} & x_1 + 2x_2 - 0.1x_3 - 3x_4\\
\mbox{subject to} & x_1, x_2, x_3, x_4 >= 0\\
                  & x_1 + x_2 <= 5\\
                  & 2x_1 - x_2 >= 0\\
				  & -x_1 + 3x_2 >= 0\\
				  & x_3 + x_4 >= 0.5\\
				  & x_3 >= 1.1\\
				  & x_3 \mbox{ is integer.}
\end{array}
$$

`CVXR` provides the `Int` and `Bool` constructors for specifying
integer and boolean variables. These can be combined with `vstack`
(analog of `rbind`) or `hstack` (analog of `cbind`) to construct
new expressions. 

The above problem now in `CVXR`.

```{r}
y1 <- Variable(2)
if (packageVersion("CVXR") > "0.99-7") {
    y2 <- Variable(1, integer = TRUE)
} else {
    y2 <- Int(1)
}
y3 <- Variable(1)
x <- vstack(y1, y2, y3) ## Create x expression
C <- matrix(c(1, 2, -0.1, -3), nrow = 1)
objective <- Maximize(C %*% x)
constraints <- list(
    x >= 0,
    x[1] + x[2] <= 5,
    2 * x[1] - x[2] >= 0,
    -x[1] + 3 * x[2] >= 0,
    x[3] + x[4] >= 0.5,
    x[3] >= 1.1)
problem <- Problem(objective, constraints)
```

We can solve this problem as usual using the default `ECOS` (actually
`ECOS_BB`) solver and obtain the optimal value as well as the
solution. 

```{r}
result <- solve(problem, solver = "ECOS_BB")
cat(sprintf("Optimal value: %.3f\n", result$value))
ecos_solution <- result$getValue(x)
```

## Alternative Solvers

We can try other solvers and compare the solutions obtained, like
`LPSOLVE` and `GLPK` provided the respective R packages are installed
as documented in the tutorial [Using Other
Solvers](/cvxr_examples/using-other-solvers/).

_Note:_ `LPSOLVE` is now orphaned on CRAN and so no longer supported.

```{r}
##result <- solve(problem, solver = "LPSOLVE")
##lpsolve_solution <- result$getValue(x)
result <- solve(problem, solver = "GLPK")
glpk_solution <- result$getValue(x)
```

Finally, we can also try a commercial solver that can handle integer
programs. 

```{r}
result <- solve(problem, solver = "GUROBI")
gurobi_solution <- result$getValue(x)
```

Below is the table is solutions from all the solvers we used.

```{r}
solutions <- data.frame(ECOS = ecos_solution,
                        ## LPSOLVE = lpsolve_solution,
                        GLPK = glpk_solution,
                        GUROBI = gurobi_solution)
row.names(solutions) <- c("$x_1$", "$x_2$", "$x_3$", "$x_4$")
knitr::kable(solutions, format = "html") %>%
    kable_styling("striped") %>%
    column_spec(1:4, background = "#ececec")
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save"), error = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(ecos = ecos_solution, glpk = glpk_solution, gurobi = gurobi_solution), file = file.path(testdata_dir, "integer-programming1.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    intprog <- readRDS(file = file.path(testdata_dir, "integer-programming1.RDS"))
    expect_identical(ecos_solution, intprog$ecos)
    expect_identical(glpk_solution, intprog$glpk)
    expect_identical(gurobi_solution, intprog$gurobi)
}
```

## Office Assignment Problem

For a slightly more involved example, we consider the [office
assignment
problem](https://www.mathworks.com/help/optim/examples/office-assignments-by-binary-integer-programming.html).

The goal is to assign six people, Marcelo, Rakesh, Peter, Tom,
Marjorie, and Mary Ann, to seven offices. Each office can have no more
than one person, and each person gets exactly one office. So there
will be one empty office. People can give preferences for the offices,
and their preferences are considered based on their seniority.  Some
offices have windows, some do not, and one window is smaller than
others. Additionally, Peter and Tom often work together, so should be
in adjacent offices. Marcelo and Rakesh often work together, and
should be in adjacent offices.

```{r, fig.height=2}
draw_office_layout()
```

The office layout is shown above. Offices 1, 2, 3, and 4 are inside
offices (no windows). Offices 5, 6, and 7 have windows, but the window
in office 5 is smaller than the other two.

We begin by recording the names of the people and offices.

```{r}
people <- c('Mary Ann', 'Marjorie', 'Tom',
            'Peter', 'Marcelo', 'Rakesh')
offices <- c('Office 1', 'Office 2', 'Office 3',
             'Office 4','Office 5', 'Office 6', 'Office 7')
```

We also have the office preferences of each person for each of the
seven offices along with seniority data which is used to scale the
office preferences.

```{r}
preference_matrix <- matrix( c(0, 0, 0, 0, 10, 40, 50,
                               0, 0, 0, 0, 20, 40, 40,
                               0, 0, 0, 0, 30, 40, 30,
                               1, 3, 3, 3, 10, 40, 40,
                               3, 4, 1, 2, 10, 40, 40,
                               10, 10, 10, 10, 20, 20, 20),
                            byrow = TRUE, nrow = length(people))
rownames(preference_matrix) <- people
colnames(preference_matrix) <- offices

seniority <- c(9, 10, 5, 3, 1.5, 2)
weightvector <- seniority / sum(seniority)
PM <- diag(weightvector) %*% preference_matrix
```

We define the the occupancy variable which indicates, using values 1
or 0, who occupies which office.

```{r}
if (packageVersion("CVXR") > "0.99-7") {
    occupy <- Variable(length(people), length(offices), integer = TRUE)
} else {
    occupy <- Int(length(people), length(offices))
}
```

The objective is to _maximize the satisfaction_ of the preferences
weighted by seniority constrained by the fact the a person can only
occupy a single office and no office can have more than 1 person.

```{r}
objective <- if (packageVersion("CVXR") > "0.99-7") {
                 Maximize(sum_entries(multiply(PM, occupy)))
             } else {
                 Maximize(sum_entries(mul_elemwise(PM, occupy)))
             }

constraints <- list(
    occupy >= 0,
    occupy <= 1,
    sum_entries(occupy, axis = 1) == 1,
    sum_entries(occupy, axis = 2) <= 1
)
```
__Version 1.0 Note__ The pre-1.0 function `mul_elemwise` is now `multiply`.

We further add the constraint that Tom (person 3) and Peter (person 4)
should be no more than one office away, and ditto for Marcelo (person
5) and Rakesh (person 6).

```{r}
tom_peter <- list(
    occupy[3, 1] + sum_entries(occupy[4, ]) - occupy[4, 2] <= 1,
    occupy[3, 2] + sum_entries(occupy[4, ]) - occupy[4, 1] - occupy[4, 3] - occupy[4, 5] <= 1,
    occupy[3, 3] + sum_entries(occupy[4, ]) - occupy[4, 2] - occupy[4, 4] - occupy[4, 6] <= 1,
    occupy[3, 4] + sum_entries(occupy[4, ]) - occupy[4, 3] - occupy[4, 7] <= 1,
    occupy[3, 5] + sum_entries(occupy[4, ]) - occupy[4, 2] - occupy[4, 6] <= 1,
    occupy[3, 6] + sum_entries(occupy[4, ]) - occupy[4, 3] - occupy[4, 5] - occupy[4, 7] <= 1,
    occupy[3, 7] + sum_entries(occupy[4, ]) - occupy[4, 4] - occupy[4, 6] <= 1
)

marcelo_rakesh <- list(
    occupy[5, 1] + sum_entries(occupy[6, ]) - occupy[6, 2] <= 1,
    occupy[5, 2] + sum_entries(occupy[6, ]) - occupy[6, 1] - occupy[6, 3] - occupy[6, 5] <= 1,
    occupy[5, 3] + sum_entries(occupy[6, ]) - occupy[6, 2] - occupy[6, 4] - occupy[6, 6] <= 1,
    occupy[5, 4] + sum_entries(occupy[6, ]) - occupy[6, 3] - occupy[6, 7] <= 1,
    occupy[5, 5] + sum_entries(occupy[6, ]) - occupy[6, 2] - occupy[6, 6] <= 1,
    occupy[5, 6] + sum_entries(occupy[6, ]) - occupy[6, 3] - occupy[6, 5] - occupy[6, 7] <= 1,
    occupy[5, 7] + sum_entries(occupy[6, ]) - occupy[6, 4] - occupy[6, 6] <= 1
)

constraints <- c(constraints, tom_peter, marcelo_rakesh)
```

We are now ready to solve the problem.

```{r}
problem <- Problem(objective, constraints)
ecos_result <- solve(problem, solver = "ECOS_BB")
ecos_soln <- round(ecos_result$getValue(occupy), 0)
rownames(ecos_soln) <- people
colnames(ecos_soln) <- offices
```

We are now ready to plot the solution (after accounting for the WC).

```{r, fig.height=2}
office_assignment <- apply(ecos_soln, 1, which.max)
office_occupants <- names(office_assignment)[match(c(5:7, 1:4), office_assignment)]
office_occupants[is.na(office_occupants)] <- "Empty"
draw_office_layout(c("WC", office_occupants))
```


```{r, echo = FALSE, eval = params$mode %in% c("test", "save"), error = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(solution = ecos_soln), file = file.path(testdata_dir, "integer-programming2.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    intprog <- readRDS(file = file.path(testdata_dir, "integer-programming2.RDS"))
    expect_identical(ecos_soln, intprog$solution)
}
```

## Session Info

```{r}
sessionInfo()
```

## Source

[R
Markdown](https://github.com/bnaras/cvxr_docs/blob/master/content/cvxr_examples/integer-programming.Rmd)

## References
