---
title: L1 Trend Filtering
author: Anqi Fu and Balasubramanian Narasimhan
date: '2017-11-02'
slug: cvxr_l1-trend-filtering
bibliography: ../bibtex/cvxr_refs.bib
link-citations: true
categories: []
tags: []
---

```{r, echo = FALSE}
library(ggplot2)
suppressMessages(suppressWarnings(library(CVXR)))
```

## Introduction

@Kim:et_al:2009 propose the $l_1$ trend filtering method for trend
 estimation. The method solves an optimization problem of the form

$$
\begin{array}{ll}
\underset{\beta}{\mbox{minimize}} & \frac{1}{2}\sum_{i=1}^m (y_i - \beta_i)^2 + \lambda ||D\beta||_1
\end{array}
$$
where the variable to be estimated is $\beta$ and we are given the
problem data $y$ and $\lambda$. The matrix $D$ is the second-order
difference matrix,

$$
D = 
\left[
\begin{matrix} 1 &  -2  &  1      &        &        &    &   \\ 
                 &   1  & -2      &  1     &        &    &   \\
                 &      &  \ddots & \ddots & \ddots &    &   \\
				 &      &         & 1      & -2     &  1 &   \\
				 &      &         &        & 1      & -2 &  1\\				 
\end{matrix}
\right].
$$

The implementation is in both C and Matlab. Hadley Wickham provides an
R interface to the C code. This is
on [GitHub](https://github.com/hadley/l1tf) and can be installed via:

```{r, eval = FALSE}
library(devtools)
install_github("hadley/l1tf")
```

## Example

We will use the example in `l1tf` to illustrate. The package provides
the function `l1tf` which computes the trend estimate for a specified
$\lambda$. 

```{r}
library(l1tf)
sp_data <- data.frame(x = sp500$date,
                      y = sp500$log,
                      l1_50 = l1tf(sp500$log, lambda = 50),
                      l1_100 = l1tf(sp500$log, lambda = 100))
```


## The `CVXR` version

`CVXR` provides all the atoms and functions necessary to formulat the
problem in a few lines. For example, the $D$ matrix above is provided
by the function `diff(..., differences = 2)`. Notice how the
formulation tracks the mathematical construct above.


```{r}
## lambda = 50
y <- sp500$log
lambda_1 <- 50 
beta <- Variable(length(y))
objective_1 <- Minimize(0.5 * p_norm(y - beta) +
                        lambda_1 * p_norm(diff(x = beta, differences = 2), 1))
p1 <- Problem(objective_1)
betaHat_50 <- solve(p1)$getValue(beta)

## lambda = 100
lambda_2 <- 100
objective_2 <- Minimize(0.5 * p_norm(y - beta) +
                        lambda_2 * p_norm(diff(x = beta, differences = 2), 1))
p2 <- Problem(objective_2)
betaHat_100 <- solve(p2)$getValue(beta)
```

_NOTE_ Of course, `CVXR` is much slower since it is not optimized just
for one problem.

## Comparison Plots

A plot of the estimates for two values of $\lambda$ is shown below
using both approaches.  First the `l1tf` plot.

```{r, fig.width=7, fig.height=5, fig.cap="$L_1$ trends for $\\lambda = 50$ (red) and $\\lambda = 100$ (blue)."}
ggplot(data = sp_data) +
    geom_line(mapping = aes(x = x, y = y), color = 'grey50') +
    labs(x = "Date", y = "SP500 log-price") +
    geom_line(mapping = aes(x = x, y = l1_50), color = 'red', size = 1) +
    geom_line(mapping = aes(x = x, y = l1_100), color = 'blue', size = 1)
```

Next the corresponding `CVXR` plots. 

```{r, fig.width=7, fig.height=5, fig.cap="`CVXR` estimated $L_1$ trends for $\\lambda = 50$ (red) and $\\lambda = 100$ (blue)."}
cvxr_data <- data.frame(x = sp500$date,
                        y = sp500$log,
                        l1_50 = betaHat_50,
                        l1_100 = betaHat_100)
ggplot(data = cvxr_data) +
    geom_line(mapping = aes(x = x, y = y), color = 'grey50') +
    labs(x = "Date", y = "SP500 log-price") +
    geom_line(mapping = aes(x = x, y = l1_50), color = 'red', size = 1) +
    geom_line(mapping = aes(x = x, y = l1_100), color = 'blue', size = 1)
```

## Notes

The `CVXR` solution is not quite exactly that of `l1tf`: on the left it shows a larger difference for the two
$\lambda$ values; in the middle, it is less flatter than `l1tf`; and
on the right, it does not have as many knots as `l1tf`. 

## Session Info

```{r}
sessionInfo()
```

## Source

[R Markdown](https://github.com/bnaras/cvxr_docs/blob/master/content/post/examples/cvxr_l1-trend-filtering.Rmd)

## References
