---
title: Near Isotonic and Near Convex Regression
author: Anqi Fu and Balasubramanian Narasimhan
date: '2017-11-02'
slug: cvxr_near-isotonic-and-near-convex-regression
bibliography: ../bibtex/cvxr_refs.bib
link-citations: true
categories: []
tags: []
---

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
library(ggplot2)
library(boot)
```

Given a set of data points $y \in {\mathbf R}^m$,
@TibshiraniHoefling:2011 fit a nearly-isotonic approximation $\beta
\in {\mathbf R}^m$ by solving

$$
\begin{array}{ll}
\underset{\beta}{\mbox{minimize}} & \frac{1}{2}\sum_{i=1}^m (y_i - \beta_i)^2 + \lambda \sum_{i=1}^{m-1}(\beta_i - \beta_{i+1})_+,
\end{array}
$$

where $\lambda \geq 0$ is a penalty parameter and $x_+
=\max(x,0)$. This can be directly formulated in `CVXR`.  As an
example, we use global warming data from
the
[Carbon Dioxide Information Analysis Center (CDIAC)](http://cdiac.ess-dive.lbl.gov/ftp/trends/temp/jonescru/). The
data points are the annual temperature anomalies relative to the
1961--1990 mean.

```{r}
data(cdiac)
str(cdiac)
```

Since we plan to fit the regression and also get some idea of the
standard errors, we write a function that computes the fit for use in
bootstrapping. We directly call the solver in this instance, to save
computational time in bootstrapping. For more on this, see [Getting
Faster Results](/cvxr_examples/speed/).

```{r}
neariso_fit <- function(y, lambda) {
    m <- length(y)
    beta <- Variable(m)
    obj <- 0.5 * sum((y - beta)^2) + lambda * sum(pos(diff(beta)))
    prob <- Problem(Minimize(obj))
    prob_data <- get_problem_data(prob, solver = "ECOS")
    solver_output <- ECOSolveR::ECOS_csolve(c = prob_data[["c"]],
                                            G = prob_data[["G"]],
                                            h = prob_data[["h"]],
                                            dims = prob_data[["dims"]],
                                            A = prob_data[["A"]],
                                            b = prob_data[["b"]])
    unpack_results(prob, "ECOS", solver_output)$getValue(beta)
}
```
The `pos` atom evaluates $x_+$ elementwise on the input expression.

The `boot` library provides all the tools for bootstrapping but
requires a statistic function that takes particular arguments: a data
frame, followed by the bootstrap indices and any other arguments
($\lambda$ for instance). This is shown below.

_NOTE_ In what follows, we use a very small number of bootstrap
samples as the fits are time consuming.

```{r}
neariso_fit_stat <- function(data, index, lambda) {
    sample <- data[index,]                  # Bootstrap sample of rows
    sample <- sample[order(sample$year),]   # Order ascending by year
    neariso_fit(sample$annual, lambda)
}
```

```{r}
set.seed(123)
boot.neariso <- boot(data = cdiac, statistic = neariso_fit_stat, R = 10, lambda = 0.44)

ci.neariso <- t(sapply(seq_len(nrow(cdiac)),
                            function(i) boot.ci(boot.out = boot.neariso, conf = 0.95,
                                                type = "norm", index = i)$normal[-1]))
data.neariso <- data.frame(year = cdiac$year, annual = cdiac$annual, est = boot.neariso$t0,
                              lower = ci.neariso[, 1], upper = ci.neariso[, 2])
```

We can now plot the fit and confidence bands for the near isotonic
fit. 

```{r}
(plot.neariso <- ggplot(data = data.neariso) +
     geom_point(mapping = aes(year, annual), color = "red") +
     geom_line(mapping = aes(year, est), color = "blue") +
     geom_ribbon(mapping = aes(x = year, ymin = lower,ymax = upper),alpha=0.3) +
     labs(x = "Year", y = "Temperature Anomalies")
)
```
The curve follows the data well, but exhibits some choppiness in
regions with a steep trend. 

For a smoother curve, we can solve for the nearly-convex fit described
in the same paper: 

$$
\begin{array}{ll}
\underset{\beta}{\mbox{minimize}} & \frac{1}{2}\sum_{i=1}^m (y_i -
\beta_i)^2 + \lambda \sum_{i=1}^{m-2}(\beta_i - 2\beta_{i+1} + \beta_{i+2})_+ \end{array} 
$$

This replaces the first difference term with an approximation to the
second derivative at $\beta_{i+1}$. In `CVXR`, the only change
necessary is the penalty line: replacing `diff(x)` by 
`diff(x, differences = 2)`.

```{r}
nearconvex_fit <- function(y, lambda) {
    m <- length(y)
    beta <- Variable(m)
    obj <- 0.5 * sum((y - beta)^2) + lambda * sum(pos(diff(beta, differences = 2)))
    prob <- Problem(Minimize(obj))
    prob_data <- get_problem_data(prob, solver = "ECOS")
    solver_output <- ECOSolveR::ECOS_csolve(c = prob_data[["c"]],
                                            G = prob_data[["G"]],
                                            h = prob_data[["h"]],
                                            dims = prob_data[["dims"]],
                                            A = prob_data[["A"]],
                                            b = prob_data[["b"]])
    unpack_results(prob, "ECOS", solver_output)$getValue(beta)
}

nearconvex_fit_stat <- function(data, index, lambda) {
    sample <- data[index,]                  # Bootstrap sample of rows
    sample <- sample[order(sample$year),]   # Order ascending by year
    nearconvex_fit(sample$annual, lambda)
}

set.seed(987)
boot.nearconvex <- boot(data = cdiac, statistic = nearconvex_fit_stat, R = 5, lambda = 0.44)

ci.nearconvex <- t(sapply(seq_len(nrow(cdiac)),
                          function(i) boot.ci(boot.out = boot.nearconvex, conf = 0.95,
                                              type = "norm", index = i)$normal[-1]))
data.nearconvex <- data.frame(year = cdiac$year, annual = cdiac$annual, est = boot.nearconvex$t0,
                              lower = ci.nearconvex[, 1], upper = ci.nearconvex[, 2])

```

The resulting curve for the near convex fit is depicted below with
95\% confidence bands generated from $R = 5$ samples. Note the jagged
staircase pattern has been smoothed out. 


```{r}
(plot.nearconvex <- ggplot(data = data.nearconvex) +
     geom_point(mapping = aes(year, annual), color = "red") +
     geom_line(mapping = aes(year, est), color = "blue") +
     geom_ribbon(mapping = aes(x = year, ymin = lower,ymax = upper),alpha=0.3) +
     labs(x = "Year", y = "Temperature Anomalies")
)
```

## Notes

We can easily extend this example to higher-order differences or
lags. To make this easy, the function `diff` takes an argument
`differences` that is 1 by default; a third-order difference is specified
as `diff(x, differences = 3)`.

## Session Info

```{r}
sessionInfo()
```

## Source

[R Markdown](https://github.com/bnaras/cvxr_docs/blob/master/content/cvxr_examples/near-isotonic-and-near-convex-regression.Rmd)

## References
