---
title: Quantile Regression
author: Anqi Fu and Balasubramanian Narasimhan
date: '2017-11-02'
slug: cvxr_quantile-regression
bibliography: ../bibtex/cvxr_refs.bib
link-citations: true
categories: []
tags: []
params:
  mode: ignore
  testdata_dir: test_data
  data_dir: quantile-regression
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir) 
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
library(ggplot2)
library(kableExtra)
library(quantreg)
```


## Introduction

Quantile regression is another variation on least squares
\citep{quantile}. The loss is the tilted $l_1$ function,

\[
	\phi(u) = \tau\max(u,0) - (1-\tau)\max(-u,0) = \frac{1}{2}|u| + \left(\tau - \frac{1}{2}\right)u,
\]

where $\tau \in (0,1)$ specifies the quantile. The problem as before
is to minimize the total residual loss. This model is commonly used in
ecology, healthcare, and other fields where the mean alone is not
enough to capture complex relationships between variables. `CVXR`
allows us to create a function to represent the loss and integrate it
seamlessly into the problem definition, as illustrated below.


## Example

We will use an example from the `quantreg` package. The vignette
provides an example of the estimation and plot.

```{r}
data(engel)
p <- ggplot(data = engel) +
    geom_point(mapping = aes(x = income, y = foodexp), color = "blue")
taus <- c(0.1, 0.25, 0.5, 0.75, 0.90, 0.95)
fits <- data.frame(
    coef(lm(foodexp ~ income, data = engel)),
    sapply(taus, function(x) coef(rq(formula = foodexp ~ income, data = engel, tau = x))))
names(fits) <- c("OLS", sprintf("$\\tau_{%0.2f}$", taus))

nf <- ncol(fits)
colors <- colorRampPalette(colors = c("black", "red"))(nf)
p <- p + geom_abline(intercept = fits[1, 1], slope = fits[2, 1], color = colors[1], size = 1.5)
for (i in seq_len(nf)[-1]) {
    p <- p + geom_abline(intercept = fits[1, i], slope = fits[2, i], color = colors[i])
}
p
```

The above plot shows the quantile regression fits for $\tau = (0.1,
0.25, 0.5, 0.75, 0.90, 0.95)$. The OLS fit is the thick black line.
    
The following is a table of the estimates.

```{r}
knitr::kable(fits, format = "html", caption = "Fits from OLS and `quantreg`") %>%
    kable_styling("striped") %>%
    column_spec(1:8, background = "#ececec")
```

The `CVXR` formulation follows. Note we make use of `model.matrix` to
get the intercept column painlessly. 

```{r}
X <- model.matrix(foodexp ~ income, data = engel)
y <- matrix(engel[, "foodexp"], ncol = 1)
beta <- Variable(2)
quant_loss <- function(u, tau) { 0.5 * abs(u) + (tau - 0.5) * u }
solutions <- sapply(taus, function(tau) {
    obj <- sum(quant_loss(y - X %*% beta, t = tau))
    prob <- Problem(Minimize(obj))
    ## THE OSQP solver returns an error for tau = 0.5
    solve(prob, solver = "ECOS")$getValue(beta)
})
solutions <- data.frame(coef(lm(foodexp ~ income, data = engel)),
                        solutions)
names(fits) <- c("OLS", sprintf("$\\tau_{%0.2f}$", taus))
```

Here is a table similar to the above with the OLS estimate added in
for easy comparison.


```{r}
knitr::kable(fits, format = "html", caption = "Fits from OLS and `CVXR`") %>%
    kable_styling("striped") %>%
    column_spec(1:8, background = "#ececec")
```

The results match.

```{r, echo = FALSE, eval = params$mode %in% c("test", "save"), error = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(solutions, file = file.path(testdata_dir, "quantile-regression.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    quantile_solutions <- readRDS(file = file.path(testdata_dir, "quantile-regression.RDS"))
    expect_equal(solutions, quantile_solutions)
}
```

## Session Info

```{r}
sessionInfo()
```

## Source

[R Markdown](https://github.com/bnaras/cvxr_docs/blob/master/content/cvxr_examples/quantile-regression.Rmd)

## References
