---
title: Saturating Hinges Fit
author: Anqi Fu and Balasubramanian Narasimhan
date: '2017-11-02'
slug: cvxr_saturating_hinges
bibliography: ../bibtex/cvxr_refs.bib
link-citations: true
categories: []
tags: []
---

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
library(ggplot2)
library(tidyr)
```

## Introduction

The following example comes from work on saturating splines in
@BoydHastie:2016. Adaptive regression splines are commonly used in
statistical modeling, but the instability they exhibit beyond their
boundary knots makes extrapolation dangerous. One way to correct this
issue for linear splines is to require they _saturate_: remain
constant outside their boundary. This problem can be solved using a
heuristic that is an extension of lasso regression, producing a
weighted sum of hinge functions, which we call a _saturating hinge_.

For simplicity, consider the univariate case with $n = 1$. Assume we
are given knots $t_1 < t_2 < \cdots < t_k$ where each
$t_j \in {\mathbf R}$. Let $h_j$ be a hinge function at knot $t_j$, \ie,
$h_j(x) = \max(x-t_j,0)$, and define
$f(x) = w_0 + \sum_{j=1}^k w_jh_j(x)$. We want to solve

$$
\begin{array}{ll} 
\underset{w_0,w}{\mbox{minimize}} & \sum_{i=1}^m \ell(y_i, f(x_i)) + \lambda\|w\|_1 \\
\mbox{subject to} & \sum_{j=1}^k w_j = 0
\end{array}
$$

for variables $(w_0,w) \in {\mathbf R} \times {\mathbf R}^k$. The function
$\ell:{\mathbf R} \times {\mathbf R} \rightarrow {\mathbf R}$ is the loss associated
with every observation, and $\lambda \geq 0$ is the penalty weight. In
choosing our knots, we set $t_1 = \min(x_i)$ and $t_k = \max(x_i)$ so
that by construction, the estimate $\hat f$ will be constant outside
$[t_1,t_k]$.

## Example

We demonstrate this technique on the bone density data for female
patients from @ESL, section 5.4. There are a total of $m = 259$
observations. Our response $y_i$ is the change in spinal bone density
between two visits, and our predictor $x_i$ is the patient's age. We
select $k = 10$ knots about evenly spaced across the range of $X$ and
fit a saturating hinge with squared error loss
$\ell(y_i, f(x_i)) = (y_i - f(x_i))^2$.

```{r}
## Import and sort data
data(bone, package = "ElemStatLearn")
X <- bone[bone$gender == "female",]$age
y <- bone[bone$gender == "female",]$spnbmd
ord <- order(X, decreasing = FALSE)
X <- X[ord]
y <- y[ord]

## Choose knots evenly distributed along domain
k <- 10
lambdas <- c(1, 0.5, 0.01)
idx <- floor(seq(1, length(X), length.out = k))
knots <- X[idx]
```

In `R`, we first define the estimation and loss functions:

```{r}
## Saturating hinge
f_est <- function(x, knots, w0, w) {
    hinges <- sapply(knots, function(t) { pmax(x - t, 0) })
    w0 + hinges %*% w
}

## Loss function
loss_obs <- function(y, f) { (y - f)^2 }
```

This allows us to easily test different losses and knot locations
later. The rest of the set-up is similar to previous examples. We
assume that `knots` is an `R` vector representing
$(t_1,\ldots,t_k)$.

```{r}
## Form problem
w0 <- Variable(1)
w <- Variable(k)
loss <- sum(loss_obs(y, f_est(X, knots, w0, w)))
constr <- list(sum(w) == 0)

xrange <- seq(min(X), max(X), length.out = 100)
splines <- matrix(0, nrow = length(xrange), ncol = length(lambdas))
```

The optimal weights are retrieved using separate calls, as shown
below.

```{r}
for (i in seq_along(lambdas)) {
    lambda <- lambdas[i]
    reg <- lambda * p_norm(w, 1)
    obj <- loss + reg
    prob <- Problem(Minimize(obj), constr)

    ## Solve problem and save spline weights
    result <- solve(prob)
    w0s <- result$getValue(w0)
    ws <- result$getValue(w)
    splines[, i] <- f_est(xrange, knots, w0s, ws)
}
```

## Results

We plot the fitted saturating hinges in Figure 1 below. As expected,
when $\lambda$ increases, the spline exhibits less variation and grows
flatter outside its boundaries. 

```{r}
d <- data.frame(xrange, splines)
names(d) <- c("x", paste0("lambda_", seq_len(length(lambdas))))
plot.data <- gather(d, key = "lambda", value = "spline",
                    "lambda_1", "lambda_2", "lambda_3", factor_key = TRUE)
ggplot() +
    geom_point(mapping = aes(x = X, y = y)) +
    geom_line(data = plot.data, mapping = aes(x = x, y = spline, color = lambda)) +
    scale_color_discrete(name = expression(lambda),
                         labels = sprintf("%0.2f", lambdas)) +
    labs(x = "Age", y = "Change in Bone Density") +
    theme(legend.position = "top")
```

The squared error loss works well in this case, but the Huber loss is
preferred when the dataset contains large outliers. To see this, we
add 50 randomly generated
outliers to the bone density data and re-estimate the saturating
hinges. 

```{r}
## Add outliers to data
set.seed(1)
nout <- 50
X_out <- runif(nout, min(X), max(X))
y_out <- runif(nout, min(y), 3*max(y)) + 0.3
X_all <- c(X, X_out)
y_all <- c(y, y_out)

## Solve with squared error loss
loss_obs <- function(y, f) { (y - f)^2 }
loss <- sum(loss_obs(y_all, f_est(X_all, knots, w0, w)))
prob <- Problem(Minimize(loss + reg), constr)
result <- solve(prob)
spline_sq <- f_est(xrange, knots, result$getValue(w0), result$getValue(w))

## Solve with Huber loss
loss_obs <- function(y, f, M) { huber(y - f, M) }
loss <- sum(loss_obs(y, f_est(X, knots, w0, w), 0.01))
prob <- Problem(Minimize(loss + reg), constr)
result <- solve(prob)
spline_hub <- f_est(xrange, knots, result$getValue(w0), result$getValue(w))
```

Figure 2 shows the results. For a Huber loss with $M = 0.01$, the
resulting spline is fairly smooth and follows the shape of the
original data, as opposed to the spline using squared error loss,
which is biased upwards by a significant amount.

```{r}
d <- data.frame(xrange, spline_hub, spline_sq)
names(d) <- c("x", "Huber", "Squared")
plot.data <- gather(d, key = "loss", value = "spline",
                    "Huber", "Squared", factor_key = TRUE)
ggplot() +
    geom_point(mapping = aes(x = X, y = y)) +
    geom_point(mapping = aes(x = X_out, y = y_out), color = "orange") +
    geom_line(data = plot.data, mapping = aes(x = x, y = spline, color = loss)) +
    scale_color_discrete(name = "Loss",
                         labels = c("Huber", "Squared")) +
    labs(x = "Age", y = "Change in Bone Density") +
    theme(legend.position = "top")
```


## Session Info

```{r}
sessionInfo()
```

## Source

[R Markdown](https://github.com/bnaras/cvxr_docs/blob/master/content/cvxr_examples/saturating-hinges.Rmd)

## References
