---
title: Solver Peculiarities
author: Anqi Fu and Balasubramanian Narasimhan
date: '2018-05-15'
slug: cvxr_solver-peculiarities
bibliography: ../bibtex/cvxr_refs.bib
link-citations: true
categories: []
tags: []
---

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
```

The default solver in `CVXR` is `ECOS`. However, it is not always the
best solver to use. As an example, let us consider again the [catenary
problem](/cvxr_examples/cvxr_catenary/). 

We will change the problem slightly to use a finer discretization from
101 points to say 501. 

```{r}
## Problem data
m <- 501
L <- 2
h <- L / (m - 1)

## Form objective
x <- Variable(m)
y <- Variable(m)
objective <- Minimize(sum(y))

## Form constraints
constraints <- list(x[1] == 0, y[1] == 1,
                    x[m] == 1, y[m] == 1,
                    diff(x)^2 + diff(y)^2 <= h^2)

## Solve the catenary problem
prob <- Problem(objective, constraints)
result <- solve(prob)
```

The solution status is no longer optimal.

```{r}
cat("Solution status is", result$status)
```

In such cases, using a different solver may give more accurate
results. Let us try `MOSEK` for example.

```{r}
result <- solve(prob, solver = "MOSEK")
cat("Solution status is", result$status)
```

This returns an optimal solution. 

Here again, even commercial solvers differ; GUROBI, for example, does
not completely solve the problem and in fact throws an error.


## Further Development

Future versions of `CVXR` will improve support for the commercial
solvers by taking advantage of problem reductions described in
@rewriting:2018 to hand the problem to solvers in a form that is
closer to what the R packages (for those solvers) expect.  This should
provide better behavior and error reporting.

## Session Info

```{r}
sessionInfo()
```

## Source

[R Markdown](https://github.com/bnaras/cvxr_docs/blob/master/content/cvxr_examples/solver-peculiarities.Rmd)

## References
