---
title: "Dealing with Solver-Specific Parameters"
author: Anqi Fu, David Kang and Balasubramanian Narasimhan
date: "2020-01-15"
slug: cvxr_solver-parameters
bibliography: ../bibtex/cvxr_refs.bib
link-citations: true
params:
  mode: ignore
  testdata_dir: test_data_1.0
  data_dir: solver_specific
---

```{r, echo = FALSE}
suppressMessages(library(CVXR))
library(tibble)
library(kableExtra)
```

## Overview

`CVXR` 1.0 has now updated how it handles its solver-specific
parameters and has diverged slightly from how `cvxpy` handles it. It
now features five standard parameters that are the default parameters
for each solve function. Here are the five parameters

* `verbose` : A parameter that deals with the verbosity of the
  solver. It is a boolean variable for all solvers, but it can also
  take integer values for the `MOSEK` solver, whose interface deals
  with an integer verbosity parameter with the default of 10. The
  default value of `verbose` for `CVXR` is set to `FALSE`.

* `reltol` : Deals with the relative tolerance of the solver. 

* `abstol` : Deals with the absolute tolerance of the solver. 

* `feastol` : Deals with feasible tolerance of the solver. 

* `num_iter` : A parameter that specifies the maximum number of
  iterations for the solver. 

The users can also pass in solver-specific parameters to the `solve`
function based on the solver they specify. On the whole, the
additional parameters given by the user will override the five values
specified above.

For each solver, the tables below show how the five standard
parameters are mapped to the solver. Cells with `NA` values indicate
no mapping and any specified value will result in a warning about
parameter being ignored.

## Solver-Specific Parameters

We will go over some examples for each solvers supported by `CVXR`,
the available parameters for each solver, and where to find the
documentation for each solver.

### OSQP

The parameters available for `OSQP` can be found in this
[link](https://osqp.org/docs/interfaces/solver_settings.html). 

```{r, echo = FALSE}
osqp <- tibble::tibble(
                    Standard = c("verbose", "reltol", "abstol", "feastol", "num_iter"),
                    OSQP = c("verbose", "eps_rel", "eps_abs", "eps_{prim/dual}_inf", "max_iter"),
                    Value = c("FALSE", "1e-05", "1e-05", "1e-04", "10000")
                )
knitr::kable(osqp,
             caption = "Default CVXR-OSQP parameter mappings",
             format = "html") %>%
    kable_styling("striped") %>%
    column_spec(2:3, background = "#ececec")
```

The parameters can be entered individually in the arguments of the
solve function or can be passed through as a list. If the user enters
in a parameter that is not specified or accepted by `OSQP`, then the
solver will ignore the parameter and proceed to solve the
problem. Here is an example.

```{r}
x <- Variable(1)
obj <- Minimize(x)
prob <- Problem(obj, list(x >= 0))

# Inside Function
result  <- solve(prob, solver = "OSQP", feastol = 1e-2, eps_prim_inf = 1e-5)

# Can also do it by passing in a list
more_params <- list(verbose = TRUE, feastol = 1e-2,
                    eps_prim_inf = 1e-5, eps_prim_inf = 1e-3)
result <- solve(prob, solver = "OSQP", verbose = FALSE, more_params)
```

In the second case that the `verbose` in the `more_params` list overrides
the value of `verbose` listed earlier in the invocation.

### SCS

The parameters available for `SCS` can be found in this
[link](https://cran.r-project.org/web/packages/scs/scs.pdf) under the
`scs_control` function. 

```{r, echo = FALSE}
scs <- tibble::tibble(
                    Standard = c("verbose", "reltol", "abstol", "feastol", "num_iter"),
                    SCS = c("verbose", NA, NA, NA, "max_iters"),
                    Value = c("FALSE", NA, NA, NA, "2500")
                )
knitr::kable(scs,
             caption = "Default CVXR-SCS parameter mappings",
             format = "html") %>%
    kable_styling("striped") %>%
    column_spec(2:3, background = "#ececec")
```


Much like `OSQP`, the parameters can be
entered individually in the arguments of the solve function or can be
passed through as a list. If the user enters in a parameter that is
not specified or accepted by `SCS`, then the solver will ignore the
parameter and proceed to solve the problem. Here is an example:

```{r}
n<-3
p<-3
C <- matrix(rnorm(n^2), ncol = n)
A <- list(matrix(rnorm(n*n), nrow = n, ncol = n), matrix(rnorm(n*n), nrow = n, ncol = n), matrix(rnorm(n*n), nrow = n, ncol = n))
b <- matrix(rnorm(p), nrow = p)

X <- Variable(c(n, n), symmetric = TRUE)
constraints <- list(X >= 0)
for(i in 1:p){
  constraints <- c(constraints, list(matrix_trace(A[[i]]%*%X) == b[i] ))
}

obj <- Minimize(matrix_trace(C %*% X))
prob <- Problem(obj, constraints)
result  <- solve(prob, solver = "SCS", verbose = FALSE, feastol = 1e-2, cg_rate = 1.5, scale = 2)
```

Notice how in this case a warning is thrown: the `SCS` solver does not
support the feasible tolerance parameter, so the solver ignores it.

### ECOS

The parameters available for `ECOS` can be found in this
[link](https://cran.r-project.org/web/packages/ECOSolveR/ECOSolveR.pdf)
under the `ecos.control` function. 

```{r, echo = FALSE}
ecos <- tibble::tibble(
                    Standard = c("verbose", "reltol", "abstol", "feastol", "num_iter"),
                    ECOS = c("verbose", "reltol", "abstol", "feastol", "max_iter"),
                    Value = c("FALSE", "1e-08", "1e-08", "1e-08", "100")
                )
knitr::kable(ecos,
             caption = "Default CVXR-ECOS parameter mappings",
             format = "html") %>%
    kable_styling("striped") %>%
    column_spec(2:3, background = "#ececec")
```

Much like `OSQP`, the parameters can be entered individually in the
arguments of the solve function or can be passed through as a list. If
the user enters in a parameter that is not specified or accepted by
`ECOS`, then the solver will ignore the parameter and proceed to solve
the problem. Here is an example:

```{r}
x <- Variable(1)
obj <- Minimize(x)
prob <- Problem(obj, list(x >= 0))
result  <- solve(prob, solver = "ECOS", verbose = FALSE, feastol = 1e-2, reltol_inacc = 1e-5)
```    

### CPLEX

The parameters available for `CPLEX` can be found in this
[link](https://cran.r-project.org/web/packages/Rcplex/Rcplex.pdf)
under the details section for the function `Rcplex` function on
page 3. 

```{r, echo = FALSE}
cplex <- tibble::tibble(
                     Standard = c("verbose", "reltol", "abstol", "feastol", "num_iter"),
                     CPLEX = c("trace", NA, NA, NA, "itlim"),
                     Value = c("FALSE", NA, NA, NA, "10000")
                )
knitr::kable(cplex,
             caption = "Default CVXR-CPLEX parameter mappings",
             format = "html") %>%
    kable_styling("striped") %>%
    column_spec(2:3, background = "#ececec")
```


Much like `OSQP`, the parameters can be entered individually
in the arguments of the solve function or can be passed through as a
list. If the user enters in a parameter that is not specified or
accepted by `CPLEX`, then the solver will give the user a warning but
will proceed to solve the problem. Here is an example:

```{r}
x <- Variable(1)
obj <- Minimize(x)
prob <- Problem(obj, list(x >= 0))
result  <- solve(prob, solver = "CPLEX", verbose = FALSE, eps_abs_tol = 1e-2, epagap = 1e-5)
```

As you can see, `eps_abs_tol` is not one of `CPLEX`'s specified
parameters, so the solver threw a warning.

### MOSEK

```{r, echo = FALSE}
mosek <- tibble::tibble(
                     Standard = c("verbose", "reltol", "abstol", "feastol", "num_iter"),
                     MOSEK = c("verbose", NA, NA, NA, "num_iter"),
                     Value = c(0, NA, NA, NA, "10000")
                )
knitr::kable(mosek,
             caption = "Default CVXR-MOSEK parameter mappings",
             format = "html") %>%
    kable_styling("striped") %>%
    column_spec(2:3, background = "#ececec")
```

The parameters available for `MOSEK` can be seen in this
[link](https://docs.mosek.com/9.1/rmosek/parameters.html#doc-all-parameter-list). As
you can see, the number of parameters is quite massive. They are also
split between three times of parameters in `Double`, `Integer`, and
`String` parameters. To pass these into the solver, they must be put
in a list beforehand. Moreover, there are also general solver
parameters that do not belong in any of the three categories. These
parameters can be found in detail in the `Rmosek` package
documentation, but here are the parameters

* `verbose` (Another important note here is while the default values
  of `TRUE` and `FALSE` work, `MOSEK` uses integer values for verbose
  with the default being 10. The `TRUE` value for verbose corresponds
  to 10 in the solver)
* `usesol`
* `useparam`
* `soldetail`
* `getinfo`
* `writebefore`
* `writeafter`

These parameters are passed into the `solve` function as normal. Here is an example:

```{r}
x <- Variable(1)
obj <- Minimize(x)
prob <- Problem(obj, list(x >= 0))
iparam <- list()
iparam$AUTO_UPDATE_SOL_INFO <- "OFF"
iparam$BI_IGNORE_NUM_ERROR <- "OFF"
iparam$BI_MAX_ITERATIONS <- 100000

dparam <- list()
dparam$BASIS_REL_TOL_S <- 1.0e-12
dparam$BASIS_TOL_S <- 1.0e-6

result  <- solve(prob, solver = "MOSEK", verbose = FALSE, soldetail = 3, dparam = dparam, iparam = iparam)
```

A quick detail in this case is that the solver threw a warning about
the `soldetail` parameter. Both the `soldetail` and `getinfo`
parameter controls the amount of data the solve returns, and some of
them might be needed in order to invert the solution. Therefore,
proceed with caution when using these variables as they can cause the
solver to not obtain the answer.

### GUROBI


```{r, echo = FALSE}
gurobi <- tibble::tibble(
                      Standard = c("verbose", "reltol", "abstol", "feastol", "num_iter"),
                      GUROBI = c("OutputFlag", NA, NA, "{Feasibility/Optimality}Tol", "IterationLimit"),
                      Value = c(0, NA, NA, "1e-06", "10000")
                  )
knitr::kable(gurobi,
             caption = "Default CVXR-GUROBI parameter mappings",
             format = "html") %>%
    kable_styling("striped") %>%
    column_spec(2:3, background = "#ececec")
```

The parameters available for `GUROBI` can be seen in this
[link](https://www.gurobi.com/documentation/6.5/refman/parameters.html#sec:Parameters). If
an unaccepted parameter is passed into the `solve` function, then the
function will not run to completion. The solver will also not let you
know which parameter is at fault unless the parameter is
continuous. Here is an example.

```{r}
x <- Variable(1)
obj <- Minimize(x)
prob <- Problem(obj, list(x >= 0))
result  <- solve(prob, solver = "GUROBI", verbose = FALSE, BarIterLimit = 20, Quad = 1)
```    
    
### GLPK


```{r, echo = FALSE}
glpk <- tibble::tibble(
                    Standard = c("verbose", "reltol", "abstol", "feastol", "num_iter"),
                    GLPK = c("verbose", NA, NA, NA, NA),
                    Value = c(0, NA, NA, NA, NA)
                )
knitr::kable(glpk,
             caption = "Default CVXR-GLPK parameter mappings",
             format = "html") %>%
    kable_styling("striped") %>%
    column_spec(2:3, background = "#ececec")
```

The parameters available for `GLPK` can be seen in this
[link](https://cran.r-project.org/web/packages/Rglpk/Rglpk.pdf) under
the `Details` section in page 4. There are only four parameters in
`verbose`, `presolve`, `tm_limit`, `canonicalize_status`. Here is an
example

```{r}
x <- Variable(1)
obj <- Minimize(x)
prob <- Problem(obj, list(x >= 0))
result  <- solve(prob, solver = "GLPK", verbose = FALSE, presolve = FALSE, tm_limit = 10, canonicalize_status = TRUE)
```    
    
### CBC


```{r, echo = FALSE}
cbc <- tibble::tibble(
                    Standard = c("verbose", "reltol", "abstol", "feastol", "num_iter"),
                    CBC = c("verbose", NA, NA, NA, NA),
                    Value = c(0, NA, NA, NA, NA)
                )
knitr::kable(cbc,
             caption = "Default CVXR-CBC parameter mappings",
             format = "html") %>%
    kable_styling("striped") %>%
    column_spec(2:3, background = "#ececec")
```

The parameters available for `CBC` can be seen in this
[link](https://projects.coin-or.org/CoinBinary/export/1059/OptimizationSuite/trunk/Installer/files/doc/cbcCommandLine.pdf). Much
like most of the solvers outlined above, if the user passes in a
unspecified parameter into the solver, then the solver just ignores
the parameter and proceeds to solve the problem. Here is an example

```{r}
x <- Variable(1)
obj <- Minimize(x)
prob <- Problem(obj, list(x >= 0))
result  <- solve(prob, solver = "CBC", verbose = FALSE, sec = 5, maxn = 10)
```
	
## Session Info

```{r}
sessionInfo()
```

## Source

[R Markdown](https://github.com/bnaras/cvxr_docs/blob/master/content/cvxr_examples/solver-parameters.Rmd)

