---
title: Getting Faster Results
author: Anqi Fu and Balasubramanian Narasimhan
date: '2017-11-02'
slug: cvxr_speed
bibliography: ../bibtex/cvxr_refs.bib
link-citations: true
categories: []
tags: []
params:
  mode: ignore
  testdata_dir: test_data
  data_dir: speed
---

## Warning

1. The solution described below is useful when you mathematically know
   a problem is DCP-compliant and none of your data inputs will change
   the nature of the problem. We recommend that users check the
   DCP-compliance of a problem (via a call to `is_dcp(prob)` for
   example) at least once to ensure this is the case.  _Not verifying
   DCP-compliance may result in garbage!_
   
2. Note also that the large speed gains in previous versions are no longer
   evident in the version 1.x because the new reduction framework has
   really made `CVXR` faster.
   
## Introduction

As was remarked in the [introduction to
`CVXR`](/cvxr_examples/intro/), its chief advantage is
_flexibility_: you can specify a problem in close to mathematical form
and `CVXR` solves it for you, if it can. Behind the scenes, `CVXR`
compiles the domain specific language and verifies the convexity of
the problem before sending it off to solvers. If the problem violates
the rules of [Disciplined Convex Programming](/post/cvxr_dcp/) it is
rejected.

Therefore, it is generally slower than tailor-made solutions to a
given problem.

## An Example

To understand the speed issues, let us consider the global warming
data from the [Carbon Dioxide Information Analysis Center
(CDIAC)](http://cdiac.ess-dive.lbl.gov/ftp/trends/temp/jonescru/)
again. The data points are the annual temperature anomalies relative
to the 1961--1990 mean. We will fit the nearly-isotonic approximation
$\beta \in {\mathbf R}^m$ by solving

$$
\begin{array}{ll}
\underset{\beta}{\mbox{Minimize}} & \frac{1}{2}\sum_{i=1}^m (y_i - \beta_i)^2 + \lambda \sum_{i=1}^{m-1}(\beta_i - \beta_{i+1})_+,
\end{array}
$$
where $\lambda \geq 0$ is a penalty parameter and $x_+ =\max(x,0)$. 

This can be solved as follows.

```{r}
suppressMessages(suppressWarnings(library(CVXR)))
data(cdiac)
y <- cdiac$annual
m <- length(y)
lambda <- 0.44
beta <- Variable(m)
obj <- 0.5 * sum((y - beta)^2) + lambda * sum(pos(diff(beta)))
prob <- Problem(Minimize(obj))
soln <- solve(prob, solver = "ECOS")
betaHat <- soln$getValue(beta)
```

This is the _recommended way_ to solve a problem. 

However, suppose we wished to construct bootstrap confidence intervals
for the estimate using 100 resamples.  It is clear that this
computation time can quickly become limiting . 

Below, we show how one can get at the problem data and directly call a
solver to get faster results.

## Profile the code 

Profiling a single fit to the model is useful to figure out where most
of the time is spent.

```{r}
library(profvis)
y <- cdiac$annual
profvis({
    beta <- Variable(m)
    obj <- Minimize(0.5 * sum((y - beta)^2) + lambda * sum(pos(diff(beta))))
    prob <- Problem(obj)
    soln <- solve(prob)
    betaHat <- soln$getValue(beta)
})
```

It is especially instructive to click on the _data_ tab and open up
the tree for `solve` to see the sequence of calls and cumulative time
used. 

The profile shows that most of the total time (2400ms for one of our
runs) time is spent in the call to `is_dcp` generic (about
2000ms). This generic is responsible to ensuring that all the problem
is DCP-compliant by checking the nature of each of the components that
make up the problem. The actual solving took a _much smaller_ fraction
of the time.

## Directly Calling the Solver

We are mathematically certain that the above is convex and so we can
avoid the `is_dcp` hit. We can obtain the the problem data for a
particular solver (like `OSQP`, or `ECOS` or `SCS`) using the function
`get_problem_data` and directly hand that data to the solver to get
the solution.

```{r}
prob_data <- get_problem_data(prob, solver = "ECOS")
```

_ASIDE_: How did we know ECOS was the solver to use? Future versions
will provide a function to match a solver to a problem. (Actually, it
is available already, but not exported yet!). For now, a single call
to `solve` with the verbose option set to `TRUE` can provide that
information. 

```{r, eval = FALSE}
soln <- solve(prob, verbose = TRUE)
```
Now that we have the problem data and know which solver to use, we can
call the ECOS solver with the right arguments. (The ECOS solver is
provided by the package `ECOSolveR`, which `CVXR` imports.)

```{r}
if (packageVersion("CVXR") > "0.99-7") {
    ECOS_dims <- ECOS.dims_to_solver_dict(prob_data$data[["dims"]])
} else {
    ECOS_dims <- prob_data$data[["dims"]]
}
solver_output <- ECOSolveR::ECOS_csolve(c = prob_data$data[["c"]],
                                        G = prob_data$data[["G"]],
                                        h = prob_data$data[["h"]],
                                        dims = ECOS_dims,
                                        A = prob_data$data[["A"]],
                                        b = prob_data$data[["b"]])
```

Finally, we can obtain the results by asking `CVXR` to _unpack_ the
solver results for us. (See `?unpack_results` for further examples.)

```{r}
if (packageVersion("CVXR") > "0.99-7") {
    direct_soln <- unpack_results(prob, solver_output, prob_data$chain, prob_data$inverse_data)
} else {
    direct_soln <- unpack_results(prob, "ECOS", solver_output)
}
```

## Profile the Direct Call

We can profile this direct call now.

```{r}
profvis({
    beta <- Variable(m)
    obj <- Minimize(0.5 * sum((y - beta)^2) + lambda * sum(pos(diff(beta))))
    prob <- Problem(obj)
    prob_data <- get_problem_data(prob, solver = "ECOS")
    if (packageVersion("CVXR") > "0.99-7") {
        ECOS_dims <- ECOS.dims_to_solver_dict(prob_data$data[["dims"]])
    } else {
        ECOS_dims <- prob_data$data[["dims"]]
    }
    solver_output <- ECOSolveR::ECOS_csolve(c = prob_data$data[["c"]],
                                            G = prob_data$data[["G"]],
                                            h = prob_data$data[["h"]],
                                            dims = ECOS_dims,
                                            A = prob_data$data[["A"]],
                                            b = prob_data$data[["b"]])
    if (packageVersion("CVXR") > "0.99-7") {
        direct_soln <- unpack_results(prob, solver_output, prob_data$chain, prob_data$inverse_data)
    } else {
        direct_soln <- unpack_results(prob, "ECOS", solver_output)
    }
})
```

For one of our runs, the total time went down from $2400ms$ to
$690ms$, more than a 3-fold speedup! In cases where the objective
function and constraints are more complex, the speedup can be more
than 10-fold. 

## Same Answer?

Of course, we should also verify that the results obtained in both
cases are same.

```{r}
identical(betaHat, direct_soln$getValue(beta))
```

## Session Info

```{r}
sessionInfo()
```

## Source

[R Markdown](https://github.com/bnaras/cvxr_docs/blob/master/content/cvxr_examples/speed.Rmd)

## References
