---
title: Using Other Solvers
author: Anqi Fu and Balasubramanian Narasimhan
date: '2018-05-15'
slug: cvxr_using-other-solvers
bibliography: ../bibtex/cvxr_refs.bib
link-citations: true
categories: []
tags: []
---

```{r, echo = FALSE}
suppressWarnings(suppressMessages(library(CVXR)))
```

## Introduction

The default installation of `CVXR` comes with two (imported) open source solvers:

- [ECOS](https://github.com/embotech/ecos) and its mixed integer
  cousin `ECOS_BB` via the CRAN package
  [ECOSolveR](https://cloud.r-project.org/package=ECOSolveR)
- [SCS](https://github.com/cvxgrp/scs) via the CRAN package
  [scs](https://cloud.r-project.org/package=scs).
  
`CVXR` can also make use of several other open source solvers
implemented in R packages. 

- The linear and mixed integer programming package
  [`lpSolve`](http://lpsolve.sourceforge.net/5.5/) via the 
  [`lpSolveAPI`](https://cloud.r-project.org/package=lpSolveAPI) package
- The linear and mixed integer programming package [`GLPK`](https://www.gnu.org/software/glpk/) via the 
  [`Rglpk`](https://cloud.r-project.org/package=Rglpk) package.
  
Since these are optional, you have to install these packages yourself.

```{r}
lapply(list(LPSOLVE = "lpSolveAPI",
            GLPK = "Rglpk"),
       function(x) x %in% installed.packages()[, 1])
```

Once the packages are installed, a call to `installed_solvers` will
indicate what solvers `CVXR` is aware of.

## Commercial Solvers

A few commercial solvers are also currently supported: [MOSEK](https://www.mosek.com) and
[GUROBI](https://www.gurobi.com).

At the moment, the support uses vendor python solver packages via
[`reticulate`](https://cran.r-project.org/package=reticulate), _not R
packages_. Future versions will provide support directly using
[_problem
reductions_](https://web.stanford.edu/~boyd/papers/cvxpy_rewriting.html),
recently implemented in [`CVXPY` version 1.0](https://www.cvxpy.org/).

Thus, one needs two pre-requisites for using these commercial solvers:

- a python installation in addition to R
- the
  [`reticulate`](https://cran.r-project.org/package=reticulate) R
  package. 
  
We also recommend that you ensure that `reticulate` is installed
correctly and working. For example `reticulate::py_eval('1+1')` should return
`2`. 

__N.B.__ If you have an Anaconda installation in your path, you have
to account for the fact that there may be interactions when using
RStudio and rendering documents.  In particular, Anaconda may include
its own version of pandoc and other tools that may conflict with what
Rstudio needs to do its work properly.

To be concrete, one problem we found was that the `MOSEK` solver was
not recognized as being available in this rendered document, even
though the command line interface showed it to be present. Ensuring an
appropriate `PATH` variable solves the problem. 


### Installing `MOSEK`

[MOSEK](https://www.mosek.com) provides an academic version that is
free of charge. As noted in the downloads page, Anaconda users can
install merely via:

```{bash, eval=FALSE}
conda install -c mosek mosek
```

Others can use the `pip` command:

```{bash, eval = FALSE}
pip install -f https://download.mosek.com/stable/wheel/index.html Mosek
```

In addition, the license for the product has to be activated per
instructions on the `Sales` section of the web page.

Once activated, you can check that `CVXR` recognizes the solver;
`installed_solvers()` should list `MOSEK`. Otherwise, rinse and repeat
until success.

### Installing `GUROBI`

[GUROBI](https://www.gurobi.com) also provides an academic version
that is free of charge. They require you to register. 

Once registered, install the _Gurobi Optimizer_ software and activate
your license as necessary.

Once activated, you can check that `CVXR::installed_solvers()` lists
`GUROBI`. Otherwise, rinse and repeat until success.

### Example session

```{r}
installed_solvers()
```

## Session Info

```{r}
sessionInfo()
```

## Source

[R Markdown](https://github.com/bnaras/cvxr_docs/blob/master/content/cvxr_examples/using-other-solvers.Rmd)

## References
