---
title: Using Other Solvers
author: Anqi Fu and Balasubramanian Narasimhan
date: '2018-05-15'
slug: cvxr_using-other-solvers
bibliography: ../bibtex/cvxr_refs.bib
link-citations: true
categories: []
tags: []
---


```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
```

## Introduction

The default installation of `CVXR` comes with several (imported) open source solvers:

- [ECOS](https://github.com/embotech/ecos) and its mixed integer
  cousin `ECOS_BB` via the CRAN package
  [ECOSolveR](https://cloud.r-project.org/package=ECOSolveR)
- [OSQP](https://www.osqp.org) 
- [SCS](https://github.com/cvxgrp/scs) via the CRAN package
  [scs](https://cran.r-project.org/package=scs).
  
`CVXR` can also make use of several other open source solvers
implemented in R packages. 

- The linear and mixed integer programming package
  [`lpSolve`](http://lpsolve.sourceforge.net/5.5/) via the 
  [`lpSolveAPI`](https://cloud.r-project.org/package=lpSolveAPI)
  package. __NOTE:__ This solver is now orphaned on CRAN and so is not
  supported any more.
- The linear and mixed integer programming package [`GLPK`](https://www.gnu.org/software/glpk/) via the 
  [`Rglpk`](https://cloud.r-project.org/package=Rglpk) package.
- The CBC solver via the R package
  [`rcbc`](https://github.com/dirkschumacher/rcbc). This is not yet on
  CRAN 

Since these are optional, you have to install these packages yourself.

```{r}
lapply(list("CBC" = "rcbc",
            GLPK = "Rglpk"),
       function(x) x %in% installed.packages()[, 1])
```

Once the packages are installed, a call to `installed_solvers` will
indicate what solvers `CVXR` is aware of.

## Commercial Solvers

A few commercial solvers are also currently supported: [MOSEK](https://www.mosek.com) and
[GUROBI](https://www.gurobi.com),
[CPLEX](https://www.ibm.com/analytics/cplex-optimizer).

Previous versions of `CVXR` used vendor python solver packages via
[`reticulate`](https://cran.r-project.org/package=reticulate), _not R
packages_. This has changed with version 1.0, where we directly call
native R packages, resulting in a cleaner interface.


### Installing `MOSEK`

[MOSEK](https://www.mosek.com) provides an academic version that is
free of charge. As noted in the downloads page, Anaconda users can
install merely via:

```{bash, eval=FALSE}
conda install -c mosek mosek
```

Others can use the `pip` command:

```{bash, eval = FALSE}
pip install -f https://download.mosek.com/stable/wheel/index.html Mosek
```

In addition, the license for the product has to be activated per
instructions on the `Sales` section of the web page.

Once activated, you can check that `CVXR` recognizes the solver;
`installed_solvers()` should list `MOSEK`. Otherwise, rinse and repeat
until success.

### Installing `GUROBI`

[GUROBI](https://www.gurobi.com) also provides an academic version
that is free of charge. They require you to register. 

Once registered, install the _Gurobi Optimizer_ software and activate
your license as necessary.

Once activated, you can check that `CVXR::installed_solvers()` lists
`GUROBI`. Otherwise, rinse and repeat until success.

### Installing `Rcplex`

[CPLEX](https://www.ibm.com/analytics/cplex-optimizer) is available in
a community edition. After installation, the
[`Rcplex`](https://cran.r-project.org/package=Rcplex) package needs to
be told where to find the libraries for linking. A sample session on
macOS is shown below for reference. 

```{bash, eval = FALSE}
R CMD INSTALL Rcplex_0.3-3.tar.gz --configure-args="--with-cplex-include=/Applications/CPLEX_Studio_Community129/cplex/include --with-cplex-lib='-L/Applications/CPLEX_Studio_Community129/cplex/lib/x86-64_osx/static_pic -lilocplex -lcplex'"
```

### Example session

```{r}
installed_solvers()
```

## Session Info

```{r}
sessionInfo()
```

## Source

[R Markdown](https://github.com/bnaras/cvxr_docs/blob/master/content/cvxr_examples/using-other-solvers.Rmd)

## References
