---
title: Elastic Net
author: Anqi Fu and Balasubramanian Narasimhan
categories: []
tags: []
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: elastic-net
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
## We restrict ourselves to solvers on CRAN
library(ggplot2)
library(tidyr)
library(kableExtra)
library(glmnet)
source(here::here("_check_status.R"))
```

## Introduction

Often in applications, we encounter problems that require
regularization to prevent overfitting, introduce sparsity, facilitate
variable selection, or impose prior distributions on parameters. Two
of the most common regularization functions are the $l_1$-norm and
squared $l_2$-norm, combined in the elastic net regression model
[@elasticnet, @glmnet],

$$
\begin{array}{ll}
\underset{\beta}{\mbox{minimize}} & \frac{1}{2m}\|y - X\beta\|_2^2 +
\lambda(\frac{1-\alpha}{2}\|\beta\|_2^2 + \alpha\|\beta\|_1).
\end{array}
$$

Here $\lambda \geq 0$ is the overall regularization weight and
$\alpha \in [0,1]$ controls the relative $l_1$ versus squared $l_2$
penalty. Thus, this model encompasses both ridge ($\alpha = 0$) and
lasso ($\alpha = 1$) regression.

## Example

To solve this problem in `CVXR`, we first define a function that
calculates the regularization term given the variable and penalty
weights.

```{r}
elastic_reg <- function(beta, lambda = 0, alpha = 0) {
    ridge <- (1 - alpha) / 2 * sum_squares(beta)
    lasso <- alpha * p_norm(beta, 1)
    lambda * (lasso + ridge)
}
```

Later, we will add it to the scaled least squares loss as shown below.

```{r, eval = FALSE}
loss <- sum_squares(y - X %*% beta) / (2 * n)
obj <- loss + elastic_reg(beta, lambda, alpha)
```

The advantage of this modular approach is that we can easily
incorporate elastic net regularization into other regression
models. For instance, if we wanted to run regularized Huber
regression, `CVXR` allows us to reuse the above code with just a
single changed line.

```{r, eval = FALSE}
loss <- huber(y - X %*% beta, M)
```

We generate some synthetic sparse data for this example. We use
correlated features (Toeplitz structure with $\rho = 0.8$) and set
$n = 100$, $p = 50$ so that regularization is essential. Correlated
features are the classic motivation for elastic net over pure lasso:
lasso tends to pick one variable from a correlated group arbitrarily,
while elastic net encourages the group to enter or leave together.

We also scale $y$ so that $\sqrt{\text{mean}(y^2)} = 1$. The reason
is explained in the `glmnet` comparison below.

```{r}
set.seed(1)

# Problem data
p <- 50
n <- 100

# Correlated features: Toeplitz covariance Sigma_ij = rho^|i-j|
rho <- 0.8
Sigma <- rho^abs(outer(seq_len(p), seq_len(p), "-"))
X <- MASS::mvrnorm(n, mu = rep(0, p), Sigma = Sigma)

# Sparse true coefficients: 10 non-zero with varying magnitudes
beta_true <- rep(0, p)
beta_true[c(1, 5, 10, 15, 20, 25, 30, 35, 40, 45)] <-
    c(3, -2.5, 2, -1.5, 1, -3, 2.5, -2, 1.5, -1)

# Noise calibrated for SNR ~ 3
signal_sd <- sqrt(drop(t(beta_true) %*% Sigma %*% beta_true))
sigma <- signal_sd / 3
y <- X %*% beta_true + rnorm(n, sd = sigma)

# Scale y so that sqrt(mean(y^2)) = 1
y <- y / sqrt(mean(y^2))
```

We fit the elastic net model for several values of $\lambda$.

```{r}
TRIALS <- 10
lambda_vals <- 10^seq(-2, 0, length.out = TRIALS)
beta <- Variable(p)
loss <- sum_squares(y - X %*% beta) / (2 * n)

## Elastic-net regression
alpha <- 0.75
beta_vals <- sapply(lambda_vals,
                    function(lambda) {
                        obj <- loss + elastic_reg(beta, lambda, alpha)
                        prob <- Problem(Minimize(obj))
                        psolve(prob)
                        check_solver_status(prob)
                        value(beta)
                    })
```

We can now get a table of the coefficients. With $p = 50$ coefficients,
we show only the 10 that correspond to the true non-zero entries.

```{r}
nz_idx <- which(beta_true != 0)
d <- as.data.frame(beta_vals[nz_idx, ])
rownames(d) <- sprintf("\\(\\beta_{%d}\\)", nz_idx)
names(d) <- sprintf("\\(\\lambda = %.3f\\)", lambda_vals)
knitr::kable(d, format = "html", escape = FALSE, caption = "Elastic net fits from <code>CVXR</code> (non-zero coefficients)", digits = 3) |>
    kable_styling("striped") |>
    column_spec(1:11, background = "#ececec")
```

We plot the regularization path for all coefficients. With correlated
features and $n$ only twice $p$, the coefficients sweep dramatically
from their unregularized values down to zero as $\lambda$ increases.

```{r}
df_path <- data.frame(lambda = lambda_vals, t(beta_vals)) |>
    pivot_longer(-lambda, names_to = "Coefficient", values_to = "Value")
ggplot(df_path, aes(x = lambda, y = Value, color = Coefficient)) +
    geom_line() +
    scale_x_log10() +
    labs(x = expression(lambda), y = expression(beta),
         title = "Regularization Path for Elastic-net Regression") +
    theme_minimal() +
    guides(color = "none")
```

## Comparison with `glmnet`

We can compare with results from `glmnet`. One subtlety: for the
Gaussian family with `intercept = FALSE`, `glmnet` internally divides
$y$ by $s = \sqrt{\text{mean}(y^2)}$ before fitting, then rescales the
returned coefficients by $s$. Because the internal fit uses $\tilde{y}
= y/s$ while the penalty is unchanged, the $l_1$ and $l_2$ terms in
the elastic net get reweighted by different powers of $s$. No single
$\lambda$-correction can undo this when $\alpha \in (0, 1)$.

The simplest fix is to ensure $s = 1$ so the standardization is a
no-op, which is why we scaled $y$ above. With that in place, the same
$\lambda$ values give both methods the same effective problem.

```{r}
model_net <- glmnet(X, y, family = "gaussian", alpha = alpha,
                    lambda = lambda_vals, standardize = FALSE,
                    intercept = FALSE)
## glmnet returns columns in decreasing lambda order; reverse to match
coef_net <- as.data.frame(as.matrix(coef(model_net)[-1, seq(TRIALS, 1, by = -1)]))
d_net <- coef_net[nz_idx, ]
rownames(d_net) <- sprintf("\\(\\beta_{%d}\\)", nz_idx)
names(d_net) <- sprintf("\\(\\lambda = %.3f\\)", lambda_vals)
knitr::kable(d_net, format = "html", escape = FALSE, digits = 3, caption = "Coefficients from <code>glmnet</code> (non-zero coefficients)") |>
    kable_styling("striped") |>
    column_spec(1:11, background = "#ececec")
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save"), }
if (params$mode == "save") {
    saveRDS(list(beta_vals = beta_vals), file = file.path(testdata_dir, "elastic-net.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    e_net <- readRDS(file = file.path(testdata_dir, "elastic-net.RDS"))
    expect_identical(beta_vals, e_net$beta_vals)
}
```


## Session Info

```{r, echo = FALSE}
sessionInfo()
```

## References
