---
title: Solver Peculiarities
author: Anqi Fu and Balasubramanian Narasimhan
date: '2018-05-15'
slug: cvxr_solver-peculiarities
bibliography: ../bibtex/cvxr_refs.bib
link-citations: true
categories: []
tags: []
params:
  mode: ignore
  testdata_dir: test_data_1.0
  data_dir: solver-peculiarities
---

```{r, echo = FALSE, message = FALSE, eval = FALSE}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir) 
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
```

The default solver in `CVXR` is `Clarabel`. However, it is not always the
best solver to use. As an example, let us consider again the [catenary
problem](/cvxr_examples/cvxr_catenary/).

We will change the problem slightly to use a finer discretization from
101 points to say 501. 

```{r}
## Problem data
m <- 501
L <- 2
h <- L / (m - 1)

## Form objective
x <- Variable(m)
y <- Variable(m)
objective <- Minimize(sum(y))

## Form constraints
constraints <- list(x[1] == 0, y[1] == 1,
                    x[m] == 1, y[m] == 1,
                    diff(x)^2 + diff(y)^2 <= h^2)

## Solve the catenary problem
prob <- Problem(objective, constraints)
result <- psolve(prob, solver = "CLARABEL")
```

The solution status is no longer optimal.

```{r}
cat("Solution status is", problem_status(prob))
```

```{r, echo = FALSE, eval = FALSE}
cat("Testthat Results: No output is good\n")
expect_identical(problem_status(prob), "optimal_inaccurate")
```


In such cases, using a different solver may give more accurate
results. Let us try `MOSEK` for example.

```{r}
result <- psolve(prob, solver = "MOSEK")
cat("Solution status is", problem_status(prob))
```

```{r, echo = FALSE, eval = FALSE}
cat("Testthat Results: No output is good\n")
expect_identical(problem_status(prob), "optimal")
```

This returns an optimal solution. 

Different solvers have varying strengths, and it is worth trying
alternatives when one solver does not produce satisfactory results.


## Session Info

```{r}
sessionInfo()
```

## Source

[R Markdown](https://github.com/bnaras/cvxr_docs/blob/master/content/cvxr_examples/solver-peculiarities.Rmd)

## References
