---
title: "Minimize Beamwidth of an Antenna Array"
author: "Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: app_antenna_array
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
library(ggplot2)
```

## Introduction

Adapted from the CVX example of the same name, by Almir Mutapcic,
2/2/2006 and Judson Wilson, 5/14/2014.

**References:**

- "Convex optimization examples" lecture notes (EE364) by S. Boyd
- "Antenna array pattern synthesis via convex optimization" by H. Lebret and S. Boyd

This algorithm designs an antenna array such that:

- it has unit sensitivity at some target direction
- it obeys a constraint on a minimum sidelobe level outside the beam
- it minimizes the beamwidth of the pattern.

This is a quasiconvex problem. Define the target direction as
$\theta_{\mathrm{tar}}$, and a beamwidth of
$\Delta \theta_{\mathrm{bw}}$. The beam occupies the angular interval

$$
\Theta_b = \left(\theta_{\mathrm{tar}}
- \frac{1}{2}\Delta \theta_{\mathrm{bw}},\;
\theta_{\mathrm{tar}}
+ \frac{1}{2}\Delta \theta_{\mathrm{bw}}\right).
$$

Solving for the minimum beamwidth $\Delta \theta_{\mathrm{bw}}$ is
performed by bisection, where the interval which contains the optimal
value is bisected according to the result of the following feasibility
problem:

$$
\begin{array}{ll}
\mbox{minimize}   &  0 \\
\mbox{subject to} & y(\theta_{\mathrm{tar}}) = 1 \\
                  &  |y(\theta)| \leq t_{\mathrm{sb}}
                         \quad \forall \theta \notin \Theta_b.
\end{array}
$$

Here $y$ is the antenna array gain pattern (a complex-valued
function), $t_{\mathrm{sb}}$ is the maximum allowed sideband gain
threshold, and the variables are $w$ (antenna array weights or shading
coefficients). The gain pattern is a linear function of $w$:
$y(\theta) = w^T a(\theta)$ for some $a(\theta)$ describing the
antenna array configuration and specs.

Once the optimal beamwidth is found, the solution $w$ is refined with:

$$
\begin{array}{ll}
\mbox{minimize}   & \|w\| \\
\mbox{subject to} & y(\theta_{\mathrm{tar}}) = 1 \\
                  & |y(\theta)| \leq t_{\mathrm{sb}}
                          \quad \forall \theta \notin \Theta_b.
\end{array}
$$

The implementation below discretizes the angular quantities.

## Problem Data

We use a random 2D positioning of antennas.

```{r}
## Problem specs
lambda_wl <- 1         # wavelength
theta_tar <- 60        # target direction (degrees)
min_sidelobe <- -20    # maximum sidelobe level in dB
max_half_beam <- 50    # starting half beamwidth (must be feasible)

## 2D random antenna locations
set.seed(1)
n <- 36
L <- 5
loc <- matrix(L * runif(n * 2), nrow = n, ncol = 2)
```

### Build Optimization Matrices

We construct the matrix $A$ that relates the weights $w$ and the gain
pattern $y(\theta)$, i.e., $y = Aw$.

```{r}
## Theta in degrees from 1 to 360
theta <- matrix(1:360, ncol = 1)

## Build matrix A: y = A * w (complex-valued)
A <- exp(2i * pi / lambda_wl *
         (cos(pi * theta / 180) %*% t(loc[, 1]) +
          sin(pi * theta / 180) %*% t(loc[, 2])))

## Target constraint
ind_closest <- which.min(abs(theta - theta_tar))
Atar <- A[ind_closest, , drop = FALSE]
```

## Solve Using Bisection

Since CVXR works with real variables, we represent complex-valued
quantities by separating real and imaginary parts. A complex vector
$w = w_R + i w_I$ is stored as a real vector $(w_R, w_I)$.

```{r}
## Set up real/imaginary representation of Atar
Atar_R <- Re(Atar)
Atar_I <- Im(Atar)
Atar_RI <- rbind(cbind(Atar_R, -Atar_I),
                 cbind(Atar_I,  Atar_R))

## Target vector: [1, 0] (unit gain, no imaginary component)
realones_ri <- c(1.0, 0.0)

## Bisection
halfbeam_bot <- 1
halfbeam_top <- max_half_beam

cat("We are only considering integer values of the half beam-width\n")
cat("(since we are sampling the angle with 1 degree resolution).\n\n")

while (halfbeam_top - halfbeam_bot > 1) {
    halfbeam_cur <- ceiling((halfbeam_top + halfbeam_bot) / 2)

    ## Indices of the stopband (outside the beam)
    ind <- which(theta <= (theta_tar - halfbeam_cur) |
                 theta >= (theta_tar + halfbeam_cur))
    As <- A[ind, , drop = FALSE]

    ## Real/imaginary representation
    As_R <- Re(As)
    As_I <- Im(As)
    As_RI_top <- cbind(As_R, -As_I)
    As_RI_bot <- cbind(As_I,  As_R)

    ## CVXR variables and constraints
    w_ri <- Variable(2 * n)
    constraints <- list(Atar_RI %*% w_ri == realones_ri)

    ## Add sidelobe constraints row by row
    sidelobe_thresh <- 10^(min_sidelobe / 20)
    for (i in seq_len(nrow(As))) {
        As_ri_row <- rbind(As_RI_top[i, , drop = FALSE],
                           As_RI_bot[i, , drop = FALSE])
        constraints <- c(constraints,
                         list(p_norm(As_ri_row %*% w_ri, 2) <= sidelobe_thresh))
    }

    ## Solve feasibility problem
    obj <- Minimize(0)
    prob <- Problem(obj, constraints)
    result <- psolve(prob, solver = "SCS")

    ## Bisection logic
    if (status(prob) == "optimal") {
        cat(sprintf("Problem is feasible for half beam-width = %g degrees\n",
                    halfbeam_cur))
        halfbeam_top <- halfbeam_cur
    } else {
        cat(sprintf("Problem is not feasible for half beam-width = %g degrees\n",
                    halfbeam_cur))
        halfbeam_bot <- halfbeam_cur
    }
}

halfbeam <- halfbeam_top
cat(sprintf("\nOptimum half beam-width for given specs is %g degrees\n", halfbeam))
```

### Minimum Noise Design

We now compute the minimum-norm weights for the optimal beamwidth.

```{r}
## Recompute stopband for optimal beamwidth
ind <- which(theta <= (theta_tar - halfbeam) |
             theta >= (theta_tar + halfbeam))
As <- A[ind, , drop = FALSE]

As_R <- Re(As)
As_I <- Im(As)
As_RI_top <- cbind(As_R, -As_I)
As_RI_bot <- cbind(As_I,  As_R)

w_ri <- Variable(2 * n)
constraints <- list(Atar_RI %*% w_ri == realones_ri)

sidelobe_thresh <- 10^(min_sidelobe / 20)
for (i in seq_len(nrow(As))) {
    As_ri_row <- rbind(As_RI_top[i, , drop = FALSE],
                       As_RI_bot[i, , drop = FALSE])
    constraints <- c(constraints,
                     list(p_norm(As_ri_row %*% w_ri, 2) <= sidelobe_thresh))
}

## Minimize the weight norm
obj <- Minimize(p_norm(w_ri, 2))
prob <- Problem(obj, constraints)
result_final <- psolve(prob, solver = "SCS")
cat(sprintf("Final objective value: %f\n", result_final))
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(halfbeam = halfbeam, obj_value = result_final),
            file = file.path(testdata_dir, "antenna.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    ant_results <- readRDS(file = file.path(testdata_dir, "antenna.RDS"))
    expect_equal(halfbeam, ant_results$halfbeam, tolerance = 1e-5)
    expect_equal(result_final, ant_results$obj_value, tolerance = 1e-5)
}
```

## Result Plots

### Antenna Locations

```{r, fig.cap = "Antenna locations in 2D plane"}
ggplot(data.frame(x = loc[, 1], y = loc[, 2])) +
    geom_point(aes(x = x, y = y), shape = 1, size = 3, color = "blue") +
    labs(title = "Antenna Locations", x = "x", y = "y") +
    coord_fixed() +
    theme_minimal()
```

### Array Pattern

```{r, fig.cap = "Array gain pattern in dB"}
## Compute full gain pattern
A_R <- Re(A)
A_I <- Im(A)
A_RI <- rbind(cbind(A_R, -A_I), cbind(A_I, A_R))

w_ri_val <- value(w_ri)
y_full <- A_RI %*% w_ri_val
n_theta <- length(theta)
y_complex <- y_full[1:n_theta] + 1i * y_full[(n_theta + 1):(2 * n_theta)]
y_mag_db <- 20 * log10(abs(y_complex))

df_pattern <- data.frame(
    theta = as.vector(theta),
    gain_db = as.vector(y_mag_db)
)

ggplot(df_pattern, aes(x = theta, y = gain_db)) +
    geom_line() +
    geom_vline(xintercept = theta_tar, linetype = "dashed", color = "green") +
    geom_vline(xintercept = theta_tar + halfbeam, linetype = "dashed", color = "red") +
    geom_vline(xintercept = theta_tar - halfbeam, linetype = "dashed", color = "red") +
    ylim(-40, 0) +
    labs(x = "Look angle (degrees)", y = "Magnitude (dB)",
         title = "Array Gain Pattern") +
    theme_minimal()
```

### Polar Pattern

```{r, fig.cap = "Polar radiation pattern"}
zerodB <- 50
dBY <- pmax(y_mag_db + zerodB, 0)
theta_rad <- as.vector(theta) * pi / 180

df_polar <- data.frame(
    x = as.vector(dBY) * cos(theta_rad),
    y = as.vector(dBY) * sin(theta_rad)
)

## 0 dB circle
circle_0dB <- data.frame(
    x = zerodB * cos(theta_rad),
    y = zerodB * sin(theta_rad)
)

## Sidelobe circle
m_sl <- min_sidelobe + zerodB
circle_sl <- data.frame(
    x = m_sl * cos(theta_rad),
    y = m_sl * sin(theta_rad)
)

ggplot() +
    geom_path(data = df_polar, aes(x = x, y = y), color = "blue") +
    geom_path(data = circle_0dB, aes(x = x, y = y), linetype = "dotted") +
    geom_path(data = circle_sl, aes(x = x, y = y), linetype = "dotted") +
    coord_fixed(xlim = c(-zerodB, zerodB), ylim = c(-zerodB, zerodB)) +
    annotate("text", x = -zerodB, y = 2, label = "0 dB") +
    annotate("text", x = -m_sl, y = 2, label = paste0(min_sidelobe, " dB")) +
    labs(title = "Polar Radiation Pattern") +
    theme_void()
```

## Session Info

```{r}
sessionInfo()
```

## References
