---
title: "Channel Capacity"
author: "CVXPY Developers and Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: app_channel_capacity
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
```

## Introduction

This example is adapted from Boyd and Vandenberghe, *Convex
Optimization*, exercise 4.57, pages 207--208.

Convex optimization can be used to find the channel capacity $C$ of a
discrete memoryless channel. Consider a communication channel with
input $X(t) \in \{1, 2, \ldots, n\}$ and output $Y(t) \in \{1, 2,
\ldots, m\}$. The random variables $X$ and $Y$ can take $n$ and $m$
different values, respectively.

In a discrete memoryless channel, the relation between the input and
the output is given by the transition probability:

$$
p_{ij} = \mathbb{P}(Y(t) = i \mid X(t) = j).
$$

These transition probabilities form the channel transition matrix $P
\in \mathbb{R}^{m \times n}$. The columns of $P$ must sum to 1 and
all elements of $P$ must be non-negative.

Assume that $X$ has a probability distribution denoted by $x \in
\mathbb{R}^n$, meaning that $x_j = \mathbb{P}(X(t) = j)$ for $j \in
\{1, \ldots, n\}$.

From Shannon's theory, the channel capacity is given by the maximum
possible mutual information $I$ between $X$ and $Y$:

$$
C = \sup_x I(X; Y),
$$

where

$$
I(X;Y) = -\sum_{i=1}^{m} y_i \log_2 y_i + \sum_{j=1}^{n} \sum_{i=1}^{m} x_j p_{ij} \log_2 p_{ij},
$$

and $y = Px$ is the output distribution.

Given that $x \log x$ is convex for $x \geq 0$, we can formulate this
as a convex optimization problem:

$$
\begin{array}{ll}
\mbox{maximize} & I(X; Y) \\
\mbox{subject to} & \sum_{i=1}^{n} x_i = 1, \quad x \succeq 0.
\end{array}
$$

Due to the entropy function in `CVXR`, this can be written quite
easily in DCP form.

## Channel Capacity Function

We write a function that computes the channel capacity for a given
transition matrix $P$ with $n$ input and $m$ output symbols.

```{r}
channel_capacity <- function(n, m, P, sum_x = 1) {
    ## x is probability distribution of the input signal X(t)
    x <- Variable(n)

    ## y is the probability distribution of the output signal Y(t)
    y <- P %*% x

    ## Compute c_j = sum_i p_{ij} log2(p_{ij})
    ## Use the convention 0*log(0) = 0 by handling zeros
    P_log <- ifelse(P > 0, log2(P), 0)
    c_vec <- colSums(P * P_log)

    ## Mutual information: I = c'x + sum(entr(y)) / log(2)
    I <- t(c_vec) %*% x + sum_entries(entr(y)) / log(2)

    ## Channel capacity is maximized by maximising the mutual information
    obj <- Maximize(I)
    constraints <- list(sum(x) == sum_x, x >= 0)

    ## Form and solve problem
    prob <- Problem(obj, constraints)
    result <- psolve(prob)

    list(status = status(prob),
         capacity = result,
         x = value(x))
}
```

## Example

In this example we consider a communication channel with two possible
inputs and outputs, so $n = m = 2$. The channel transition matrix we
use is:

$$
P = \begin{pmatrix} 0.75 & 0.25 \\ 0.25 & 0.75 \end{pmatrix}.
$$

Note that the columns of $P$ must sum to 1 and all elements of $P$
must be positive.

```{r, results = "hold"}
n <- 2
m <- 2
P <- matrix(c(0.75, 0.25,
              0.25, 0.75), nrow = m, ncol = n, byrow = TRUE)

result <- channel_capacity(n, m, P)
cat("Problem status:", result$status, "\n")
cat(sprintf("Optimal value of C = %.4g\n", result$capacity))
cat("Optimal variable x =\n")
print(round(result$x, 4))
```

The optimal input distribution is uniform ($x = [0.5, 0.5]$) and the
channel capacity is approximately $C \approx 0.1887$ bits.

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(capacity = result$capacity, x = result$x),
            file = file.path(testdata_dir, "channel_capacity.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    saved <- readRDS(file = file.path(testdata_dir, "channel_capacity.RDS"))
    expect_equal(result$capacity, saved$capacity, tolerance = 1e-5)
    expect_equal(result$x, saved$x, tolerance = 1e-5)
}
```

## A Larger Example

We can also consider a non-symmetric channel with $n = 4$ inputs and
$m = 3$ outputs.

```{r, results = "hold"}
n2 <- 4
m2 <- 3
set.seed(42)
## Generate a random transition matrix (columns sum to 1)
P2 <- matrix(runif(m2 * n2), nrow = m2, ncol = n2)
P2 <- sweep(P2, 2, colSums(P2), "/")

result2 <- channel_capacity(n2, m2, P2)
cat("Problem status:", result2$status, "\n")
cat(sprintf("Optimal value of C = %.4g\n", result2$capacity))
cat("Optimal variable x =\n")
print(round(result2$x, 4))
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(capacity = result2$capacity, x = result2$x),
            file = file.path(testdata_dir, "channel_capacity2.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    saved2 <- readRDS(file = file.path(testdata_dir, "channel_capacity2.RDS"))
    expect_equal(result2$capacity, saved2$capacity, tolerance = 1e-5)
    expect_equal(result2$x, saved2$x, tolerance = 1e-5)
}
```

## Session Info

```{r, echo = FALSE}
sessionInfo()
```

## References

- Boyd, S. and Vandenberghe, L. (2004). *Convex Optimization*. Cambridge University Press. Exercise 4.57, pages 207--208.
- Shannon, C.E. (1948). A Mathematical Theory of Communication. *Bell System Technical Journal*, 27, 379--423, 623--656.
