---
title: "Consensus Optimization"
author: "Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: app_consensus_opt
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
library(ggplot2)
```

## Introduction

This example is adapted from the CVXPY documentation on consensus
optimization.

Suppose we have a convex optimization problem with $N$ terms in the
objective:

$$
\begin{array}{ll}
\mbox{minimize} & \sum_{i=1}^N f_i(x).
\end{array}
$$

For example, we might be fitting a model to data and $f_i$ is the
loss function for the $i$th block of training data.

We can convert this problem into **consensus form**:

$$
\begin{array}{ll}
\mbox{minimize} & \sum_{i=1}^N f_i(x_i) \\
\mbox{subject to} & x_i = z,
\end{array}
$$

where we interpret the $x_i$ as local variables (particular to a
given $f_i$) and $z$ as the global variable. The constraints $x_i =
z$ enforce consistency, or consensus.

## ADMM Algorithm

We can solve a problem in consensus form using the Alternating
Direction Method of Multipliers (ADMM). Each iteration of ADMM
reduces to the following updates:

$$
\begin{array}{lll}
x^{k+1}_i & := & \arg\min_{x_i} \left( f_i(x_i) + \frac{\rho}{2} \left\| x_i - \bar{x}^k + u^k_i \right\|^2_2 \right) \\
u^{k+1}_i & := & u^{k}_i + x^{k+1}_i - \bar{x}^{k+1}
\end{array}
$$

where $\bar{x}^k = \frac{1}{N} \sum_{i=1}^N x^k_i$.

## Example: Consensus Least Squares

We illustrate the consensus ADMM approach on a simple least-squares
problem where the data is split across $N$ blocks. The original
problem is:

$$
\mbox{minimize} \quad \|A x - b\|_2^2,
$$

where $A \in \mathbb{R}^{m \times n}$ and $b \in \mathbb{R}^m$.

We split $A$ and $b$ into $N$ blocks row-wise, so $f_i(x) = \|A_i x
- b_i\|_2^2$.

```{r}
set.seed(1)

## Problem dimensions
n <- 10        # variable dimension
m <- 400       # total measurements
N <- 4         # number of blocks
block_size <- m / N

## Generate problem data
A_full <- matrix(rnorm(m * n), nrow = m, ncol = n)
x_true <- rnorm(n)
b_full <- A_full %*% x_true + 0.1 * rnorm(m)

## Split data into N blocks
A_blocks <- lapply(1:N, function(i) {
    rows <- ((i-1) * block_size + 1):(i * block_size)
    A_full[rows, ]
})
b_blocks <- lapply(1:N, function(i) {
    rows <- ((i-1) * block_size + 1):(i * block_size)
    b_full[rows]
})
```

## Centralized Solution

First, we solve the problem in the standard (non-distributed) way for
comparison.

```{r}
x_central <- Variable(n)
obj_central <- Minimize(sum_squares(A_full %*% x_central - b_full))
prob_central <- Problem(obj_central)
result_central <- psolve(prob_central)

x_central_val <- as.numeric(value(x_central))
cat(sprintf("Centralized solution status: %s\n", status(prob_central)))
cat(sprintf("Centralized objective: %.6f\n", result_central))
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(x_central = x_central_val, obj_central = result_central),
            file = file.path(testdata_dir, "consensus_central.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    saved <- readRDS(file = file.path(testdata_dir, "consensus_central.RDS"))
    expect_equal(x_central_val, saved$x_central, tolerance = 1e-5)
    expect_equal(result_central, saved$obj_central, tolerance = 1e-5)
}
```

## ADMM Consensus Solution

Now we implement the consensus ADMM algorithm. In each iteration, we
solve $N$ local subproblems using `CVXR` and then average to get the
global update.

```{r}
rho <- 1.0
MAX_ITER <- 50

## Initialize
x_vars <- matrix(0, nrow = n, ncol = N)   # local variables
u_vars <- matrix(0, nrow = n, ncol = N)   # dual variables
xbar <- rep(0, n)                          # global average

## Track convergence
obj_history <- numeric(MAX_ITER)

for (iter in seq_len(MAX_ITER)) {
    ## x-update: solve each local subproblem
    for (i in 1:N) {
        x_i <- Variable(n)
        f_i <- sum_squares(A_blocks[[i]] %*% x_i - b_blocks[[i]])
        augmented <- f_i + (rho / 2) * sum_squares(x_i - xbar + u_vars[, i])
        prob_i <- Problem(Minimize(augmented))
        result_i <- psolve(prob_i)
        x_vars[, i] <- as.numeric(value(x_i))
    }

    ## z-update: average the x_i
    xbar <- rowMeans(x_vars)

    ## u-update: update dual variables
    for (i in 1:N) {
        u_vars[, i] <- u_vars[, i] + x_vars[, i] - xbar
    }

    ## Compute objective at current average
    obj_history[iter] <- sum((A_full %*% xbar - b_full)^2)
}

cat(sprintf("ADMM objective after %d iterations: %.6f\n", MAX_ITER, obj_history[MAX_ITER]))
cat(sprintf("Centralized objective: %.6f\n", result_central))
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(xbar = xbar, obj_final = obj_history[MAX_ITER]),
            file = file.path(testdata_dir, "consensus_admm.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    saved <- readRDS(file = file.path(testdata_dir, "consensus_admm.RDS"))
    expect_equal(xbar, saved$xbar, tolerance = 1e-3)
    expect_equal(obj_history[MAX_ITER], saved$obj_final, tolerance = 1e-3)
}
```

## Convergence Plot

```{r, fig.cap = "ADMM convergence: objective value over iterations"}
df_conv <- data.frame(
    iteration = 1:MAX_ITER,
    objective = obj_history
)

ggplot(df_conv, aes(x = iteration, y = objective)) +
    geom_line(color = "steelblue") +
    geom_hline(yintercept = result_central, linetype = "dashed",
               color = "red") +
    annotate("text", x = MAX_ITER * 0.7, y = result_central * 1.05,
             label = "Centralized optimum", color = "red") +
    labs(x = "Iteration", y = "Objective Value",
         title = "ADMM Consensus Convergence") +
    theme_minimal()
```

The ADMM consensus solution converges to match the centralized
optimal solution.

## Comparison of Solutions

```{r}
cat("Max absolute difference between ADMM and centralized solution:\n")
cat(sprintf("  %.6e\n", max(abs(xbar - x_central_val))))
```

## Session Info

```{r}
sessionInfo()
```

## References

- Boyd, S., Parikh, N., Chu, E., Peleato, B. and Eckstein, J. (2011). "Distributed Optimization and Statistical Learning via the Alternating Direction Method of Multipliers." *Foundations and Trends in Machine Learning*, 3(1), 1--122.
- Boyd, S. and Vandenberghe, L. (2004). *Convex Optimization*. Cambridge University Press.
