---
title: "Fault Detection"
author: "Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: app_fault_detection
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
library(ggplot2)
```

## Introduction

We consider a problem of identifying faults that have occurred in a
system based on sensor measurements of system performance. This
example is adapted from the CVXPY documentation.

**Topic reference:** Samar, S., Gorinevsky, D. and Boyd, S.
"Likelihood Bounds for Constrained Estimation with Uncertainty."
*44th IEEE Conference on Decision and Control*, 2005.

## Problem Statement

Each of $n$ possible faults occurs independently with probability
$p$. The vector $x \in \{0, 1\}^n$ encodes the fault occurrences,
with $x_i = 1$ indicating that fault $i$ has occurred. System
performance is measured by $m$ sensors. The sensor output is

$$
y = Ax + v = \sum_{i=1}^n a_i x_i + v,
$$

where $A \in \mathbb{R}^{m \times n}$ is the sensing matrix with
column $a_i$ being the **fault signature** of fault $i$, and $v \in
\mathbb{R}^m$ is a noise vector where $v_j$ is Gaussian with mean 0
and variance $\sigma^2$.

The objective is to guess $x$ (which faults have occurred) given $y$
(sensor measurements).

We are interested in the setting where $n > m$, that is, when we have
more possible faults than measurements. In this setting, we can expect
a good recovery when the vector $x$ is sparse. This is the subject of
compressed sensing.

## Solution Approach

To identify the faults, one reasonable approach is to choose $x \in
\{0, 1\}^n$ to minimize the negative log-likelihood function

$$
\ell(x) = \frac{1}{2\sigma^2} \|Ax - y\|_2^2 + \log(1/p - 1) \mathbf{1}^T x + c.
$$

However, this problem is nonconvex and NP-hard, due to the constraint
that $x$ must be Boolean.

To make this problem tractable, we relax the Boolean constraints and
instead constrain $x_i \in [0, 1]$. The optimization problem

$$
\begin{array}{ll}
\mbox{minimize} & \|Ax - y\|_2^2 + 2\sigma^2 \log(1/p - 1) \mathbf{1}^T x \\
\mbox{subject to} & 0 \leq x_i \leq 1, \quad i = 1, \ldots, n
\end{array}
$$

is convex. We refer to the solution as the **relaxed ML** estimate.

By rounding the relaxed ML estimate to the nearest of 0 or 1, we
recover a Boolean estimate of the fault occurrences.

## Example

We generate an example with $n = 2000$ possible faults, $m = 200$
measurements, and fault probability $p = 0.01$. We choose $\sigma^2$
so that the signal-to-noise ratio is 5, i.e.,

$$
\sqrt{\frac{\mathbf{E}\|Ax\|_2^2}{\mathbf{E}\|v\|_2^2}} = 5.
$$

```{r}
set.seed(1)

n <- 2000
m <- 200
p <- 0.01
snr <- 5

sigma <- sqrt(p * n / (snr^2))
A <- matrix(rnorm(m * n), nrow = m, ncol = n)

x_true <- as.integer(runif(n) <= p)
v <- sigma * rnorm(m)

y <- A %*% x_true + v
```

Below, we show the true fault vector $x$, the signal $Ax$, and the
noise $v$.

```{r, fig.cap = "True fault vector x"}
df_x <- data.frame(index = 1:n, value = x_true)
ggplot(df_x, aes(x = index, y = value)) +
    geom_line() +
    labs(x = "Index", y = "x_true", title = "True Faults") +
    theme_minimal()
```

```{r, fig.cap = "Signal Ax and noise v"}
Ax <- as.vector(A %*% x_true)
df_signal <- data.frame(
    index = rep(1:m, 2),
    value = c(Ax, v),
    type = rep(c("Ax", "v"), each = m)
)
ggplot(df_signal, aes(x = index, y = value, color = type)) +
    geom_line() +
    labs(x = "Index", y = "Value", title = "Signal and Noise", color = "") +
    theme_minimal() +
    theme(legend.position = "bottom")
```

## Recovery

We solve the relaxed maximum likelihood problem with `CVXR` and then
round the result to get a Boolean solution.

```{r}
x_var <- Variable(n)
tau <- 2 * log(1/p - 1) * sigma^2
obj <- Minimize(sum_squares(A %*% x_var - y) + tau * sum(x_var))
constraints <- list(x_var >= 0, x_var <= 1)
prob <- Problem(obj, constraints)
result <- psolve(prob)

cat(sprintf("Problem status: %s\n", status(prob)))
cat(sprintf("Final objective value: %.4f\n", result))

## Relaxed ML estimate
x_rml <- as.numeric(value(x_var))

## Rounded solution
x_rnd <- as.integer(x_rml >= 0.5)
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(objective = result, x_rml = x_rml, x_rnd = x_rnd),
            file = file.path(testdata_dir, "fault_detection.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    saved <- readRDS(file = file.path(testdata_dir, "fault_detection.RDS"))
    expect_equal(result, saved$objective, tolerance = 1e-5)
    expect_equal(x_rml, saved$x_rml, tolerance = 1e-5)
}
```

## Evaluation

We compute the estimation errors and plot the true, relaxed ML, and
rounded solutions.

```{r}
errors <- function(x_true, x_est, threshold = 0.5) {
    k <- sum(x_true)
    false_pos <- sum(x_true < threshold & x_est >= threshold)
    false_neg <- sum(x_true >= threshold & x_est < threshold)
    list(true_faults = k, false_positives = false_pos, false_negatives = false_neg)
}

err <- errors(x_true, x_rml)
cat(sprintf("True faults: %d\n", err$true_faults))
cat(sprintf("False positives: %d\n", err$false_positives))
cat(sprintf("False negatives: %d\n", err$false_negatives))
```

```{r, fig.cap = "True, relaxed ML, and rounded fault estimates", fig.width = 10, fig.height = 4}
df_plot <- data.frame(
    index = rep(1:n, 3),
    value = c(x_true, x_rml, x_rnd),
    type = factor(rep(c("x_true", "x_rml (relaxed ML)", "x_rnd (rounded)"),
                      each = n),
                  levels = c("x_true", "x_rml (relaxed ML)", "x_rnd (rounded)"))
)

ggplot(df_plot, aes(x = index, y = value)) +
    geom_line() +
    facet_wrap(~ type, nrow = 1) +
    coord_cartesian(ylim = c(0, 1)) +
    labs(x = "Index", y = "Value") +
    theme_minimal()
```

The rounded solution gives perfect recovery with 0 false negatives
and 0 false positives.

## Session Info

```{r}
sessionInfo()
```

## References

- Samar, S., Gorinevsky, D. and Boyd, S. (2005). "Likelihood Bounds for Constrained Estimation with Uncertainty." *44th IEEE Conference on Decision and Control (CDC-ECC'05)*, IEEE.
