---
title: "Chebyshev Design of an FIR Filter"
author: "Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: app_fir_filter
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
library(ggplot2)
```

## Introduction

This example is adapted from the CVX example of the same name, by
Almir Mutapcic (2/2/2006) and the CVXPY adaptation by Judson Wilson
(5/27/2014).

This program designs an FIR filter, given a desired frequency response
$H_{\mathrm{des}}(\omega)$. The design is judged by the maximum
absolute error (Chebyshev norm). This is a convex problem (after
sampling it can be formulated as an SOCP), which may be written in the
form:

$$
\begin{array}{ll}
\mbox{minimize} & \max |H(\omega) - H_{\mathrm{des}}(\omega)| \quad \text{for } 0 \leq \omega \leq \pi,
\end{array}
$$

where the variable $H$ is the frequency response function,
corresponding to an impulse response $h$.

**Topic reference:** "Filter design" lecture notes (EE364) by S. Boyd.

## Problem Data

```{r}
## Number of FIR coefficients (including the zeroth one)
n <- 20

## Rule-of-thumb frequency discretization (Cheney's Approx. Theory book)
m <- 15 * n
w <- seq(0, pi, length.out = m)

## Construct the desired filter: fractional delay
D <- 8.25              # Delay value
Hdes <- exp(-1i * D * w)  # Desired frequency response
```

## Solve the Minimax (Chebyshev) Design Problem

The frequency response from a vector of filter coefficients $h$ is
computed as $H(\omega) = \sum_{k=0}^{n-1} h_k e^{-j k \omega}$,
which can be expressed in matrix form as $H = A h$ where
$A_{\omega,k} = e^{-j k \omega}$.

Since `CVXR` works with real-valued math, we split the problem into
real and imaginary parts. The objective becomes:

$$
\mbox{minimize} \quad \max_\omega \left[ (\mathrm{Re}(A) h - \mathrm{Re}(H_{\mathrm{des}}))^2 + (\mathrm{Im}(A) h - \mathrm{Im}(H_{\mathrm{des}}))^2 \right].
$$

```{r}
## A is the matrix used to compute the frequency response
## A[w,:] = [1, exp(-j*w), exp(-j*2*w), ..., exp(-j*(n-1)*w)]
A <- exp(-1i * outer(w, 0:(n-1)))

## Split into real and imaginary parts
Hdes_r <- Re(Hdes)
Hdes_i <- Im(Hdes)
A_R <- Re(A)
A_I <- Im(A)

## h is the (real) FIR coefficient vector
h <- Variable(n)

## The objective minimizes max(|A*h - Hdes|^2)
## = max( (Re(A)*h - Re(Hdes))^2 + (Im(A)*h - Im(Hdes))^2 )
obj <- Minimize(
    max_entries(
        (A_R %*% h - Hdes_r)^2 + (A_I %*% h - Hdes_i)^2
    )
)

## Solve problem
prob <- Problem(obj)
result <- psolve(prob)

cat(sprintf("Problem status: %s\n", status(prob)))
cat(sprintf("Final objective value: %f\n", result))
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(objective = result, h = value(h)),
            file = file.path(testdata_dir, "fir_filter.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    saved <- readRDS(file = file.path(testdata_dir, "fir_filter.RDS"))
    expect_equal(result, saved$objective, tolerance = 1e-5)
    expect_equal(value(h), saved$h, tolerance = 1e-5)
}
```

## Result Plots

We plot the FIR impulse response, and the frequency response magnitude
and phase.

```{r, fig.cap = "FIR filter impulse response"}
h_val <- value(h)
df_impulse <- data.frame(n = 0:(n-1), h = as.numeric(h_val))

ggplot(df_impulse, aes(x = n, y = h)) +
    geom_segment(aes(xend = n, yend = 0)) +
    geom_point() +
    labs(x = "n", y = "h(n)", title = "FIR Filter Impulse Response") +
    theme_minimal()
```

```{r, fig.cap = "FIR filter frequency response magnitude"}
## Compute the frequency response
H <- as.vector(A %*% h_val)

df_mag <- data.frame(
    omega = rep(w, 2),
    magnitude = c(20 * log10(Mod(H)), 20 * log10(Mod(Hdes))),
    type = rep(c("Optimized", "Desired"), each = m)
)

ggplot(df_mag, aes(x = omega, y = magnitude, color = type, linetype = type)) +
    geom_line() +
    scale_linetype_manual(values = c("Desired" = "dashed", "Optimized" = "solid")) +
    coord_cartesian(xlim = c(0, pi), ylim = c(-30, 10)) +
    labs(x = expression(omega), y = expression(paste("|H(", omega, ")| in dB")),
         title = "FIR Filter Frequency Response Magnitude",
         color = "", linetype = "") +
    theme_minimal() +
    theme(legend.position = "bottom")
```

```{r, fig.cap = "FIR filter frequency response phase"}
df_phase <- data.frame(omega = w, phase = Arg(H))

ggplot(df_phase, aes(x = omega, y = phase)) +
    geom_line() +
    coord_cartesian(xlim = c(0, pi), ylim = c(-pi, pi)) +
    labs(x = expression(omega), y = expression(paste(symbol("\xd0"), " H(", omega, ")")),
         title = "FIR Filter Frequency Response Phase") +
    theme_minimal()
```

## Session Info

```{r}
sessionInfo()
```

## References

- Boyd, S. "Filter design" lecture notes (EE364), Stanford University.
- Mutapcic, A. (2006). CVX example: Chebyshev design of an FIR filter.
