---
title: "Interdiction"
author: "Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: app_interdiction
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
library(ggplot2)
library(igraph)
```

## Introduction

This example is adapted from the CVXPY interdiction notebook. We
explore a game between a smuggler moving through a directed graph and
a security team trying to catch him.

**Topic references:**

- Boyd, S. and Vandenberghe, L. *Convex Optimization*. Cambridge
  University Press.
- Additional Exercise 15.4: Allocation of interdiction effort.
- Shortest path problem in graph theory.

## Problem Setup

The smuggler moves along a directed graph with $n$ nodes and $m$
edges from a source node to a destination node. Edge $j$ has evasion
probability $p_j$, which is the probability that the smuggler moves
along that edge undetected. The probability that the smuggler makes
it to the destination undetected is $\prod_{j \in \mathcal{P}} p_j$,
where $\mathcal{P}$ is the set of edges on the smuggler's path.

The security team can control the evasion probabilities through
allocation of resources. Their goal is to choose evasion
probabilities to minimize the chance that the smuggler gets through
undetected.

## Formulation

Taking the logarithm of all probabilities, let $c_j = \log(p_j)$.
The smuggler's path evasion probability in log-space is $\sum_{j \in
\mathcal{P}} c_j$.

We denote possible paths with a decision variable $x \in [0, 1]^m$,
where $x_j = 1$ denotes that edge $j$ is on the path. The smuggler's
problem (as an LP relaxation) is:

$$
\begin{array}{ll}
\mbox{maximize}   & c^T x \\
\mbox{subject to} & A_{\mathrm{inc}} x = b, \quad 0 \leq x \leq 1,
\end{array}
$$

where $A_{\mathrm{inc}}$ is the node-edge incidence matrix and $b$
encodes the source-to-sink flow constraint.

Using LP duality, the security team's problem becomes:

$$
\begin{array}{ll}
\mbox{minimize}   & \nu^T b \\
\mbox{subject to} & A_{\mathrm{inc}}^T \nu \geq c, \quad c \in \mathcal{C},
\end{array}
$$

where $\nu$ is the dual variable and $\mathcal{C}$ represents
constraints on the edge evasion probabilities.

## Helper Functions

We define helper functions for generating random geometric graphs and
for finding optimal paths using `igraph`.

```{r}
## Form an N x N grid of nodes, perturb positions, and connect nearby nodes
form_graph <- function(N, mu, eta, seed = NULL) {
    if (!is.null(seed)) set.seed(seed)
    mu <- mu / (N - 1)
    eta <- eta / (N - 1)

    ## Generate perturbed grid positions
    grid <- expand.grid(
        x = seq(0, 1, length.out = N),
        y = seq(1, 0, length.out = N)
    )
    pos <- data.frame(
        x = grid$x + mu * rnorm(nrow(grid)),
        y = grid$y + mu * rnorm(nrow(grid))
    )
    n <- nrow(pos)

    ## Connect nodes within distance eta (random geometric graph)
    edges <- c()
    for (i in 1:(n-1)) {
        for (j in (i+1):n) {
            d <- sqrt((pos$x[i] - pos$x[j])^2 + (pos$y[i] - pos$y[j])^2)
            if (d <= eta) {
                ## Add both directions for a directed graph
                edges <- c(edges, i, j, j, i)
            }
        }
    }

    G <- make_graph(edges, directed = TRUE, n = n)
    list(G = G, pos = pos, n = n)
}

## Generate random evasion probabilities for edges
## (same probability in both directions for symmetric pairs)
get_edge_probs <- function(G, seed = NULL) {
    if (!is.null(seed)) set.seed(seed)
    el <- as_edgelist(G)
    m <- nrow(el)
    a <- numeric(m)
    assigned <- list()

    for (k in 1:m) {
        e <- el[k, ]
        key <- paste(min(e[1], e[2]), max(e[1], e[2]), sep = "-")
        if (is.null(assigned[[key]])) {
            assigned[[key]] <- runif(1)
        }
        a[k] <- assigned[[key]]
    }
    a
}

## Find optimal smuggler path using shortest path (Dijkstra)
opt_path <- function(G, p) {
    p <- pmin(p, 1)
    w <- -log(p)
    E(G)$weight <- w
    sp <- shortest_paths(G, from = 1, to = vcount(G), output = "epath")
    path_edges <- sp$epath[[1]]

    x <- rep(0, ecount(G))
    x[as.integer(path_edges)] <- 1
    x
}
```

## Smuggler's Path Example

We create a $10 \times 10$ grid graph with small perturbations. The
evasion probabilities are distributed as $p_j = e^{-a_j}$ where
$a_j \sim \mathrm{Uniform}(0, 1)$.

```{r}
N <- 10
graph_data <- form_graph(N, mu = 0.12, eta = 1.2, seed = 5)
G <- graph_data$G
pos <- graph_data$pos
n <- graph_data$n

a <- get_edge_probs(G, seed = 2)
p <- exp(-a)
```

We solve the smuggler's relaxed LP to find the optimal path from node
1 (upper-left) to node $n$ (lower-right).

```{r}
## Build the incidence matrix
el <- as_edgelist(G)
m <- nrow(el)
A_inc <- matrix(0, nrow = n, ncol = m)
for (k in 1:m) {
    A_inc[el[k, 1], k] <- -1   # tail (outgoing)
    A_inc[el[k, 2], k] <-  1   # head (incoming)
}

b <- rep(0, n)
b[1] <- -1       # source
b[n] <- 1        # sink

c_log <- log(p)

## Solve smuggler's LP
x_var <- Variable(m)
constraints <- list(A_inc %*% x_var == b, x_var >= 0, x_var <= 1)
prob <- Problem(Maximize(t(c_log) %*% x_var), constraints)
result_smuggler <- psolve(prob)

x_smuggler <- as.numeric(value(x_var))

cat(sprintf("Problem status: %s\n", status(prob)))
evasion_prob_lp <- exp(sum(x_smuggler * c_log))
cat(sprintf("Evasion probability of smuggler's optimal path: %e (%.3f%%)\n",
            evasion_prob_lp, evasion_prob_lp * 100))
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(objective = result_smuggler,
                 evasion_prob = evasion_prob_lp),
            file = file.path(testdata_dir, "smuggler_lp.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    saved <- readRDS(file = file.path(testdata_dir, "smuggler_lp.RDS"))
    expect_equal(result_smuggler, saved$objective, tolerance = 1e-3)
}
```

We verify using Dijkstra's algorithm (via `igraph`) on the same
graph.

```{r}
x_dijkstra <- opt_path(G, p)
evasion_dijkstra <- exp(sum(x_dijkstra * c_log))
cat(sprintf("Evasion probability via Dijkstra: %e (%.3f%%)\n",
            evasion_dijkstra, evasion_dijkstra * 100))
```

The LP relaxation and the graph-theoretic shortest path give the same
optimal evasion probability.

## Visualization

We plot the graph with edges colored by evasion probability, and
highlight the smuggler's optimal path.

```{r, fig.cap = "Graph with evasion probabilities and smuggler's optimal path", fig.width = 8, fig.height = 6}
## Create a data frame for edges
edge_df <- data.frame(
    x1 = pos$x[el[, 1]],
    y1 = pos$y[el[, 1]],
    x2 = pos$x[el[, 2]],
    y2 = pos$y[el[, 2]],
    prob = p,
    on_path = x_dijkstra > 0.1
)

ggplot() +
    ## Draw all edges (faded)
    geom_segment(data = edge_df,
                 aes(x = x1, y = y1, xend = x2, yend = y2, color = prob),
                 linewidth = 0.5, alpha = 0.3) +
    ## Draw path edges (bold)
    geom_segment(data = edge_df[edge_df$on_path, ],
                 aes(x = x1, y = y1, xend = x2, yend = y2, color = prob),
                 linewidth = 2) +
    scale_color_gradient(low = "red", high = "green",
                         name = "Evasion\nProbability") +
    ## Draw nodes
    geom_point(data = pos, aes(x = x, y = y), color = "white",
               size = 2, shape = 21, fill = "white", stroke = 0.5) +
    ## Highlight source and destination
    geom_point(data = pos[1, ], aes(x = x, y = y),
               color = "green", size = 5, alpha = 0.4) +
    geom_point(data = pos[n, ], aes(x = x, y = y),
               color = "red", size = 5, alpha = 0.4) +
    coord_equal() +
    theme_void() +
    theme(legend.position = "right")
```

## Security's Objective

The security team's goal is to allocate resources to minimize the
smuggler's evasion probability. With evasion probabilities modeled as
$p_j = e^{-a_j r_j}$, where $r_j \geq 0$ is the effort allocated to
edge $j$, the security problem (using LP duality) is:

$$
\begin{array}{ll}
\mbox{minimize}   & \nu^T b \\
\mbox{subject to} & A_{\mathrm{inc}}^T \nu \geq -a \circ r \\
                   & \sum_j r_j = m, \quad 0 \leq r_j \leq R \\
                   & Br = 0,
\end{array}
$$

where $B$ enforces that symmetric edges have equal evasion
probabilities, and $R$ is the maximum per-edge budget.

```{r}
## Build B matrix for symmetry constraints
## For each pair of edges (i,j) and (j,i), enforce equal allocation
pairs <- list()
count <- 0
B_rows <- list()
for (k in 1:m) {
    i <- el[k, 1]; j <- el[k, 2]
    if (i < j) {
        ## Find the reverse edge
        rev_idx <- which(el[, 1] == j & el[, 2] == i)
        if (length(rev_idx) > 0) {
            count <- count + 1
            row <- rep(0, m)
            row[k] <- 1
            row[rev_idx[1]] <- -1
            B_rows[[count]] <- row
        }
    }
}

if (count > 0) {
    B <- do.call(rbind, B_rows)
} else {
    B <- matrix(0, nrow = 1, ncol = m)
}

## Solve security team's problem with R_max = 5
R_max <- 5
nu <- Variable(n)
r <- Variable(m)
constraints_sec <- list(
    t(A_inc) %*% nu >= -multiply(a, r),
    sum(r) == m,
    r >= 0,
    B %*% r == 0,
    r <= R_max
)
prob_sec <- Problem(Minimize(t(b) %*% nu), constraints_sec)
result_sec <- psolve(prob_sec)

nu_val <- as.numeric(value(nu))
r_val <- as.numeric(value(r))

evasion_security <- exp(sum(nu_val * b))
cat(sprintf("Evasion probability with optimal resource allocation: %e\n", evasion_security))
cat(sprintf("Improvement factor over uniform allocation: %.2f\n",
            evasion_prob_lp / evasion_security))
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(objective_sec = result_sec,
                 evasion_security = evasion_security),
            file = file.path(testdata_dir, "security.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    saved <- readRDS(file = file.path(testdata_dir, "security.RDS"))
    expect_equal(result_sec, saved$objective_sec, tolerance = 1e-2)
}
```

## Optimized Evasion Probabilities

```{r, fig.cap = "Edge evasion probabilities after optimal resource allocation", fig.width = 8, fig.height = 6}
p_opt <- exp(-a * r_val)

edge_df_opt <- data.frame(
    x1 = pos$x[el[, 1]],
    y1 = pos$y[el[, 1]],
    x2 = pos$x[el[, 2]],
    y2 = pos$y[el[, 2]],
    prob = p_opt
)

ggplot() +
    geom_segment(data = edge_df_opt,
                 aes(x = x1, y = y1, xend = x2, yend = y2, color = prob),
                 linewidth = 1) +
    scale_color_gradient(low = "red", high = "green",
                         name = "Evasion\nProbability") +
    geom_point(data = pos, aes(x = x, y = y), color = "white",
               size = 2, shape = 21, fill = "white", stroke = 0.5) +
    geom_point(data = pos[1, ], aes(x = x, y = y),
               color = "green", size = 5, alpha = 0.4) +
    geom_point(data = pos[n, ], aes(x = x, y = y),
               color = "red", size = 5, alpha = 0.4) +
    coord_equal() +
    theme_void() +
    theme(legend.position = "right")
```

## Smuggler's Optimal Path Under Optimized Security

We use Dijkstra's algorithm to find the smuggler's optimal path under
the optimized resource allocation.

```{r}
x_opt_path <- opt_path(G, p_opt)
c_opt <- log(p_opt)
evasion_opt_path <- exp(sum(x_opt_path * c_opt))
cat(sprintf("Evasion probability of smuggler's optimal path: %e\n", evasion_opt_path))
```

```{r, fig.cap = "Smuggler's optimal path under optimized security allocation", fig.width = 8, fig.height = 6}
edge_df_opt$on_path <- x_opt_path > 0.1

ggplot() +
    geom_segment(data = edge_df_opt,
                 aes(x = x1, y = y1, xend = x2, yend = y2, color = prob),
                 linewidth = 0.5, alpha = 0.3) +
    geom_segment(data = edge_df_opt[edge_df_opt$on_path, ],
                 aes(x = x1, y = y1, xend = x2, yend = y2, color = prob),
                 linewidth = 2) +
    scale_color_gradient(low = "red", high = "green",
                         name = "Evasion\nProbability") +
    geom_point(data = pos, aes(x = x, y = y), color = "white",
               size = 2, shape = 21, fill = "white", stroke = 0.5) +
    geom_point(data = pos[1, ], aes(x = x, y = y),
               color = "green", size = 5, alpha = 0.4) +
    geom_point(data = pos[n, ], aes(x = x, y = y),
               color = "red", size = 5, alpha = 0.4) +
    coord_equal() +
    theme_void() +
    theme(legend.position = "right")
```

The optimal resource allocation makes the smuggler's evasion
probability dramatically smaller than the uniform allocation.

## Session Info

```{r}
sessionInfo()
```

## References

- Boyd, S. and Vandenberghe, L. (2004). *Convex Optimization*. Cambridge University Press.
- Boyd, S. and Vandenberghe, L. Additional Exercise 15.4: Allocation of interdiction effort.
