---
title: "Entropy Maximization"
author: "CVXPY Developers and Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: app_max_entropy
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
source(here::here("_check_status.R"))
```

## Introduction

Adapted from the CVX example of the same name, by Joelle Skaf,
4/24/2008.

Consider the linear inequality constrained entropy maximization
problem:

$$
\begin{array}{ll}
    \mbox{maximize}   & -\sum_{i=1}^n x_i \log(x_i) \\
    \mbox{subject to} & \sum_{i=1}^n x_i = 1 \\
                      & Fx \preceq g,
\end{array}
$$

where the variable is $x \in \mathbf{R}^n$.

This problem can be formulated in CVXR using the `entr` atom, which
computes the elementwise entropy $-x_i \log(x_i)$.

## Generate Problem Data

```{r}
## Make random input repeatable
set.seed(0)

## Matrix size parameters
n <- 20
m <- 10
p <- 5

## Generate random problem data
tmp <- runif(n)
A_mat <- matrix(rnorm(m * n), nrow = m, ncol = n)
b_vec <- A_mat %*% tmp
F_mat <- matrix(rnorm(p * n), nrow = p, ncol = n)
g_vec <- F_mat %*% tmp + runif(p)
```

## Formulate and Solve

```{r, results = "hold"}
## Entropy maximization
x <- Variable(n)
obj <- Maximize(sum(entr(x)))
constraints <- list(
    A_mat %*% x == b_vec,
    F_mat %*% x <= g_vec
)
prob <- Problem(obj, constraints)
result <- psolve(prob, verbose = TRUE)
check_solver_status(prob)

## Print result
cat("\nThe optimal value is:", result, "\n")
cat("\nThe optimal solution is:\n")
print(round(value(x), 6))
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(value = result, x = as.vector(value(x))),
            file = file.path(testdata_dir, "max_entropy.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    me_results <- readRDS(file = file.path(testdata_dir, "max_entropy.RDS"))
    expect_equal(result, me_results$value, tolerance = 1e-5)
    expect_equal(as.vector(value(x)), me_results$x, tolerance = 1e-5)
}
```

## Session Info

```{r, echo = FALSE}
sessionInfo()
```

## References
