---
title: "Nonnegative Matrix Factorization"
author: "CVXPY Developers and Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: app_nonneg_matrix_fact
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
library(ggplot2)
```

## Introduction

Adapted from the CVX example of the same name, by Argyris Zymnis,
Joelle Skaf, and Stephen Boyd.

We are given a matrix $A \in \mathbf{R}^{m \times n}$ and are
interested in solving the problem:

$$
\begin{array}{ll}
    \mbox{minimize}   & \| A - YX \|_F \\
    \mbox{subject to} & Y \succeq 0 \\
                      & X \succeq 0,
\end{array}
$$

where $Y \in \mathbf{R}^{m \times k}$ and
$X \in \mathbf{R}^{k \times n}$.

This example generates a random matrix $A$ and obtains an
*approximate* solution to the above problem by first generating a
random initial guess for $Y$ and then alternatively minimizing over
$X$ and $Y$ for a fixed number of iterations.

## Generate Problem Data

```{r}
## Ensure repeatably random problem data
set.seed(0)

## Generate random data matrix A
m <- 10
n <- 10
k <- 5
A <- matrix(runif(m * k), nrow = m, ncol = k) %*%
     matrix(runif(k * n), nrow = k, ncol = n)

## Initialize Y randomly
Y_init <- matrix(runif(m * k), nrow = m, ncol = k)
```

## Perform Alternating Minimization

We alternate between optimizing over $X$ (with $Y$ fixed) and
optimizing over $Y$ (with $X$ fixed). In each sub-problem, the
nonnegative variable is optimized while the other factor is held
constant as a numeric matrix.

```{r}
## Ensure same initial random Y
Y_val <- Y_init

MAX_ITERS <- 30
residual <- numeric(MAX_ITERS)

for (iter_num in seq_len(MAX_ITERS)) {
    ## For odd iterations, treat Y constant, optimize over X
    if (iter_num %% 2 == 1) {
        X_var <- Variable(c(k, n))
        constraint <- list(X_var >= 0)
        obj <- Minimize(p_norm(vec(A - Y_val %*% X_var), 2))
        prob <- Problem(obj, constraint)
        result <- psolve(prob, solver = "SCS", max_iters = 10000L)
    } else {
        ## For even iterations, treat X constant, optimize over Y
        Y_var <- Variable(c(m, k))
        constraint <- list(Y_var >= 0)
        obj <- Minimize(p_norm(vec(A - Y_var %*% X_val), 2))
        prob <- Problem(obj, constraint)
        result <- psolve(prob, solver = "SCS", max_iters = 10000L)
    }

    if (!(status(prob) %in% c("optimal", "optimal_inaccurate"))) {
        stop(sprintf("Solver did not converge at iteration %d! Status: %s",
                     iter_num, status(prob)))
    }

    cat(sprintf("Iteration %d, residual norm %.6f\n", iter_num, result))
    residual[iter_num] <- result

    ## Convert variable to numeric for next iteration
    if (iter_num %% 2 == 1) {
        X_val <- value(X_var)
    } else {
        Y_val <- value(Y_var)
    }
}
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(residual = residual),
            file = file.path(testdata_dir, "nnmf.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    nnmf_results <- readRDS(file = file.path(testdata_dir, "nnmf.RDS"))
    expect_equal(residual, nnmf_results$residual, tolerance = 1e-3)
}
```

## Output Results

### Residual Plot

```{r, fig.cap = "Convergence of alternating minimization for NMF"}
df_resid <- data.frame(
    iteration = seq_len(MAX_ITERS),
    residual = residual
)

ggplot(df_resid, aes(x = iteration, y = residual)) +
    geom_line() +
    geom_point(size = 1) +
    labs(x = "Iteration Number", y = "Residual Norm",
         title = "Nonnegative Matrix Factorization Convergence") +
    theme_minimal()
```

### Factor Matrices

```{r, results = "hold"}
cat("Original matrix A:\n")
print(round(A, 4))

cat("\nLeft factor Y:\n")
print(round(Y_val, 4))

cat("\nRight factor X:\n")
print(round(X_val, 4))

cat("\nResidual A - Y %*% X:\n")
print(round(A - Y_val %*% X_val, 6))

cat(sprintf("\nFinal residual after %d iterations: %.6f\n",
            MAX_ITERS, residual[MAX_ITERS]))
```

## Session Info

```{r, echo = FALSE}
sessionInfo()
```

## References
