---
title: "Robust Kalman Filtering for Vehicle Tracking"
author: "CVXPY Developers and Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: app_robust_kalman
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
library(ggplot2)
source(here::here("_check_status.R"))
```

## Introduction

We will try to pinpoint the location of a moving vehicle with high
accuracy from noisy sensor data. We do this by modeling the vehicle
state as a discrete-time linear dynamical system. Standard **Kalman
filtering** can be used to approach this problem when the sensor noise
is assumed to be Gaussian. We use **robust Kalman filtering** to get
a more accurate estimate of the vehicle state for a non-Gaussian case
with outliers.

## Problem Statement

A discrete-time linear dynamical system consists of a sequence of state
vectors $x_t \in \mathbf{R}^n$, indexed by time
$t \in \{0, \ldots, N-1\}$ and dynamics equations

$$
\begin{align}
x_{t+1} &= Ax_t + Bw_t\\
y_t &= Cx_t + v_t,
\end{align}
$$

where $w_t \in \mathbf{R}^m$ is an input to the dynamical system
(say, a drive force on the vehicle), $y_t \in \mathbf{R}^r$ is a
state measurement, $v_t \in \mathbf{R}^r$ is noise, $A$ is
the drift matrix, $B$ is the input matrix, and $C$ is the
observation matrix.

Given $A$, $B$, $C$, and $y_t$ for $t = 0, \ldots, N-1$, the goal is
to estimate $x_t$ for $t = 0, \ldots, N-1$.

## Kalman Filtering

A Kalman filter estimates $x_t$ by solving the optimization problem

$$
\begin{array}{ll}
\mbox{minimize} & \sum_{t=0}^{N-1} \left(
\|w_t\|_2^2 + \tau \|v_t\|_2^2\right)\\
\mbox{subject to} & x_{t+1} = Ax_t + Bw_t,\quad t=0,\ldots, N-1\\
& y_t = Cx_t+v_t,\quad t = 0, \ldots, N-1,
\end{array}
$$

where $\tau$ is a tuning parameter. This problem is actually a least
squares problem, and can be solved via linear algebra, without the
need for more general convex optimization.

This model performs well when $w_t$ and $v_t$ are Gaussian. However,
the quadratic objective can be influenced by large outliers, which
degrades the accuracy of the recovery. To improve estimation in the
presence of outliers, we can use **robust Kalman filtering**.

## Robust Kalman Filtering

To handle outliers in $v_t$, robust Kalman filtering replaces the
quadratic cost with a Huber cost, which results in the convex
optimization problem

$$
\begin{array}{ll}
\mbox{minimize} & \sum_{t=0}^{N-1} \left( \|w_t\|^2_2 + \tau \phi_\rho(v_t) \right)\\
\mbox{subject to} & x_{t+1} = Ax_t + Bw_t,\quad t=0,\ldots, N-1\\
& y_t = Cx_t+v_t,\quad t=0,\ldots, N-1,
\end{array}
$$

where $\phi_\rho$ is the Huber function

$$
\phi_\rho(a) = \begin{cases} \|a\|_2^2 & \|a\|_2 \leq \rho\\
2\rho \|a\|_2 - \rho^2 & \|a\|_2 > \rho.
\end{cases}
$$

The Huber penalty function penalizes estimation error linearly outside
of a ball of radius $\rho$, whereas in standard Kalman filtering,
all errors are penalized quadratically. Thus, large errors are penalized
less harshly, making this model more robust to outliers.

## Vehicle Tracking Example

We apply standard and robust Kalman filtering to a vehicle tracking
problem with state $x_t \in \mathbf{R}^4$, where $(x_{t,1}, x_{t,2})$
is the position of the vehicle in two dimensions, and $(x_{t,3}, x_{t,4})$
is the vehicle velocity. The vehicle has unknown drive force $w_t$, and
we observe noisy measurements of the vehicle's position,
$y_t \in \mathbf{R}^2$.

The matrices for the dynamics are

$$
A = \begin{bmatrix}
1 & 0 & \left(1-\frac{\gamma}{2}\Delta t\right) \Delta t & 0 \\
0 & 1 & 0 & \left(1-\frac{\gamma}{2} \Delta t\right) \Delta t\\
0 & 0 & 1-\gamma \Delta t & 0 \\
0 & 0 & 0 & 1-\gamma \Delta t
\end{bmatrix},
$$

$$
B = \begin{bmatrix}
\frac{1}{2}\Delta t^2 & 0 \\
0 & \frac{1}{2}\Delta t^2 \\
\Delta t & 0 \\
0 & \Delta t \\
\end{bmatrix},
\qquad
C = \begin{bmatrix}
1 & 0 & 0 & 0 \\
0 & 1 & 0 & 0
\end{bmatrix},
$$

where $\gamma$ is a velocity damping parameter.

### Problem Data

We generate the data for the vehicle tracking problem. We have
$N = 1000$, $w_t$ a standard Gaussian, and $v_t$ a standard Gaussian,
except $20\%$ of the points will be outliers with $\sigma = 20$.

```{r}
N <- 1000  # number of timesteps
T_end <- 50  # time will vary from 0 to T_end
ts <- seq(0, T_end, length.out = N)
delt <- ts[2] - ts[1]
gamma_damp <- 0.05  # damping, 0 is no damping

A <- matrix(0, 4, 4)
B <- matrix(0, 4, 2)
C <- matrix(0, 2, 4)

A[1,1] <- 1
A[2,2] <- 1
A[1,3] <- (1 - gamma_damp * delt / 2) * delt
A[2,4] <- (1 - gamma_damp * delt / 2) * delt
A[3,3] <- 1 - gamma_damp * delt
A[4,4] <- 1 - gamma_damp * delt

B[1,1] <- delt^2 / 2
B[2,2] <- delt^2 / 2
B[3,1] <- delt
B[4,2] <- delt

C[1,1] <- 1
C[2,2] <- 1
```

### Simulation

We seed $x_0 = 0$ (starting at the origin with zero velocity) and
simulate the system forward in time. The results are the true vehicle
positions `x_true` and the observed positions `y`.

```{r}
sigma <- 20
p <- 0.20
set.seed(6)

x_sim <- matrix(0, nrow = 4, ncol = N + 1)
y <- matrix(0, nrow = 2, ncol = N)

## Generate random input and noise vectors
w_sim <- matrix(rnorm(2 * N), nrow = 2, ncol = N)
v_sim <- matrix(rnorm(2 * N), nrow = 2, ncol = N)

## Add outliers to v
set.seed(0)
inds <- runif(N) <= p
v_sim[, inds] <- sigma * matrix(rnorm(2 * N), nrow = 2, ncol = N)[, inds]

## Simulate the system forward in time
for (t in 1:N) {
    y[, t] <- C %*% x_sim[, t] + v_sim[, t]
    x_sim[, t + 1] <- A %*% x_sim[, t] + B %*% w_sim[, t]
}

x_true <- x_sim
w_true <- w_sim
```

```{r, fig.cap="True and observed vehicle positions", fig.width=10, fig.height=4}
library(gridExtra)

df_true <- data.frame(x = x_true[1, 1:N], y = x_true[2, 1:N])
df_obs <- data.frame(x = y[1, ], y = y[2, ])

xlims <- range(c(df_true$x, df_obs$x))
ylims <- range(c(df_true$y, df_obs$y))

p1 <- ggplot(df_true, aes(x = x, y = y)) +
    geom_point(color = "black", alpha = 0.4, size = 0.5) +
    coord_fixed(xlim = xlims, ylim = ylims) +
    labs(title = "True positions") + theme_minimal()

p2 <- ggplot(df_obs, aes(x = x, y = y)) +
    geom_point(color = "firebrick", alpha = 0.15, size = 0.5) +
    coord_fixed(xlim = xlims, ylim = ylims) +
    labs(title = "Observed positions") + theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```

## Kalman Filtering Recovery

The code below solves the standard Kalman filtering problem using
CVXR. Note that the recovery is distorted by outliers in the
measurements.

```{r}
x <- Variable(c(4, N + 1))
w <- Variable(c(2, N))
v <- Variable(c(2, N))

tau <- 0.08

obj_kf <- sum_squares(w) + tau * sum_squares(v)
obj_kf <- Minimize(obj_kf)

constr <- list()
for (t in 1:N) {
    constr <- c(constr,
                list(x[, t + 1] == A %*% x[, t] + B %*% w[, t]),
                list(y[, t] == C %*% x[, t] + v[, t]))
}

prob_kf <- Problem(obj_kf, constr)
result_kf <- psolve(prob_kf, solver = "OSQP", verbose = TRUE)
check_solver_status(prob_kf)
cat("Optimal objective value:", result_kf, "\n")
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(value = result_kf), file = file.path(testdata_dir, "kf.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    kf_results <- readRDS(file = file.path(testdata_dir, "kf.RDS"))
    expect_equal(result_kf, kf_results$value, tolerance = 1e-5)
}
```

```{r, fig.cap = "True and Kalman filter recovered positions", fig.width=10, fig.height=4}
x_kf <- value(x)
df_kf <- data.frame(x = x_kf[1, 1:N], y = x_kf[2, 1:N])

p1 <- ggplot(df_true, aes(x = x, y = y)) +
    geom_point(color = "black", alpha = 0.4, size = 0.5) +
    coord_fixed(xlim = xlims, ylim = ylims) +
    labs(title = "True positions") + theme_minimal()

p2 <- ggplot(df_kf, aes(x = x, y = y)) +
    geom_point(color = "steelblue", alpha = 0.4, size = 0.5) +
    coord_fixed(xlim = xlims, ylim = ylims) +
    labs(title = "Kalman filter recovery") + theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```

## Robust Kalman Filtering Recovery

Here we implement robust Kalman filtering with CVXR, replacing the
quadratic penalty on the noise with a Huber penalty. We get a better
recovery than the standard Kalman filtering, as shown in the plots
below.

```{r}
x <- Variable(c(4, N + 1))
w <- Variable(c(2, N))
v <- Variable(c(2, N))

tau <- 2
rho <- 2

obj_rkf <- sum_squares(w)
huber_cost <- 0
for (t in 1:N) {
    huber_cost <- huber_cost + tau * huber(p_norm(v[, t], 2), rho)
}
obj_rkf <- Minimize(obj_rkf + huber_cost)

constr <- list()
for (t in 1:N) {
    constr <- c(constr,
                list(x[, t + 1] == A %*% x[, t] + B %*% w[, t]),
                list(y[, t] == C %*% x[, t] + v[, t]))
}

prob_rkf <- Problem(obj_rkf, constr)
result_rkf <- psolve(prob_rkf, solver = "ECOS", verbose = TRUE)
check_solver_status(prob_rkf)
cat("Optimal objective value:", result_rkf, "\n")
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(value = result_rkf), file = file.path(testdata_dir, "rkf.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    rkf_results <- readRDS(file = file.path(testdata_dir, "rkf.RDS"))
    expect_equal(result_rkf, rkf_results$value, tolerance = 1e-5)
}
```

```{r, fig.cap = "True and robust Kalman filter recovered positions", fig.width=10, fig.height=4}
x_rkf <- value(x)
df_rkf <- data.frame(x = x_rkf[1, 1:N], y = x_rkf[2, 1:N])

p1 <- ggplot(df_true, aes(x = x, y = y)) +
    geom_point(color = "black", alpha = 0.4, size = 0.5) +
    coord_fixed(xlim = xlims, ylim = ylims) +
    labs(title = "True positions") + theme_minimal()

p2 <- ggplot(df_rkf, aes(x = x, y = y)) +
    geom_point(color = "darkgreen", alpha = 0.4, size = 0.5) +
    coord_fixed(xlim = xlims, ylim = ylims) +
    labs(title = "Robust Kalman filter recovery") + theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```

```{r, fig.cap = "Comparison of all four position estimates", fig.width=10, fig.height=8}
p1 <- ggplot(df_true, aes(x = x, y = y)) +
    geom_point(color = "black", alpha = 0.4, size = 0.5) +
    coord_fixed(xlim = xlims, ylim = ylims) +
    labs(title = "True") + theme_minimal()

p2 <- ggplot(df_obs, aes(x = x, y = y)) +
    geom_point(color = "firebrick", alpha = 0.15, size = 0.5) +
    coord_fixed(xlim = xlims, ylim = ylims) +
    labs(title = "Observed") + theme_minimal()

p3 <- ggplot(df_kf, aes(x = x, y = y)) +
    geom_point(color = "steelblue", alpha = 0.4, size = 0.5) +
    coord_fixed(xlim = xlims, ylim = ylims) +
    labs(title = "Kalman filter") + theme_minimal()

p4 <- ggplot(df_rkf, aes(x = x, y = y)) +
    geom_point(color = "darkgreen", alpha = 0.4, size = 0.5) +
    coord_fixed(xlim = xlims, ylim = ylims) +
    labs(title = "Robust Kalman filter") + theme_minimal()

grid.arrange(p1, p2, p3, p4, ncol = 2, nrow = 2)
```

## Session Info

```{r, echo = FALSE}
sessionInfo()
```

## References
