---
title: "Sparse Solution of Linear Inequalities"
author: "Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: app_sparse_solution
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
library(ggplot2)
```

## Introduction

This example is adapted from the CVX example of the same name, by
Almir Mutapcic (2/28/2006) and the CVXPY adaptation by Judson Wilson
(5/11/2014).

**Topic references:**

- Section 6.2, Boyd & Vandenberghe, *Convex Optimization*.
- Tropp, J. A. "Just relax: Convex programming methods for subset
  selection and sparse approximation."

We consider a set of linear inequalities $Ax \preceq b$ which are
feasible. We apply two heuristics to find a sparse point $x$ that
satisfies these inequalities.

## The $\ell_1$-norm Heuristic

The standard $\ell_1$-norm heuristic for finding a sparse solution is:

$$
\begin{array}{ll}
\mbox{minimize}   & \|x\|_1 \\
\mbox{subject to} & Ax \preceq b.
\end{array}
$$

## The Iterative Log Heuristic

The log-based heuristic is an iterative method for finding a sparse
solution, by finding a local optimal point for the problem:

$$
\begin{array}{ll}
\mbox{minimize}   & \sum_i \log(\delta + |x_i|) \\
\mbox{subject to} & Ax \preceq b,
\end{array}
$$

where $\delta$ is a small threshold value. Since this is a
minimization of a concave function, it is not convex. However, we can
apply a heuristic in which we linearize the objective, solve, and
re-iterate. This becomes a weighted $\ell_1$-norm heuristic:

$$
\begin{array}{ll}
\mbox{minimize}   & \sum_i W_i |x_i| \\
\mbox{subject to} & Ax \preceq b,
\end{array}
$$

which in each iteration re-adjusts the weights $W_i$ based on the
rule $W_i = 1 / (\delta + |x_i|)$.

This algorithm is described in:

- Rao, B. D. and Kreutz-Delgado, K. "An affine scaling methodology
  for best basis selection."
- Lobo, M. S., Fazel, M. and Boyd, S. "Portfolio optimization with
  linear and fixed transaction costs."

## Generate Problem Data

```{r}
## Fix random number generator so we can repeat the experiment
set.seed(1)

## The threshold value below which we consider an element to be zero
delta <- 1e-8

## Problem dimensions (m inequalities in n-dimensional space)
m <- 100
n <- 50

## Construct a feasible set of inequalities
## (This system is feasible for the x0 point.)
A <- matrix(rnorm(m * n), nrow = m, ncol = n)
x0 <- rnorm(n)
b <- as.vector(A %*% x0) + runif(m)
```

## $\ell_1$-norm Heuristic

```{r}
## Create variable
x_l1 <- Variable(n)

## Create constraint
constraints <- list(A %*% x_l1 <= b)

## Form objective
obj <- Minimize(p_norm(x_l1, 1))

## Form and solve problem
prob <- Problem(obj, constraints)
result_l1 <- psolve(prob)
cat(sprintf("Status: %s\n", status(prob)))

## Number of nonzero elements in the solution
x_l1_val <- as.numeric(value(x_l1))
nnz_l1 <- sum(abs(x_l1_val) > delta)
cat(sprintf("Found a feasible x in R^%d that has %d nonzeros.\n", n, nnz_l1))
cat(sprintf("Optimal objective value: %.6f\n", result_l1))
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(objective_l1 = result_l1, nnz_l1 = nnz_l1, x_l1 = x_l1_val),
            file = file.path(testdata_dir, "sparse_l1.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    saved <- readRDS(file = file.path(testdata_dir, "sparse_l1.RDS"))
    expect_equal(result_l1, saved$objective_l1, tolerance = 1e-5)
    expect_equal(nnz_l1, saved$nnz_l1)
}
```

## Iterative Log Heuristic

```{r}
## Do 15 iterations
NUM_RUNS <- 15
nnzs_log <- numeric(NUM_RUNS)

## Store W as a numeric vector for weights
W <- rep(1, n)
x_log <- Variable(n)

for (k in seq_len(NUM_RUNS)) {
    ## Form and solve the weighted l1 problem
    obj <- Minimize(t(W) %*% abs(x_log))
    constraints <- list(A %*% x_log <= b)
    prob <- Problem(obj, constraints)

    ## Solve
    result_log <- psolve(prob)

    prob_status <- status(prob)
    if (!prob_status %in% c("optimal", "optimal_inaccurate")) {
        stop(sprintf("Solver did not converge (status: %s)", prob_status))
    }

    ## Get current solution
    x_log_val <- as.numeric(value(x_log))

    ## Display new number of nonzeros in the solution vector
    nnz <- sum(abs(x_log_val) > delta)
    nnzs_log[k] <- nnz
    cat(sprintf("Iteration %d: Found a feasible x in R^%d with %d nonzeros...\n",
                k, n, nnz))

    ## Adjust the weights elementwise and re-iterate
    W <- 1 / (delta + abs(x_log_val))
}
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(nnzs_log = nnzs_log, x_log = x_log_val),
            file = file.path(testdata_dir, "sparse_log.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    saved <- readRDS(file = file.path(testdata_dir, "sparse_log.RDS"))
    expect_equal(nnzs_log, saved$nnzs_log, tolerance = 1e-5)
}
```

## Result Plot

The following plot shows the result of the $\ell_1$-norm heuristic, as
well as the result for each iteration of the log heuristic.

```{r, fig.cap = "Number of non-zeros vs. iteration for both heuristics"}
df_plot <- data.frame(
    iteration = 1:NUM_RUNS,
    log_heuristic = nnzs_log
)

ggplot(df_plot, aes(x = iteration, y = log_heuristic)) +
    geom_line(aes(color = "Log heuristic")) +
    geom_point(aes(color = "Log heuristic")) +
    geom_hline(aes(yintercept = nnz_l1, color = "l1-norm heuristic"),
               linetype = "dashed") +
    coord_cartesian(xlim = c(1, NUM_RUNS), ylim = c(0, n)) +
    labs(x = "Iteration", y = "Number of non-zeros (cardinality)",
         color = "") +
    theme_minimal() +
    theme(legend.position = "bottom")
```

The iterative log heuristic converges to a sparser solution than the
$\ell_1$-norm heuristic.

## Session Info

```{r}
sessionInfo()
```

## References

- Boyd, S. and Vandenberghe, L. (2004). *Convex Optimization*. Cambridge University Press. Section 6.2.
- Tropp, J. A. (2006). "Just relax: Convex programming methods for subset selection and sparse approximation." *IEEE Transactions on Information Theory*, 52(3), 1030--1051.
- Rao, B. D. and Kreutz-Delgado, K. (1999). "An affine scaling methodology for best basis selection." *IEEE Transactions on Signal Processing*, 47(1), 187--200.
