---
title: "Sparse Covariance Estimation"
author: "Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: app_sparse_covariance
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
library(ggplot2)
library(Matrix)
library(expm)
```

## Introduction

This example is adapted from the CVX example of the same name, by
Joelle Skaf (4/24/2008) and the CVXPY adaptation by Judson Wilson
(5/22/2014).

**Topic reference:** Section 7.1.1, Boyd & Vandenberghe, *Convex
Optimization*.

Suppose $y \in \mathbb{R}^n$ is a Gaussian random variable with zero
mean and covariance matrix $R = \mathbf{E}[yy^T]$, with sparse
inverse $S = R^{-1}$ ($S_{ij} = 0$ means that $y_i$ and $y_j$ are
conditionally independent). We want to estimate the covariance matrix
$R$ based on $N$ independent samples $y_1, \ldots, y_N$ drawn from
the distribution, using prior knowledge that $S$ is sparse.

A good heuristic for estimating $R$ is to solve the problem

$$
\begin{array}{ll}
\mbox{maximize}   & \log \det(S) - \mathrm{tr}(SY) \\
\mbox{subject to} & \sum_{i=1}^n \sum_{j=1}^n |S_{ij}| \leq \alpha \\
                   & S \succeq 0,
\end{array}
$$

where $Y$ is the sample covariance of $y_1, \ldots, y_N$, and
$\alpha$ is a sparsity parameter to be chosen or tuned.

## Generate Problem Data

```{r}
set.seed(0)

## Dimension of matrix
n <- 10

## Number of samples
N <- 1000

## Create sparse, symmetric PSD matrix S
A <- matrix(rnorm(n * n), nrow = n, ncol = n)

## Sparsen A: zero out ~85% of entries
mask <- matrix(rbinom(n * n, 1, prob = 0.15), nrow = n, ncol = n)
A <- A * mask

Strue <- A %*% t(A) + 0.05 * diag(n)   # Force strict positive definiteness

## Create the covariance matrix associated with S
R <- solve(Strue)

## Create samples y_i from the distribution with covariance R
y_sample <- expm::sqrtm(R) %*% matrix(rnorm(n * N), nrow = n, ncol = N)

## Calculate the sample covariance matrix
Y <- cov(t(y_sample))
```

## Solve for Several $\alpha$ Values

```{r}
alphas <- c(10, 2, 1)
S_results <- list()

for (i in seq_along(alphas)) {
    alpha <- alphas[i]

    ## Create a variable constrained to the positive semidefinite cone
    S <- Variable(c(n, n), PSD = TRUE)

    ## Form the log_det(S) - tr(S %*% Y) objective
    obj <- Maximize(log_det(S) - matrix_trace(S %*% Y))

    ## Set constraint
    constraints <- list(sum(abs(S)) <= alpha)

    ## Form and solve optimization problem
    prob <- Problem(obj, constraints)
    result <- psolve(prob, solver = "SCS")

    if (status(prob) != "optimal" && status(prob) != "optimal_inaccurate") {
        warning(sprintf("Problem with alpha = %g did not solve optimally: %s",
                        alpha, status(prob)))
    }

    ## Get the estimated inverse covariance
    S_val <- value(S)

    ## Threshold to enforce exact zeros
    S_val[abs(S_val) <= 1e-4] <- 0

    S_results[[i]] <- S_val

    cat(sprintf("Completed optimization for alpha = %g, obj value = %.6f\n",
                alpha, result))
}
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(S_results = S_results),
            file = file.path(testdata_dir, "sparse_covariance.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    saved <- readRDS(file = file.path(testdata_dir, "sparse_covariance.RDS"))
    for (i in seq_along(S_results)) {
        expect_equal(S_results[[i]], saved$S_results[[i]], tolerance = 1e-3)
    }
}
```

## Result Plots

We plot the sparsity pattern for the true inverse covariance matrix
and the estimated inverse covariance matrices for each value of
$\alpha$.

```{r, fig.cap = "Sparsity patterns of true and estimated inverse covariance matrices", fig.width = 10, fig.height = 6}
## Helper function to create a sparsity pattern data frame
spy_df <- function(mat, title) {
    n <- nrow(mat)
    df <- expand.grid(j = 1:n, i = 1:n)
    df$nonzero <- as.factor(as.numeric(mat[cbind(df$i, df$j)] != 0))
    df$title <- title
    df
}

df_all <- rbind(
    spy_df(as.matrix(Strue), "True Inverse Cov."),
    spy_df(S_results[[1]], sprintf("alpha = %g", alphas[1])),
    spy_df(S_results[[2]], sprintf("alpha = %g", alphas[2])),
    spy_df(S_results[[3]], sprintf("alpha = %g", alphas[3]))
)
df_all$title <- factor(df_all$title,
                        levels = c("True Inverse Cov.",
                                   sprintf("alpha = %g", alphas)))

ggplot(df_all, aes(x = j, y = i, fill = nonzero)) +
    geom_tile(color = "gray70") +
    scale_fill_manual(values = c("0" = "white", "1" = "steelblue")) +
    scale_y_reverse() +
    facet_wrap(~ title, nrow = 1) +
    coord_equal() +
    labs(x = "", y = "") +
    theme_minimal() +
    theme(legend.position = "none",
          axis.text = element_blank(),
          panel.grid = element_blank())
```

As $\alpha$ decreases, the estimated inverse covariance matrix becomes
sparser. At $\alpha = 10$, the estimate is relatively dense. At
$\alpha = 1$, most off-diagonal entries are zero.

## Session Info

```{r}
sessionInfo()
```

## References

- Boyd, S. and Vandenberghe, L. (2004). *Convex Optimization*. Cambridge University Press. Section 7.1.1.
