---
title: "Total Variation In-Painting"
author: "CVXPY Developers and Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: app_tv_inpainting
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
library(png)
library(ggplot2)
library(grid)
library(gridExtra)
source(here::here("_check_status.R"))
```

## Introduction

A grayscale image is represented as an $m \times n$ matrix of
intensities $U^{\mathrm{orig}}$ (typically between the values $0$ and
$255$). We are given the values $U^{\mathrm{orig}}_{ij}$, for
$(i,j) \in \mathcal{K}$, where
$\mathcal{K} \subset \{1,\ldots,m\} \times \{1,\ldots,n\}$ is the
set of indices corresponding to known pixel values. Our job is to
*in-paint* the image by guessing the missing pixel values, i.e., those
with indices not in $\mathcal{K}$. The reconstructed image will be
represented by $U \in \mathbf{R}^{m \times n}$, where $U$ matches the
known pixels, i.e., $U_{ij} = U^{\mathrm{orig}}_{ij}$ for
$(i,j) \in \mathcal{K}$.

The reconstruction $U$ is found by minimizing the total variation of
$U$, subject to matching the known pixel values. We use the
$\ell_2$ total variation, defined as

$$
\mathop{\bf tv}(U) =
\sum_{i=1}^{m-1} \sum_{j=1}^{n-1}
\left\| \begin{pmatrix}
U_{i+1,j} - U_{ij} \\ U_{i,j+1} - U_{ij}
\end{pmatrix} \right\|_2.
$$

Note that the norm of the discretized gradient is *not* squared.

## Grayscale In-Painting

We load a $128 \times 128$ grayscale image and a corrupted version
where text has been overlaid on the original. The corrupted image has
white text obscuring parts of the image. We construct the *known*
matrix by comparing the two images: a pixel is known if it was not
affected by the text overlay.

```{r, results = "hold"}
## Load the images
u_orig <- readPNG(here::here("images", "loki128.png"))
u_corr <- readPNG(here::here("images", "loki128_corrupted.png"))

## If images are RGB, convert to grayscale by averaging channels
if (length(dim(u_orig)) == 3) {
    u_orig <- (u_orig[, , 1] + u_orig[, , 2] + u_orig[, , 3]) / 3
}
if (length(dim(u_corr)) == 3) {
    u_corr <- (u_corr[, , 1] + u_corr[, , 2] + u_corr[, , 3]) / 3
}

rows <- nrow(u_orig)
cols <- ncol(u_orig)

## known is 1 if the pixel is known, 0 if corrupted
known <- 1 * (u_orig == u_corr)
cat("Image size:", rows, "x", cols, "\n")
cat("Known pixels:", sum(known), "out of", rows * cols,
    sprintf("(%.0f%%)\n", 100 * sum(known) / (rows * cols)))
```

```{r}
#| fig-cap: "Original (left) and corrupted (right) grayscale images"
#| fig-width: 8
#| fig-height: 4
p1 <- ggplot() +
    annotation_raster(matrix(gray(u_orig), rows, cols),
                      xmin = 0, xmax = 1, ymin = 0, ymax = 1,
                      interpolate = FALSE) +
    coord_fixed() + theme_void() + ggtitle("Original Image")

p2 <- ggplot() +
    annotation_raster(matrix(gray(u_corr), rows, cols),
                      xmin = 0, xmax = 1, ymin = 0, ymax = 1,
                      interpolate = FALSE) +
    coord_fixed() + theme_void() + ggtitle("Corrupted Image")

grid.arrange(p1, p2, ncol = 2)
```

Now we solve the total variation in-painting problem using CVXR. We
use the SCS solver, which scales well to larger problems.

```{r}
U <- Variable(c(rows, cols))
obj <- Minimize(tv(U))
constraints <- list(multiply(known, U) == multiply(known, u_corr))
prob <- Problem(obj, constraints)

result <- psolve(prob, solver = "SCS", verbose = TRUE)
check_solver_status(prob)
cat("Optimal objective value:", result, "\n")
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(value = result), file = file.path(testdata_dir, "tv_gray.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    tv_results <- readRDS(file = file.path(testdata_dir, "tv_gray.RDS"))
    expect_equal(result, tv_results$value, tolerance = 1e-5)
}
```

After solving the problem, the in-painted image is stored in the value
of `U`. We display the in-painted image and the intensity difference
between the original and in-painted images (magnified by a factor of
10 for visibility).

```{r}
#| fig-cap: "In-painted (left) and difference (right) grayscale images"
#| fig-width: 8
#| fig-height: 4
U_val <- pmin(pmax(value(U), 0), 1)

p3 <- ggplot() +
    annotation_raster(matrix(gray(U_val), rows, cols),
                      xmin = 0, xmax = 1, ymin = 0, ymax = 1,
                      interpolate = FALSE) +
    coord_fixed() + theme_void() + ggtitle("In-Painted Image")

img_diff <- pmin(10 * abs(u_orig - U_val), 1)
p4 <- ggplot() +
    annotation_raster(matrix(gray(img_diff), rows, cols),
                      xmin = 0, xmax = 1, ymin = 0, ymax = 1,
                      interpolate = FALSE) +
    coord_fixed() + theme_void() + ggtitle("Difference Image (10x)")

grid.arrange(p3, p4, ncol = 2)
```

## Color In-Painting

For color images, the in-painting problem is similar to the grayscale
case. A color image is represented as an $m \times n \times 3$ array
of RGB values $U^{\mathrm{orig}}$ (typically between the values $0$
and $1$). We solve the problem by creating three CVXR variables (one
per color channel) and minimizing the total variation jointly across
all channels.

We load a $128 \times 128$ color image and randomly discard $70\%$ of
the pixels.

```{r}
set.seed(1)

## Load color image
u_orig_color <- readPNG(here::here("images", "loki128color.png"))
## Drop alpha channel if present
if (dim(u_orig_color)[3] == 4) {
    u_orig_color <- u_orig_color[, , 1:3]
}
rows_c <- dim(u_orig_color)[1]
cols_c <- dim(u_orig_color)[2]
colors <- dim(u_orig_color)[3]

## known is 1 if the pixel is known, 0 if corrupted
## Randomly keep 30% of pixels
pixel_mask <- matrix(as.numeric(runif(rows_c * cols_c) > 0.7), rows_c, cols_c)
known_color <- array(rep(pixel_mask, colors), dim = c(rows_c, cols_c, colors))

u_corr_color <- known_color * u_orig_color
cat("Color image size:", rows_c, "x", cols_c, "x", colors, "\n")
```

```{r}
#| fig-cap: "Original (left) and corrupted (right) color images"
#| fig-width: 8
#| fig-height: 4
## Helper to convert array to raster string matrix for ggplot
arr_to_raster <- function(arr) {
    arr <- pmin(pmax(arr, 0), 1)
    nr <- dim(arr)[1]; nc <- dim(arr)[2]
    matrix(rgb(arr[, , 1], arr[, , 2], arr[, , 3]), nr, nc)
}

p5 <- ggplot() +
    annotation_raster(arr_to_raster(u_orig_color),
                      xmin = 0, xmax = 1, ymin = 0, ymax = 1,
                      interpolate = FALSE) +
    coord_fixed() + theme_void() + ggtitle("Original Image")

p6 <- ggplot() +
    annotation_raster(arr_to_raster(u_corr_color),
                      xmin = 0, xmax = 1, ymin = 0, ymax = 1,
                      interpolate = FALSE) +
    coord_fixed() + theme_void() + ggtitle("Corrupted Image")

grid.arrange(p5, p6, ncol = 2)
```

```{r}
## Solve the TV in-painting problem for each channel
channels <- lapply(seq_len(colors), function(k) {
    U_k <- Variable(c(rows_c, cols_c))
    constr <- multiply(known_color[, , k], U_k) ==
              multiply(known_color[, , k], u_corr_color[, , k])
    list(var = U_k, constr = constr)
})
variables <- lapply(channels, `[[`, "var")
constraints_c <- lapply(channels, `[[`, "constr")

prob_color <- Problem(Minimize(tv(variables[[1]], variables[[2]], variables[[3]])),
                      constraints_c)
result_color <- psolve(prob_color, solver = "SCS", verbose = TRUE)
check_solver_status(prob_color)
cat("Optimal objective value:", result_color, "\n")
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(value = result_color), file = file.path(testdata_dir, "tv_color.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    tv_color_results <- readRDS(file = file.path(testdata_dir, "tv_color.RDS"))
    expect_equal(result_color, tv_color_results$value, tolerance = 1e-5)
}
```

After solving, we reconstruct the color image and compare with the
original.

```{r}
#| fig-cap: "Original, corrupted, in-painted, and difference color images"
#| fig-width: 10
#| fig-height: 5
rec_arr <- array(
    unlist(lapply(variables, function(v) pmin(pmax(value(v), 0), 1))),
    dim = c(rows_c, cols_c, colors)
)

p7 <- ggplot() +
    annotation_raster(arr_to_raster(u_orig_color),
                      xmin = 0, xmax = 1, ymin = 0, ymax = 1,
                      interpolate = FALSE) +
    coord_fixed() + theme_void() + ggtitle("Original")

p8 <- ggplot() +
    annotation_raster(arr_to_raster(u_corr_color),
                      xmin = 0, xmax = 1, ymin = 0, ymax = 1,
                      interpolate = FALSE) +
    coord_fixed() + theme_void() + ggtitle("Corrupted")

p9 <- ggplot() +
    annotation_raster(arr_to_raster(rec_arr),
                      xmin = 0, xmax = 1, ymin = 0, ymax = 1,
                      interpolate = FALSE) +
    coord_fixed() + theme_void() + ggtitle("In-Painted")

diff_arr <- pmin(10 * abs(u_orig_color - rec_arr), 1)
p10 <- ggplot() +
    annotation_raster(arr_to_raster(diff_arr),
                      xmin = 0, xmax = 1, ymin = 0, ymax = 1,
                      interpolate = FALSE) +
    coord_fixed() + theme_void() + ggtitle("Difference (10x)")

grid.arrange(p7, p8, p9, p10, ncol = 2)
```

## Session Info

```{r, echo = FALSE}
sessionInfo()
```

## References
