---
title: "Total Variation In-Painting"
author: "Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: app_tv_inpainting
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
library(png)
```

## Introduction

A grayscale image is represented as an $m \times n$ matrix of
intensities $U^{\mathrm{orig}}$ (typically between the values $0$ and
$255$). We are given the values $U^{\mathrm{orig}}_{ij}$, for
$(i,j) \in \mathcal{K}$, where
$\mathcal{K} \subset \{1,\ldots,m\} \times \{1,\ldots,n\}$ is the
set of indices corresponding to known pixel values. Our job is to
*in-paint* the image by guessing the missing pixel values, i.e., those
with indices not in $\mathcal{K}$. The reconstructed image will be
represented by $U \in \mathbf{R}^{m \times n}$, where $U$ matches the
known pixels, i.e., $U_{ij} = U^{\mathrm{orig}}_{ij}$ for
$(i,j) \in \mathcal{K}$.

The reconstruction $U$ is found by minimizing the total variation of
$U$, subject to matching the known pixel values. We use the
$\ell_2$ total variation, defined as

$$
\mathop{\bf tv}(U) =
\sum_{i=1}^{m-1} \sum_{j=1}^{n-1}
\left\| \begin{pmatrix}
U_{i+1,j} - U_{ij} \\ U_{i,j+1} - U_{ij}
\end{pmatrix} \right\|_2.
$$

Note that the norm of the discretized gradient is *not* squared.

## Grayscale In-Painting

We demonstrate total variation in-painting on a small grayscale image.
Since loading specific image files may not be available, we create a
synthetic $32 \times 32$ test image and corrupt it by removing $50\%$
of the pixels randomly.

```{r}
set.seed(123)

## Create a synthetic grayscale test image (32 x 32)
rows <- 32
cols <- 32
u_orig <- matrix(0, rows, cols)

## Create a pattern: horizontal and vertical bars with gradients
for (i in 1:rows) {
    for (j in 1:cols) {
        u_orig[i, j] <- 0.5 * sin(2 * pi * i / rows) + 0.5 * cos(2 * pi * j / cols)
    }
}
## Rescale to [0, 1]
u_orig <- (u_orig - min(u_orig)) / (max(u_orig) - min(u_orig))

## Create known matrix: keep 50% of pixels
known <- matrix(rbinom(rows * cols, 1, 0.5), rows, cols)

## Corrupted image (unknown pixels set to 0)
u_corr <- known * u_orig
```

```{r, fig.cap = "Original (left) and corrupted (right) images"}
par(mfrow = c(1, 2))
image(t(u_orig[nrow(u_orig):1, ]), col = grey.colors(256),
      main = "Original Image", axes = FALSE)
image(t(u_corr[nrow(u_corr):1, ]), col = grey.colors(256),
      main = "Corrupted Image", axes = FALSE)
```

Now we solve the total variation in-painting problem using CVXR. We
use the SCS solver, which scales well to larger problems.

```{r}
U <- Variable(c(rows, cols))
obj <- Minimize(tv(U))
constraints <- list(multiply(known, U) == multiply(known, u_corr))
prob <- Problem(obj, constraints)

result <- psolve(prob, solver = "SCS", verbose = TRUE)
cat("Optimal objective value:", result, "\n")
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(value = result), file = file.path(testdata_dir, "tv_gray.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    tv_results <- readRDS(file = file.path(testdata_dir, "tv_gray.RDS"))
    expect_equal(result, tv_results$value, tolerance = 1e-5)
}
```

After solving the problem, the in-painted image is stored in the value
of `U`. We display the in-painted image and the intensity difference
between the original and in-painted images (magnified by a factor of
10 for visibility).

```{r, fig.cap = "In-painted (left) and difference (right) images"}
U_val <- value(U)
par(mfrow = c(1, 2))
image(t(U_val[nrow(U_val):1, ]), col = grey.colors(256),
      main = "In-Painted Image", axes = FALSE)
img_diff <- 10 * abs(u_orig - U_val)
image(t(img_diff[nrow(img_diff):1, ]), col = grey.colors(256),
      main = "Difference Image (10x)", axes = FALSE)
```

## Color In-Painting

For color images, the in-painting problem is similar to the grayscale
case. A color image is represented as an $m \times n \times 3$ array
of RGB values $U^{\mathrm{orig}}$ (typically between the values $0$
and $1$). We solve the problem by creating three CVXR variables (one
per color channel) and minimizing the total variation jointly across
all channels.

We create a synthetic $32 \times 32 \times 3$ color test image and
randomly discard $70\%$ of the pixels.

```{r}
set.seed(1)

## Create a synthetic color image
rows_c <- 32
cols_c <- 32
colors <- 3

u_orig_color <- array(0, dim = c(rows_c, cols_c, colors))
for (i in 1:rows_c) {
    for (j in 1:cols_c) {
        u_orig_color[i, j, 1] <- 0.5 + 0.5 * sin(2 * pi * i / rows_c)
        u_orig_color[i, j, 2] <- 0.5 + 0.5 * cos(2 * pi * j / cols_c)
        u_orig_color[i, j, 3] <- 0.5 + 0.25 * sin(2 * pi * (i + j) / rows_c)
    }
}
## Rescale each channel to [0, 1]
for (k in 1:colors) {
    ch <- u_orig_color[, , k]
    u_orig_color[, , k] <- (ch - min(ch)) / (max(ch) - min(ch))
}

## Create known matrix: keep 30% of pixels
known_color <- array(0, dim = c(rows_c, cols_c, colors))
for (i in 1:rows_c) {
    for (j in 1:cols_c) {
        if (runif(1) > 0.7) {
            known_color[i, j, ] <- 1
        }
    }
}

u_corr_color <- known_color * u_orig_color
```

```{r}
## Solve the TV in-painting problem for each channel
variables <- list()
constraints_c <- list()
for (k in 1:colors) {
    U_k <- Variable(c(rows_c, cols_c))
    variables[[k]] <- U_k
    constraints_c <- c(constraints_c,
        list(multiply(known_color[, , k], U_k) ==
             multiply(known_color[, , k], u_corr_color[, , k])))
}

prob_color <- Problem(Minimize(tv(variables[[1]], variables[[2]], variables[[3]])),
                      constraints_c)
result_color <- psolve(prob_color, solver = "SCS", verbose = TRUE)
cat("Optimal objective value:", result_color, "\n")
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(value = result_color), file = file.path(testdata_dir, "tv_color.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    tv_color_results <- readRDS(file = file.path(testdata_dir, "tv_color.RDS"))
    expect_equal(result_color, tv_color_results$value, tolerance = 1e-5)
}
```

After solving, we reconstruct the color image and compare with the
original.

```{r, fig.cap = "Original (left), corrupted (center), and in-painted (right) color images"}
rec_arr <- array(0, dim = c(rows_c, cols_c, colors))
for (k in 1:colors) {
    rec_arr[, , k] <- pmin(pmax(value(variables[[k]]), 0), 1)
}

par(mfrow = c(1, 3))
## R's image() doesn't directly handle RGB, so we use rasterImage
plot(c(0, 1), c(0, 1), type = "n", axes = FALSE, xlab = "", ylab = "",
     main = "Original Image")
rasterImage(u_orig_color, 0, 0, 1, 1, interpolate = FALSE)

plot(c(0, 1), c(0, 1), type = "n", axes = FALSE, xlab = "", ylab = "",
     main = "Corrupted Image")
rasterImage(u_corr_color, 0, 0, 1, 1, interpolate = FALSE)

plot(c(0, 1), c(0, 1), type = "n", axes = FALSE, xlab = "", ylab = "",
     main = "In-Painted Image")
rasterImage(rec_arr, 0, 0, 1, 1, interpolate = FALSE)
```

## Session Info

```{r}
sessionInfo()
```

## References
