---
title: "Water Filling in Communications"
author: "Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: app_water_filling
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
library(ggplot2)
```

## Introduction

From Boyd and Vandenberghe, *Convex Optimization*, Example 5.2 (page
145).

Convex optimization can be used to solve the classic water filling
problem. This problem is where a total amount of power $P$ has to be
assigned to $n$ communication channels, with the objective of
maximizing the total communication rate. The communication rate of the
$i$th channel is given by

$$
\log(\alpha_i + x_i)
$$

where $x_i$ represents the power allocated to channel $i$ and
$\alpha_i$ represents the floor above the baseline at which power can
be added to the channel. Since $-\log(X)$ is convex, we can write the
water-filling problem as a convex optimization problem:

$$
\begin{array}{ll}
\mbox{minimize} & \sum_{i=1}^{N} -\log(\alpha_i + x_i) \\
\mbox{subject to} & x_i \geq 0, \quad \sum_{i=1}^{N} x_i = P
\end{array}
$$

This form is straightforward to express in DCP format and thus can be
simply solved using CVXR.

## Water Filling Function

We define a function that solves the water-filling problem for given
channel parameters and total power budget.

```{r}
water_filling <- function(n, a, sum_x = 1) {
    ## Boyd and Vandenberghe, Convex Optimization, example 5.2 page 145
    ## Water-filling.
    ##
    ## This problem arises in information theory, in allocating power to a
    ## set of n communication channels in order to maximize the total
    ## channel capacity. The variable x_i represents the transmitter power
    ## allocated to the ith channel, and log(alpha_i + x_i) gives the
    ## capacity or maximum communication rate of the channel. The objective
    ## is to maximize sum(log(alpha_i + x_i)) subject to sum(x_i) = sum_x.

    ## Declare variables and parameters
    x <- Variable(n)
    alpha <- a

    ## Objective: maximize the total communication rate
    obj <- Maximize(sum(log(alpha + x)))

    ## Constraints
    constraints <- list(x >= 0, sum(x) == sum_x)

    ## Solve
    prob <- Problem(obj, constraints)
    result <- psolve(prob)

    prob_status <- status(prob)
    if (prob_status == "optimal") {
        list(status = prob_status,
             value = result,
             x = as.vector(value(x)))
    } else {
        list(status = prob_status,
             value = NA,
             x = NA)
    }
}
```

## Example

As a simple example, we set $N = 3$, $P = 1$ and
$\boldsymbol{\alpha} = (0.8, 1.0, 1.2)$.

The function outputs the problem status, the maximum communication
rate, and the power allocation required to achieve this maximum
communication rate.

```{r}
buckets <- 3
alpha <- c(0.8, 1.0, 1.2)

sol <- water_filling(buckets, alpha)
cat("Problem status:", sol$status, "\n")
cat(sprintf("Optimal communication rate = %.4g\n", sol$value))
cat("Transmitter powers:\n")
print(round(sol$x, 3))
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(value = sol$value, x = sol$x),
            file = file.path(testdata_dir, "water_filling.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    wf_results <- readRDS(file = file.path(testdata_dir, "water_filling.RDS"))
    expect_equal(sol$value, wf_results$value, tolerance = 1e-5)
    expect_equal(sol$x, wf_results$x, tolerance = 1e-5)
}
```

## Visualization

To illustrate the water filling principle, we plot $\alpha_i + x_i$
and check that this level is flat where power has been allocated.

```{r, fig.cap = "Water filling solution showing power allocation across channels"}
x_vals <- sol$x
y_vals <- alpha + x_vals

df <- data.frame(
    bucket = seq_len(buckets),
    alpha = alpha,
    total = y_vals,
    power = x_vals
)

ggplot(df) +
    geom_col(aes(x = bucket, y = alpha), fill = "steelblue", width = 0.8) +
    geom_col(aes(x = bucket, y = total), fill = "lightblue", alpha = 0.5, width = 0.8) +
    geom_hline(yintercept = max(y_vals), linetype = "dashed", color = "red") +
    scale_x_continuous(breaks = seq_len(buckets)) +
    labs(x = "Bucket Number", y = "Power Level",
         title = "Water Filling Solution") +
    ylim(0, 1.5) +
    theme_minimal()
```

## Larger Example

We now solve a larger water filling problem with $N = 6$ channels and
varying floor levels.

```{r}
n_large <- 6
alpha_large <- c(0.2, 0.5, 0.8, 1.0, 1.2, 1.5)
P_large <- 2

sol_large <- water_filling(n_large, alpha_large, sum_x = P_large)
cat("Problem status:", sol_large$status, "\n")
cat(sprintf("Optimal communication rate = %.4g\n", sol_large$value))
cat("Transmitter powers:\n")
print(round(sol_large$x, 3))
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(value = sol_large$value, x = sol_large$x),
            file = file.path(testdata_dir, "water_filling_large.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    wf_large_results <- readRDS(file = file.path(testdata_dir, "water_filling_large.RDS"))
    expect_equal(sol_large$value, wf_large_results$value, tolerance = 1e-5)
    expect_equal(sol_large$x, wf_large_results$x, tolerance = 1e-5)
}
```

```{r, fig.cap = "Water filling solution for 6 channels"}
x_vals_l <- sol_large$x
y_vals_l <- alpha_large + x_vals_l

df_l <- data.frame(
    bucket = seq_len(n_large),
    alpha = alpha_large,
    total = y_vals_l,
    power = x_vals_l
)

ggplot(df_l) +
    geom_col(aes(x = bucket, y = alpha), fill = "steelblue", width = 0.8) +
    geom_col(aes(x = bucket, y = total), fill = "lightblue", alpha = 0.5, width = 0.8) +
    geom_hline(yintercept = max(y_vals_l), linetype = "dashed", color = "red") +
    scale_x_continuous(breaks = seq_len(n_large)) +
    labs(x = "Bucket Number", y = "Power Level",
         title = "Water Filling Solution (6 Channels)") +
    theme_minimal()
```

## Session Info

```{r}
sessionInfo()
```

## References
