---
title: "Dealing with Solver-Specific Parameters"
author: Anqi Fu, David Kang and Balasubramanian Narasimhan
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: solver_specific
---

```{r, echo = FALSE}
suppressMessages(library(CVXR))
library(tibble)
library(kableExtra)
source(here::here("_check_status.R"))
```

## Overview

`CVXR` features five standard parameters that can be passed to the
`psolve` function. Here are the five parameters:

* `verbose` : A parameter that deals with the verbosity of the
  solver. It is a boolean variable for all solvers, but it can also
  take integer values for the `MOSEK` solver, whose interface deals
  with an integer verbosity parameter with the default of 10. The
  default value of `verbose` for `CVXR` is set to `FALSE`.

* `reltol` : Deals with the relative tolerance of the solver.

* `abstol` : Deals with the absolute tolerance of the solver.

* `feastol` : Deals with feasible tolerance of the solver.

* `num_iter` : A parameter that specifies the maximum number of
  iterations for the solver.

These standard parameters are automatically translated to
solver-specific parameter names by `psolve()`. You can also pass
solver-specific parameters directly via `...`. If both a standard
parameter and its solver-native equivalent are given, the
solver-native name in `...` takes priority.

The mapping table is available as `SOLVER_DEFAULT_PARAM`.

## Solver-Specific Parameters

We will go over some examples for each solver supported by `CVXR`,
the available parameters for each solver, and where to find the
documentation for each solver.

### CLARABEL

The parameters available for `CLARABEL` can be found in this
[link](https://oxfordcontrol.github.io/clarabel-r/reference/clarabel_control.html).

```{r, echo = FALSE}
clarabel_params <- tibble::tibble(
                    Standard = c("verbose", "reltol", "abstol", "feastol", "num_iter"),
                    CLARABEL = c("verbose", "tol_gap_rel", "tol_gap_abs", "tol_feas", "max_iter"),
                    Value = c("FALSE", "1e-08", "1e-08", "1e-08", "200")
                )
knitr::kable(clarabel_params,
             caption = "Default CVXR-CLARABEL parameter mappings",
             format = "html") |>
    kable_styling("striped") |>
    column_spec(2:3, background = "#ececec")
```

The parameters can be entered individually in the arguments of the
`psolve` function. You can use either standard names or
solver-specific names:

```{r}
#| message: false
x <- Variable(1)
obj <- Minimize(x)
prob <- Problem(obj, list(x >= 0))

# Using standard parameter names
result  <- psolve(prob, solver = "CLARABEL", feastol = 1e-2, verbose = FALSE)
check_solver_status(prob)

# Using solver-specific parameter names (equivalent)
result  <- psolve(prob, solver = "CLARABEL", tol_feas = 1e-2, verbose = FALSE)
check_solver_status(prob)
```

### OSQP

The parameters available for `OSQP` can be found in this
[link](https://osqp.org/docs/interfaces/solver_settings.html).

```{r, echo = FALSE}
osqp_params <- tibble::tibble(
                    Standard = c("verbose", "reltol", "abstol", "feastol", "num_iter"),
                    OSQP = c("verbose", "eps_rel", "eps_abs", "eps_prim_inf", "max_iter"),
                    Value = c("FALSE", "1e-05", "1e-05", "1e-04", "10000")
                )
knitr::kable(osqp_params,
             caption = "Default CVXR-OSQP parameter mappings",
             format = "html") |>
    kable_styling("striped") |>
    column_spec(2:3, background = "#ececec")
```

```{r}
#| message: false
x <- Variable(1)
obj <- Minimize(x)
prob <- Problem(obj, list(x >= 0))

# Using standard parameter name
result  <- psolve(prob, solver = "OSQP", feastol = 1e-5, verbose = FALSE)
check_solver_status(prob)

# Using solver-specific parameter name (equivalent)
result  <- psolve(prob, solver = "OSQP", eps_prim_inf = 1e-5, verbose = FALSE)
check_solver_status(prob)
```

### SCS

The parameters available for `SCS` can be found in this
[link](https://cran.r-project.org/web/packages/scs/scs.pdf) under the
`scs_control` function.

```{r, echo = FALSE}
scs_params <- tibble::tibble(
                    Standard = c("verbose", "reltol", "abstol", "feastol", "num_iter"),
                    SCS = c("verbose", "eps_rel", "eps_abs", NA, "max_iters"),
                    Value = c("FALSE", "1e-05", "1e-05", NA, "2500")
                )
knitr::kable(scs_params,
             caption = "Default CVXR-SCS parameter mappings",
             format = "html") |>
    kable_styling("striped") |>
    column_spec(2:3, background = "#ececec")
```

Note that `SCS` does not support a `feastol` parameter --- only
`verbose`, `reltol`, `abstol`, and `num_iter` map to standard names.
Additional solver-specific parameters (like `scale`) can be passed
directly.

```{r}
#| message: false
set.seed(1)
n <- 3
p <- 3
C <- matrix(rnorm(n^2), ncol = n)
A <- lapply(1:p, function(i) matrix(rnorm(n^2), nrow = n))
b <- rnorm(p)

X <- Variable(c(n, n), symmetric = TRUE)
trace_constraints <- lapply(1:p, function(i) matrix_trace(A[[i]] %*% X) == b[i])

obj <- Minimize(matrix_trace(C %*% X))
prob <- Problem(obj, c(list(X >= 0), trace_constraints))
result  <- psolve(prob, solver = "SCS", verbose = FALSE, num_iter = 5000)
check_solver_status(prob)
```

### HiGHS

The HiGHS solver is used for linear and mixed-integer programs.
The parameters available for `HiGHS` can be found in this
[link](https://cran.r-project.org/web/packages/highs/highs.pdf).

```{r, echo = FALSE}
highs_params <- tibble::tibble(
                    Standard = c("verbose", "reltol", "abstol", "feastol", "num_iter"),
                    HIGHS = c("output_flag", NA, NA, NA, "simplex_iteration_limit"),
                    Value = c("FALSE", NA, NA, NA, "2147483647")
                )
knitr::kable(highs_params,
             caption = "Default CVXR-HiGHS parameter mappings",
             format = "html") |>
    kable_styling("striped") |>
    column_spec(2:3, background = "#ececec")
```

```{r}
#| message: false
x <- Variable(1)
obj <- Minimize(x)
prob <- Problem(obj, list(x >= 0))
result  <- psolve(prob, solver = "HIGHS", verbose = FALSE)
check_solver_status(prob)
```

### MOSEK

```{r, echo = FALSE}
mosek_params <- tibble::tibble(
                     Standard = c("verbose", "reltol", "abstol", "feastol", "num_iter"),
                     MOSEK = c("verbose", NA, NA, NA, "num_iter"),
                     Value = c(0, NA, NA, NA, "10000")
                )
knitr::kable(mosek_params,
             caption = "Default CVXR-MOSEK parameter mappings",
             format = "html") |>
    kable_styling("striped") |>
    column_spec(2:3, background = "#ececec")
```

The parameters available for `MOSEK` can be seen in this
[link](https://docs.mosek.com/latest/rmosek/parameters.html). As
you can see, the number of parameters is quite massive. They are also
split between three types of parameters in `Double`, `Integer`, and
`String` parameters. To pass these into the solver, they must be put
in a list beforehand. Moreover, there are also general solver
parameters. Here is an example:

```{r, eval = FALSE}
x <- Variable(1)
obj <- Minimize(x)
prob <- Problem(obj, list(x >= 0))
iparam <- list()
iparam$AUTO_UPDATE_SOL_INFO <- "OFF"
iparam$BI_IGNORE_NUM_ERROR <- "OFF"
iparam$BI_MAX_ITERATIONS <- 100000

dparam <- list()
dparam$BASIS_REL_TOL_S <- 1.0e-12
dparam$BASIS_TOL_S <- 1.0e-6

result  <- psolve(prob, solver = "MOSEK", verbose = FALSE, dparam = dparam, iparam = iparam)
```

Note: MOSEK requires a license (free academic licenses are available).

### Gurobi

```{r, echo = FALSE}
gurobi_params <- tibble::tibble(
                    Standard = c("verbose", "reltol", "abstol", "feastol", "num_iter"),
                    GUROBI = c("verbose", NA, NA, "FeasibilityTol", "IterationLimit"),
                    Value = c("FALSE", NA, NA, "1e-06", "10000")
                )
knitr::kable(gurobi_params,
             caption = "Default CVXR-Gurobi parameter mappings",
             format = "html") |>
    kable_styling("striped") |>
    column_spec(2:3, background = "#ececec")
```

Gurobi parameters are documented at
[https://www.gurobi.com/documentation/](https://www.gurobi.com/documentation/).

```{r, eval = FALSE}
x <- Variable(1)
obj <- Minimize(x)
prob <- Problem(obj, list(x >= 0))
result  <- psolve(prob, solver = "GUROBI", feastol = 1e-4, verbose = FALSE)
```

Note: Gurobi requires a license (free academic licenses are available).

### GLPK / GLPK_MI

```{r, echo = FALSE}
glpk_params <- tibble::tibble(
                    Standard = c("verbose", "reltol", "abstol", "feastol", "num_iter"),
                    GLPK = c("verbose", NA, NA, NA, NA),
                    Value = c("FALSE", NA, NA, NA, NA)
                )
knitr::kable(glpk_params,
             caption = "Default CVXR-GLPK parameter mappings",
             format = "html") |>
    kable_styling("striped") |>
    column_spec(2:3, background = "#ececec")
```

GLPK and GLPK_MI have no standard parameter mappings. Solver-specific
parameters can be passed directly via `...`.

### ECOS

```{r, echo = FALSE}
ecos_params <- tibble::tibble(
                    Standard = c("verbose", "reltol", "abstol", "feastol", "num_iter"),
                    ECOS = c("verbose", "RELTOL", "ABSTOL", "FEASTOL", "MAXIT"),
                    Value = c("FALSE", "1e-08", "1e-08", "1e-08", "100")
                )
knitr::kable(ecos_params,
             caption = "Default CVXR-ECOS parameter mappings",
             format = "html") |>
    kable_styling("striped") |>
    column_spec(2:3, background = "#ececec")
```

```{r}
#| message: false
x <- Variable(1)
obj <- Minimize(x)
prob <- Problem(obj, list(x >= 0))
result  <- psolve(prob, solver = "CLARABEL", feastol = 1e-6, verbose = FALSE)
check_solver_status(prob)
```

### ECOS_BB

```{r, echo = FALSE}
ecos_bb_params <- tibble::tibble(
                    Standard = c("verbose", "reltol", "abstol", "feastol", "num_iter"),
                    ECOS_BB = c("verbose", "RELTOL", "ABSTOL", "FEASTOL", "MI_MAX_ITERS"),
                    Value = c("FALSE", "1e-03", "1e-06", "1e-06", "1000")
                )
knitr::kable(ecos_bb_params,
             caption = "Default CVXR-ECOS_BB parameter mappings",
             format = "html") |>
    kable_styling("striped") |>
    column_spec(2:3, background = "#ececec")
```

ECOS_BB is the mixed-integer extension of ECOS.

### CPLEX

```{r, echo = FALSE}
cplex_params <- tibble::tibble(
                    Standard = c("verbose", "reltol", "abstol", "feastol", "num_iter"),
                    CPLEX = c("verbose", NA, NA, NA, "itlim"),
                    Value = c("FALSE", NA, NA, NA, "10000")
                )
knitr::kable(cplex_params,
             caption = "Default CVXR-CPLEX parameter mappings",
             format = "html") |>
    kable_styling("striped") |>
    column_spec(2:3, background = "#ececec")
```

Note: CPLEX requires a license (free academic licenses are available).

## Session Info

```{r, echo = FALSE}
sessionInfo()
```

