---
title: Getting Faster Results
author: Anqi Fu and Balasubramanian Narasimhan
categories: []
tags: []
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: speed
---

```{r prereqs, echo = FALSE, message = FALSE}
source(here::here("_check_status.R"))
```

## Introduction

As was remarked in the [introduction to
`CVXR`](../../examples/intro.qmd), its chief advantage is
_flexibility_: you can specify a problem in close to mathematical form
and `CVXR` solves it for you, if it can. Behind the scenes, `CVXR`
compiles the domain specific language and verifies the convexity of
the problem before sending it off to solvers. If the problem violates
the rules of [Disciplined Convex Programming](../../reference/dcp.qmd) it is
rejected.

Therefore, it is generally slower than tailor-made solutions to a
given problem.

## Recommended: Use DPP for Repeated Solves

If you need to solve the same problem structure repeatedly with
different data (e.g., tuning a regularization parameter, computing
bootstrap confidence intervals), the recommended approach is
**Disciplined Parametrized Programming (DPP)**. By using `Parameter`
objects, `CVXR` compiles the problem once and caches the mapping from
parameters to problem data. Subsequent solves skip re-compilation and
can be significantly faster.

See the [DPP tutorial](../dpp/tutorial.qmd) for a detailed example
showing roughly a 2x speed-up on a regularized least-squares problem.

## Advanced: Directly Calling the Solver

If DPP is not sufficient for your performance needs, you can bypass
the DSL entirely and call the solver directly. This avoids all
overhead from `CVXR`'s compilation and DCP verification on each
solve, but requires working with low-level solver data structures.

`CVXR` provides a three-step public API for this, mirroring the
`CVXPY` interface:

1. `problem_data()` — compile the problem into solver-ready data
2. `solve_via_data()` — call the solver on the compiled data
3. `problem_unpack_results()` — invert the compilation chain and
   populate variable values

::: {.callout-warning}
We recommend checking DCP-compliance at least once (via
`is_dcp(prob)`) before bypassing verification. _Not verifying
DCP-compliance may result in garbage!_
:::

## An Example

To understand the speed issues, let us consider the global warming
data from the [Carbon Dioxide Information Analysis Center
(CDIAC)](http://cdiac.ess-dive.lbl.gov/ftp/trends/temp/jonescru/)
again. The data points are the annual temperature anomalies relative
to the 1961--1990 mean. We will fit the nearly-isotonic approximation
$\beta \in {\mathbf R}^m$ by solving

$$
\begin{array}{ll}
\underset{\beta}{\mbox{Minimize}} & \frac{1}{2}\sum_{i=1}^m (y_i - \beta_i)^2 + \lambda \sum_{i=1}^{m-1}(\beta_i - \beta_{i+1})_+,
\end{array}
$$
where $\lambda \geq 0$ is a penalty parameter and $x_+ =\max(x,0)$.

This can be solved as follows.

```{r}
suppressMessages(suppressWarnings(library(CVXR)))
data(cdiac)
y <- cdiac$annual
m <- length(y)
lambda <- 0.44
beta <- Variable(m)
obj <- 0.5 * sum((y - beta)^2) + lambda * sum(pos(diff(beta)))
prob <- Problem(Minimize(obj))
soln <- psolve(prob, solver = "CLARABEL")
check_solver_status(prob)
betaHat <- value(beta)
```

This is the _recommended way_ to solve a problem. After solving, you
can inspect timing via `solver_stats()`:

```{r}
stats <- solver_stats(prob)
stats
```

The `SolverStats` object reports the solver's own wall-clock time,
setup time, and iteration count.

However, suppose we wished to construct bootstrap confidence intervals
for the estimate using 100 resamples. It is clear that this
computation time can quickly become limiting.

Below, we show how one can get at the problem data and directly call a
solver to get faster results.

## Profile the Code

Profiling a single fit to the model is useful to figure out where most
of the time is spent.

```{r}
library(profvis)
y <- cdiac$annual
profvis({
    beta <- Variable(m)
    obj <- Minimize(0.5 * sum((y - beta)^2) + lambda * sum(pos(diff(beta))))
    prob <- Problem(obj)
    soln <- psolve(prob, solver = "CLARABEL")
    check_solver_status(prob)
    betaHat <- value(beta)
})
```

It is especially instructive to click on the _data_ tab and open up
the tree for `psolve` to see the sequence of calls and cumulative time
used.

## Directly Calling the Solver

We are mathematically certain that the above is convex and so we can
avoid the repeated DCP checks. The three steps are:

**Step 1.** Extract the compiled problem data using `problem_data()`.
This returns the low-level solver data, the compilation chain, and
inverse data needed to reconstruct the solution.

```{r}
prob_data <- problem_data(prob, solver = "CLARABEL")
```

**Step 2.** Call the solver directly via `solve_via_data()` on the
compilation chain. This bypasses DCP verification entirely.

```{r}
solver_output <- solve_via_data(prob_data$chain, prob_data$data,
                                warm_start = FALSE, verbose = FALSE,
                                solver_opts = list())
```

**Step 3.** Unpack the solver results back into the problem's
variables using `problem_unpack_results()`. This reverses the
compilation chain and populates `value(beta)` etc.

```{r}
problem_unpack_results(prob, solver_output,
                       prob_data$chain, prob_data$inverse_data)
```

## Profile the Direct Call

We can profile this direct call now.

```{r}
profvis({
    beta <- Variable(m)
    obj <- Minimize(0.5 * sum((y - beta)^2) + lambda * sum(pos(diff(beta))))
    prob <- Problem(obj)
    prob_data <- problem_data(prob, solver = "CLARABEL")
    solver_output <- solve_via_data(prob_data$chain, prob_data$data,
                                    warm_start = FALSE, verbose = FALSE,
                                    solver_opts = list())
    problem_unpack_results(prob, solver_output,
                           prob_data$chain, prob_data$inverse_data)
})
```

## Same Answer?

Of course, we should also verify that the results obtained in both
cases are the same.

```{r}
identical(betaHat, value(beta))
```

## Session Info

```{r, echo = FALSE}
sessionInfo()
```

## References
