---
title: Getting Faster Results
author: Anqi Fu and Balasubramanian Narasimhan
categories: []
tags: []
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: speed
---

```{r prereqs, echo = FALSE, message = FALSE}
source(here::here("_check_status.R"))
```

## Warning

1. The solution described below is useful when you mathematically know
   a problem is DCP-compliant and none of your data inputs will change
   the nature of the problem. We recommend that users check the
   DCP-compliance of a problem (via a call to `is_dcp(prob)` for
   example) at least once to ensure this is the case.  _Not verifying
   DCP-compliance may result in garbage!_

## Introduction

As was remarked in the [introduction to
`CVXR`](/cvxr_examples/intro/), its chief advantage is
_flexibility_: you can specify a problem in close to mathematical form
and `CVXR` solves it for you, if it can. Behind the scenes, `CVXR`
compiles the domain specific language and verifies the convexity of
the problem before sending it off to solvers. If the problem violates
the rules of [Disciplined Convex Programming](../../reference/dcp.qmd) it is
rejected.

Therefore, it is generally slower than tailor-made solutions to a
given problem.

## An Example

To understand the speed issues, let us consider the global warming
data from the [Carbon Dioxide Information Analysis Center
(CDIAC)](http://cdiac.ess-dive.lbl.gov/ftp/trends/temp/jonescru/)
again. The data points are the annual temperature anomalies relative
to the 1961--1990 mean. We will fit the nearly-isotonic approximation
$\beta \in {\mathbf R}^m$ by solving

$$
\begin{array}{ll}
\underset{\beta}{\mbox{Minimize}} & \frac{1}{2}\sum_{i=1}^m (y_i - \beta_i)^2 + \lambda \sum_{i=1}^{m-1}(\beta_i - \beta_{i+1})_+,
\end{array}
$$
where $\lambda \geq 0$ is a penalty parameter and $x_+ =\max(x,0)$.

This can be solved as follows.

```{r}
suppressMessages(suppressWarnings(library(CVXR)))
data(cdiac)
y <- cdiac$annual
m <- length(y)
lambda <- 0.44
beta <- Variable(m)
obj <- 0.5 * sum((y - beta)^2) + lambda * sum(pos(diff(beta)))
prob <- Problem(Minimize(obj))
soln <- psolve(prob, solver = "CLARABEL")
check_solver_status(prob)
betaHat <- value(beta)
```

This is the _recommended way_ to solve a problem.

However, suppose we wished to construct bootstrap confidence intervals
for the estimate using 100 resamples.  It is clear that this
computation time can quickly become limiting .

Below, we show how one can get at the problem data and directly call a
solver to get faster results.

## Profile the code

Profiling a single fit to the model is useful to figure out where most
of the time is spent.

```{r}
library(profvis)
y <- cdiac$annual
profvis({
    beta <- Variable(m)
    obj <- Minimize(0.5 * sum((y - beta)^2) + lambda * sum(pos(diff(beta))))
    prob <- Problem(obj)
    soln <- psolve(prob, solver = "CLARABEL")
    check_solver_status(prob)
    betaHat <- value(beta)
})
```

It is especially instructive to click on the _data_ tab and open up
the tree for `solve` to see the sequence of calls and cumulative time
used.

## Directly Calling the Solver

We are mathematically certain that the above is convex and so we can
avoid the repeated DCP checks. We can obtain the problem data for a
particular solver using the function `get_problem_data` and directly
hand that data to the solver to get the solution.

```{r}
prob_data <- get_problem_data(prob, solver = "CLARABEL")
```

Now that we have the problem data, we can call the Clarabel solver
directly with the right arguments.

```{r}
solver_data <- prob_data$data
## Convert ConeDims to a list format that clarabel::clarabel expects
cone_dims <- solver_data[["dims"]]
cones <- list()
if (cone_dims@zero > 0L) cones$z <- cone_dims@zero
if (cone_dims@nonneg > 0L) cones$l <- cone_dims@nonneg
for (q in cone_dims@soc) cones <- c(cones, list(q = q))
if (cone_dims@exp > 0L) {
    for (i in seq_len(cone_dims@exp)) cones <- c(cones, list(ep = 0L))
}
solver_output <- clarabel::clarabel(
    A = solver_data[["A"]],
    b = solver_data[["b"]],
    q = solver_data[["c"]],
    cones = cones
)
```

Finally, we can obtain the results by using the internal chain to
unpack the solver results.

```{r}
solver_solution <- CVXR:::problem_unpack_results(prob, solver_output,
                                                  prob_data$chain, prob_data$inverse_data)
```

## Profile the Direct Call

We can profile this direct call now.

```{r}
profvis({
    beta <- Variable(m)
    obj <- Minimize(0.5 * sum((y - beta)^2) + lambda * sum(pos(diff(beta))))
    prob <- Problem(obj)
    prob_data <- get_problem_data(prob, solver = "CLARABEL")
    solver_data <- prob_data$data
    cone_dims <- solver_data[["dims"]]
    cones <- list()
    if (cone_dims@zero > 0L) cones$z <- cone_dims@zero
    if (cone_dims@nonneg > 0L) cones$l <- cone_dims@nonneg
    for (q in cone_dims@soc) cones <- c(cones, list(q = q))
    if (cone_dims@exp > 0L) {
        for (i in seq_len(cone_dims@exp)) cones <- c(cones, list(ep = 0L))
    }
    solver_output <- clarabel::clarabel(
        A = solver_data[["A"]],
        b = solver_data[["b"]],
        q = solver_data[["c"]],
        cones = cones
    )
    CVXR:::problem_unpack_results(prob, solver_output,
                                  prob_data$chain, prob_data$inverse_data)
})
```

## Same Answer?

Of course, we should also verify that the results obtained in both
cases are same.

```{r}
identical(betaHat, value(beta))
```

## Session Info

```{r}
sessionInfo()
```

## References
