---
title: Using Other Solvers
author: Anqi Fu and Balasubramanian Narasimhan
categories: []
tags: []
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: using-other-solvers
---


```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
library(kableExtra)
```

## Introduction

`CVXR` supports 13 solvers, including several bundled open source
solvers and optional commercial ones.

### Open Source Solvers (on CRAN)

- [Clarabel](https://oxfordcontrol.github.io/clarabel-r/), a versatile
  conic solver written in Rust (**default solver**).
- [SCS](https://github.com/cvxgrp/scs), a conic solver using operator
  splitting, effective for large-scale problems.
- [OSQP](https://www.osqp.org), for quadratic and linear programs.
- [HiGHS](https://highs.dev/), for linear, quadratic, and
  mixed-integer programs.
- [ECOS / ECOS_BB](https://github.com/embotech/ecos), an interior-point
  solver for second-order cone programs (ECOS_BB adds
  mixed-integer support).
- [GLPK / GLPK_MI](https://www.gnu.org/software/glpk/), for linear
  programs and mixed-integer linear programs.
- [CVXOPT](https://cvxopt.org/), for linear and second-order cone programs.
- [PIQP](https://predict-epfl.github.io/piqp/), a proximal
  interior-point solver for quadratic programs.

### Commercial Solvers

- [MOSEK](https://www.mosek.com/), supporting LP, QP, SOCP, SDP, and
  exponential/power cone programs. Free academic licenses available.
- [Gurobi](https://www.gurobi.com/), supporting LP, QP, SOCP, and
  mixed-integer programs. Free academic licenses available.
- [CPLEX](https://www.ibm.com/products/ilog-cplex-optimization-studio),
  supporting LP, QP, and mixed-integer programs.

## Solver Capabilities

The table below shows each solver, its R package dependency, and the
problem classes it supports.

```{r, echo = FALSE}
check <- "\U2713"
d <- tibble::tibble(
    Solver = c("CLARABEL", "SCS", "MOSEK", "ECOS", "ECOS_BB",
               "GUROBI", "GLPK", "GLPK_MI", "HIGHS",
               "CVXOPT", "OSQP", "CPLEX", "PIQP"),
    `R Package` = c(
        '<a href="https://cran.r-project.org/package=clarabel"><code>clarabel</code></a>',
        '<a href="https://cran.r-project.org/package=scs"><code>scs</code></a>',
        '<a href="https://docs.mosek.com/latest/rmosek/index.html"><code>Rmosek</code></a>',
        '<a href="https://cran.r-project.org/package=ECOSolveR"><code>ECOSolveR</code></a>',
        '<a href="https://cran.r-project.org/package=ECOSolveR"><code>ECOSolveR</code></a>',
        '<a href="https://www.gurobi.com/"><code>gurobi</code></a>',
        '<a href="https://cran.r-project.org/package=Rglpk"><code>Rglpk</code></a>',
        '<a href="https://cran.r-project.org/package=Rglpk"><code>Rglpk</code></a>',
        '<a href="https://cran.r-project.org/package=highs"><code>highs</code></a>',
        '<a href="https://cran.r-project.org/package=cccp"><code>cccp</code></a>',
        '<a href="https://cran.r-project.org/package=osqp"><code>osqp</code></a>',
        '<a href="https://cran.r-project.org/package=Rcplex"><code>Rcplex</code></a>',
        '<a href="https://cran.r-project.org/package=piqp"><code>piqp</code></a>'),
    LP   = rep(check, 13),
    QP   = c(check, check, check, " ", " ",
             check, " ", " ", check,
             " ", check, check, check),
    SOCP = c(check, check, check, check, check,
             check, " ", " ", " ",
             check, " ", " ", " "),
    SDP  = c(check, check, check, " ", " ",
             " ", " ", " ", " ",
             " ", " ", " ", " "),
    EXP  = c(check, check, check, check, check,
             " ", " ", " ", " ",
             " ", " ", " ", " "),
    MIP  = c(" ", " ", " ", " ", check,
             check, " ", check, check,
             " ", " ", check, " ")
)
knitr::kable(d, format = "html", escape = FALSE,
             caption = "Solver Capabilities") |>
    kable_styling("striped") |>
    column_spec(1:8, background = "#ececec")
```

## Installing Commercial Solvers

### MOSEK

[MOSEK](https://www.mosek.com) provides an academic version that is
free of charge: one can obtain the free academic license after
registering. Once the license for the product has been activated, the
[Rmosek documentation](https://docs.mosek.com/latest/rmosek/index.html)
provides all the details for installation. A quick check to ensure
things are working is to run the example:

```{r, eval = FALSE}
library(Rmosek)
example(mosek)
```

### Gurobi

[Gurobi](https://www.gurobi.com/) offers free academic licenses. After
installing Gurobi and the `gurobi` R package (bundled with the Gurobi
installation), verify with:

```{r, eval = FALSE}
library(gurobi)
model <- list(A = matrix(c(1, 1), nrow = 1),
              obj = c(1, 1), sense = "<=", rhs = 1)
gurobi(model)
```

### CPLEX

[IBM CPLEX](https://www.ibm.com/products/ilog-cplex-optimization-studio)
is available through IBM's academic initiative. Install the `Rcplex`
package from CRAN after installing the CPLEX library.

Once a commercial solver is installed, you can use it by passing
`solver = "MOSEK"`, `solver = "GUROBI"`, or `solver = "CPLEX"` to
`psolve()`.

## Session Info

```{r}
sessionInfo()
```

## References
