---
title: Using Other Solvers
author: Anqi Fu and Balasubramanian Narasimhan
categories: []
tags: []
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: using-other-solvers
---


```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
library(kableExtra)
```

## Introduction

`CVXR` (version `packageVersion("CVXR")`) supports `length(installed_solvers)` solvers, including several bundled open source
solvers and optional commercial ones.

### Open Source Solvers (on CRAN)

- [Clarabel](https://oxfordcontrol.github.io/clarabel-r/), a versatile
  conic solver written in Rust (**default solver**).
- [SCS](https://github.com/cvxgrp/scs), a conic solver using operator
  splitting, effective for large-scale problems.
- [OSQP](https://www.osqp.org), for quadratic and linear programs.
- [HiGHS](https://highs.dev/), for linear, quadratic, and
  mixed-integer programs.
- [ECOS / ECOS_BB](https://github.com/embotech/ecos), an interior-point
  solver for second-order cone programs (ECOS_BB adds
  mixed-integer support).
- [GLPK / GLPK_MI](https://www.gnu.org/software/glpk/), for linear
  programs and mixed-integer linear programs.
- [CVXOPT](https://cvxopt.org/), for linear and second-order cone programs.
- [PIQP](https://predict-epfl.github.io/piqp/), a proximal
  interior-point solver for quadratic programs.
- [SCIP](https://www.scipopt.org/) for linear, second-order cone
  programs and mixed-integer programs.

### Commercial Solvers

- [MOSEK](https://www.mosek.com/), supporting LP, QP, SOCP, SDP, and
  exponential/power cone programs. Free academic licenses available.
- [Gurobi](https://www.gurobi.com/), supporting LP, QP, SOCP, and
  mixed-integer programs. Free academic licenses available.
- [CPLEX](https://www.ibm.com/products/ilog-cplex-optimization-studio),
  supporting LP, QP, and mixed-integer programs.
- [XPRESS](https://www.fico.com/en/products/fico-xpress-optimization)
  supporting LP, QP, SOCP and mixed-integer programs

## Solver Capabilities

The table below shows each solver, its R package dependency, and the
problem classes it supports.

```{r, echo = FALSE}
source(here::here("_solver_table.R"))
knitr::kable(solver_tbl, format = "html", escape = FALSE,
             caption = "Solver Capabilities") |>
    kableExtra::kable_styling("striped") |>
    kableExtra::column_spec(1:8, background = "#ececec")
```

## Installing Commercial Solvers

Once a commercial solver is installed, it will be auto-detected
depending on the type of problem you solve. However, you can force its
use by passing `solver = "MOSEK"`, `solver = "GUROBI"`, or `solver =
"CPLEX"` to `psolve()`.

Below we provide some hints for installing commercial solvers. 

### MOSEK

[MOSEK](https://www.mosek.com) provides an academic version that is
free of charge: one can obtain the free academic license after
registering. Once the license for the product has been activated, the
[Rmosek documentation](https://docs.mosek.com/latest/rmosek/index.html)
provides all the details for installation. 

To be concrete, on my Apple Silicon Mac, the personal use installation
lands in `~/mosek/11.1`. So all one has to do is to
`source("~/mosek/11.1/tools/platform/osxaarch64/rmosek.R")` and follow
the directions given.

A quick check to ensure things are working is to run the example:

```{r, eval = FALSE}
library(Rmosek)
example(mosek)
```

### Gurobi

[Gurobi](https://www.gurobi.com/) offers free academic licenses and
includes a binary R package in the installation bundle. On my Apple
Silicon mac, this binary package may be found at
`/Library/gurobi1301/macos_universal2/R` along with a `README`
containing instructions to follow. Verify with:

```{r, eval = FALSE}
library(gurobi)
model <- list(A = matrix(c(1, 1), nrow = 1),
              obj = c(1, 1), sense = "<=", rhs = 1)
gurobi(model)
```

### CPLEX

[IBM
CPLEX](https://www.ibm.com/products/ilog-cplex-optimization-studio) is
available through IBM's academic initiative. The hardest part is
finding the community edition and IBM doesn't make this easy, but a
free community edition **is** available. There is also a specific
version for Apple Silicon and you have to ensure that you don't
mistakenly use the `x86_64` version; that will not work on the Apple
Silicon platform, even if you use rosetta.

Once CPLEX is installed, the R package installation is from source and
so you need compilation tools.  An example R installation script for
Apple Silicon is below.

```{r, eval = FALSE}
cplex_location <- "/Applications/CPLEX_Studio_Community2212"
arch <- "arm64"
cplex_configure_args <- c(Rcplex = sprintf("--with-cplex-include='%s/cplex/include' --with-cplex-lib='-L%s/cplex/lib/%s_osx/static_pic -lilocplex -lcplex'", cplex_location, cplex_location, arch))
install.packages("Rcplex", configure.args = cplex_configure_args)
```

### XPRESS

[XPRESS](https://www.fico.com/en/products/fico-xpress-optimization)
can be downloaded from the website after surrendering some information
on who you are etc. Once again, a community license is automatically
included with the download. The annoyance is that the license needs to
be available any time you use the solver. So I usually copy the first
section below into `~/.Rprofile` after verifying that it works in an
initial R session.

```{r, eval = FALSE}
## Test that this code chunk doesn't warn in an R session and then copy into ~/.Rprofile
XPRESSDIR <- "/Applications/FICO Xpress/xpressmp"
XPRESS_LICENSE_FILE <- "xpauth.xpr" ## sometimes named "community-xpauth.xpr"
XPAUTH_PATH <- sprintf("%s/bin/%s", XPRESSDIR, XPRESS_LICENSE_FILE)
if (!file.exists(XPAUTH_PATH)) {
  warning(sprintf("XPRESS license file not located at %s", XPAUTH_PATH))
}
## Set environment variables
Sys.setenv(
  XPRESSDIR = XPRESSDIR,
  XPAUTH_PATH = XPAUTH_PATH
)
```

Install the solver from source (so you need compilation tools) via:

```{r, eval = FALSE}
install.packages(sprintf("%s/R/xpress.tar.gz", XPRESSDIR))
```

Finally, verify that this runs.

```{r, eval = FALSE}
library(xpress)
prob <- createprob()
print(prob)
```

## Session Info

```{r, echo = FALSE}
sessionInfo()
```

## References
