---
title: Using Other Solvers
author: Anqi Fu and Balasubramanian Narasimhan
categories: []
tags: []
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: using-other-solvers
---


```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
library(kableExtra)
```

## Introduction

`CVXR` comes with several open source solvers:

- [Clarabel](https://oxfordcontrol.github.io/clarabel-r/), a versatile
  conic solver written in Rust (default solver).
- [OSQP](https://www.osqp.org), for quadratic and linear programs.
- [SCS](https://github.com/cvxgrp/scs), a conic solver using operator splitting.
- [HiGHS](https://highs.dev/), for linear and mixed-integer programs.

`CVXR` can also make use of the commercial solver:

- [MOSEK](https://www.mosek.com/), supporting LP, QP, SOCP, SDP, and
  exponential/power cone programs.

__NOTE__: `MOSEK` requires a license to use but free academic licenses
are available.

## Solver capabilities

Table 1 below shows the capabilities of various solvers.

```{r, echo = FALSE}
check <- "\U2713"
d  <- tibble::tibble(
                  Solver = c("CLARABEL", "OSQP", "SCS", "HIGHS", "MOSEK"),
                  `R package` = c("[`clarabel`](https://cran.r-project.org/package=clarabel)",
                                  "[`osqp`](https://www.osqp.org)",
                                  "[`scs`](https://cran.r-project.org/package=scs)",
                                  "[`highs`](https://cran.r-project.org/package=highs)",
                                  "[`Rmosek`](https://docs.mosek.com/latest/rmosek/index.html)"),
                  LP = c(check, check, check, check, check),
                  QP = c(check, check, check, check, check),
                  SOCP = c(check, " " , check, " " , check),
                  SDP = c(check, " " , check, " " , check),
                  EXP = c(check, " " , check, " " , check),
                  MIP = c(" " , " " , " " , check, " ")
              )
knitr::kable(d, format = "html", caption = "Solver Capabilities") |>
    kable_styling("striped") |>
    column_spec(1:8, background = "#ececec")

```

## Commercial Solvers

### Installing `MOSEK`

[MOSEK](https://www.mosek.com) provides an academic version that is
free of charge: one can obtain the free academic license after
registering. Once the license for the product has been activated, the
[Rmosek documentation](https://docs.mosek.com/latest/rmosek/index.html)
provides all the details for installation. A quick check to ensure
things are working is to run the example:

```{r, eval = FALSE}
library(Rmosek)
example(mosek)
```

Once everything is working, you can check that `CVXR` recognizes the
solver by trying to solve a problem with `solver = "MOSEK"`.

## Session Info

```{r}
sessionInfo()
```

## References
