---
title: Warm Starts
author: Anqi Fu, David Kang and Balasubramanian Narasimhan
categories:
  - Version 1.0 features
params:
  mode: ignore
---

Starting with version 0.11.1, `CVXR` supports `Parameter` objects
that allow you to change problem data and re-solve without
reconstructing the problem from scratch. This uses the `EvalParams`
reduction under the hood to substitute parameter values before each
solve.

## Lasso Example

We demonstrate the parametric workflow in a simple lasso problem:

$$
	\begin{array}{ll} \underset{x}{\mbox{minimize}} & \frac{1}{2}\|Ax - b\|_2^2 + \gamma \|x\|_1\\
	\mbox{subject to} & x \geq 0
	\end{array}
$$

where $\gamma > 0$ is a regularization parameter.

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
library(Matrix)
source(here::here("_check_status.R"))
```

```{r, results = "hold"}
## Problem data
set.seed(1)
m <- 2000
n <- 1000
A <- matrix(rnorm(m*n), nrow = m, ncol = n)
b <- rnorm(m)

## Define the problem once using a Parameter
x <- Variable(n)
gamma <- Parameter(nonneg = TRUE)
obj <- Minimize(.5 * sum_squares(A %*% x - b) + gamma * norm1(x))
constraint <- list(x >= 0)
prob <- Problem(obj, constraint)

## Solve for gamma = 1
value(gamma) <- 1
result1 <- psolve(prob, solver = "OSQP")
check_solver_status(prob)
cat(sprintf("gamma = 1, optimal value: %f\n", result1))

## Re-solve for gamma = 2 (much faster — reuses cached compilation)
value(gamma) <- 2
result2 <- psolve(prob, solver = "OSQP")
check_solver_status(prob)
cat(sprintf("gamma = 2, optimal value: %f\n", result2))
```

## Session Info

```{r, echo = FALSE}
sessionInfo()
```

## References
