---
title: Warm Starts
author: Anqi Fu, David Kang and Balasubramanian Narasimhan
categories:
  - Version 1.0 features
params:
  mode: ignore
---

Starting with version 0.11.1, `CVXR` supports `Parameter` objects
that allow you to change problem data and re-solve without
reconstructing the problem from scratch. This uses the `EvalParams`
reduction under the hood to substitute parameter values before each
solve.

## Lasso Example

We demonstrate the parametric workflow in a simple lasso problem:

$$
	\begin{array}{ll} \underset{x}{\mbox{minimize}} & \frac{1}{2}\|Ax - b\|_2^2 + \gamma \|x\|_1\\
	\mbox{subject to} & x \geq 0
	\end{array}
$$

where $\gamma > 0$ is a regularization parameter.

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
library(Matrix)
source(here::here("_check_status.R"))
```

```{r, message = FALSE}
## Problem data
set.seed(1)
m <- 200
n <- 100
A <- matrix(rnorm(m*n), nrow = m, ncol = n)
b <- rnorm(m)

## Define the problem once using a Parameter
x <- Variable(n)
gamma <- Parameter(nonneg = TRUE)
obj <- Minimize(.5 * sum_squares(A %*% x - b) + gamma * norm1(x))
constraint <- list(x >= 0)
prob <- Problem(obj, constraint)

## Sweep over gamma values
gamma_vals <- c(0.1, 0.5, 1, 2, 5, 10)
timings <- numeric(length(gamma_vals))
opt_vals <- numeric(length(gamma_vals))

for (i in seq_along(gamma_vals)) {
    value(gamma) <- gamma_vals[i]
    timings[i] <- system.time(opt_vals[i] <- psolve(prob, solver = "OSQP"))["elapsed"]
    check_solver_status(prob)
}

results <- data.frame(
    gamma = gamma_vals,
    optimal_value = round(opt_vals, 4),
    elapsed = round(timings, 3)
)
```

The first solve includes one-time compilation overhead; subsequent
re-solves skip this step entirely:

```{r}
knitr::kable(results, col.names = c("$\\gamma$", "Optimal Value", "Elapsed (s)"))
```

## Session Info

```{r, echo = FALSE}
sessionInfo()
```

## References
