---
title: Warm Starts
author: Anqi Fu, David Kang and Balasubramanian Narasimhan
categories:
  - Version 1.0 features
params:
  mode: ignore
---

Starting with version 0.11.1, `CVXR` supports `Parameter` objects
that allow you to change problem data and re-solve without
reconstructing the problem from scratch. This uses the `EvalParams`
reduction under the hood to substitute parameter values before each
solve.

## Lasso Example

We demonstrate the parametric workflow in a simple lasso problem:

$$
	\begin{array}{ll} \underset{x}{\mbox{minimize}} & \frac{1}{2}\|Ax - b\|_2^2 + \gamma \|x\|_1\\
	\mbox{subject to} & x \geq 0
	\end{array}
$$

where $\gamma > 0$ is a regularization parameter.

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
library(Matrix)
```

::: {.callout-note}
DPP re-solve with parameter updates is planned for a future release.
The code below demonstrates solving the same problem structure for
multiple parameter values by constructing a new problem for each
value.
:::

```{r}
## Problem data
set.seed(1)
m <- 2000
n <- 1000
A <- matrix(rnorm(m*n), nrow = m, ncol = n)
b <- rnorm(m)

## Solve for gamma = 1
x <- Variable(n)
gamma_val <- 1
obj <- Minimize(.5 * sum_squares(A %*% x - b) + gamma_val * norm1(x))
constraint <- list(x >= 0)
prob <- Problem(obj, constraint)
result1 <- psolve(prob, solver = "OSQP")
cat(sprintf("gamma = 1, optimal value: %f\n", result1))

## Re-solve for gamma = 2
gamma_val <- 2
obj2 <- Minimize(.5 * sum_squares(A %*% x - b) + gamma_val * norm1(x))
prob2 <- Problem(obj2, constraint)
result2 <- psolve(prob2, solver = "OSQP")
cat(sprintf("gamma = 2, optimal value: %f\n", result2))
```

Once DPP re-solve is available, you will be able to define the
problem once with a `Parameter` for `gamma` and re-solve for different
parameter values without reconstructing the problem.

## Session Info

```{r}
sessionInfo()
```

## References
