---
title: Disciplined Nonlinear Programming
author: "CVXPY Developers and Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: dnlp_tutorial
  min_version: 1.9.0
---

```{r feature-gate, echo = FALSE}
## Show feature code regardless of version, but only EVALUATE it when the
## installed CVXR implements DNLP (params$min_version). On an older CVXR the
## code is displayed but not run, so the page still renders.
run_ok <- utils::packageVersion("CVXR") >= numeric_version(params$min_version) &&
          all(vapply(c("sparsediff", "ipopt", "Uno"), requireNamespace,
                     logical(1), quietly = TRUE))
knitr::opts_chunk$set(eval = run_ok)
```

```{r, echo = FALSE, message = FALSE, eval = run_ok && params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
```

## Introduction

**Disciplined Nonlinear Programming (DNLP)**, introduced in `CVXR` 1.9.0,
extends `CVXR` beyond convex optimization to **smooth nonlinear programs** —
problems whose objective and constraints are built from differentiable atoms,
but which need not be convex. A DNLP is solved by handing exact first and
second derivatives to a general nonlinear-programming solver.

The disciplined grammar for DNLP is broader than DCP: any DCP problem is also a
DNLP, but a DNLP may also use smooth atoms in ways DCP forbids (for example, a
product of two expressions that each involve a variable). You check membership
with `is_dnlp()`, and you solve with `psolve(prob, nlp = TRUE)`.

The DNLP path is powered by automatic derivatives from the optional
[`sparsediff`](https://github.com/bnaras/sparsediff) package and one of the
nonlinear solvers in the optional [`ipopt`](https://github.com/bnaras/ipopt)
or [`Uno`](https://github.com/bnaras/Uno) packages.

```{r availability}
packageVersion("CVXR")
c(sparsediff = requireNamespace("sparsediff", quietly = TRUE),
  ipopt      = requireNamespace("ipopt", quietly = TRUE),
  Uno        = requireNamespace("Uno", quietly = TRUE))
```

## A first smooth program

Every DCP problem is a DNLP, so the simplest way to meet the NLP path is to send
a convex problem down it. We minimize a quadratic and confirm the problem is a
DNLP. We pass `print_level = 0L` and `sb = "yes"` (Ipopt options) to silence the
solver's iteration log and its one-time startup banner.

```{r first}
x <- Variable(2)
prob <- Problem(Minimize(sum_squares(x - c(1, 2))))
is_dnlp(prob)

opt <- psolve(prob, nlp = TRUE, print_level = 0L, sb = "yes")
opt                 # optimal value (0)
value(x)            # (1, 2)
```

```{r first-test, echo = FALSE, eval = run_ok && params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(opt = opt, x = value(x)), file = file.path(testdata_dir, "first.RDS"))
} else {
    saved <- readRDS(file.path(testdata_dir, "first.RDS"))
    expect_equal(opt, saved$opt, tolerance = 1e-10)
    expect_equal(value(x), saved$x, tolerance = 1e-10)
}
```

## Smooth atoms

DNLP adds a family of classically differentiable atoms that are *not* part of
the convex grammar, so they can appear anywhere in a smooth objective or
constraint:

| Atom | Meaning |
|------|---------|
| `sin(x)`, `cos(x)`, `tan(x)` | Trigonometric functions |
| `sinh(x)`, `tanh(x)` | Hyperbolic functions |
| `asinh(x)`, `atanh(x)` | Inverse hyperbolic functions |
| `normcdf(x)` | Standard normal CDF |
| `prod(x)` | Product of entries (also `prod_entries(X, axis = ...)`) |

These join the smooth atoms `CVXR` already had — affine maps, `exp`, `log`,
`entr`, `logistic`, `kl_div`, `rel_entr`, `xexp`, `power` (constant exponent),
`geo_mean`, `log_sum_exp`, `quad_form`, and `quad_over_lin` — which DNLP can now
use in either curvature direction.

## Beyond DCP: a nonconvex form of a convex problem

Consider the **maximum-entropy distribution** on $\{1,\dots,K\}$ with a
prescribed mean. The negative entropy $\sum_i p_i \log p_i$ is convex, but if we
write it literally as a sum of products `p * log(p)` — rather than with the
`entr` atom — the expression is *not* DCP, because it multiplies two
variable-dependent terms. It is, however, a perfectly good **smooth** objective,
so it is a DNLP.

```{r entropy}
K <- 5; support <- 1:K; mu <- 2
p <- Variable(K, nonneg = TRUE)

objective   <- Minimize(sum(p * log(p)))          # not DCP, but smooth
constraints <- list(sum(p) == 1, sum(support * p) == mu)
prob <- Problem(objective, constraints)

c(is_dcp = is_dcp(prob), is_dnlp = is_dnlp(prob))

neg_entropy <- psolve(prob, nlp = TRUE, print_level = 0L, sb = "yes")
round(value(p), 5)
c(total = sum(value(p)), mean = sum(support * value(p)))
```

Maximum-entropy theory says the solution is a discrete exponential (Gibbs)
distribution $p_i \propto e^{-\theta i}$ — equivalently, $\log p_i$ is affine in
$i$. The solution bears this out: the successive differences of $\log p_i$ are
constant, and that common value is $-\theta$.

```{r entropy-gibbs}
log_p <- log(as.numeric(value(p)))
round(diff(log_p), 5)                 # constant => Gibbs form
theta <- -mean(diff(log_p))
cf <- exp(-theta * support); cf <- cf / sum(cf)
round(cf, 5)                          # matches value(p)
```

## Duals: IPOPT vs UNO

The two NLP backends differ in one user-visible way: **dual variables**. The
`IPOPT` path returns *no* duals, matching CVXPY's NLP interface — so after an
IPOPT solve, `dual_value()` is `NULL`:

```{r dual-ipopt}
psolve(prob, solver = "IPOPT", nlp = TRUE, print_level = 0L, sb = "yes")
dual_value(constraints[[2]])          # NULL: IPOPT path exposes no duals
```

The `UNO` path, as a `CVXR` extension beyond CVXPY, *does* recover the
constraint duals. The multiplier on the mean constraint is exactly the Gibbs
parameter $\theta$ we read off the solution above:

```{r dual-uno}
psolve(prob, solver = "UNO", nlp = TRUE, logger = "SILENT")
dual_value(constraints[[2]])          # the mean-constraint multiplier (= theta)
```

```{r dual-test, echo = FALSE, eval = run_ok && params$mode %in% c("test", "save")}
uno_theta <- as.numeric(dual_value(constraints[[2]]))
if (params$mode == "save") {
    saveRDS(list(theta = theta, uno_theta = uno_theta), file = file.path(testdata_dir, "duals.RDS"))
} else {
    saved <- readRDS(file.path(testdata_dir, "duals.RDS"))
    expect_equal(uno_theta, saved$uno_theta, tolerance = 1e-10)   # same solver/machine: bit-exact
    ## cross-solver agreement (UNO dual vs theta from the IPOPT solution): two
    ## different algorithms, so this agrees to ~1e-8, not to 1e-10.
    expect_equal(uno_theta, theta, tolerance = 1e-6)
}
```

```{r entropy-test, echo = FALSE, eval = run_ok && params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(val = neg_entropy, p = value(p)), file = file.path(testdata_dir, "entropy.RDS"))
} else {
    saved <- readRDS(file.path(testdata_dir, "entropy.RDS"))
    expect_equal(neg_entropy, saved$val, tolerance = 1e-10)
    expect_equal(value(p), saved$p, tolerance = 1e-10)
}
```

## Nonconvex problems and random restarts

DNLP problems can be genuinely nonconvex, with multiple local minima. A
nonlinear solver returns a *local* optimum that depends on the starting point.
For example, minimizing $\sin(x) + \tfrac{1}{10}x^2$ over $[-10, 10]$ has several
local minima.

`psolve(prob, nlp = TRUE, best_of = n)` solves from `n` random initial points
and keeps the best result. Each starting point is drawn from the variable's
finite bounds, or from a sampling region you set with `sample_bounds()`.

```{r bestof}
set.seed(1)
x <- Variable(bounds = list(-10, 10))
prob <- Problem(Minimize(sin(x) + 0.1 * x^2))

best <- psolve(prob, nlp = TRUE, best_of = 5, print_level = 0L, sb = "yes")
best                                  # best objective over the 5 restarts
value(x)

## the per-restart objectives are available for inspection
solver_stats(prob)@extra_stats$all_objs_from_best_of
```

```{r bestof-test, echo = FALSE, eval = run_ok && params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(best = best, x = value(x)), file = file.path(testdata_dir, "bestof.RDS"))
} else {
    saved <- readRDS(file.path(testdata_dir, "bestof.RDS"))
    expect_equal(best, saved$best, tolerance = 1e-10)
    expect_equal(value(x), saved$x, tolerance = 1e-10)
}
```

For a variable with no finite bounds, supply a sampling region explicitly:

```{r samplebounds, eval = FALSE}
x <- Variable(2)
sample_bounds(x) <- c(-5, 5)          # draw restarts from [-5, 5]^2
```

## Choosing an NLP solver

With `nlp = TRUE`, `CVXR` picks an installed NLP solver automatically — `IPOPT`
first when the `ipopt` package is available, matching CVXPY's preference order.
You can name one explicitly:

```{r solvers, eval = FALSE}
psolve(prob, nlp = TRUE, solver = "IPOPT")     # Ipopt (interior point, via MUMPS)
psolve(prob, nlp = TRUE, solver = "UNO")       # Uno
psolve(prob, nlp = TRUE, solver = "uno_sqp")   # Uno's filterSQP preset
```

A note on `UNO`: the `CVXR` interface defaults `UNO` to its interior-point
`ipopt` preset (with MUMPS), which differs from CVXPY's `filtersqp` default. The
bundled `Uno` build is HiGHS-only, and HiGHS solves only *convex* quadratic
subproblems, so `filtersqp` is reliable only for problems whose subproblem
Hessians stay positive semidefinite; the interior-point preset is robust on
nonconvex problems too.

## Notes

* DNLP requires the `sparsediff` package (for derivatives) and at least one of
  `ipopt` or `Uno` (for the solver). With none installed, `psolve(nlp = TRUE)`
  raises an informative error.
* `is_dnlp(prob)` reports whether a problem satisfies the disciplined nonlinear
  grammar; since DCP $\subset$ DNLP, every DCP problem qualifies.
* Because the problems may be nonconvex, the returned point is a local optimum;
  use `best_of` to improve the odds of finding the global optimum.

## References
