---
title: Disciplined Nonlinear Programming
author: "CVXPY Developers and Balasubramanian Narasimhan"
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: dnlp_tutorial
  min_version: 1.9.1
---

```{r feature-gate, echo = FALSE}
## Show feature code regardless of version, but only EVALUATE it when DNLP is
## available: CVXR >= 1.9.1, the sparsediff derivative package, and an NLP
## solver. Uno is the realistic R solver (it is on CRAN and bundles its
## own MUMPS/HiGHS); the `ipopt` package is optional (not on CRAN), so any
## IPOPT-specific chunk below is additionally guarded with requireNamespace.
run_ok <- utils::packageVersion("CVXR") >= numeric_version(params$min_version) &&
          requireNamespace("sparsediff", quietly = TRUE) &&
          requireNamespace("Uno", quietly = TRUE)
knitr::opts_chunk$set(eval = run_ok)
```

```{r, echo = FALSE, message = FALSE, eval = run_ok && params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
```

## Introduction

**Disciplined Nonlinear Programming (DNLP)**, introduced in `CVXR` 1.9.1,
extends `CVXR` beyond convex optimization to **smooth nonlinear programs**, which
need not be convex. You build the problem from differentiable atoms, check it
with `is_dnlp()`, and solve it with `psolve(prob, nlp = TRUE)`. Every DCP problem
is also a DNLP, and the disciplined nonlinear grammar additionally allows smooth
atoms in forms DCP forbids (for example, a product of two variable-dependent
expressions).

The DNLP path needs two optional pieces: the
[`sparsediff`](https://cran.r-project.org/package=sparsediff) package for automatic
derivatives, and a nonlinear solver. These live in `Enhances` (not installed by
default); `CVXR` checks for them at solve time and errors informatively if one is
missing, so you do not need to guard your own calls. You can still see what is
available:

```{r availability}
have_dnlp <- requireNamespace("sparsediff", quietly = TRUE) &&
             requireNamespace("Uno", quietly = TRUE)
c(sparsediff = requireNamespace("sparsediff", quietly = TRUE),
  Uno        = requireNamespace("Uno", quietly = TRUE),
  ipopt      = requireNamespace("ipopt", quietly = TRUE))
```

This vignette uses **Uno** — the solver most R users will have, since it is
installable from CRAN and bundles its own linear-algebra backend. We name it
explicitly with `solver = "UNO"` and quiet it with `logger = "SILENT"`. (The
optional `ipopt` package provides IPOPT, which `nlp = TRUE` prefers when present;
see *Solvers* below.)

## A first smooth program

Every DCP problem is a DNLP, so the simplest way to meet the NLP path is to send
a convex problem down it.

```{r first}
x <- Variable(2)
prob <- Problem(Minimize(sum_squares(x - c(1, 2))))
is_dnlp(prob)

opt <- psolve(prob, nlp = TRUE, solver = "UNO", logger = "SILENT")
opt                 # optimal value (0)
value(x)            # (1, 2)
```

```{r first-test, echo = FALSE, eval = run_ok && params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(opt = opt, x = value(x)), file = file.path(testdata_dir, "first.RDS"))
} else {
    saved <- readRDS(file.path(testdata_dir, "first.RDS"))
    expect_equal(opt, saved$opt, tolerance = 1e-10)
    expect_equal(value(x), saved$x, tolerance = 1e-10)
}
```

## A nonconvex example: maximizing a quadratic

Unlike DCP, DNLP can express genuinely nonconvex problems. A classic example
maximizes a quadratic form over the unit sphere,
$$\text{maximize}\ \ x^\top A x \quad\text{subject to}\quad \lVert x \rVert_2^2 = 1,$$
whose optimal value is the **largest eigenvalue of $A$**. Maximizing a convex
quadratic is nonconvex, so this is not a DCP problem — but it *is* a DNLP. We set
the starting point with `value(x) <- ...`.

```{r quadform}
set.seed(0)
n <- 3
A <- matrix(rnorm(n * n), n); A <- t(A) %*% A      # a random PSD matrix

x <- Variable(n)
prob <- Problem(Maximize(quad_form(x, A)), list(sum_squares(x) == 1))
is_dnlp(prob)

value(x) <- rep(1, n)                              # initial point
opt_nlp <- psolve(prob, nlp = TRUE, solver = "UNO", logger = "SILENT")

c(dnlp = opt_nlp, max_eigenvalue = max(eigen(A, only.values = TRUE)$values))
```

```{r quadform-test, echo = FALSE, eval = run_ok && params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(opt = opt_nlp), file = file.path(testdata_dir, "quadform.RDS"))
} else {
    saved <- readRDS(file.path(testdata_dir, "quadform.RDS"))
    expect_equal(opt_nlp, saved$opt, tolerance = 1e-10)
    ## cross-check against the analytic optimum (max eigenvalue): different
    ## computations, so they agree to solver tolerance, not 1e-10.
    expect_equal(opt_nlp, max(eigen(A, only.values = TRUE)$values), tolerance = 1e-6)
}
```

::: {.callout-warning}
Unlike convex optimization, an NLP solver gives **no global guarantee**. On a
nonconvex problem it may return a local (not global) optimum, give a result that
depends on the starting point, or even fail to find a feasible point. Supplying a
good starting point with `value(x) <- ...` — or trying several with `best_of`
(below) — improves the odds of finding the global optimum.
:::

## Smooth atoms

DNLP adds a family of classically differentiable atoms that are *not* part of
the convex grammar, so they can appear anywhere in a smooth objective or
constraint. Each is twice continuously differentiable on the interior of its
domain:

| Atom | Meaning | Domain |
|------|---------|--------|
| `sin(x)`, `cos(x)` | Trigonometric functions | all $x$ |
| `tan(x)` | Tangent | $x \in (-\pi/2,\ \pi/2)$ |
| `sinh(x)`, `tanh(x)` | Hyperbolic functions | all $x$ |
| `asinh(x)` | Inverse hyperbolic sine | all $x$ |
| `atanh(x)` | Inverse hyperbolic tangent | $x \in (-1,\ 1)$ |
| `normcdf(x)` | Standard normal CDF | all $x$ |
| `prod(x)` | Product of entries (also `prod_entries(X, axis = ...)`) | all $x$ |

These join the smooth atoms `CVXR` already had — affine maps, `exp`, `log`,
`entr`, `logistic`, `kl_div`, `rel_entr`, `xexp`, `power` (constant exponent),
`geo_mean`, `log_sum_exp`, `quad_form`, and `quad_over_lin` — which DNLP can now
use in either curvature direction.

**Atoms with new meaning.** A few existing atoms become smooth in DNLP and may be
used in ways DCP forbids: elementwise multiplication `x * y` and `quad_form(x, Q)`
accept *two* variable arguments (in DCP, `x * y` requires one operand to be
constant). Nonsmooth convex or
concave atoms (`abs`, `max`, `min`, `norm1`, …) keep their curvature and may
appear wherever a convex or concave expression is allowed. (When dividing, as in
`1/x`, DNLP assumes the denominator is nonnegative.)

### Inspecting an expression

Three predicates classify an individual expression under the DNLP grammar,
mirroring CVXPY's `expr.is_smooth()` family. `is_smooth(e)` is `TRUE` when `e`
is twice continuously differentiable on the interior of its domain;
`is_linearizable_convex(e)` and `is_linearizable_concave(e)` report whether `e`
becomes convex (respectively concave) once every smooth subexpression is
linearized — the composition rule `is_dnlp()` applies problem-wide.

```{r predicates}
x <- Variable()
classify <- function(e) c(smooth      = is_smooth(e),
                          lin_convex  = is_linearizable_convex(e),
                          lin_concave = is_linearizable_concave(e))
rbind(`sin(x)`  = classify(sin(x)),    # smooth atom
      `abs(x)`  = classify(abs(x)),    # nonsmooth, convex only
      `-abs(x)` = classify(-abs(x)))   # nonsmooth, concave only
```

A smooth atom such as `sin(x)` is linearizable in *both* directions (it
linearizes to an affine expression), so it may appear anywhere in a smooth
objective or constraint. A nonsmooth convex atom like `abs(x)` is
linearizable-convex only, and its negation is linearizable-concave only — they
keep the curvature DCP assigns them.

```{r predicates-test, echo = FALSE, eval = run_ok && params$mode %in% c("test", "save")}
pred_tbl <- rbind(`sin(x)`  = classify(sin(x)),
                  `abs(x)`  = classify(abs(x)),
                  `-abs(x)` = classify(-abs(x)))
if (params$mode == "save") {
    saveRDS(pred_tbl, file = file.path(testdata_dir, "predicates.RDS"))
} else {
    expect_identical(pred_tbl, readRDS(file.path(testdata_dir, "predicates.RDS")))
}
```

## Beyond DCP: a nonconvex form of a convex problem

Consider the **maximum-entropy distribution** on $\{1,\dots,K\}$ with a
prescribed mean. The negative entropy $\sum_i p_i \log p_i$ is convex, but if we
write it literally as a sum of products `p * log(p)` — rather than with the
`entr` atom — the expression is *not* DCP, because it multiplies two
variable-dependent terms. It is, however, a perfectly good **smooth** objective,
so it is a DNLP.

```{r entropy}
K <- 5; support <- 1:K; mu <- 2
p <- Variable(K, nonneg = TRUE)

objective   <- Minimize(sum(p * log(p)))          # not DCP, but smooth
constraints <- list(sum(p) == 1, sum(support * p) == mu)
prob <- Problem(objective, constraints)

c(is_dcp = is_dcp(prob), is_dnlp = is_dnlp(prob))

neg_entropy <- psolve(prob, nlp = TRUE, solver = "UNO", logger = "SILENT")
round(value(p), 5)
c(total = sum(value(p)), mean = sum(support * value(p)))
```

Maximum-entropy theory says the solution is a discrete exponential (Gibbs)
distribution $p_i \propto e^{-\theta i}$ — equivalently, $\log p_i$ is affine in
$i$. The solution bears this out: the successive differences of $\log p_i$ are
constant, and that common value is $-\theta$.

```{r entropy-gibbs}
log_p <- log(as.numeric(value(p)))
round(diff(log_p), 5)                 # constant => Gibbs form
theta <- -mean(diff(log_p))
cf <- exp(-theta * support); cf <- cf / sum(cf)
round(cf, 5)                          # matches value(p)
```

## Duals on the NLP path

As a `CVXR` extension beyond CVXPY, the **UNO** solver recovers constraint duals
after an NLP solve. For the entropy problem, the multiplier on the mean
constraint is exactly the Gibbs parameter $\theta$ found above:

```{r dual-uno}
dual_value(constraints[[2]])          # mean-constraint multiplier (= theta)
```

```{r dual-test, echo = FALSE, eval = run_ok && params$mode %in% c("test", "save")}
uno_theta <- as.numeric(dual_value(constraints[[2]]))
if (params$mode == "save") {
    saveRDS(list(uno_theta = uno_theta), file = file.path(testdata_dir, "duals.RDS"))
} else {
    saved <- readRDS(file.path(testdata_dir, "duals.RDS"))
    expect_equal(uno_theta, saved$uno_theta, tolerance = 1e-10)   # same solver/machine
    expect_equal(uno_theta, theta, tolerance = 1e-6)              # UNO dual == Gibbs theta
}
```

The **IPOPT** path, by contrast, returns no duals — matching CVXPY's NLP
interface. If the optional `ipopt` package is installed, you can see the
difference (`dual_value()` is `NULL` after an IPOPT solve):

```{r dual-ipopt, eval = run_ok && requireNamespace("ipopt", quietly = TRUE)}
psolve(prob, nlp = TRUE, solver = "IPOPT", print_level = 0L, sb = "yes")
dual_value(constraints[[2]])          # NULL: IPOPT path exposes no duals
```

## Nonconvex problems and random restarts

DNLP problems can be genuinely nonconvex, with multiple local minima. A
nonlinear solver returns a *local* optimum that depends on the starting point.
For example, minimizing $\sin(x) + \tfrac{1}{10}x^2$ over $[-10, 10]$ has several
local minima.

`psolve(prob, nlp = TRUE, best_of = n)` solves from `n` random initial points
and keeps the best result. Each starting point is drawn from the variable's
finite bounds, or from a sampling region you set with `sample_bounds()`.

```{r bestof}
set.seed(1)
x <- Variable(bounds = list(-10, 10))
prob <- Problem(Minimize(sin(x) + 0.1 * x^2))

best <- psolve(prob, nlp = TRUE, solver = "UNO", logger = "SILENT", best_of = 5)
best                                  # best objective over the 5 restarts
value(x)

## the per-restart objectives are available for inspection
solver_stats(prob)@extra_stats$all_objs_from_best_of
```

```{r bestof-test, echo = FALSE, eval = run_ok && params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(best = best, x = value(x)), file = file.path(testdata_dir, "bestof.RDS"))
} else {
    saved <- readRDS(file.path(testdata_dir, "bestof.RDS"))
    expect_equal(best, saved$best, tolerance = 1e-10)
    expect_equal(value(x), saved$x, tolerance = 1e-10)
}
```

For a variable with no finite bounds, supply a sampling region explicitly:

```{r samplebounds, eval = FALSE}
x <- Variable(2)
sample_bounds(x) <- c(-5, 5)          # draw restarts from [-5, 5]^2
```

## Solvers

`CVXR` supports two open-source nonlinear solvers for the DNLP path, each from an
optional R package (both in `Enhances`, so not installed by default):

| Solver | R package | Notes |
|--------|-----------|-------|
| **UNO** | [`Uno`](https://cran.r-project.org/package=Uno) | On CRAN; self-contained (bundles MUMPS/HiGHS); also recovers `dual_value()`. The practical default for R users. |
| **IPOPT** | [`ipopt`](https://bnaras.github.io/ipopt/) | Interior point; **not on CRAN** (license), so installed only by determined users. When present, `nlp = TRUE` prefers it (matching CVXPY). |

(`KNITRO` and `COPT` solver names are registered for CVXPY parity but require
bindings not yet available in R.)

With `nlp = TRUE` and no explicit `solver`, `CVXR` picks an installed NLP solver
automatically — IPOPT first if available, otherwise UNO. Because IPOPT is rarely
installed, most R users get UNO; we name it explicitly above for reproducibility.
You can always request one:

```{r solvers, eval = FALSE}
psolve(prob, nlp = TRUE, solver = "UNO")       # Uno (recommended for R)
psolve(prob, nlp = TRUE, solver = "uno_sqp")   # Uno's filterSQP preset
psolve(prob, nlp = TRUE, solver = "IPOPT")     # Ipopt, if the ipopt package is installed
```

A note on `UNO`: the `CVXR` interface defaults `UNO` to its interior-point
`ipopt` preset (with MUMPS), which differs from CVXPY's `filtersqp` default. The
bundled `Uno` build is HiGHS-only, and HiGHS solves only *convex* quadratic
subproblems, so `filtersqp` is reliable only for problems whose subproblem
Hessians stay positive semidefinite; the interior-point preset is robust on
nonconvex problems too.

## Notes

* DNLP needs the `sparsediff` package (for derivatives) and an NLP solver —
  `Uno` or the optional `ipopt`. All are in `Enhances` (not installed by
  default). `CVXR` checks for them at solve time, so you need not guard your own
  calls: if anything is missing, `psolve(nlp = TRUE)` raises an informative error
  naming exactly what to install (`sparsediff` and `Uno` from CRAN; `ipopt` from
  <https://bnaras.github.io/ipopt/>).
* `is_dnlp(prob)` reports whether a problem satisfies the disciplined nonlinear
  grammar; since DCP $\subset$ DNLP, every DCP problem qualifies.
* Because the problems may be nonconvex, the returned point is a local optimum;
  use `best_of` to improve the odds of finding the global optimum.

## References
