---
title: Least Squares
author: Anqi Fu and Balasubramanian Narasimhan
categories: []
tags: []
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: least-squares
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
```

## Introduction

In a least-squares, or linear regression, problem, we have
measurements $A \in \mathcal{R}^{m \times n}$ and
$b \in \mathcal{R}^m$ and seek a vector $x \in \mathcal{R}^{n}$ such
that $Ax$ is close to $b$. Closeness is defined as the sum of the
squared differences:

$$
\sum_{i=1}^m (a_i^Tx - b_i)^2,
$$

also known as the $\ell_2$-norm squared, $\|Ax - b\|_2^2$.

For example, we might have a dataset of $m$ users, each represented
by $n$ features. Each row $a_i^T$ of $A$ is the features for user
$i$, while the corresponding entry $b_i$ of $b$ is the measurement we
want to predict from $a_i^T$, such as ad spending. The prediction is
given by $a_i^Tx$.

We find the optimal $x$ by solving the optimization problem

$$
\begin{array}{ll}
\mbox{minimize} & \|Ax - b\|_2^2.
\end{array}
$$

Let $x^\star$ denote the optimal $x$. The quantity
$r = Ax^\star - b$ is known as the residual. If $\|r\|_2 = 0$, we
have a perfect fit.

## Example

In the following code, we solve a least-squares problem with `CVXR`.

```{r}
## Generate data
set.seed(1)
m <- 20
n <- 15
A <- matrix(rnorm(m * n), nrow = m, ncol = n)
b <- rnorm(m)

## Define and solve the CVXR problem
x <- Variable(n)
cost <- sum_squares(A %*% x - b)
prob <- Problem(Minimize(cost))
result <- psolve(prob)
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(optimal_value = result, x_value = value(x)),
            file = file.path(testdata_dir, "least-squares.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    ls_data <- readRDS(file = file.path(testdata_dir, "least-squares.RDS"))
    expect_equal(result, ls_data$optimal_value, tolerance = 1e-5)
    expect_equal(value(x), ls_data$x_value, tolerance = 1e-5)
}
```

```{r}
## Print result
cat(sprintf("The optimal value is %f\n", result))
cat("The optimal x is\n")
print(value(x))
cat(sprintf("The norm of the residual is %f\n",
            value(norm2(A %*% x - b))))
```


## Session Info

```{r}
sessionInfo()
```

## References
