---
title: Linear Program
author: Anqi Fu and Balasubramanian Narasimhan
categories: []
tags: []
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: linear-program
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
source(here::here("_check_status.R"))
```

## Introduction

A linear program is an optimization problem with a linear objective
and affine inequality constraints. A common standard form is the
following:

$$
\begin{array}{ll}
\mbox{minimize}   & c^Tx \\
\mbox{subject to} & Ax \leq b.
\end{array}
$$

Here $A \in \mathcal{R}^{m \times n}$, $b \in \mathcal{R}^m$, and
$c \in \mathcal{R}^n$ are problem data and $x \in \mathcal{R}^{n}$ is
the optimization variable. The inequality constraint $Ax \leq b$ is
elementwise.

For example, we might have $n$ different products, each constructed
out of $m$ components. Each entry $A_{ij}$ is the amount of component
$i$ required to build one unit of product $j$. Each entry $b_i$ is
the total amount of component $i$ available. We lose $c_j$ for each
unit of product $j$ ($c_j < 0$ indicates profit). Our goal then is to
choose how many units of each product $j$ to make, $x_j$, in order to
minimize loss without exceeding our budget for any component.

In addition to a solution $x^\star$, we obtain a dual solution
$\lambda^\star$. A positive entry $\lambda^\star_i$ indicates that the
constraint $a_i^Tx \leq b_i$ holds with equality for $x^\star$ and
suggests that changing $b_i$ would change the optimal value.

## Example

In the following code, we solve a linear program with `CVXR`.

```{r}
## Generate a random non-trivial linear program
set.seed(1)
m <- 15
n <- 10
s0 <- rnorm(m)
lamb0 <- pmax(-s0, 0)
s0 <- pmax(s0, 0)
x0 <- rnorm(n)
A <- matrix(rnorm(m * n), nrow = m, ncol = n)
b <- A %*% x0 + s0
c_vec <- -t(A) %*% lamb0

## Define and solve the CVXR problem
x <- Variable(n)
constraints <- list(A %*% x <= b)
prob <- Problem(Minimize(t(c_vec) %*% x), constraints)
result <- psolve(prob)
check_solver_status(prob)
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(optimal_value = result, x_value = value(x),
                 dual_value = dual_value(constraints[[1]])),
            file = file.path(testdata_dir, "linear-program.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    lp_data <- readRDS(file = file.path(testdata_dir, "linear-program.RDS"))
    expect_equal(result, lp_data$optimal_value, tolerance = 1e-5)
    expect_equal(value(x), lp_data$x_value, tolerance = 1e-5)
}
```

```{r}
## Print result
cat(sprintf("The optimal value is %f\n", result))
cat("A solution x is\n")
print(value(x))
cat("A dual solution is\n")
print(dual_value(constraints[[1]]))
```


## Session Info

```{r}
sessionInfo()
```

## References
