---
title: Mixed-Integer Quadratic Program
author: Anqi Fu and Balasubramanian Narasimhan
categories: []
tags: []
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: miqp
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
```

## Introduction

A mixed-integer quadratic program (MIQP) is an optimization problem of
the form

$$
\begin{array}{ll}
\mbox{minimize}   & x^T Q x + q^T x + r \\
\mbox{subject to} & x \in \mathcal{C}\\
                   & x \in \mathbf{Z}^n,
\end{array}
$$

where $x \in \mathbf{Z}^n$ is the optimization variable
($\mathbf{Z}^n$ is the set of $n$-dimensional vectors with
integer-valued components), $Q \in \mathbf{S}_+^n$ (the set of
$n \times n$ symmetric positive semidefinite matrices),
$q \in \mathbf{R}^n$, and $r \in \mathbf{R}$ are problem data, and
$\mathcal{C}$ is some convex set.

An example of an MIQP is mixed-integer least squares, which has the
form

$$
\begin{array}{ll}
\mbox{minimize}   & \|Ax-b\|_2^2 \\
\mbox{subject to} & x \in \mathbf{Z}^n,
\end{array}
$$

where $x \in \mathbf{Z}^n$ is the optimization variable, and
$A \in \mathbf{R}^{m \times n}$ and $b \in \mathbf{R}^{m}$ are the
problem data. A solution $x^{\star}$ of this problem will be a vector
in $\mathbf{Z}^n$ that minimizes $\|Ax-b\|_2^2$.

## Example

In the following code, we solve a mixed-integer least-squares problem
with `CVXR`. We use the `GUROBI` solver which supports
mixed-integer quadratic programs. (Alternatively, `CPLEX` can also
be used.)

::: {.callout-note}
MIQP problems require a commercial solver such as GUROBI or CPLEX.
Free solvers like GLPK_MI support mixed-integer _linear_ programs but
not mixed-integer _quadratic_ programs.
:::

```{r}
## Generate a random problem
set.seed(0)
m <- 40
n <- 25

A <- matrix(runif(m * n), nrow = m, ncol = n)
b <- rnorm(m)
```

```{r}
## Construct and solve a CVXR problem
x <- Variable(n, integer = TRUE)
objective <- Minimize(sum_squares(A %*% x - b))
prob <- Problem(objective)
result <- psolve(prob, solver = "GUROBI")
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(optimal_value = result, x_value = value(x),
                 status = status(prob)),
            file = file.path(testdata_dir, "miqp.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    miqp_data <- readRDS(file = file.path(testdata_dir, "miqp.RDS"))
    expect_equal(result, miqp_data$optimal_value, tolerance = 1e-3)
    expect_equal(value(x), miqp_data$x_value, tolerance = 1e-3)
}
```

```{r, results = "hold"}
cat(sprintf("Status: %s\n", status(prob)))
cat(sprintf("The optimal value is %f\n", result))
cat("A solution x is\n")
print(value(x))
```


## Session Info

```{r, echo = FALSE}
sessionInfo()
```

## References
