---
title: Quadratic Program
author: Anqi Fu and Balasubramanian Narasimhan
categories: []
tags: []
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: quadratic-program
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
```

## Introduction

A quadratic program is an optimization problem with a quadratic
objective and affine equality and inequality constraints. A common
standard form is the following:

$$
\begin{array}{ll}
\mbox{minimize}   & (1/2)x^TPx + q^Tx\\
\mbox{subject to} & Gx \leq h \\
                   & Ax = b.
\end{array}
$$

Here $P \in \mathcal{S}^{n}_+$, $q \in \mathcal{R}^n$,
$G \in \mathcal{R}^{m \times n}$, $h \in \mathcal{R}^m$,
$A \in \mathcal{R}^{p \times n}$, and $b \in \mathcal{R}^p$ are
problem data and $x \in \mathcal{R}^{n}$ is the optimization
variable. The inequality constraint $Gx \leq h$ is elementwise.

A simple example of a quadratic program arises in finance. Suppose we
have $n$ different stocks, an estimate $r \in \mathcal{R}^n$ of the
expected return on each stock, and an estimate
$\Sigma \in \mathcal{S}^{n}_+$ of the covariance of the returns. Then
we solve the optimization problem

$$
\begin{array}{ll}
\mbox{minimize}   & (1/2)x^T\Sigma x - r^Tx\\
\mbox{subject to} & x \geq 0 \\
                   & \mathbf{1}^Tx = 1,
\end{array}
$$

to find a portfolio allocation $x \in \mathcal{R}^n_+$ that optimally
balances expected return and variance of return.

When we solve a quadratic program, in addition to a solution
$x^\star$, we obtain a dual solution $\lambda^\star$ corresponding to
the inequality constraints. A positive entry $\lambda^\star_i$
indicates that the constraint $g_i^Tx \leq h_i$ holds with equality
for $x^\star$ and suggests that changing $h_i$ would change the
optimal value.

## Example

In the following code, we solve a quadratic program with `CVXR`.

```{r}
## Generate a random non-trivial quadratic program
set.seed(1)
m <- 15
n <- 10
p <- 5
P <- matrix(rnorm(n * n), nrow = n, ncol = n)
P <- t(P) %*% P
q <- rnorm(n)
G <- matrix(rnorm(m * n), nrow = m, ncol = n)
h <- G %*% rnorm(n)
A <- matrix(rnorm(p * n), nrow = p, ncol = n)
b <- rnorm(p)

## Define and solve the CVXR problem
x <- Variable(n)
ineq_con <- G %*% x <= h
eq_con <- A %*% x == b
prob <- Problem(Minimize(0.5 * quad_form(x, P) + t(q) %*% x),
                constraints = list(ineq_con, eq_con))
result <- psolve(prob)
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(optimal_value = result, x_value = value(x),
                 dual_value = dual_value(ineq_con)),
            file = file.path(testdata_dir, "quadratic-program.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    qp_data <- readRDS(file = file.path(testdata_dir, "quadratic-program.RDS"))
    expect_equal(result, qp_data$optimal_value, tolerance = 1e-5)
    expect_equal(value(x), qp_data$x_value, tolerance = 1e-5)
}
```

```{r}
## Print result
cat(sprintf("The optimal value is %f\n", result))
cat("A solution x is\n")
print(value(x))
cat("A dual solution corresponding to the inequality constraints is\n")
print(dual_value(ineq_con))
```


## Session Info

```{r}
sessionInfo()
```

## References
