---
title: Semidefinite Program
author: Anqi Fu and Balasubramanian Narasimhan
categories: []
tags: []
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: sdp
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
source(here::here("_check_status.R"))
```

## Introduction

A semidefinite program (SDP) is an optimization problem of the form

$$
\begin{array}{ll}
\mbox{minimize}   & \mathbf{tr}(CX) \\
\mbox{subject to} & \mathbf{tr}(A_iX) = b_i, \quad i=1,\ldots,p \\
                   & X \succeq 0,
\end{array}
$$

where $\mathbf{tr}$ is the trace function,
$X \in \mathcal{S}^{n}$ is the optimization variable and
$C, A_1, \ldots, A_p \in \mathcal{S}^{n}$, and
$b_1, \ldots, b_p \in \mathcal{R}$ are problem data, and
$X \succeq 0$ is a matrix inequality. Here $\mathcal{S}^{n}$ denotes
the set of $n$-by-$n$ symmetric matrices.

An example of an SDP is to complete a covariance matrix
$\tilde \Sigma \in \mathcal{S}^{n}_+$ with missing entries
$M \subset \{1,\ldots,n\} \times \{1,\ldots,n\}$:

$$
\begin{array}{ll}
\mbox{minimize}   & 0 \\
\mbox{subject to} & \Sigma_{ij} = \tilde \Sigma_{ij}, \quad (i,j) \notin M \\
                   & \Sigma \succeq 0,
\end{array}
$$

## Example

In the following code, we solve a SDP with `CVXR`.

```{r}
## Problem dimensions
set.seed(4)
n <- 3   # matrix size
p <- 3   # number of equality constraints

## Generate random problem data
C <- matrix(rnorm(n^2), nrow = n)
sdp_data <- lapply(1:p, function(i) {
    list(A = matrix(rnorm(n^2), nrow = n), b = rnorm(1))
})

## Define and solve the CVXR problem
X <- Variable(c(n, n), symmetric = TRUE)
## Trace constraints: tr(A_i X) = b_i
trace_constraints <- lapply(sdp_data, function(s) {
    matrix_trace(s$A %*% X) == s$b
})
prob <- Problem(Minimize(matrix_trace(C %*% X)),
                constraints = c(list(PSD(X)), trace_constraints))
result <- psolve(prob)
check_solver_status(prob)
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(optimal_value = result, X_value = value(X)),
            file = file.path(testdata_dir, "sdp.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    sdp_data <- readRDS(file = file.path(testdata_dir, "sdp.RDS"))
    expect_equal(result, sdp_data$optimal_value, tolerance = 1e-5)
    expect_equal(value(X), sdp_data$X_value, tolerance = 1e-5)
}
```

```{r, results = "hold"}
## Print result
cat(sprintf("The optimal value is %f\n", result))
cat("A solution X is\n")
print(value(X))
```


## Session Info

```{r, echo = FALSE}
sessionInfo()
```

## References
