---
title: Second-Order Cone Program
author: Anqi Fu and Balasubramanian Narasimhan
categories: []
tags: []
params:
  mode: ignore
  testdata_dir: test_data_1.8
  data_dir: socp
---

```{r, echo = FALSE, message = FALSE, eval = params$mode %in% c("test", "save")}
library(here)
testdata_dir  <- here("static", params$testdata_dir, params$data_dir)
if (params$mode == "test") {
    library(testthat)
} else {
    if (!dir.exists(testdata_dir)) dir.create(testdata_dir)
}
```

```{r prereqs, message = FALSE, echo = FALSE}
library(CVXR)
source(here::here("_check_status.R"))
```

## Introduction

A second-order cone program (SOCP) is an optimization problem of the
form

$$
\begin{array}{ll}
\mbox{minimize}   & f^Tx\\
\mbox{subject to} & \|A_ix + b_i\|_2 \leq c_i^Tx + d_i, \quad i=1,\ldots,m \\
                   & Fx = g,
\end{array}
$$

where $x \in \mathcal{R}^{n}$ is the optimization variable and
$f \in \mathcal{R}^n$, $A_i \in \mathcal{R}^{n_i \times n}$,
$b_i \in \mathcal{R}^{n_i}$, $c_i \in \mathcal{R}^n$,
$d_i \in \mathcal{R}$, $F \in \mathcal{R}^{p \times n}$, and
$g \in \mathcal{R}^p$ are problem data.

An example of an SOCP is the robust linear program

$$
\begin{array}{ll}
\mbox{minimize}   & c^Tx\\
\mbox{subject to} & (a_i + u_i)^Tx \leq b_i \textrm{ for all } \|u_i\|_2 \leq 1, \quad i=1,\ldots,m,
\end{array}
$$

where the problem data $a_i$ are known within an $\ell_2$-norm ball of
radius one. The robust linear program can be rewritten as the SOCP

$$
\begin{array}{ll}
\mbox{minimize}   & c^Tx\\
\mbox{subject to} & a_i^Tx + \|x\|_2 \leq b_i, \quad i=1,\ldots,m,
\end{array}
$$

When we solve a SOCP, in addition to a solution $x^\star$, we obtain a
dual solution $\lambda_i^\star$ corresponding to each second-order
cone constraint. A non-zero $\lambda_i^\star$ indicates that the
constraint $\|A_ix + b_i\|_2 \leq c_i^Tx + d_i$ holds with equality
for $x^\star$ and suggests that changing $d_i$ would change the
optimal value.

## Example

In the following code, we solve a SOCP with `CVXR`. The second-order
cone constraint $\|Ax + b\|_2 \leq t$ is expressed as
`norm2(A %*% x + b) <= t` in `CVXR`.

```{r}
## Generate a random feasible SOCP
set.seed(2)
m <- 3
n <- 10
p <- 5
n_i <- 5
f <- rnorm(n)
A <- list()
b_vec <- list()
c_list <- list()
d <- list()
x0 <- rnorm(n)
for (i in 1:m) {
    A[[i]] <- matrix(rnorm(n_i * n), nrow = n_i, ncol = n)
    b_vec[[i]] <- rnorm(n_i)
    c_list[[i]] <- rnorm(n)
    d[[i]] <- norm(A[[i]] %*% x0 + b_vec[[i]], type = "2") -
        sum(c_list[[i]] * x0)
}
F_mat <- matrix(rnorm(p * n), nrow = p, ncol = n)
g <- F_mat %*% x0

## Define and solve the CVXR problem
x <- Variable(n)
## SOC constraints: ||A_i x + b_i||_2 <= c_i^T x + d_i
soc_constraints <- lapply(1:m, function(i) {
    norm2(A[[i]] %*% x + b_vec[[i]]) <= sum(c_list[[i]] * x) + d[[i]]
})
prob <- Problem(Minimize(t(f) %*% x),
                constraints = c(soc_constraints, list(F_mat %*% x == g)))
result <- psolve(prob)
check_solver_status(prob)
```

```{r, echo = FALSE, eval = params$mode %in% c("test", "save")}
if (params$mode == "save") {
    saveRDS(list(optimal_value = result, x_value = value(x)),
            file = file.path(testdata_dir, "socp.RDS"))
} else {
    cat("Testthat Results: No output is good\n")
    socp_data <- readRDS(file = file.path(testdata_dir, "socp.RDS"))
    expect_equal(result, socp_data$optimal_value, tolerance = 1e-5)
    expect_equal(value(x), socp_data$x_value, tolerance = 1e-5)
}
```

```{r, results = "hold"}
## Print result
cat(sprintf("The optimal value is %f\n", result))
cat("A solution x is\n")
print(value(x))
for (i in 1:m) {
    cat(sprintf("SOC constraint %d dual variable solution\n", i))
    print(dual_value(soc_constraints[[i]]))
}
```


## Session Info

```{r, echo = FALSE}
sessionInfo()
```

## References
